### 2017-09-05 analyses/Loading_data_example.Rmd
Initial commit

---
title: "quick pipeline for getting data from discrete matrices"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 6
    fig_height: 6
---

Quick pipeline for getting data from discrete matrices.
===============

Packages
-----------------------------------

Make sure you work with `dispRity` 0.4 and my version of `Claddis` (that is much faster than Graeme's).

```{r, eval = TRUE}
# The latest version of devtools and ape
if(!require(devtools)) install.packages("devtools")
if(!require(ape)) install.packages("ape")

## The latest version of Claddis (on development on my own branch)
install_github("TGuillerme/Claddis")

## The latest version of dispRity 
install_github("TGuillerme/dispRity", ref = "release")
```

Loading the data
-----------------------------------

Loading the packages and setting the seed:

```{r}
set.seed(123)
library(ape); library(Claddis) ; library(dispRity) 
```

Loading the data.
It's based on a chain name (here "Beck2014") and should work with different chain names if their files are in the exact same format (it's not the case yet!).

```{r}
## Setting up the chain name
chain <- "Beck2014"

## Loading the discrete morphological matrix
matrix <- ReadMorphNexus(paste0("../Data/Matrices/", chain, ".nex"))

## Loading the FADLAD
FADLAD <- read.csv(paste0("../Data/FADLAD/", chain, ".csv"), row.names = 1, header = TRUE)

## Loading the tree
tree <- read.nexus(paste0("../Data/Trees/", chain, ".tre"))

## Adding node labels
tree <- makeNodeLabel(tree, method = "number", prefix = "n")

## Adding a root age (max tree age date)
tree$root.time <- max(tree.age(tree)[,1])
```

We'll also clean the data to match the tree and the matrix:

```{r}
## Cleaning the data 
cleaned_data <- clean.data(matrix$matrix, tree)

## Cleaned matrix
matrix$matrix <- cleaned_data$data

## Cleaned tree
tree <- cleaned_data$tree

## Removed taxa:
as.vector(na.omit(c(cleaned_data$dropped_tips, cleaned_data$dropped_rows)))
```

Ordination
===============

Ancestral states reconstruction
-----------------------------------

First let's recreate the nodal states but excluding any estimation with less than 0.95 scaled likelihood value.
This allows us to discard any dodgy/ambiguous reconstruction (note that NAs are not a problem since we use Gower's distances further down the road).

This have some running time!

```{r, eval = TRUE}
## Estimating the ancestral states (with a uncertainty threshold of 0.95)
matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
                                  uncertainty.threshold = 0.95)

## Saving the matrix
save(matrix_nodes, file = paste0("../data/backups/ancestral_states.", chain ,".Rda"))
```

This takes some time (couple of hours) so it can be skipped for the demo.

After that we can add the node states to the overall matrix:

```{r}
## Loading the matrix
load(paste0("../data/backups/ancestral_states.", chain ,".Rda"))

## Adding the node names as row names
row.names(matrix_nodes) <- tree$node.label

## Combining the tips and the nodes in the same matrix
matrix$matrix <- rbind(matrix$matrix, matrix_nodes) 
```

Distance matrix
-----------------------------------

Let's then calculate the distance matrix (using Gower's distance).
Same thing, the step takes some time (couple of minutes) so can be skipped for convenience.

```{r, eval = TRUE}
## Calculating the Gower distance
matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")

## Saving the matrix
save(matrix_dist, file = paste0("../data/backups/distance_matrix.", chain ,".Rda"))
```

Ordination
-----------------------------------

Ordination is a classic NMDS including only the non-zero variance axis.

```{r}
## Loading the distance matrices
load(paste0("../data/backups/distance_matrix.", chain ,".Rda"))

## Ordinating the matrix
morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = T)$points
```

And that's it... You know have all the arguments for running the `dispRity` package properly: the space (`morphospace`), the tree (`tree`) and the FADLAD (`FADLAD`).
Remember that for the `time.subsamples` function, you can use the `inc.nodes` logical option to include nodes or not!

### 2017-09-05 manuscript/time-slice.tex
Initial commit

% Preamble
\documentclass[12pt,a4paper]{article}
\usepackage{enumerate} 	
\usepackage{setspace}						
\usepackage{authblk}	
\usepackage{graphicx} 	
%\usepackage[nomarkers, nolists]{endfloat} 
\usepackage{pdflscape}	
\usepackage{mathtools}	
\usepackage[osf]{mathpazo} 
\usepackage{lineno} 
\usepackage{ms}    	
\usepackage{hyperref}
\usepackage[round]{natbib} 
\usepackage{setspace}

\setcounter{secnumdepth}{0} 
\raggedright 			
\pagenumbering{arabic}	
\linenumbers

% First order headings upper case bold
\usepackage{titlesec}
\titleformat*{\section}{\small\bfseries\uppercase}

% Second order headings normal case italics
\titleformat*{\subsection}{\small\itshape}

% Third order, italics, paragraph style
\titleformat*{\paragraph}{\small\itshape}

% lists - arabic in (1; (2); (3) .

% Title page information

\title{Time for a rethink: disparity-through-time}

\author{
	Thomas Guillerme$^{1,2*}$ and Natalie Cooper$^{2}$
}

\date{}

\affiliation{\noindent{\footnotesize
	$^1$ Imperial\\ 
	$^2$Department of Life Sciences, Natural History Museum, Cromwell Road, London, SW7 5BD, UK. natalie.cooper@nhm.ac.uk}\\
	$^*$Corresponding author\\}

\vfill

\runninghead{Time slicing}
\keywords{??} 
% up to 6

\begin{document}

\mstitlepage
\parindent = 1.5em
\addtolength{\parskip}{.3em}

\section{Abstract}
% 300 words max

\newpage
\raggedright
\doublespacing
\setlength{\parindent}{1cm}

\section{Introduction}
Disparity-through-time analyes are common in palaeontology.
Examples
They often provide us with exciting insights into mass extinctions, competitive replacements etc.
However, our methods for analysing diversity-through-time have not changed all that much in the XX years they have been popular. 
The way we perfrom these analyses may have profound effects on our conclusions.

One aspect that is not always carefully considered, is how to divide taxa into subsets through time. The nature of disparity, i.e. a diversity metric, means it cannot be calculated using just one individual, so some way of subsetting taxa is required.

Stratigraphy
Time bins
Time slicing (requires a phylogeny)

Changes in disparity through time are generally investigated by calculating the disparity of taxa that occupy the cladisto-space during specific time intervals \citep[e.g][]{cisneros2010,prentice2011,Hughes20082013,hopkinsdecoupling2013,bentonmodels2014,bensonfaunal2014}.
These time intervals are usually defined based on biostratigraphy \citep[e.g.][]{cisneros2010,prentice2011,Hughes20082013,bentonmodels2014} but can also be arbitrarily chosen time periods of equal duration \citep{Butler2012,hopkinsdecoupling2013,bensonfaunal2014}.
However, this approach suffers from two main biases. 
First, if biostratigraphy is used to determine the time intervals, disparity may be distorted towards higher differences between time intervals because biostratigraphical periods are geologically defined based on differences in the morphology of fossils found in the different strata.
Second, this approach assumes that all characters evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring between intervals, rather than also allowing for gradual changes within intervals \citep{Hunt21042015}.

To address these issues, we used a ``time-slicing'' approach that considers subsets of taxa in the cladisto-space at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
This results in even-sampling of the cladisto-space across time and permits us to define the underlying model of character evolution (punctuated or gradual).  
In practice, time-slicing considers the disparity of any element present in the phylogeny (branches, nodes and tips) at any point in time.

We present this using data from Beck \& Lee and Halliday???

\section{Materials and Methods}
\subsection{Overview}
\label{overview-section}
To test the different time subsampling methods, we first need to prepare the data for the disparity

\begin{enumerate}
  \item Collate data. Disparity-through-time analysis requires a matrix containing morphological characters for each taxon, and information on the time range over which each taxon existed. Time slicing methods also require a phylogeny.
  \item Estimate ancestral character states at the nodes of the phylogeny. This step is required for the time slicing methods.
  \item Build cladisto-spaces.  
    \begin{enumerate}
      \item Construct distance matrices.
      \item ordiantion
    \end{enumerate}
  \item time subsample
    \begin{enumerate}
      \item stratigraphy
      \item time bins
        \begin{enumerate}
          \item 5 million years
          \item 10 million years
          \item 15 million years
        \end{enumerate}  
      \item time slicing
        \begin{enumerate}
          \item punctuated
          \item gradual
          \item random
        \end{enumerate}  
    \end{enumerate}
  \item Estimate disparity for each time subsample
  \item Compare different methods
\end{enumerate}

\subsection{Example datasets}
\label{datasets}
To test the different time slicing methods we selected X datasets. 
Note that although disparity-through-time analyses are possible without a phylogeny, we only use examples with phylogenies here so that we can compare results of time slicing methods, which do require a phylogeny.

\paragraph{\textit{Mammal data - Halliday.}}
We used the cladistic morphological matrices and the Total Evidence tip-dated trees from \citet[][103 taxa with 446 morphological characters;]{Slater2012MEE} and \citet[][102 taxa with 421 morphological characters]{beckancient2014}.
%We chose these two datasets because they have a similar number of taxa and morphological characters.
%\cite{Slater2012MEE} ranges from 310 million years ago (Ma; Late Carboniferous) to the present and focuses on the clade Mammaliaformes at the family-level and is called hereafter the Mammaliaformes dataset.
%\cite{beckancient2014} ranges from 170 Ma (Middle Jurassic) to the present and focuses on Eutheria at the genus-level and is called hereafter the Eutheria dataset.
%We used the first and last occurrences reported in \cite{Slater2012MEE} and \cite{beckancient2014} as the temporal range of each taxon in our analysis.
%Both phylogenies are illustrated in figures \ref{fig:SlaterTree} and \ref{fig:BeckTree}.
%Both trees contain few taxa compared to the overall species richness of living and fossil mammals \citep{bininda-emondsthe2007,archibald2011extinction}.
%This is because Total Evidence trees need a lot of data, particularly morphological data for living taxa that can be hard to locate \citep[Chapter \ref{chap:TEM_paper} and \ref{chap:missing_mammals};][]{GuillermeCooperMissing,GuillermeCooper}.
%Therefore, most Total Evidence studies to date contain one or two orders of magnitude fewer taxa than phylogenies based solely on molecular data (e.g. thousands of taxa in \citealt{bininda-emondsthe2007,meredithimpacts2011} \textit{vs.} hundreds in \citealt{ronquista2012,Slater2012MEE,beckancient2014}).

\subsection{Estimating ancestral character states}
\label{ace}
For each dataset we used the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny, using the \texttt{rerootingMethod} function from the \texttt{R} package \texttt{phytools} version 0.4-45 \citep{phytools,R}. % Version still the same?
Where there were missing character data for a taxon we followed the method of \cite{Claddis} and treated missing data as any possible observed state for each character.
For example, if a character had two observed states ($0$ and $1$) across all taxa, we attributed the multi-state ``$0$\&$1$" value to the taxon with missing data, representing an equal probability of being either $0$ or $1$.
This allows the ancestral node of a taxon with missing data to be estimated with no assumptions other than that the taxon has one of the observed character states.
To prevent poor ancestral state reconstructions from biasing our results, especially when a lot of error is associated with the reconstruction, we only included ancestral state reconstructions with a scaled Likelihood $\geq$ $0.95$.
Ancestral state reconstructions with scaled Likelihoods below this threshold were recoded as missing data (``?'').

\subsection{Building cladisto-spaces} % Maybe reduce detail here!
To explore disparity-through-time in our datasets we used a cladisto-space approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
This approach is similar to constructing a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010}, except a cladisto-space is an approximation of the morphospace based on cladistic data (i.e. the discrete morphological characters used to build a phylogenetic tree).
Mathematically, a cladisto-space is an $n$ dimensional object that summarises the cladistic distances between the taxa present in a cladistic matrix (see details below).
Although empirically inter-taxon distances are the same in a morphospace or a cladisto-space \citep{foth2012different,hetherington2015cladistic}, we prefer the term cladisto-space to make it clear that this space is estimated using cladistic data and not morphometric data. %and because both objects have slightly different properties.
%For example, because of its inherent combinatory properties, a cladisto-space is a finite theoretical object limited by the product of the number of character states, whereas a morphospace is an infinite theoretical object.
%Thus a cladisto-space will be overloaded if the number of taxa is higher than the product of the number of character states, although this is rarely an issue with empirical data (our cladisto-spaces have maximal capacities of $1.9$$\times$$10^{181}$ taxa for the Mammaliaformes dataset, i.e. 101 orders of magnitude more taxa than the number of particles in the universe; and $4.5$$\times$$10^{159}$ taxa for the Eutheria dataset).

\paragraph{Constructing distance matrices.}
To estimate the cladisto-spaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
For each dataset separately, we calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the Euclidean distance between two taxa divided by the number of shared characters. 
This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e. because some pairs of taxa share more characters in common than others, they are more likely to be similar).
For cladistic matrices, using this corrected distance is preferable to the raw Euclidean distance because of its ability to deal with discrete or/and ordinated characters as well as with missing data \citep{anderson2012using}.
However, the Gower distance cannot calculate distances when taxa have no overlapping data.
Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis} to remove pairs of taxa with no cladistic characters in common.
%This led to us removing 11 taxa from the Mammaliaformes dataset but none from the Eutheria dataset.
% NC: Add info on species we have to remove.

\paragraph{Ordination.}
After constructing our distance matrices we transformed them using classical multidimensional scaling \citep[MDS;][]{torgerson1965multidimensional,GOWER01121966,cailliez1983analytical}.
This method (also referred to as PCO; e.g. \citealt{Brusatte2015}; or PCoA; e.g. \citealt{paradisape:2004}) is an eigen decomposition of the distance matrix.
Because we used Gower distances instead of raw Euclidean distances, negative eigenvalues can be calculated.
To avoid this problem, we first transformed the distance matrices by applying the Cailliez correction \citep{cailliez1983analytical} which adds a constant $c^*$ to the values in a distance matrix (apart from the diagonal) so that all the Gower distances become Euclidean ($d_{Gower}+c^*=d_{Euclidean}$; \citealt{cailliez1983analytical}). 
We were then able to extract $n$ eigenvectors for each matrix (representing the $n$ dimensions of the cladisto-space) where $n$ is equal to $k-2$, i.e. the number of taxa in the matrix ($k$) minus the last two eigenvectors that are always null after applying the Cailliez correction.
Contrary to previous studies \citep[e.g][]{brusatte50,cisneros2010,prentice2011,anderson2012using,Hughes20082013,bentonmodels2014}, we use all $n$ dimensions of our cladisto-spaces and not a subsample representing the majority of the variance in the distance matrix (e.g. selecting only $m$ dimensions that represent up to 90\% of the variance in the distance matrix; \citealt{Brusatte12092008,toljagictriassic-jurassic2013}). % NC: Hmmm should we do this the usual way???

Note that our cladisto-spaces represent an ordination of all possible morphologies coded in each study through time.
It is unlikely that all morphologies will co-occur at each time point, therefore, the disparity of the whole cladisto-space is expected to be greater than the disparity at any specific point in time.

\subsection{Calculating disparity}
\label{disparity_calc}
Disparity can be estimated in many different ways \citep[e.g.][]{Wills1994,Ciampaglio2004,thorneresetting2011,hopkinsdecoupling2013,huang2015origins}, however most studies estimate disparity using four metrics: the sum and products of ranges and variances, each of which gives a slightly different estimate of how the data fits within the cladisto-space \citep{Foote01071994,Wills1994,brusatte50,Brusatte12092008,cisneros2010,thorneresetting2011,prentice2011,brusattedinosaur2012,toljagictriassic-jurassic2013,ruta2013,bentonmodels2014,bensonfaunal2014}.
However, these metrics have limitations. 
First, the range metrics are affected by the uneven sampling of the fossil record \citep{Butler2012}.
Second, because we include all $n$ dimensions in the analysis (see above), the products of ranges and variances will tend towards zero since the scores of the last dimension are usually really close to zero themselves. 
We therefore only use the sum of variances metric.

\subsection{Estimating disparity through time} 
\label{time_slicing}

% Might make a table with all the subsets???

% Just tips or tips and nodes????
\paragraph{Stratigraphic time bins.}
We extracted the taxa found in each geological stage from each cladisto-space.
These consisted of the Middle Jurassic to the present (XX subsamples) for dataset 1, the Middle Jurassic to the present (XX subsamples) for dataset 2 and the Middle Jurassic to the present (XX subsamples) X for dataset 3.
For each subsample, we estimated its disparity.
To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement by randomly resampling with replacement a new subsample of taxa from the observed taxa in the subsample 1000 times.
We then calculated the median disparity value for each subsample along with the 50\% and 95\% confidence intervals.
We also recorded the number of taxa in each subsample as a proxy for taxonomic diversity.

\paragraph{Equally sized time bins.}
We extracted the taxa found in each time bin for each cladisto-space.
We varied the size of the time bins from 5 million year, 10 million years and 20 million years, to determine if the length of the time bin affected our results.
There were XX subsamples for dataset 1, XX subsamples for dataset 2 and XX subsamples for dataset 3.
For each subsample, we estimated its disparity.
To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement by randomly resampling with replacement a new subsample of taxa from the observed taxa in the subsample 1000 times.
We then calculated the median disparity value for each subsample along with the 50\% and 95\% confidence intervals.
We also recorded the number of taxa in each subsample as a proxy for taxonomic diversity.

\paragraph{Time slicing.}
The ``time-slicing'' approach considers subsets of taxa in the cladisto-space at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
This results in even-sampling of the cladisto-space across time and permits us to define the underlying model of character evolution (punctuated or gradual).  
In practice, time-slicing considers the disparity of any element present in the phylogeny (branches, nodes and tips) at any point in time.
When the phylogenetic elements are nodes or tips, the eigenvector scores for the nodes (estimated using ancestral state reconstruction as described in section \ref{ace}) or tips are directly used for estimating disparity.
When the phylogenetic elements are branches we chose the eigenvector score for the branch using one of two evolutionary models:
\begin{enumerate}
    \item{\textbf{Punctuated evolution.}} 
    This model selects the eigenvector score from either the ancestral node or the descendant node/tip of the branch regardless of the position of the slice along the branch. 
    Similarly to the time interval approach, this reflects a model of punctuated evolution where changes in disparity occur either at the start or at the end of a branch over a relatively short time period and clades undergo long periods of stasis during their evolution \citep{Gould1977,Hunt20112007}.
    We applied this model in three ways: 
    \begin{enumerate}[(i)]
      \item selecting the eigenvector score of the ancestral node of the branch (ACCTRAN).
      \item selecting the eigenvector score of the descendant node/tip of the branch (DELTRAN).
      \item randomly selecting either the eigenvector score of the ancestral node or the descendant node/tip of the branch (random).
    \end{enumerate}
    Method (i) assumes that changes always occur early on the branch (accelerated transition, ACCTRAN) and (ii) assumes that changes always occur later (delayed transition, DELTRAN).
    We prefer not to make either assumption so we report the results from (iii), although the ACCTRAN and DELTRAN results are available in the appendix \ref{chap:Appendix_STD} (Fig. \ref{Supp_disparity_all_Mammaliaformes}, \ref{Supp_disparity_all_Eutheria}, \ref{Supp_disparity_all_Mammaliaformes_rarefied}, \ref{Supp_disparity_all_Eutheria_rarefied}).
    \item{\textbf{Gradual evolution.}}
    This model also selects the eigenvector score from either the ancestral node or the descendant node/tip of the branch, but the choice depends on the distance between the sampling time point and the end of the branch.
    If the sampling time point falls in the first half of the branch length the eigenvector score is taken from the ancestral node, conversely, if the sampling time point falls in the second half of the branch length the eigenvector score is taken from the descendant node/tip.
    This reflects a model of gradual evolution where changes in disparity are gradual and cumulative along the branch.
    Under this model, the gradual changes could be either directional or random, however, directional evolution has been empirically shown to be rare \citep[only 5\% of the time][]{Hunt20112007}.
    We therefore considered that changes from a character state A to B were only dependent on the branch length.
\end{enumerate}

We applied our time-slicing approach separately to each cladisto-space, time-slicing the phylogeny every five million years.
% Do we want to vary the time slice timings too?
This resulted in XX subsamples for dataset 1, XX for dataset 2 and XX for dataset 3.
For each subsample, we estimated its disparity assuming punctuated (ACCTRAN, DELTRAN and random) and gradual evolution as described above.
To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement by randomly resampling with replacement a new subsample of taxa from the observed taxa in the subsample 1000 times.
We then calculated the median disparity value for each subsample along with the 50\% and 95\% confidence intervals.
We also recorded the number of phylogenetic elements (nodes and tips) in each subsample as a proxy for taxonomic diversity.

\subsection{Comparing the results}
%Testing the difference section

\subsection{Rarefaction}

Finally, disparity may be higher in subsamples with more phylogenetic elements simply because there are more taxa represented.
To test whether this influenced our results, we repeated the analyses using rarefied disparities.
In dataset 1, the minimum number of taxa in each stratigraphy subsample was eight, time bin XX and time slice was XX.

\section{Results}

%Fig. or Figs
%Fig width 80mm single column
%110mm - 2/3 width
%166mm full page

%PCA figure
	%\begin{figure}[!htbp]
	%\centering
	%\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/PCA_threeplot.png}
	%\caption[Morphospace (principal components) plot of morphological diversity in tenrec and golden mole skulls.]
	%	{Principal components plots of the morphospaces occupied by tenrecs (triangles, n=31 species) and golden moles (circles, n=12 species) for skulls in dorsal (top left), ventral (top right) and lateral (bottom left) views. Each point represents the average skull shape of an individual species. Axes are principal component 1 (PC1) and principal component 2 (PC2) of the average scores from principal components analyses of mean Procrustes shape coordinates for each species.}
	%\label{fig:threePCA}
	%\end{figure}

% Results table
	%\begin{table}[!htbp]			
		%\caption[Comparing morphological diversity in tenrecs and golden moles.]
		%{Morphological diversity in tenrecs compared to golden moles (12 species). N is the number of tenrec species: 31 species or 17 species including just five representatives of the \textit{Microgale} Genus. Morphological diversity of the Family is the mean Euclidean distance from each species to the Family centroid. Significant differences between the two Families (p$<$0.05) from two-tailed t-tests are highlighted in bold.}
		%\input{tables/diversity.results} 
		%\label{tab:diversity}  
	%\end{table}

\section{Discussion}

\section{Conclusions}

\section{Acknowledgments}
	We thank all the PalAss folk

%\bibliographystyle{sysbio} 
%\bibliography{time-refs}

\end{document}

### 2017-09-05 functions/data.load_fun.R

@@ -1,0 +1,65 @@
+#' @title Load FADLAD
+#'
+#' @description Loading and formatting the First and Last Apparition Data
+#'
+#' @param chain the chain name
+#' @param header logical, whether there is a header to the file
+#' @param row.names numeric, which row names to use
+#' 
+#' @author Thomas Guillerme
+
+load.FADLAD <- function(chain, header = TRUE, row.names = 1) {
+    ## Set up the path
+    current_wd <- strsplit(getwd(), split = "time-slice")[[1]]
+    current <- current_wd[2]
+    if(is.na(current)) {
+        stop("Wrong directory root (should be '.../time-slice')")
+    } else {
+        path <- paste0(current_wd[1], "time-slice/")
+    }
+
+    ## Get the right file name
+    file <- list.files(path = paste0(path, "Data/FADLAD/"), pattern = chain)
+    if(length(file) == 0) {
+        stop(paste("No FADLAD file found for", chain, "in", path))
+    }
+
+    ## Read the file
+    FADLAD_table <- read.csv(paste0(path, "Data/FADLAD/", file), header = header, row.names = row.names)
+
+    ## Check if the FADLAD name is correct
+    if(class(rownames(FADLAD_table)) != "character") {
+        warning("Rownames of the FADLAD object don't have the right class (should be character).")
+    }
+
+    if(colnames(FADLAD_table)[1] != "FAD" && colnames(FADLAD_table)[1] != "LAD") {
+        warning("The two first column names are in the wrong format (should be 'FAD' and 'LAD').")
+    }
+
+    return(FADLAD_table)
+}
+
+#' @title Load space
+#'
+#' @description Loading and formatting the multidimensional space
+#'
+#' @param chain the chain name
+#' 
+#' @author Thomas Guillerme
+
+load.space <- function(chain) {
+    return()
+}
+
+#' @title Load tree
+#'
+#' @description Loading and formatting the tree
+#'
+#' @param chain the chain name
+#' 
+#' @author Thomas Guillerme
+
+load.tree <- function(chain) {
+    return()
+}
+

### 2017-09-05 analyses/01-extract-data-for-analyses.Rmd
Fix name to remove space

---
title: "Extracting data for time slice analyses"
author: "Natalie Cooper and Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 6
    fig_height: 6
---

Quick pipeline for getting data from discrete matrices.
===============

This code loads the data and extracts matrices and other files needed to run the analyses in the paper. Note that these analyses can take time, so we save the outputs to a `data/processed` folder and use these in later analyses so that the analyses can be replicated more quickly. Below we lay out the process in detail for one dataset, and then provide the code for the other datasets below this.

```{r}
set.seed(123)
```

Load packages
-----------------------------------

Make sure you work with `dispRity` 0.4 and TG's version of `Claddis` (much faster than GL's).

```{r, eval = TRUE}
# The latest version of devtools and ape
library(devtools)
library(ape)

## The latest version of Claddis (on development on my own branch)
install_github("TGuillerme/Claddis")
library(Claddis)

## The latest version of dispRity 
install_github("TGuillerme/dispRity", ref = "release")
library(dispRity) 
```

Loading the data
-----------------------------------

This is based on a slug (here "Beck2014") and should work with different slugs, assuming the files are in the exact same format (by slug here we are referring to the fact that each of our files have the name of the dataset within them, so we can easily identify each one using regular expression matching).

The analyses require a NEXUS matrix of characters, a csv file of first and last occurrence dates (FADLAD) and a phylogeny.

```{r}
## Setting up the slug name
slug <- "Beck2014"

## Loading the discrete morphological matrix
matrix <- ReadMorphNexus(paste0("data/matrices/", slug, ".nex"))

## Loading the FADLAD
FADLAD <- read.csv(paste0("data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)

## Loading the tree
tree <- read.nexus(paste0("data/trees/", slug, ".tre"))

## Adding node labels
tree <- makeNodeLabel(tree, method = "number", prefix = "n")

## Adding a root age (max tree age date)
tree$root.time <- max(tree.age(tree)[, 1])
```

We'll also clean the data so that the species match in the tree and the matrix:

```{r}
## Cleaning the data 
cleaned_data <- clean.data(matrix$matrix, tree)

## Reassign the cleaned matrix and tree
matrix$matrix <- cleaned_data$data
tree <- cleaned_data$tree

## Removed taxa, i.e. those in the tree but not the matrix and vice versa
as.vector(na.omit(c(cleaned_data$dropped_tips, cleaned_data$dropped_rows)))
```

Ancestral states reconstruction
-----------------------------------

Next we need to recreate the nodal states but excluding any ancestral state estimation with less than a 0.95 scaled Likelihood value. This allows us to discard any dodgy/ambiguous reconstructions (note that NAs are not a problem since we use Gower's distances later on).

This takes some time to run (could be a few hours)! Therefore this is set to `eval = FALSE`. The results are available in `data/processed`, or you can replace `FALSE` with `TRUE` to run this yourself.

```{r, eval = FALSE}
## Estimating the ancestral states (with a uncertainty threshold of 0.95)
matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
                                  uncertainty.threshold = 0.95)

## Saving the matrix
save(matrix_nodes, file = paste0("data/processed/ancestral_states.", slug ,".Rda"))
```

After this we can add the node states to the overall matrix:

```{r}
## Loading the matrix
load(paste0("data/processed/ancestral_states.", slug ,".Rda"))

## Adding the node names as row names
row.names(matrix_nodes) <- tree$node.label

## Combining the tips and the nodes in the same matrix
matrix$matrix <- rbind(matrix$matrix, matrix_nodes) 
```

Constructing the distance matrix
-----------------------------------

Again this takes some time to run (a couple of minutes). Therefore this is set to `eval = FALSE`. The results are available in `data/processed`, or you can replace `FALSE` with `TRUE` to run this yourself.

```{r, eval = FALSE}
## Calculating the Gower distance
matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")

## Saving the matrix
save(matrix_dist, file = paste0("data/processed/distance_matrix.", slug ,".Rda"))
```

Ordinating the matrix
----------------

Ordination is a classic NMDS (multidimensional scaling) including only the non-zero variance axes.

```{r}
## Loading the distance matrices
load(paste0("data/processed/distance_matrix.", slug ,".Rda"))

## Ordinating the matrix
morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
```

And that's it... we now have all the arguments for running the `dispRity` package properly: the space (`morphospace`), the tree (`tree`) and the FADLAD (`FADLAD`). Remember that for the `time.subsamples` function, you can use the `inc.nodes` logical option to include nodes or not!

Running these extractions for other datasets
===================================================

```{r}
## Setting up the slug names
slugs <- c("Bapst2016", "Hopkins", "Other")

# Big old loop (too lazy to make functions sorry!)
for(slug in 1: length(slugs)) {

## Loading the data
matrix <- ReadMorphNexus(paste0("data/matrices/", slugs[slug], ".nex"))
FADLAD <- read.csv(paste0("data/FADLAD/", slugs[slug], ".csv"), row.names = 1, header = TRUE)
tree <- read.nexus(paste0("data/trees/", slugs[slug], ".tre"))

# Tidy the tree and the data
tree <- makeNodeLabel(tree, method = "number", prefix = "n")
tree$root.time <- max(tree.age(tree)[, 1])
cleaned_data <- clean.data(matrix$matrix, tree)
matrix$matrix <- cleaned_data$data
tree <- cleaned_data$tree

## Estimating the ancestral states (with a uncertainty threshold of 0.95)
matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
                                  uncertainty.threshold = 0.95)
save(matrix_nodes, file = paste0("data/processed/ancestral_states.", slugs[slug] ,".Rda"))

## Add ancestral states to nodes
load(paste0("data/processed/ancestral_states.", slugs[slug] ,".Rda"))
row.names(matrix_nodes) <- tree$node.label
matrix$matrix <- rbind(matrix$matrix, matrix_nodes)

## Constructing the distance matrix
matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
save(matrix_dist, file = paste0("data/processed/distance_matrix.", slugs[slug] ,".Rda"))

## Ordinating the matrix - no need to run this here, we will do this in the next script
## load(paste0("data/processed/distance_matrix.", slugs[slug] ,".Rda"))
## morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points

}
```

### 2017-09-05 analyses/02-time-slice-analyses.Rmd
Move to analyses folder

---
title: "Time slice analyses"
author: "Natalie Cooper and Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 6
    fig_height: 6
---

TITLE?
===============

DETAILS
This code loads the data and extracts matrices and other files needed to run the analyses in the paper. Note that these analyses can take time, so we save the outputs to a `data/backups` folder and use these in other `.Rmd` files so that the analyses can be replicated more quickly. Below we lay out the process in detail for one dataset, and then provide the code for the other datasets below this.

```{r}
set.seed(123)
```

Load packages
-----------------------------------

Make sure you work with `dispRity` 0.4 and TG's version of `Claddis` (much faster than GL's).

```{r, eval = TRUE}
# The latest version of devtools and ape
library(devtools)
library(ape)

## The latest version of Claddis (on development on my own branch)
install_github("TGuillerme/Claddis")
library(Claddis)

## The latest version of dispRity 
install_github("TGuillerme/dispRity", ref = "release")
library(dispRity) 
```

Loading the data
-----------------------------------

To run `dispRity` analyses we need the **morphospace**, the **phylogeny** (with a root age and node labels) and the **first and last occurrence dates** (FADLAD). The tree and FADLAD are in the `data/` folder. To construct the morphospace we need to use the `distance_matrix` files we created in `01-extract-data-for analyses.Rmd` and stored in `data/processed`.

```{r}
## Choose a slug
slug <- "Beck2014"

## Loading the discrete morphological matrix
## This is just to match the names with the tree
matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))

## Loading the FADLAD
FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)

## Loading the tree
tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
## Adding node labels and a root age (max tree age date)
tree <- makeNodeLabel(tree, method = "number", prefix = "n")
tree$root.time <- max(tree.age(tree)[, 1])

## Cleaning the tree to remove things that aren't in the matrix 
cleaned_data <- clean.data(matrix$matrix, tree)
tree <- cleaned_data$tree

## Loading the distance matrices
## See "01-extract-data-for analyses.Rmd" for how we made these
load(paste0("data/processed/distance_matrix.", slug ,".Rda"))

## Ordinating the matrix
morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points

```

time slicing
-----------------------------------

### 2017-09-05 analyses/01-extract-data-for-analyses.Rmd
Paths corrected again

@@
 ## Loading the discrete morphological matrix
-matrix <- ReadMorphNexus(paste0("data/matrices/", slug, ".nex"))
+matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
 
 ## Loading the FADLAD
-FADLAD <- read.csv(paste0("data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
 
 ## Loading the tree
-tree <- read.nexus(paste0("data/trees/", slug, ".tre"))
+tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
 
@@
 ## Saving the matrix
-save(matrix_nodes, file = paste0("data/processed/ancestral_states.", slug ,".Rda"))
+save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 ```
@@
 ## Loading the matrix
-load(paste0("data/processed/ancestral_states.", slug ,".Rda"))
+load(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 
@@
 ## Saving the matrix
-save(matrix_dist, file = paste0("data/processed/distance_matrix.", slug ,".Rda"))
+save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
 ```
@@
 ## Loading the distance matrices
-load(paste0("data/processed/distance_matrix.", slug ,".Rda"))
+load(paste0("../data/processed/distance_matrix.", slug ,".Rda"))
 
@@
 ## Loading the data
-matrix <- ReadMorphNexus(paste0("data/matrices/", slugs[slug], ".nex"))
-FADLAD <- read.csv(paste0("data/FADLAD/", slugs[slug], ".csv"), row.names = 1, header = TRUE)
-tree <- read.nexus(paste0("data/trees/", slugs[slug], ".tre"))
+matrix <- ReadMorphNexus(paste0("../data/matrices/", slugs[slug], ".nex"))
+FADLAD <- read.csv(paste0("../data/FADLAD/", slugs[slug], ".csv"), row.names = 1, header = TRUE)
+tree <- read.nexus(paste0("../data/trees/", slugs[slug], ".tre"))
 
@@
                                   uncertainty.threshold = 0.95)
-save(matrix_nodes, file = paste0("data/processed/ancestral_states.", slugs[slug] ,".Rda"))
+save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slugs[slug] ,".Rda"))
 
 ## Add ancestral states to nodes
-load(paste0("data/processed/ancestral_states.", slugs[slug] ,".Rda"))
+load(paste0("../data/processed/ancestral_states.", slugs[slug] ,".Rda"))
 row.names(matrix_nodes) <- tree$node.label
@@
 matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
-save(matrix_dist, file = paste0("data/processed/distance_matrix.", slugs[slug] ,".Rda"))
+save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slugs[slug] ,".Rda"))
 
 ## Ordinating the matrix - no need to run this here, we will do this in the next script
-## load(paste0("data/processed/distance_matrix.", slugs[slug] ,".Rda"))
+## load(paste0("../data/processed/distance_matrix.", slugs[slug] ,".Rda"))
 ## morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points

### 2017-09-05 analyses/02-time-slice-analyses.Rmd
Add notes for disparity phase

@@
 cleaned_data <- clean.data(matrix$matrix, tree)
+matrix$matrix <- cleaned_data$matrix
 tree <- cleaned_data$tree
@@
 ## See "01-extract-data-for analyses.Rmd" for how we made these
-load(paste0("data/processed/distance_matrix.", slug ,".Rda"))
+load(paste0("../data/processed/distance_matrix.", slug ,".Rda"))
 
@@
 
-time slicing
+Stratigraphy
 -----------------------------------
+
+How to go about that????
+Need to collate data ourselves or can we use strap?
+
+Time bins (equal sized)
+-----------------------------------
+
+What do we want to vary? Metrics? Number of bins?
+
+Note we might want to exclude node but current won't work due to mismatch - ooh maybe use clean.data to fix??? Yes that works
+
+no.nodes <- clean.data(morphospace, tree)
+m<- no.nodes$data
+
+```{r, eval = FALSE}
+time.bin.sim <- function(times, metric) {
+
+  sub <- time.subsamples(data = morphospace, tree = tree, method = "discrete",
+                time = times, FADLAD = FADLAD, inc.nodes = TRUE)
+
+  disparity <- summary(dispRity(sub, metric = metric),
+                      round = 10)
+}
+```
+
+What outputs do we want?
+
+Time slicing 
+-----------------------------------
+
+What do we want to vary? Metrics? Number of slices? Model of evolution
+
+Note we might want to exclude node but current won't work due to mismatch - ooh maybe use clean.data to fix???
+
+```{r, eval = FALSE}
+time.slice.sim <- function(times, metric) {
+
+  sub <- time.subsamples(data = morphospace, tree = tree, 
+    method = "continuous", model = "gradual", time = times,
+    FADLAD = FADLAD, inc.nodes = TRUE)
+
+  disparity <- summary(dispRity(sub, metric = metric),
+                      round = 10)
+}
+```

### 2017-09-07 analyses/02-time-slice-analyses.Rmd
Small progress toward getting the outputs?

--- 2017-09-05 analyses/02-time-slice-analyses.Rmd
@@
 
-time slicing
+Stratigraphy
 -----------------------------------
+
+How to go about that????
+Need to collate data ourselves or can we use strap?
+
+```{r, eval = FALSE}
+
+## Run the function for Beck data
+
+slug <- "BeckLee"
+metric <- c(sum, variances)
+metric.name <- "sumvar"
+
+for (bins in seq_along(c(5,10,15,20))){
+ run.all.sims(morphospace, tree, FADLAD, bins, metric, 
+              metric.name, inc.nodes = TRUE, slug)
+}
+
+slug <- "BeckLee"
+metric <- c(median, centroids)
+metric.name <- "medcentroid"
+
+for (bins in seq_along(c(5,10,15,20))){
+ run.all.sims(morphospace, tree, FADLAD, bins, metric, 
+              metric.name, inc.nodes = TRUE, slug)
+}
+
+```

### 2017-09-07 functions/simulations_fun.R
Writing shitty functions for simulations

@@ -1,0 +1,136 @@
+# Functions to run the simulations and extract data
+#---------------------------------------------------
+
+#---------------------------------------------------
+# 1. STRATIGRAPHY BINS
+#---------------------------------------------------
+
+
+#---------------------------------------------------
+# 2. TIME BINS
+# Allows you to vary the metric, bins and whether
+# to include nodes
+# Outputs to a slug specific text file
+# Bootstraps and rarefies
+#---------------------------------------------------
+time.bin.sims <- function(morphospace, tree, FADLAD, bins, metric, 
+                          metric.name, inc.nodes = TRUE, slug) {
+  
+  # Need to remove the nodes if not using them
+  if(inc.nodes == FALSE){
+    remove.nodes <- clean.data(morphospace, tree)
+    morphospace <- remove.nodes$data  
+  }
+  
+  # Subset samples by required number of time bins
+  subsamples <- time.subsamples(data = morphospace, tree = tree, method = "discrete",
+                                time = bins, FADLAD = FADLAD, inc.nodes = inc.nodes)
+  
+  # Obtain a bootstrap matrix for the subsamples
+  bootstrapped_data <- boot.matrix(subsamples, bootstraps = 1, rarefaction = FALSE)
+  
+  # Estimate disparity - observed and bootstrapped
+  disparity <- summary(dispRity(bootstrapped_data, metric = metric), round = 10)
+  
+  # Obtain a rarified bootstrap matrix for the subsamples
+  rarefied_data <- boot.matrix(subsamples, bootstraps = 1, rarefaction = TRUE)
+  
+  # Estimate disparity - observed and bootstrapped with rarefaction
+  disparity_rarefied <- summary(dispRity(rarefied_data, metric = metric), round = 10)
+  
+  # Outputs
+  output <- as.data.frame(disparity)
+  output$bins <- bins
+  output$metric <- metric.name
+  output$inc.nodes <- inc.nodes
+  readr::write_csv(output, path = paste0("../outputs/timebins.", slug ,".csv"), 
+                   append = TRUE)
+  
+  output2 <- as.data.frame(disparity_rarefied)
+  output2$bins <- bins
+  output2$metric <- metric.name
+  output2$inc.nodes <- inc.nodes
+  readr::write_csv(output2, path = paste0("../outputs/timebins_rarefied.", slug ,".csv"), 
+                   append = TRUE)
+    
+}
+
+#---------------------------------------------------
+# 3. TIME SLICES
+# Allows you to vary the metric, slices and whether
[…]

### 2017-09-07 functions/simulations_fun.R
Add comments on rarefaction

@@ -32,6 +32,7 @@ time.bin.sims <- function(morphospace, tree, FADLAD, bins, metric,
   # Estimate disparity - observed and bootstrapped
   disparity <- summary(dispRity(bootstrapped_data, metric = metric), round = 10)
   
+  # Rarefaction appears to be broken...
   # Obtain a rarified bootstrap matrix for the subsamples
   rarefied_data <- boot.matrix(subsamples, bootstraps = 1, rarefaction = TRUE)
   

### 2017-09-15 analyses/02-time-slice-analyses.Rmd
Updates on the time bins stuff

@@
 ## The latest version of Claddis (on development on my own branch)
-install_github("TGuillerme/Claddis")
+if(!require(Claddis)) install_github("TGuillerme/Claddis")
 library(Claddis)
@@
 ## The latest version of dispRity 
-install_github("TGuillerme/dispRity", ref = "release")
+if(!require(dispRity)) install_github("TGuillerme/dispRity", ref = "master")
 library(dispRity) 
 ```
+<!-- TG: I've switched to master branch for the new functions -->
 
@@
 Need to collate data ourselves or can we use strap?
+<!-- TG: got all that covered with the new get.bin.ages function! -->
+
+```{r}
+## Get the stratigraphic ages
+time_ages <- get.bin.ages(tree, type = "Age")
+
+## Get the stratigraphic ages
+time_epoch <- get.bin.ages(tree, type = "Epoch")
+
+## Get the median age duration (to create equal time bins)
+age_duration <- median(abs(diff(time_ages)))
+
+## Get the number of ages (to create equal time bins)
+age_length <- length(time_ages)
+
+## Get the median epoch duration (to create equal time bins)
+epoch_duration <- median(abs(diff(time_epoch)))
+
+## Get the number of epochs (to create equal time bins)
+epoch_length <- length(time_epoch)
+```
+
+With this data, I suggest we can explore these different scenarios:
+
+Name | Type | Scenario | Based on
+------------|----------|----------
+un_bin_age  | unequal bins | stratigraphic ages | Ages durations
+un_bin_epo  | unequal bins | stratigraphic epochs | Epoch durations
+eq_bin_age_med  | equal bins   | median age bins | Median age duration
+eq_bin_epo_med  | equal bins   | median epoch bins | Median epoch duration
+eq_bin_age_len  | equal bins   | number of age bins | same number of ages but equal duration
+eq_bin_epo_len  | equal bins   | number epoch bins | same number of epoch but equal duration
+slice_age_med  | time slicing | median age slices | median age duration (+ our 4 "models" of evolution)
+slice_epo_med  | time slicing | median epoch slices| median epoch duration (+ our 4 "models" of evolution)
+slice_age_len  | time slicing | number of age slices | same number of ages but slices (+ our 4 "models" of evolution)
+slice_epo_len  | time slicing | number of epoch slices | same number of epoch but slices (+ our 4 "models" of evolution)
+
+What do you think?
 
@@
 ```
+
+<!-- TG: Suggestion for a "testing the effect of binning/slicing" pipeline -->
+
+```{r, eval = FALSE}
+
+## Run the function for Beck data
+
+slug <- "BeckLee"
+metric <- c(sum, variances)
+metric.name <- "sumvar"
+
+for (bins in seq_along(c(5,10,15,20))){
+ run.all.sims(morphospace, tree, FADLAD, bins, metric, 
+              metric.name, inc.nodes = TRUE, slug)
+}
+
+slug <- "BeckLee"
+metric <- c(median, centroids)
+metric.name <- "medcentroid"
+
+for (bins in seq_along(c(5,10,15,20))){
+ run.all.sims(morphospace, tree, FADLAD, bins, metric, 
+              metric.name, inc.nodes = TRUE, slug)
+}
+
+```

### 2017-09-19 analyses/02-time-slice-analyses.Rmd
Comment out installs

@@
 
-## The latest version of Claddis (on development on my own branch)
-if(!require(Claddis)) install_github("TGuillerme/Claddis")
+## The latest version of Claddis (on development on TG branch)
+## install_github("TGuillerme/Claddis")
 library(Claddis)
 
-## The latest version of dispRity 
-if(!require(dispRity)) install_github("TGuillerme/dispRity", ref = "master")
+## The latest version of dispRity on master branch
+## install_github("TGuillerme/dispRity", ref = "master")
 library(dispRity) 
 ```
-<!-- TG: I've switched to master branch for the new functions -->
 

### 2017-09-19 functions/sim_fun.R
Initial commit of tidying attempt for existing simualtions functions

@@ -1,0 +1,61 @@
+## Functions
+
+# Output data
+
+
+## Extract outputs needed for results
+get.outputs <- function(disparity.object, bins, metric.name, inc.nodes){
+  output <- as.data.frame(disparity.object)
+  output$bins <- bins
+  output$metric <- metric.name
+  output$inc.nodes <- inc.nodes
+  return(output)
+ }
+
+## Save outputs to correct folder and file
+save.outputs <- function(output, path.to.folder, slug){
+  readr::write_csv(output, path = paste0(path.to.folder, "_", slug, ".csv"), 
+                   append = TRUE)
+}
+
+## Combine output extraction and saving
+output.time.results <- function(disparity.object, bins, metric.name, inc.nodes, 
+	                            path.to.folder, slug){
+  output <- get.outputs(disparity.object, bins, metric.name, inc.nodes)
+  save.outputs(output, )
+
+
+
+
+
+
+
+
+
+
+
+
+# Bootstrap and rarefy
+
+boot.rarefy.data <- function(subsamples, bootstraps, metric){
+  # Obtain a bootstrap matrix for the subsamples
+  bootstrapped_data <- boot.matrix(subsamples, bootstraps = bootstraps, rarefaction = FALSE)
+  # Estimate disparity - observed and bootstrapped
+  disparity <- summary(dispRity(bootstrapped_data, metric = metric), round = 10)
+  
+  # Obtain a rarified bootstrap matrix for the subsamples
+  rarefied_data <- boot.matrix(subsamples, bootstraps = bootstraps, rarefaction = TRUE)
+  # Estimate disparity - observed and bootstrapped with rarefaction
+  disparity_rarefied <- summary(dispRity(rarefied_data, metric = metric), round = 10)
+  
+  # Return both as list
+  return(list(disparity, disparity_rarefied))
+}
+
+
+  
+
+
+
+
+

### 2017-09-19 functions/sim_fun.R
Complete output functions

@@ -1,7 +1,8 @@
 ## Functions
 
-# Output data
-
+##------------------------------------------
+## Output disparity results from simulations
+##------------------------------------------
 
 ## Extract outputs needed for results
 get.outputs <- function(disparity.object, bins, metric.name, inc.nodes){
@@ -19,11 +20,11 @@ save.outputs <- function(output, path.to.folder, slug){
 }
 
 ## Combine output extraction and saving
-output.time.results <- function(disparity.object, bins, metric.name, inc.nodes, 
+output.results <- function(disparity.object, bins, metric.name, inc.nodes, 
 	                            path.to.folder, slug){
   output <- get.outputs(disparity.object, bins, metric.name, inc.nodes)
-  save.outputs(output, )
-
+  save.outputs(output, path.to.folder, slug)
+}
 
 
 

### 2017-09-19 functions/sim_fun.R
Finish boot and rarefy all subsamples functions

@@ -26,36 +26,76 @@ output.results <- function(disparity.object, bins, metric.name, inc.nodes,
   save.outputs(output, path.to.folder, slug)
 }
 
+##-------------------------------------------
+## Creating subsamples for disparity analyses
+##-------------------------------------------
 
+## Extract all types of subsample for both methods and four continuous models
+get.subsamples <- function(morphospace, tree, bins, FADLAD, inc.nodes){
 
+  # Subset samples by required number of time bins
+  subsamples.bins <- time.subsamples(data = morphospace, tree = tree, 
+  	                                 method = "discrete", time = bins, 
+  	                                 FADLAD = FADLAD, inc.nodes = inc.nodes)
+  
+  # Subset samples by required number of time slices with acctran model
+  subsamples.acctran <- time.subsamples(data = morphospace, tree = tree, 
+  	                                    method = "continuous", time = bins, 
+  	                                    FADLAD = FADLAD, inc.nodes = inc.nodes,
+                                        model = "acctran")
 
+  # Subset samples by required number of time slices with deltran model
+  subsamples.deltran <- time.subsamples(data = morphospace, tree = tree, 
+  	                                    method = "continuous", time = bins, 
+  	                                    FADLAD = FADLAD, inc.nodes = inc.nodes,
+                                        model = "deltran")
+  
+  # Subset samples by required number of time slices with punctuated model
+  subsamples.punctuated <- time.subsamples(data = morphospace, tree = tree, 
+  	                                       method = "continuous", time = bins, 
+  	                                       FADLAD = FADLAD, inc.nodes = inc.nodes,
+                                           model = "punctuated")
+  
+  # Subset samples by required number of time slices with gradual model
+  subsamples.gradual <- time.subsamples(data = morphospace, tree = tree, 
+  	                                    method = "continuous", time = bins, 
+  	                                    FADLAD = FADLAD, inc.nodes = inc.nodes,
+  	                                    model = "gradual")
 
+  return(list(subsamples.bins, subsamples.acctran, subsamples.deltran, 
+  	          subsamples.punctuated, subsamples.gradual))
 
+}
 
-
-
-
-
-# Bootstrap and rarefy
-
[…]

### 2017-09-19 functions/sim_fun.R
Move putput functions down

@@ -1,30 +1,5 @@
 ## Functions
 
-##------------------------------------------
-## Output disparity results from simulations
-##------------------------------------------
-
-## Extract outputs needed for results
-get.outputs <- function(disparity.object, bins, metric.name, inc.nodes){
-  output <- as.data.frame(disparity.object)
-  output$bins <- bins
-  output$metric <- metric.name
-  output$inc.nodes <- inc.nodes
-  return(output)
- }
-
-## Save outputs to correct folder and file
-save.outputs <- function(output, path.to.folder, slug){
-  readr::write_csv(output, path = paste0(path.to.folder, "_", slug, ".csv"), 
-                   append = TRUE)
-}
-
-## Combine output extraction and saving
-output.results <- function(disparity.object, bins, metric.name, inc.nodes, 
-	                            path.to.folder, slug){
-  output <- get.outputs(disparity.object, bins, metric.name, inc.nodes)
-  save.outputs(output, path.to.folder, slug)
-}
 
 ##-------------------------------------------
 ## Creating subsamples for disparity analyses
@@ -68,6 +43,7 @@ get.subsamples <- function(morphospace, tree, bins, FADLAD, inc.nodes){
 }
 
 ## Bootstrap and rarefy subsamples
+## subsamples is output from get.subsamples as a list
 boot.rarefy <- function(subsamples, bootstraps){
   # Obtain a bootstrap matrix for the subsamples
   bootstrapped_data <- boot.matrix(subsamples, bootstraps = bootstraps, rarefaction = FALSE)
@@ -83,10 +59,11 @@ boot.rarefy.all <- function(subsamples, bootstraps){
   purrr::map(subsamples, boot.rarefy, bootstraps = bootstraps)
 }
 
+##-------------------------------------------
+## Extract disparity for all subsamples
+##-------------------------------------------
 
-## Disparity for all
-
-    # Estimate disparity - observed and bootstrapped
+  # Estimate disparity - observed and bootstrapped
   disparity <- summary(dispRity(bootstrapped_data, metric = metric), round = 10)
   
 
@@ -99,4 +76,28 @@ boot.rarefy.all <- function(subsamples, bootstraps){
 
 
 
+##------------------------------------------
+## Output disparity results from simulations
+##------------------------------------------
 
+## Extract outputs needed for results
+get.outputs <- function(disparity.object, bins, metric.name, inc.nodes){
+  output <- as.data.frame(disparity.object)
[…]

### 2017-09-19 functions/sim_fun.R
Complete disparity list extraction functions

@@ -8,30 +8,30 @@
 ## Extract all types of subsample for both methods and four continuous models
 get.subsamples <- function(morphospace, tree, bins, FADLAD, inc.nodes){
 
-  # Subset samples by required number of time bins
+  ## Subset samples by required number of time bins
   subsamples.bins <- time.subsamples(data = morphospace, tree = tree, 
   	                                 method = "discrete", time = bins, 
   	                                 FADLAD = FADLAD, inc.nodes = inc.nodes)
   
-  # Subset samples by required number of time slices with acctran model
+  ## Subset samples by required number of time slices with acctran model
   subsamples.acctran <- time.subsamples(data = morphospace, tree = tree, 
   	                                    method = "continuous", time = bins, 
   	                                    FADLAD = FADLAD, inc.nodes = inc.nodes,
                                         model = "acctran")
 
-  # Subset samples by required number of time slices with deltran model
+  ## Subset samples by required number of time slices with deltran model
   subsamples.deltran <- time.subsamples(data = morphospace, tree = tree, 
   	                                    method = "continuous", time = bins, 
   	                                    FADLAD = FADLAD, inc.nodes = inc.nodes,
                                         model = "deltran")
   
-  # Subset samples by required number of time slices with punctuated model
+  ## Subset samples by required number of time slices with punctuated model
   subsamples.punctuated <- time.subsamples(data = morphospace, tree = tree, 
   	                                       method = "continuous", time = bins, 
   	                                       FADLAD = FADLAD, inc.nodes = inc.nodes,
                                            model = "punctuated")
   
-  # Subset samples by required number of time slices with gradual model
+  ## Subset samples by required number of time slices with gradual model
   subsamples.gradual <- time.subsamples(data = morphospace, tree = tree, 
   	                                    method = "continuous", time = bins, 
   	                                    FADLAD = FADLAD, inc.nodes = inc.nodes,
@@ -63,18 +63,17 @@ boot.rarefy.all <- function(subsamples, bootstraps){
 ## Extract disparity for all subsamples
 ##-------------------------------------------
 
-  # Estimate disparity - observed and bootstrapped
-  disparity <- summary(dispRity(bootstrapped_data, metric = metric), round = 10)
-  
-
-  # Estimate disparity - observed and bootstrapped with rarefaction
-  disparity_rarefied <- summary(dispRity(rarefied_data, metric = metric), round = 10)
[…]

### 2017-09-19 functions/sim_fun.R
Add full function - still some issues

@@ -82,7 +82,7 @@ get.disparity.all <- function(subsampled_data, metric){
 ## Extract outputs needed for results
 get.outputs <- function(disparity.object, bins, metric.name, inc.nodes){
   output <- as.data.frame(disparity.object)
-  output$bins <- bins
+  output$bins <- paste(as.character(bins), collapse = "")
   output$metric <- metric.name
   output$inc.nodes <- inc.nodes
   return(output)
@@ -99,4 +99,44 @@ output.results <- function(disparity.object, bins, metric.name, inc.nodes,
 	                            path.to.folder, slug){
   output <- get.outputs(disparity.object, bins, metric.name, inc.nodes)
   save.outputs(output, path.to.folder, slug)
+}
+
+## Apply to list
+output.all.results <- function(disparity.object, bins, metric.name, inc.nodes, 
+	                            path.to.folder, slug){
+
+  purrr::map(disparity.object, output.results, bins = bins, metric.name = metric.name, 
+  	         inc.nodes = inc.nodes, path.to.folder = path.to.folder, slug = slug)
+}
+
+##### Still not outputting to file
+#### Also need some way to name lists/extract models to put into the output somehow???
+
+##------------------------------------
+## Putting it all together
+##------------------------------------
+#path.to.folder = ../outputs/timebins_rarefied
+
+run.all.disparity <- function(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
+                              bootstraps = 1, metric, metric.name, path.to.folder, slug){
+
+  # Need to remove the nodes if not using them
+  if(inc.nodes == FALSE){
+    remove.nodes <- clean.data(morphospace, tree)
+    morphospace <- remove.nodes$data  
+  }
+
+  ## Extract all types of subsample for both methods and four continuous models
+  subsamples <- get.subsamples(morphospace, tree, bins, FADLAD, inc.nodes)
+
+  ## Bootstrap and rarefy all types of subsamples
+  subsampled_data <- boot.rarefy.all(subsamples, bootstraps)
+
+  ## Get disparity for all types of subsamples
+  disparity.object <- get.disparity.all(subsampled_data, metric)
+
+  ## Outputs
+  output.results(disparity.object, bins, metric.name, inc.nodes, 
+	             path.to.folder, slug)
+
 }

### 2017-09-20 functions/sim_fun.R
Remove rarefy step as it's possible to do both at once

@@ -43,14 +43,12 @@ get.subsamples <- function(morphospace, tree, bins, FADLAD, inc.nodes){
 }
 
 ## Bootstrap and rarefy subsamples
-## subsamples is output from get.subsamples as a list
+## subsamples is an individual subsample
+## bootstraps is the number of bootstraps to run
 boot.rarefy <- function(subsamples, bootstraps){
-  # Obtain a bootstrap matrix for the subsamples
-  bootstrapped_data <- boot.matrix(subsamples, bootstraps = bootstraps, rarefaction = FALSE)
-  # Obtain a rarefied bootstrap matrix for the subsamples
-  rarefied_data <- boot.matrix(subsamples, bootstraps = bootstraps, rarefaction = TRUE)
-}  
-  return(list(bootstrapped_data, rarefied_data))
+  # Obtain a rarefied bootstrap matrix matrix for the subsamples
+  bootstrapped_data <- boot.matrix(subsamples, bootstraps = bootstraps, rarefaction = TRUE)
+  return(bootstrapped_data)
 }
 
 ## Bootstrap and rarefy all types of subsamples

### 2017-09-20 functions/sim_fun.R
Tidy and remove writing outputs to file from function

@@ -6,6 +6,11 @@
 ##-------------------------------------------
 
 ## Extract all types of subsample for both methods and four continuous models
+## morphospace is the cleaned morphospace
+## tree is the phylogeny matched to the morphospace
+## bins is a vector of bin times
+## FADLAD is first and last occurrence data
+## inc.nodes is logical for whether to estimate diparity using nodal values
 get.subsamples <- function(morphospace, tree, bins, FADLAD, inc.nodes){
 
   ## Subset samples by required number of time bins
@@ -43,34 +48,22 @@ get.subsamples <- function(morphospace, tree, bins, FADLAD, inc.nodes){
 }
 
 ## Bootstrap and rarefy subsamples
-## subsamples is an individual subsample
+## subsamples is a subsample output from time.subsamples
 ## bootstraps is the number of bootstraps to run
 boot.rarefy <- function(subsamples, bootstraps){
   # Obtain a rarefied bootstrap matrix matrix for the subsamples
-  bootstrapped_data <- boot.matrix(subsamples, bootstraps = bootstraps, rarefaction = TRUE)
-  return(bootstrapped_data)
-}
-
-## Bootstrap and rarefy all types of subsamples
-## subsamples is output from get.subsamples as a list
-boot.rarefy.all <- function(subsamples, bootstraps){
-  purrr::map(subsamples, boot.rarefy, bootstraps = bootstraps)
+  boot.matrix(subsamples, bootstraps = bootstraps, rarefaction = TRUE)
 }
 
 ##-------------------------------------------
 ## Extract disparity for all subsamples
 ##-------------------------------------------
 
-## Estimate disparity - observed and bootstrapped
-## subsampled_data is the output of boot.rarefy.all
+## Estimate disparity - observed and bootstrapped and rarefied
+## subsampled_data is output from boot.matrix
+## metric is the chosen metric - see dispRity documentation for details
 get.disparity <- function(subsampled_data, metric){
   summary(dispRity(subsampled_data, metric = metric), round = 10)
-}
-
-## Get disparity for all types of subsamples
-## subsampled_data is the output of boot.rarefy.all as a list
-get.disparity.all <- function(subsampled_data, metric){
-  purrr::map(subsampled_data, get.disparity, metric = metric)
 }
 
 ##------------------------------------------
@@ -78,6 +71,10 @@ get.disparity.all <- function(subsampled_data, metric){
 ##------------------------------------------
 
[…]

### 2017-09-20 functions/sim_fun.R
Tidy up function documentation

@@ -1,16 +1,15 @@
-## Functions
+## Functions for time slice analyses
+## Natalie Cooper 2017
 
-
-##-------------------------------------------
-## Creating subsamples for disparity analyses
-##-------------------------------------------
-
+##---------------------------------------------------------------------------
 ## Extract all types of subsample for both methods and four continuous models
 ## morphospace is the cleaned morphospace
 ## tree is the phylogeny matched to the morphospace
 ## bins is a vector of bin times
 ## FADLAD is first and last occurrence data
 ## inc.nodes is logical for whether to estimate diparity using nodal values
+##--------------------------------------------------------------------------
+
 get.subsamples <- function(morphospace, tree, bins, FADLAD, inc.nodes){
 
   ## Subset samples by required number of time bins
@@ -47,34 +46,36 @@ get.subsamples <- function(morphospace, tree, bins, FADLAD, inc.nodes){
 
 }
 
+##--------------------------------------------------------------------------
 ## Bootstrap and rarefy subsamples
 ## subsamples is a subsample output from time.subsamples
 ## bootstraps is the number of bootstraps to run
+##--------------------------------------------------------------------------
+
 boot.rarefy <- function(subsamples, bootstraps){
   # Obtain a rarefied bootstrap matrix matrix for the subsamples
   boot.matrix(subsamples, bootstraps = bootstraps, rarefaction = TRUE)
 }
 
-##-------------------------------------------
+##--------------------------------------------------------------------------
 ## Extract disparity for all subsamples
-##-------------------------------------------
-
 ## Estimate disparity - observed and bootstrapped and rarefied
 ## subsampled_data is output from boot.matrix
 ## metric is the chosen metric - see dispRity documentation for details
+##--------------------------------------------------------------------------
+
 get.disparity <- function(subsampled_data, metric){
   summary(dispRity(subsampled_data, metric = metric), round = 10)
 }
 
-##------------------------------------------
-## Output disparity results from simulations
-##------------------------------------------
-
+##--------------------------------------------------------------------------
 ## Extract outputs needed for results
 ## disparity.object is the output of disparity
 ## bins is a vector of bin times
 ## metricname is a string specifying the name of the metric being used
 ## inc.nodes is logical for whether diparity was estimated using nodal values
+##--------------------------------------------------------------------------
+
[…]

### 2017-09-20 functions/sim_fun.R
Fix to documentation

@@ -70,7 +70,7 @@ get.disparity <- function(subsampled_data, metric){
 
 ##--------------------------------------------------------------------------
 ## Extract outputs needed for results
-## disparity.object is the output of disparity
+## disparity.object is the output of get.disparity
 ## bins is a vector of bin times
 ## metricname is a string specifying the name of the metric being used
 ## inc.nodes is logical for whether diparity was estimated using nodal values

### 2017-09-20 functions/sim_fun.R
Outputs now include model names and bin numbers

@@ -76,14 +76,16 @@ get.disparity <- function(subsampled_data, metric){
 ## inc.nodes is logical for whether diparity was estimated using nodal values
 ##--------------------------------------------------------------------------
 
-get.outputs <- function(disparity.object, bins, metric.name, inc.nodes){
+get.outputs <- function(disparity.object, bins, metric.name, inc.nodes, model.name){
   output <- as.data.frame(disparity.object)
-  output$bins <- paste(as.character(bins), collapse = "")
+  output$n_bins <- length(bins)
   output$metric <- metric.name
   output$inc.nodes <- inc.nodes
   return(output)
  }
 
+
+mapply(cbind, disparity.object, "SampleID"=ID, SIMPLIFY=F)
 #### Also need some way to name lists/extract models to put into the output somehow???
 
 ##--------------------------------------------------------------------------
@@ -114,7 +116,17 @@ run.all.disparity <- function(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
   disparity.object <- purrr::map(subsampled_data, get.disparity, metric = metric)
 
   ## Outputs
+  ## Extract output information
   disparity.outputs <- purrr::map(disparity.object, get.outputs, bins = bins, 
                                   metric.name = metric.name, inc.nodes = inc.nodes)
-
+  
+  ## Add model designations
+  models <- c("equalbins", "acctran", "deltran", "punctuated", "gradual")
+  disparity.outputs <- mapply(cbind, disparity.object, "model"= models, SIMPLIFY = FALSE)
+  ## Flatten list to a dataframe
+  ## This will throw warnings as factor levels of subsamples are converted to characters
+  disparity.outputs <- purrr:map_dfr(disparity.object, cbind)
+  
+  return(disparity.outputs)
 }
+

### 2017-09-20 functions/sim_fun.R
Remove comment

@@ -84,10 +84,6 @@ get.outputs <- function(disparity.object, bins, metric.name, inc.nodes, model.na
   return(output)
  }
 
-
-mapply(cbind, disparity.object, "SampleID"=ID, SIMPLIFY=F)
-#### Also need some way to name lists/extract models to put into the output somehow???
-
 ##--------------------------------------------------------------------------
 ## Putting it all together
 ## morphospace is the cleaned morphospace

### 2017-09-20 functions/sim_fun.R
Typo

@@ -11,7 +11,7 @@
 ##--------------------------------------------------------------------------
 
 get.subsamples <- function(morphospace, tree, bins, FADLAD, inc.nodes){
-
+  
   ## Subset samples by required number of time bins
   subsamples.bins <- time.subsamples(data = morphospace, tree = tree, 
   	                                 method = "discrete", time = bins, 

### 2017-09-20 functions/sim_fun.R
Typo

@@ -118,7 +118,7 @@ run.all.disparity <- function(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
   
   ## Add model designations
   models <- c("equalbins", "acctran", "deltran", "punctuated", "gradual")
-  disparity.outputs <- mapply(cbind, disparity.object, "model"= models, SIMPLIFY = FALSE)
+  disparity.outputs <- mapply(cbind, disparity.object, "model" = models, SIMPLIFY = FALSE)
   ## Flatten list to a dataframe
   ## This will throw warnings as factor levels of subsamples are converted to characters
   disparity.outputs <- purrr:map_dfr(disparity.object, cbind)

### 2017-09-20 functions/sim_fun.R
Prep to add new strat functions

@@ -43,7 +43,6 @@ get.subsamples <- function(morphospace, tree, bins, FADLAD, inc.nodes){
 
   return(list(subsamples.bins, subsamples.acctran, subsamples.deltran, 
   	          subsamples.punctuated, subsamples.gradual))
-
 }
 
 ##--------------------------------------------------------------------------
@@ -85,7 +84,7 @@ get.outputs <- function(disparity.object, bins, metric.name, inc.nodes, model.na
  }
 
 ##--------------------------------------------------------------------------
-## Putting it all together
+## Putting it all together for four models plus equal bins
 ## morphospace is the cleaned morphospace
 ## tree is the phylogeny matched to the morphospace
 ## bins is a vector of bin times

### 2017-09-20 functions/sim_fun.R
Add get bins function

@@ -2,6 +2,23 @@
 ## Natalie Cooper 2017
 
 ##---------------------------------------------------------------------------
+## Define sets of bins based on stratigraphy
+## tree is the phylogeny
+## type is "Age" or "Epoch"
+##---------------------------------------------------------------------------
+
+get.bins <- function(tree, type){
+  ## Get the stratigraphic ages
+  time_strat <- get.bin.ages(tree, type = type)
+  ## Get the median duration of the stratigraphic period (to create equal time bins)
+  strat_duration <- median(abs(diff(time_strat)))
+  ## Get the number of stratigraphic periods (to create equal time bins)
+  strat_number <- length(time_strat)
+  
+  return(list(time_strat, strat_duration, strat_number))
+}
+
+##---------------------------------------------------------------------------
 ## Extract all types of subsample for both methods and four continuous models
 ## morphospace is the cleaned morphospace
 ## tree is the phylogeny matched to the morphospace

### 2017-09-20 functions/sim_fun.R
Add function to extract subsamples by stratigraphy

@@ -19,12 +19,27 @@ get.bins <- function(tree, type){
 }
 
 ##---------------------------------------------------------------------------
+## Extract stratigraphy subsamples
+## morphospace is the cleaned morphospace
+## tree is the phylogeny matched to the morphospace
+## FADLAD is first and last occurrence data
+## inc.nodes is logical for whether to estimate disparity using nodal values
+## type is "Age" or "Epoch"
+##--------------------------------------------------------------------------
+
+get.subsamples.stratigraphy <- function(morphospace, tree, FADLAD, inc.nodes, type){
+  time.subsamples(data = morphospace, tree = tree, 
+                  method = "discrete", time = get.bins(tree, type)[[1]], 
+                  FADLAD = FADLAD, inc.nodes = inc.nodes)
+}  
+  
+##---------------------------------------------------------------------------
 ## Extract all types of subsample for both methods and four continuous models
 ## morphospace is the cleaned morphospace
 ## tree is the phylogeny matched to the morphospace
 ## bins is a vector of bin times
 ## FADLAD is first and last occurrence data
-## inc.nodes is logical for whether to estimate diparity using nodal values
+## inc.nodes is logical for whether to estimate disparity using nodal values
 ##--------------------------------------------------------------------------
 
 get.subsamples <- function(morphospace, tree, bins, FADLAD, inc.nodes){

### 2017-09-20 functions/sim_fun.R
Update doc

@@ -4,7 +4,7 @@
 ##---------------------------------------------------------------------------
 ## Define sets of bins based on stratigraphy
 ## tree is the phylogeny
-## type is "Age" or "Epoch"
+## type is "Age" or "Epoch" - see get.bin.ages documentation
 ##---------------------------------------------------------------------------
 
 get.bins <- function(tree, type){
@@ -24,7 +24,7 @@ get.bins <- function(tree, type){
 ## tree is the phylogeny matched to the morphospace
 ## FADLAD is first and last occurrence data
 ## inc.nodes is logical for whether to estimate disparity using nodal values
-## type is "Age" or "Epoch"
+## type is "Age" or "Epoch" - see get.bin.ages documentation
 ##--------------------------------------------------------------------------
 
 get.subsamples.stratigraphy <- function(morphospace, tree, FADLAD, inc.nodes, type){

### 2017-09-20 functions/sim_fun.R
Complete wrapper for stratigraphy bins

@@ -109,14 +109,17 @@ get.disparity <- function(subsampled_data, metric){
 
 get.outputs <- function(disparity.object, bins, metric.name, inc.nodes, model.name){
   output <- as.data.frame(disparity.object)
-  output$n_bins <- length(bins)
   output$metric <- metric.name
   output$inc.nodes <- inc.nodes
+  if(length(bins) == 1){output$n_bins <- bins}
+  else{output$n_bins <- length(bins)}
   return(output)
  }
 
 ##--------------------------------------------------------------------------
 ## Putting it all together for four models plus equal bins
+## This function will run models of various time bin/slice numbers ignoring
+## stratigraphic information
 ## morphospace is the cleaned morphospace
 ## tree is the phylogeny matched to the morphospace
 ## bins is a vector of bin times
@@ -157,3 +160,51 @@ run.all.disparity <- function(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
   return(disparity.outputs)
 }
 
+##--------------------------------------------------------------------------
+## Putting it all together for four models plus equal bins
+## morphospace is the cleaned morphospace
+## tree is the phylogeny matched to the morphospace
+## bins is a vector of bin times
+## FADLAD is first and last occurrence data
+## inc.nodes is logical for whether to estimate diparity using nodal values
+##--------------------------------------------------------------------------
+
+run.all.disparity.strat <- function(morphospace, tree, type, FADLAD, inc.nodes = TRUE,
+                              bootstraps = 1, metric, metric.name){
+  
+  # Need to remove the nodes if not using them
+  if(inc.nodes == FALSE){
+    remove.nodes <- clean.data(morphospace, tree)
+    morphospace <- remove.nodes$data  
+  }
+  
+  # Extract bins to go with particular stratigraphy
+  strat.bins <-get.bins(tree, type)
+  
+  ## STRATIGRAPHY
+  ## Extract subsamples for stratigraphy
+  subsamples <- get.subsamples.stratigraphy(morphospace, tree, FADLAD, inc.nodes, type)
+  ## Bootstrap and rarefy
+  subsampled_data <- boot.rarefy(subsamples, bootstraps = bootstraps)
+  ## Get disparity 
+  disparity.object <- purrr::map(subsampled_data, get.disparity, metric = metric)
+  ## Extract output information
[…]

### 2017-09-20 functions/sim_fun.R
Typo

@@ -155,7 +155,7 @@ run.all.disparity <- function(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
   disparity.outputs <- mapply(cbind, disparity.object, "model" = models, SIMPLIFY = FALSE)
   ## Flatten list to a dataframe
   ## This will throw warnings as factor levels of subsamples are converted to characters
-  disparity.outputs <- purrr:map_dfr(disparity.object, cbind)
+  disparity.outputs <- purrr::map_dfr(disparity.object, cbind)
   
   return(disparity.outputs)
 }

### 2017-09-20 functions/sim_fun.R
Fix cbind code

@@ -187,21 +187,21 @@ run.all.disparity.strat <- function(morphospace, tree, type, FADLAD, inc.nodes =
   ## Bootstrap and rarefy
   subsampled_data <- boot.rarefy(subsamples, bootstraps = bootstraps)
   ## Get disparity 
-  disparity.object <- purrr::map(subsampled_data, get.disparity, metric = metric)
+  disparity.object <- get.disparity(subsampled_data, metric = metric)
   ## Extract output information
-  disparity.stratigraphy <- purrr::map(disparity.object, get.outputs, bins = bins, 
-                                  metric.name = metric.name, inc.nodes = inc.nodes)
-  disparity.stratigraphy <- cbind(disparity.number, statigraphy = type, bin_type = "unequal")
+  disparity.stratigraphy <- get.outputs(disparity.object, bins = strat.bins[[1]], 
+                                        metric.name = metric.name, inc.nodes = inc.nodes)
+  disparity.stratigraphy <- cbind(disparity.stratigraphy, "statigraphy" = type, "bin_type" = "unequal")
   
   ## BIN/SLICES of same DURATION as STRATIGRAPHY
   disparity.duration <- run.all.disparity(morphospace, tree, strat.bins[[2]], FADLAD, inc.nodes = inc.nodes,
                                           bootstraps = bootstraps, metric, metric.name)
-  disparity.duration <- cbind(disparity.duration, statigraphy = type, bin_type = "duration")
+  disparity.duration <- cbind(disparity.duration, "statigraphy" = type, "bin_type" = "duration")
   
   ## BIN/SLICES of same NUMBER as STRATIGRAPHY
   disparity.number <- run.all.disparity(morphospace, tree, strat.bins[[3]], FADLAD, inc.nodes = inc.nodes,
                                           bootstraps = bootstraps, metric, metric.name)
-  disparity.number <- cbind(disparity.number, statigraphy = type, bin_type = "number")
+  disparity.number <- cbind(disparity.number, "statigraphy" = type, "bin_type" = "number")
 
   ## Outputs
   disparity.outputs <- cbind(disparity.stratigraphy, disparity.duration, disparity.number)

### 2017-09-20 functions/sim_fun.R
Fix documentation and tidy

@@ -37,7 +37,7 @@ get.subsamples.stratigraphy <- function(morphospace, tree, FADLAD, inc.nodes, ty
 ## Extract all types of subsample for both methods and four continuous models
 ## morphospace is the cleaned morphospace
 ## tree is the phylogeny matched to the morphospace
-## bins is a vector of bin times
+## bins is a vector of bin times or number of bins
 ## FADLAD is first and last occurrence data
 ## inc.nodes is logical for whether to estimate disparity using nodal values
 ##--------------------------------------------------------------------------
@@ -102,7 +102,7 @@ get.disparity <- function(subsampled_data, metric){
 ##--------------------------------------------------------------------------
 ## Extract outputs needed for results
 ## disparity.object is the output of get.disparity
-## bins is a vector of bin times
+## bins is a vector of bin times or number of bins
 ## metricname is a string specifying the name of the metric being used
 ## inc.nodes is logical for whether diparity was estimated using nodal values
 ##--------------------------------------------------------------------------
@@ -120,15 +120,19 @@ get.outputs <- function(disparity.object, bins, metric.name, inc.nodes, model.na
 ## Putting it all together for four models plus equal bins
 ## This function will run models of various time bin/slice numbers ignoring
 ## stratigraphic information
+
 ## morphospace is the cleaned morphospace
 ## tree is the phylogeny matched to the morphospace
-## bins is a vector of bin times
+## bins is a vector of bin times or number of bins
 ## FADLAD is first and last occurrence data
 ## inc.nodes is logical for whether to estimate diparity using nodal values
+## bootstraps is the number of bootstraps
+## metric is the chosen metric - see dispRity documentation for details 
+## metricname is a string specifying the name of the metric being used
 ##--------------------------------------------------------------------------
 
 run.all.disparity <- function(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
-                              bootstraps = 1, metric, metric.name){
+                              bootstraps = bootstraps, metric, metric.name){
 
   # Need to remove the nodes if not using them
   if(inc.nodes == FALSE){
@@ -161,16 +165,20 @@ run.all.disparity <- function(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
 }
 
 ##--------------------------------------------------------------------------
-## Putting it all together for four models plus equal bins
[…]

### 2017-09-20 analyses/02-time-slice-analyses.Rmd
Add details so code can be run when fixed

--- 2017-09-07 analyses/02-time-slice-analyses.Rmd
@@
 
-TITLE?
-===============
-
-DETAILS
-This code loads the data and extracts matrices and other files needed to run the analyses in the paper.
-Note that these analyses can take time, so we save the outputs to a `data/backups` folder and use these in other `.Rmd` files so that the analyses can be replicated more quickly.
+This code runs the analyses in the paper.
 Below we lay out the process in detail for one dataset, and then provide the code for the other datasets below this.
+Note that much of this uses functions in the `simulations.R` file.
 
@@
 
-Load packages
+Load packages and functions
 -----------------------------------
@@
 
-```{r, eval = TRUE}
+```{r, eval = TRUE, message = FALSE}
 # The latest version of devtools and ape
@@
 
-## The latest version of Claddis (on development on my own branch)
-install_github("TGuillerme/Claddis")
+## The latest version of Claddis (on development on TG branch)
+## install_github("TGuillerme/Claddis")
 library(Claddis)
 
-## The latest version of dispRity 
-install_github("TGuillerme/dispRity", ref = "release")
+## The latest version of dispRity on master branch
+## install_github("TGuillerme/dispRity", ref = "master")
 library(dispRity)
+
+## Load functions using source
+# Need to fix this!
+# source("../functions/simulations_fun.R")
 ```
@@
 
-Stratigraphy
------------------------------------
+Scenarios
+--------------------------------
 
-How to go about that????
-Need to collate data ourselves or can we use strap?
+We used the data to explore these different scenarios:
+
+1. Stratigraphy. This is the traditional method, where all the species within each stratigraphic period are included in the disparity calculation. This often leads to bins of unequal duration.
+
+2. Equally sized time bins. 
+    i. Where the *duration* of the bin is equal to the median duration of the stratigraphic period.
+    ii. Where the *number* of bins is equal to the number of stratigraphic periods.
+    iii. Where the *number* of bins is either 2, 5, 10, 15, or 20.
+
+3. Time slicing. 
+    i. Where the *interval* between slices is equal to the median duration of the stratigraphic period.
+    ii. Where the *number* of slices is equal to the number of stratigraphic periods.
+    iii. Where the *number* of slices is either 2, 5, 10, 15, or 20.
+
+Running the analyses and saving the outputs
+--------------------------------------------
+
+These analyses use a series of functions in `functions\`. These essentially wrap the existing `dispRity` functions so that running lots of models and extracting lots of data is not such a big issue.
+
+```{r, eval = FALSE, warning = FALSE}
+## Run the function for Beck data
+slug <- "BeckLee"
+
+## Ages
+out1 <- run.all.disparity.strat(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
+                                bootstraps = 100, metric = c(sum, variances), metric.name = sum_var)
+readr::write_csv(out1, path = paste0("../outputs/age", "_", slug, ".csv"))
+
+## Epochs
+out2 <- run.all.disparity.strat(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
+                        bootstraps = 100, metric = c(sum, variances), metric.name = sum_var)
+readr::write_csv(out2, path = paste0("../outputs/epoch", "_", slug, ".csv"))
+```
+
+We also ran the models using varying numbers of bins as follows.
 
 ```{r, eval = FALSE}
+## Run the function for Beck data
+slug <- "BeckLee"
 
-## Run the function for Beck data
+for (bins in seq_along(c(2, 5, 10, 15, 20))){
+  output <- run.all.disparity(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
+                              bootstraps = 100, = c(sum, variances), metric.name = sum_var){
 
-slug <- "BeckLee"
-metric <- c(sum, variances)
-metric.name <- "sumvar"
-
-for (bins in seq_along(c(5,10,15,20))){
- run.all.sims(morphospace, tree, FADLAD, bins, metric, 
-              metric.name, inc.nodes = TRUE, slug)
-}
-
-slug <- "BeckLee"
-metric <- c(median, centroids)
-metric.name <- "medcentroid"
-
-for (bins in seq_along(c(5,10,15,20))){
- run.all.sims(morphospace, tree, FADLAD, bins, metric, 
-              metric.name, inc.nodes = TRUE, slug)
+  readr::write_csv(output, path = paste0("../outputs/bins", "_", slug, ".csv"), append = TRUE)
 }
@@
 ```
+
+<!-- TG: Suggestion for a "testing the effect of binning/slicing" pipeline -->
+
+Examining the output 
+--------------------
+
+Plots
+Statistics

### 2017-09-20 functions/simulations_fun.R
Remove defunct functions

@@ -1,137 +1,0 @@
-# Functions to run the simulations and extract data
-#---------------------------------------------------
-
-#---------------------------------------------------
-# 1. STRATIGRAPHY BINS
-#---------------------------------------------------
-
-
-#---------------------------------------------------
-# 2. TIME BINS
-# Allows you to vary the metric, bins and whether
-# to include nodes
-# Outputs to a slug specific text file
-# Bootstraps and rarefies
-#---------------------------------------------------
-time.bin.sims <- function(morphospace, tree, FADLAD, bins, metric, 
-                          metric.name, inc.nodes = TRUE, slug) {
-  
-  # Need to remove the nodes if not using them
-  if(inc.nodes == FALSE){
-    remove.nodes <- clean.data(morphospace, tree)
-    morphospace <- remove.nodes$data  
-  }
-  
-  # Subset samples by required number of time bins
-  subsamples <- time.subsamples(data = morphospace, tree = tree, method = "discrete",
-                                time = bins, FADLAD = FADLAD, inc.nodes = inc.nodes)
-  
-  # Obtain a bootstrap matrix for the subsamples
-  bootstrapped_data <- boot.matrix(subsamples, bootstraps = 1, rarefaction = FALSE)
-  
-  # Estimate disparity - observed and bootstrapped
-  disparity <- summary(dispRity(bootstrapped_data, metric = metric), round = 10)
-  
-  # Rarefaction appears to be broken...
-  # Obtain a rarified bootstrap matrix for the subsamples
-  rarefied_data <- boot.matrix(subsamples, bootstraps = 1, rarefaction = TRUE)
-  
-  # Estimate disparity - observed and bootstrapped with rarefaction
-  disparity_rarefied <- summary(dispRity(rarefied_data, metric = metric), round = 10)
-  
-  # Outputs
-  output <- as.data.frame(disparity)
-  output$bins <- bins
-  output$metric <- metric.name
-  output$inc.nodes <- inc.nodes
-  readr::write_csv(output, path = paste0("../outputs/timebins.", slug ,".csv"), 
-                   append = TRUE)
-  
-  output2 <- as.data.frame(disparity_rarefied)
-  output2$bins <- bins
-  output2$metric <- metric.name
-  output2$inc.nodes <- inc.nodes
-  readr::write_csv(output2, path = paste0("../outputs/timebins_rarefied.", slug ,".csv"), 
-                   append = TRUE)
-    
-}
-
-#---------------------------------------------------
-# 3. TIME SLICES
[…]

### 2017-11-01 analyses/02-time-slice-analyses.Rmd
Fix bins and source

@@
 ## Load functions using source
-# Need to fix this!
-# source("../functions/simulations_fun.R")
+# Doesn't always work... Rmd problem?
+source("../functions/simulations_fun.R")
 ```
@@
 
-for (bins in seq_along(c(2, 5, 10, 15, 20))){
+for (bins in c(2, 5, 10, 15, 20)){
   output <- run.all.disparity(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,

### 2017-11-01 analyses/02-time-slice-analyses.Rmd
Add " to metric name

@@
 out1 <- run.all.disparity.strat(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
-                                bootstraps = 100, metric = c(sum, variances), metric.name = sum_var)
+                                bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var")
 readr::write_csv(out1, path = paste0("../outputs/age", "_", slug, ".csv"))
@@
 out2 <- run.all.disparity.strat(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
-                        bootstraps = 100, metric = c(sum, variances), metric.name = sum_var)
+                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var")
 readr::write_csv(out2, path = paste0("../outputs/epoch", "_", slug, ".csv"))
@@
   output <- run.all.disparity(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
-                              bootstraps = 100, = c(sum, variances), metric.name = sum_var){
+                              bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var")
 

### 2017-11-06 analyses/02-time-slice-analyses.Rmd
Add comments on issues

@@
 
+<!-- Both the above throw this error:
+
+Error in if (is.na(data$subsamples[subsamples_check[subsample]][[1]]$elements[1, : argument is of length zero -->
+
 We also ran the models using varying numbers of bins as follows.
@@
 ```
+<!-- Also throws Error in if (is.na(data$subsamples[subsamples_check[subsample]][[1]]$elements[1, : argument is of length zero -->
+
+<!-- Need to think more about the outputs here, currently they don't make much sense as there are no column headers -->
 

### 2017-11-07 functions/simulations_fun.R
changing the simulations fun to use `boot.matrix` with `rarefaction` arg rather than default `boot.rarefy`
This will allow to run the functions without the full rarefaction
analysis each time. We can first just use the non-rarefied data and do
a rarefied analysis later for the supplementary (or simply fix the
rarefaction to a specific number of taxa)

@@ -132,7 +132,7 @@ get.outputs <- function(disparity.object, bins, metric.name, inc.nodes, model.na
 ##--------------------------------------------------------------------------
 
 run.all.disparity <- function(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
-                              bootstraps = bootstraps, metric, metric.name){
+                              bootstraps = bootstraps, metric, metric.name, rarefaction){
 
   # Need to remove the nodes if not using them
   if(inc.nodes == FALSE){
@@ -144,7 +144,7 @@ run.all.disparity <- function(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
   subsamples <- get.subsamples(morphospace, tree, bins, FADLAD, inc.nodes)
 
   ## Bootstrap and rarefy all types of subsamples
-  subsampled_data <- purrr::map(subsamples, boot.rarefy, bootstraps = bootstraps)
+  subsampled_data <- purrr::map(subsamples, boot.matrix, bootstraps = bootstraps, rarefaction = rarefaction)
 
   ## Get disparity for all types of subsamples
   disparity.object <- purrr::map(subsampled_data, get.disparity, metric = metric)
@@ -178,7 +178,7 @@ run.all.disparity <- function(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
 ##--------------------------------------------------------------------------
 
 run.all.disparity.strat <- function(morphospace, tree, type, FADLAD, inc.nodes = TRUE,
-                              bootstraps = bootstraps, metric, metric.name){
+                              bootstraps = bootstraps, metric, metric.name, rarefaction = FALSE){
   
   # Need to remove the nodes if not using them
   if(inc.nodes == FALSE){
@@ -194,7 +194,8 @@ run.all.disparity.strat <- function(morphospace, tree, type, FADLAD, inc.nodes =
   ## Extract subsamples for stratigraphy
   subsamples <- get.subsamples.stratigraphy(morphospace, tree, FADLAD, inc.nodes, type)
   ## Bootstrap and rarefy
-  subsampled_data <- boot.rarefy(subsamples, bootstraps = bootstraps)
+  subsampled_data <- boot.matrix(subsamples, bootstraps = bootstraps, rarefaction = rarefaction)
+
   ## Get disparity 
   disparity.object <- get.disparity(subsampled_data, metric = metric)
   ## Extract output information
@@ -205,13 +206,13 @@ run.all.disparity.strat <- function(morphospace, tree, type, FADLAD, inc.nodes =
[…]

### 2017-11-07 functions/simulations_fun.R
Running wrapper

@@ -12,8 +12,12 @@ get.bins <- function(tree, type){
   time_strat <- get.bin.ages(tree, type = type)
   ## Get the median duration of the stratigraphic period (to create equal time bins)
   strat_duration <- median(abs(diff(time_strat)))
+  ## Get the range of the tree age
+  min_max <- range(time_strat)
+  ## Create the sequence of bins of equal duration
+  strat_duration <- rev(seq(from = min_max[1], to = min_max[2], by = strat_duration))
   ## Get the number of stratigraphic periods (to create equal time bins)
-  strat_number <- length(time_strat)
+  strat_number <- length(time_strat)-1 #TG: -1 because they are bin boundaries, not bin numbers
   
   return(list(time_strat, strat_duration, strat_number))
 }
@@ -49,27 +53,42 @@ get.subsamples <- function(morphospace, tree, bins, FADLAD, inc.nodes){
   	                                 method = "discrete", time = bins, 
   	                                 FADLAD = FADLAD, inc.nodes = inc.nodes)
   
+
+  #TG: For the time slicing, should we use the same boundaries of the time bins or should we go with the centre of the bins?
+  #TG: for example, if the median strat length is 5, the time bins will be 0-5, 5-10, 10-15 etc...
+  #TG: The slices, using the same data will be at 5, 10, 15. Is that what we want or we want them in the centre of the bins? I.e. 2.5, 7.5, 12.5, etc...
+  #TG: If so (I would argue yes) it's doing it below. Else, we can comment this out.
+
+  if(length(bins) != 1) {
+    ## Removing the last bin (that would be a negative slice)
+    slices <- bins[-length(bins)]
+    ## Getting the centre of the bins as slicing values
+    slices <- slices - (diff(bins)/2)
+  } else {
+    slices <- bins
+  }
+
   ## Subset samples by required number of time slices with acctran model
   subsamples.acctran <- time.subsamples(data = morphospace, tree = tree, 
-  	                                    method = "continuous", time = bins, 
+  	                                    method = "continuous", time = slices, 
   	                                    FADLAD = FADLAD, inc.nodes = inc.nodes,
                                         model = "acctran")
 
   ## Subset samples by required number of time slices with deltran model
   subsamples.deltran <- time.subsamples(data = morphospace, tree = tree, 
-  	                                    method = "continuous", time = bins, 
[…]

### 2017-11-07 analyses/02-time-slice-analyses.Rmd
pipelined analysis (test on test data not real!) and plots

--- 2017-11-01 analyses/02-time-slice-analyses.Rmd
@@
 
+> TG: for the following pipeline, I've reduced the number of rarefaction that highly increased the calculation time.
+We now only calculate the disparity for the non-rarefied subsamples and the subsamples rarefied to contain only 3 taxa (the lowest level of rarefaction).
+We can use the full rarefaction if we want to plot rarefaction curves later on.
+
+```{r}
+## Testing data: TO REMOVE
+#TG: it's just faster this way
+#TG: also, I've only tested it on "Epoch" since it take ages to run, should be pipelined all right though
+data(BeckLee_mat99)
+data(BeckLee_tree)
+data(BeckLee_ages)
+
+morphospace <- BeckLee_mat99
+tree <- BeckLee_tree
+FADLAD <- BeckLee_ages
+
+```
+
 ```{r, eval = FALSE, warning = FALSE}
@@
 
-## Ages
-out1 <- run.all.disparity.strat(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
-                                bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var")
-readr::write_csv(out1, path = paste0("../outputs/age", "_", slug, ".csv"))
+## Ages (rarefied to 3 taxa only)
+out1 <- run.all.disparity.wrapper(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
+                                bootstraps = 100, metric = c(sum, variances),
+                                metric.name = "sum_var", rarefaction = 3)
+#readr::write_csv(out1, path = paste0("../outputs/age", "_", slug, ".csv"))
 
-## Epochs
-out2 <- run.all.disparity.strat(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
-                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var")
-readr::write_csv(out2, path = paste0("../outputs/epoch", "_", slug, ".csv"))
+## Epochs (rarefied to 3 taxa only)
+out2 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
+                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var",
+                        rarefaction = FALSE)
+#readr::write_csv(out2, path = paste0("../outputs/epoch", "_", slug, ".csv"))
 ```
@@
 ```
+<!-- Need to think more about the outputs here, currently they don't make much sense as there are no column headers -->
 
@@
 
-Plots
-Statistics
+The results as a table
+
+```{r, eval = FALSE}
+knitr::kable(out2$results$stratigraphy, caption = "Disparity per stratigraphic time: unequal time bins and non-equidistant time slices")
+knitr::kable(out2$results$duration, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the median stratigraphic epoch duration")
+knitr::kable(out2$results$number, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the number of stratigraphic epochs")
+```
+
+This is not really digest and we should find a better way to present that.
+Anyway, these boring tables are always good for the supplementaries.
+
+Plotting wise, we have to solutions, either using `dispRity`
+
+```{r, eval = FALSE}
+op <- par(bty = "n")
+colors <- palette() #TG: that can be made fancier
+
+## STRATIGRAPHY
+## Plotting the continuous data
+plot(out2$object$stratigraphy[[2]], quantile = 1, col = colors[1], ylim = c(1,4), main = "Equal duration and stratigraphys equal to stratigraphy", xlab = "sum of ranges")
+plot(out2$object$stratigraphy[[3]], quantile = 1, col = colors[2], ylim = c(1,4), add = TRUE)
+plot(out2$object$stratigraphy[[4]], quantile = 1, col = colors[3], ylim = c(1,4), add = TRUE)
+plot(out2$object$stratigraphy[[5]], quantile = 1, col = colors[4], ylim = c(1,4), add = TRUE)
+## Plotting the discrete data
+plot(out2$object$stratigraphy[[1]], quantile = 1, type = "c", col = colors[5], ylim = c(1,4), add = TRUE, density = 0)
+#TG: density = 0 is to not show the huge confidence interval, even for quantile = 1!
+legend(x = "bottomright", legend = c("acctran", "deltran", "punctuated", "gradual", "time bins"), col = colors[1:5], lty = 1)
+```
+
+This on is interesting with no time bins after around 39 Mya and a drastic drop!
+
+```{r, eval = FALSE}
+op <- par(bty = "n")
+colors <- palette() #TG: that can be made fancier
+## DURATION
+## Plotting the continuous data
+plot(out2$object$duration[[2]], quantile = 1, col = colors[1], ylim = c(1,4), main = "Equal duration and durations equal to stratigraphy", xlab = "sum of ranges")
+plot(out2$object$duration[[3]], quantile = 1, col = colors[2], ylim = c(1,4), add = TRUE)
+plot(out2$object$duration[[4]], quantile = 1, col = colors[3], ylim = c(1,4), add = TRUE)
+plot(out2$object$duration[[5]], quantile = 1, col = colors[4], ylim = c(1,4), add = TRUE)
+## Plotting the discrete data
+plot(out2$object$duration[[1]], quantile = 1, type = "c", col = colors[5], ylim = c(1,4), add = TRUE, density = 0)
+legend(x = "bottomright", legend = c("acctran", "deltran", "punctuated", "gradual", "time bins"), col = colors[1:5], lty = 1)
+```
+
+Same absence of data after 39 but no drop!
+
+```{r, eval = FALSE}
+op <- par(bty = "n")
+colors <- palette() #TG: that can be made fancier
+## NUMBERS
+## Plotting the continuous data
+plot(out2$object$number[[2]], quantile = 1, col = colors[1], ylim = c(1,4), main = "Equal duration and numbers equal to stratigraphy", xlab = "sum of ranges")
+plot(out2$object$number[[3]], quantile = 1, col = colors[2], ylim = c(1,4), add = TRUE)
+plot(out2$object$number[[4]], quantile = 1, col = colors[3], ylim = c(1,4), add = TRUE)
+plot(out2$object$number[[5]], quantile = 1, col = colors[4], ylim = c(1,4), add = TRUE)
+## Plotting the discrete data
+plot(out2$object$number[[1]], quantile = 1, type = "c", col = colors[5], ylim = c(1,4), add = TRUE)
+legend(x = "bottomright", legend = c("acctran", "deltran", "punctuated", "gradual", "time bins"), col = colors[1:5], lty = 1)
+
+```
+
+This one is fairly similar but starts earlier in time.
+
+Another way would be to do histograms the same way we drew it when we designed the project but I'll work on that tomorrow I guess.

### 2017-11-09 functions/simulations_fun.R
Added some plotting functions

@@ -253,3 +253,89 @@ run.all.disparity.wrapper <- function(morphospace, tree, type, FADLAD, inc.nodes
   
   return(list("results" = disparity.outputs, "objects" = disparity.objects))
 }
+
+
+##--------------------------------------------------------------------------
+## Plotting the results using dispRity objects
+## 
+## slug.object The chain object element (e.g. ou1$object)
+## method Which method: 1 = stratigraphy, 2 = duration, 3 = number
+## colors The central tendency colors
+## colors_CI The central CI colors
+## main The plot title
+## ylim The Y axis limits
+## legend Whether to display the legend or not
+## CI Whether to display the CIs or not
+## time.bins.line Whether to display the time bins as lines or not
+##--------------------------------------------------------------------------
+
+plot.results.dispRity <- function(slug.object, method, colors, colors_CI, main, ylim, legend = FALSE, CI = TRUE, time.bins.line = FALSE) {
+
+    bins_type <- ifelse(time.bins.line, "c", "l")
+
+    ## STRATIGRAPHY
+    if(CI) {
+      ## Plotting the continuous data CIs
+      plot(slug.object[[method]][[2]], density = 70, col = c("white", colors_CI[1]), quantile = c(95), ylim = ylim, ylab = "Sum of variance", main = main)
+      plot(slug.object[[method]][[3]], density = 70, col = c("white", colors_CI[2]), quantile = c(95), ylim = ylim, add = TRUE, xlab = "", ylab = "")
+      plot(slug.object[[method]][[4]], density = 70, col = c("white", colors_CI[3]), quantile = c(95), ylim = ylim, add = TRUE, xlab = "", ylab = "")
+      plot(slug.object[[method]][[5]], density = 70, col = c("white", colors_CI[4]), quantile = c(95), ylim = ylim, add = TRUE, xlab = "", ylab = "")
+
+      ## Plotting the continuous data median
+      plot(slug.object[[method]][[2]], density = 0, col = colors[1], ylim = ylim, add = TRUE, xlab = "", ylab = "")
+    } else {
+      ## Plotting the continuous data (first plot)
+      plot(slug.object[[method]][[2]], density = 0, col = colors[1], ylim = ylim, ylab = "Sum of variance", main = main)
+    }
+
+
+    ## Plotting the continuous data medians (other plots)
[…]

### 2017-11-09 analyses/02-time-slice-analyses.Rmd
Updated pipelined analysis (compiles!)

--- 2017-11-01 analyses/02-time-slice-analyses.Rmd
@@
 
-```{r, eval = FALSE, warning = FALSE}
+> TG: for the following pipeline, I've reduced the number of rarefaction that highly increased the calculation time.
+We now only calculate the disparity for the non-rarefied subsamples and the subsamples rarefied to contain only 3 taxa (the lowest level of rarefaction).
+We can use the full rarefaction if we want to plot rarefaction curves later on.
+
+```{r, eval = FALSE, echo = FALSE}
+## Testing data: TO REMOVE
+#TG: it's just faster this way
+#TG: also, I've only tested it on "Epoch" since it take ages to run, should be pipelined all right though
+data(BeckLee_mat99)
+data(BeckLee_tree)
+data(BeckLee_ages)
+
+morphospace <- BeckLee_mat99
+tree <- BeckLee_tree
+FADLAD <- BeckLee_ages
+
+```
+
+```{r, eval = TRUE, warning = FALSE}
 ## Run the function for Beck data
@@
 
-## Ages
-out1 <- run.all.disparity.strat(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
-                                bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var")
-readr::write_csv(out1, path = paste0("../outputs/age", "_", slug, ".csv"))
+## Ages (rarefied to 3 taxa only)
+out1 <- run.all.disparity.wrapper(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
+                                bootstraps = 100, metric = c(sum, variances),
+                                metric.name = "sum_var", rarefaction = FALSE)
+#readr::write_csv(out1, path = paste0("../outputs/age", "_", slug, ".csv"))
 
-## Epochs
-out2 <- run.all.disparity.strat(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
-                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var")
-readr::write_csv(out2, path = paste0("../outputs/epoch", "_", slug, ".csv"))
+## Epochs (rarefied to 3 taxa only)
+out2 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
+                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var",
+                        rarefaction = FALSE)
+#readr::write_csv(out2, path = paste0("../outputs/epoch", "_", slug, ".csv"))
 ```
 
-We also ran the models using varying numbers of bins as follows.
+Note that you can use the previous output format (the data tables) as before by accessing them through `out$results$stratigraphy/duration/number`.
+Writing them as a csv will require some thinking since they don't have the same number of columns (though it's still doable I guess).
+Also, if you want to plot things through `ggplot`, you can use the values in these data tables!
 
-```{r, eval = FALSE}
+<!-- TG: I've not run this analysis below -->
+<!-- We also ran the models using varying numbers of bins as follows. -->
+
+```{r, eval = FALSE, echo = FALSE}
 ## Run the function for Beck data
@@
 
-<!-- TG: Suggestion for a "testing the effect of binning/slicing" pipeline -->
+<!-- NC: Need to think more about the outputs here, currently they don't make much sense as there are no column headers -->
 
@@
 
-Plots
-Statistics
+The results as a table
+
+```{r, eval = TRUE, echo = FALSE}
+knitr::kable(out2$results$stratigraphy, caption = "Disparity per stratigraphic time: unequal time bins and non-equidistant time slices")
+knitr::kable(out2$results$duration, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the median stratigraphic epoch duration")
+knitr::kable(out2$results$number, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the number of stratigraphic epochs")
+```
+
+This is not really digest and we should find a better way to present that.
+Anyway, these boring tables are always good for the supplementaries.
+
+## Plotting the results `dispRity` style
+
+Here is an example of the results using `plot.dispRity`.
+The upper panel is with the CIs and the lower one is the same without the CIs.
+
+### The results for the Age type
+
+```{r, eval = TRUE, fig.width = 12, fig.height = 18, echo = TRUE}
+op <- par(bty = "n", mfrow = c(3,2))
+## Principal colors
+colors <- palette()[1:5][-1] #TG: that can be made fancier
+colors[5] <- "black"
+## Color combination (for the CI)
+colors_CI <- c("pink", "lightgreen", "lightblue", "lightcyan", "lightgray")
+## y limits
+ylim <- range(unlist(lapply(out1$results, function(X) return(c(min(X[,5], na.rm = TRUE), max(X[,5],
+  na.rm = TRUE))))))
+## Zooming
+ylim_zoom <- c(20, 35)
+
+## Plotting the different methods (with/without CIs)
+plot.results.dispRity(out1$object, method = 1, colors = colors, colors_CI = colors_CI,
+  main = "Stratigraphy", ylim = ylim, CI = TRUE, legend = TRUE)
+plot.results.dispRity(out1$object, method = 1, colors = colors, colors_CI = colors_CI,
+  main = "Stratigraphy (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)
+
+plot.results.dispRity(out1$object, method = 2, colors = colors, colors_CI = colors_CI,
+  main = "Duration", ylim = ylim, CI = TRUE)
+plot.results.dispRity(out1$object, method = 2, colors = colors, colors_CI = colors_CI,
+  main = "Duration (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)
+
+plot.results.dispRity(out1$object, method = 3, colors = colors, colors_CI = colors_CI,
+  main = "Number", ylim = ylim, CI = TRUE)
+plot.results.dispRity(out1$object, method = 3, colors = colors, colors_CI = colors_CI,
+  main = "Number (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)
+```
+
+### The results for the Epoch type
+
+```{r, eval = TRUE, fig.width = 12, fig.height = 18, echo = TRUE}
+op <- par(bty = "n", mfrow = c(3,2))
+## Principal colors
+colors <- palette()[1:5][-1] #TG: that can be made fancier
+colors[5] <- "black"
+## Color combination (for the CI)
+colors_CI <- c("pink", "lightgreen", "lightblue", "lightcyan", "lightgray")
+## y limits
+ylim <- range(unlist(lapply(out2$results, function(X) return(c(min(X[,5], na.rm = TRUE),
+  max(X[,5], na.rm = TRUE))))))
+
+## Zooming
+ylim_zoom <- c(20, 35)
+
+## Plotting the different methods (with/without CIS)
+plot.results.dispRity(out2$object, method = 1, colors = colors, colors_CI = colors_CI,
+  main = "Stratigraphy", ylim = ylim, CI = TRUE, legend = TRUE)
+plot.results.dispRity(out2$object, method = 1, colors = colors, colors_CI = colors_CI,
+  main = "Stratigraphy (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)
+
+plot.results.dispRity(out2$object, method = 2, colors = colors, colors_CI = colors_CI,
+  main = "Duration", ylim = ylim, CI = TRUE)
+plot.results.dispRity(out2$object, method = 2, colors = colors, colors_CI = colors_CI,
+  main = "Duration (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)
+
+plot.results.dispRity(out2$object, method = 3, colors = colors, colors_CI = colors_CI,
+  main = "Number", ylim = ylim, CI = TRUE)
+plot.results.dispRity(out2$object, method = 3, colors = colors, colors_CI = colors_CI,
+  main = "Number(no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)
+
+```
+
+## Plotting the results histogram style
+
+Another way for the plots would be kind of our histogram representation (you can certainly use `ggplot` here, below is just an example):
+
+### The results per method
+
+```{r, fig.width = 12, fig.height = 18}
+## Set the list of available models
+models <- list("equalbins", "acctran", "deltran", "punctuated", "gradual")
+
+## Get the stratigraphic, duration and number data for the ages
+strat_data_ages <- lapply(models, make.class.histogram, out1$results, method = 1)
+durat_data_ages <- lapply(models, make.class.histogram, out1$results, method = 2)
+numbe_data_ages <- lapply(models, make.class.histogram, out1$results, method = 3)
+
+## Get the stratigraphic, duration and number data for the epochs
+strat_data_epoc <- lapply(models, make.class.histogram, out2$results, method = 1)
+durat_data_epoc <- lapply(models, make.class.histogram, out2$results, method = 2)
+numbe_data_epoc <- lapply(models, make.class.histogram, out2$results, method = 3)
+
+## The lower age and the lower disparity limit are arbitrarily fixed to 0 and 15 respectively.
+lower_age <- 0; lower_lim <- 15
+#TG: might want to change that for other datasets!
+
+## limits functions
+get.break <- function(x) return(x$breaks)
+get.count <- function(x) return(x$counts)
+## Ages limits
+xlim_ages <- c(max(unlist(c(sapply(strat_data_ages, get.break), sapply(durat_data_ages, get.break),
+  sapply(numbe_data_ages, get.break)))), lower_age)
+ylim_ages <- c(lower_lim, max(unlist(c(sapply(strat_data_ages, get.count),
+  sapply(durat_data_ages, get.count), sapply(numbe_data_ages, get.count)))))
+## Epoch limits
+xlim_epoc <- c(max(unlist(c(sapply(strat_data_epoc, get.break), sapply(durat_data_epoc, get.break),
+  sapply(numbe_data_epoc, get.break)))), lower_age)
+ylim_epoc <- c(lower_lim, max(unlist(c(sapply(strat_data_epoc, get.count),
+  sapply(durat_data_epoc, get.count), sapply(numbe_data_epoc, get.count)))))
+
+## Colors
+colors <- palette()[1:5]
+colors[1] <- "grey"
+
+## Graphics
+op <- par(bty = "n", mfrow = c(3,2))
+
+## Plot the stratigraphy - ages
+plot(strat_data_ages[[1]], xlim = xlim_ages, ylim = ylim_ages , xlab = "Time (Mya)",
+  ylab = "Sum of variance", col = colors[1], main = "Stratigraphy - ages")
+## Plot the slicing
+for(slice in 2:5) {
+  lines(strat_data_ages[[slice]]$breaks, strat_data_ages[[slice]]$counts, col = colors[slice])
+}
+
+## Adding the legend
+legend(x = "bottomleft", legend = c("time bins", "acctran", "deltran", "punctuated", "gradual"),
+  col = colors[1:5], lty = 1, bg = "white")
+
+## Plot the stratigraphy - epoch
+plot(strat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
+  ylab = "Sum of variance", col = colors[1], main = "Stratigraphy - Epoch")
+## Plot the slicing
+for(slice in 2:5) {
+  lines(strat_data_epoc[[slice]]$breaks, strat_data_epoc[[slice]]$counts, col = colors[slice])
+}
+
+## Plot the duration - ages
+plot(durat_data_ages[[1]], xlim = xlim_ages, ylim = ylim_ages , xlab = "Time (Mya)",
+  ylab = "Sum of variance", col = colors[1], main = "Duration - ages")
+## Plot the slicing
+for(slice in 2:5) {
+  lines(durat_data_ages[[slice]]$breaks, durat_data_ages[[slice]]$counts, col = colors[slice])
+}
+
+## Plot the duration - epoch
+plot(durat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
+  ylab = "Sum of variance", col = colors[1], main = "Duration - Epoch")
+## Plot the slicing
+for(slice in 2:5) {
+  lines(durat_data_epoc[[slice]]$breaks, durat_data_epoc[[slice]]$counts, col = colors[slice])
+}
+
+## Plot the number - ages
+plot(numbe_data_ages[[1]], xlim = xlim_ages, ylim = ylim_ages , xlab = "Time (Mya)",
+  ylab = "Sum of variance", col = colors[1], main = "Number - ages")
+## Plot the slicing
+for(slice in 2:5) {
+  lines(numbe_data_ages[[slice]]$breaks, numbe_data_ages[[slice]]$counts, col = colors[slice])
+}
+
+## Plot the number - epoch
+plot(numbe_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
+  ylab = "Sum of variance", col = colors[1], main = "Number - Epoch")
+## Plot the slicing
+for(slice in 2:5) {
+  lines(numbe_data_epoc[[slice]]$breaks, numbe_data_epoc[[slice]]$counts, col = colors[slice])
+}
+```
+
+Some of the plots are a bit off set and everything but it gives us an idea
+
+### The results per model
+
+Something like that but with transparency:
+
+```{r, fig.width = 6, fig.height = 6}
+plot(strat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
+  ylab = "Sum of variance", col = colors[1], main = "Epoch")
+plot(durat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
+  ylab = "Sum of variance", col = colors[2], main = "", add = TRUE)
+plot(numbe_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
+  ylab = "Sum of variance", col = colors[3], main = "", add = TRUE)
+```

### 2017-11-22 analyses/01-extract-data-for-analyses.Rmd
change dispRity version to master branch
Add Wright2017 data extractions

@@
 ## The latest version of dispRity 
-install_github("TGuillerme/dispRity", ref = "release")
+install_github("TGuillerme/dispRity", ref = "master")
 library(dispRity) 
@@
 
+Wright 2017
+-----------
+
+```{r}
+## Setting up the slug names
+slug <- "Wright2017"
+
+## Loading the data
+matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
+FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
+
+# Note from Wright 2017: MCC tree.
+Note that to properly timescale the tree [e.g., in R] you must first divide the branch lengths by the corresponding clock rate, (= 0.03517385), and then set the root.time to 485.4]
+
+# Replace zero branches with minimu branch length
+tree$edge.length[which(tree$edge.length == 0)] <- 
+  min(tree$edge.length[-which(tree$edge.length == 0)])
+
+# Multiply branches by 0.03517385
+tree$edge.length <- tree$edge.length * 0.03517385
+
+# Add root time: 485.4
+tree$root.time <- 485.4
+
+# Tidy the tree and the data
+tree <- makeNodeLabel(tree, method = "number", prefix = "n")
+cleaned_data <- clean.data(matrix$matrix, tree)
+matrix$matrix <- cleaned_data$data
+tree <- cleaned_data$tree
+
+## Estimating the ancestral states (with a uncertainty threshold of 0.95)
+matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
+                                  uncertainty.threshold = 0.95)
+save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+
+## Add ancestral states to nodes
+load(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+row.names(matrix_nodes) <- tree$node.label
+matrix$matrix <- rbind(matrix$matrix, matrix_nodes)
+
+## Constructing the distance matrix
+matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
+save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug,".Rda"))
+
+## Ordinating the matrix - no need to run this here, we will do this in the next script
+## load(paste0("../data/processed/distance_matrix.", slug,".Rda"))
+## morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
+```
+
+# Loop
 ```{r}
@@
 tree <- read.nexus(paste0("../data/trees/", slugs[slug], ".tre"))
+
+[MCC tree. Note that to properly timescale the tree [e.g., in R] you must first divide the branch lengths by the corresponding clock rate, (= 0.03517385), and then set the root.time to 485.4]
 

### 2017-11-22 analyses/02-time-slice-analyses-Wright2017.Rmd
Analysis file for Wright 2017

--- 2017-11-09 analyses/02-time-slice-analyses.Rmd
@@
 ## Choose a slug
-slug <- "Beck2014"
+slug <- "Wright2017"
 
@@
 tree <- makeNodeLabel(tree, method = "number", prefix = "n")
-tree$root.time <- max(tree.age(tree)[, 1])
+tree$root.time <- 485.4 # specific to Wright 2017 tree
 
@@
 
-```{r, eval = FALSE, echo = FALSE}
-## Testing data: TO REMOVE
-#TG: it's just faster this way
-#TG: also, I've only tested it on "Epoch" since it take ages to run, should be pipelined all right though
-data(BeckLee_mat99)
-data(BeckLee_tree)
-data(BeckLee_ages)
-
-morphospace <- BeckLee_mat99
-tree <- BeckLee_tree
-FADLAD <- BeckLee_ages
-
-```
-
 ```{r, eval = TRUE, warning = FALSE}
-## Run the function for Beck data
-slug <- "BeckLee"
-
 ## Ages (rarefied to 3 taxa only)
-out1 <- run.all.disparity.wrapper(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
+out1 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
                                 bootstraps = 100, metric = c(sum, variances),

### 2017-11-23 analyses/02-time-slice-analyses.Rmd
cleaned up former analysis drafts

@@
 
-## The latest version of dispRity on master branch
-## install_github("TGuillerme/dispRity", ref = "master")
+## The latest version of dispRity on release branch #TG: master is under many changes!
+## install_github("TGuillerme/dispRity", ref = "release")
 library(dispRity)
@@
 
-```{r, eval = FALSE, echo = FALSE}
-## Testing data: TO REMOVE
-#TG: it's just faster this way
-#TG: also, I've only tested it on "Epoch" since it take ages to run, should be pipelined all right though
-data(BeckLee_mat99)
-data(BeckLee_tree)
-data(BeckLee_ages)
-
-morphospace <- BeckLee_mat99
-tree <- BeckLee_tree
-FADLAD <- BeckLee_ages
-
-```
-
 ```{r, eval = TRUE, warning = FALSE}
@@
 Also, if you want to plot things through `ggplot`, you can use the values in these data tables!
-
-<!-- TG: I've not run this analysis below -->
-<!-- We also ran the models using varying numbers of bins as follows. -->
-
-```{r, eval = FALSE, echo = FALSE}
-## Run the function for Beck data
-slug <- "BeckLee"
-
-for (bins in c(2, 5, 10, 15, 20)){
-  output <- run.all.disparity(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
-                              bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var")
-
-  readr::write_csv(output, path = paste0("../outputs/bins", "_", slug, ".csv"), append = TRUE)
-}
-
-```
-
-<!-- NC: Need to think more about the outputs here, currently they don't make much sense as there are no column headers -->
 
@@
 ```
+
+## Comparing the disparity through time curves
+
+```{r}
+curve1 <- out2$object[[1]][[1]]
+curve2 <- out2$object[[1]][[2]]
+curve3 <- out2$object[[1]][[3]]
+curve4 <- out2$object[[1]][[4]]
+curve5 <- out2$object[[1]][[5]]
+
+```

### 2017-11-23 analyses/01-extract-data-for-analyses.Rmd
Remove unneeded comment
Add Bapst et al data

@@
 ## The latest version of dispRity 
-install_github("TGuillerme/dispRity", ref = "master")
+install_github("TGuillerme/dispRity", ref = "release")
 library(dispRity) 
@@
 ===================================================
+Ideally we'd have written functions to do this, and/or a big old loop. But each dataset was slightly different so we've dealt with each individually.
 
@@
 ```{r}
-## Setting up the slug names
 slug <- "Wright2017"
@@
 
-# Loop
+Bapst et al 2016 Biol Letters
+-----------
+
 ```{r}
-## Setting up the slug names
-slugs <- c("Bapst2016", "Hopkins", "Other")
-
-# Big old loop (too lazy to make functions sorry!)
-for(slug in 1: length(slugs)) {
+slug <- "Bapst2016"
 
 ## Loading the data
-matrix <- ReadMorphNexus(paste0("../data/matrices/", slugs[slug], ".nex"))
-FADLAD <- read.csv(paste0("../data/FADLAD/", slugs[slug], ".csv"), row.names = 1, header = TRUE)
-tree <- read.nexus(paste0("../data/trees/", slugs[slug], ".tre"))
-
-[MCC tree. Note that to properly timescale the tree [e.g., in R] you must first divide the branch lengths by the corresponding clock rate, (= 0.03517385), and then set the root.time to 485.4]
+matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
+FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+# Note that these are the upper bound for FAD and lower bound for LAD
+tree <- read.nexus(paste0("../data/trees/", slug, ".nex"))
 
@@
                                   uncertainty.threshold = 0.95)
-save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slugs[slug] ,".Rda"))
+save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 
 ## Add ancestral states to nodes
-load(paste0("../data/processed/ancestral_states.", slugs[slug] ,".Rda"))
+load(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 row.names(matrix_nodes) <- tree$node.label
@@
 matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
-save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slugs[slug] ,".Rda"))
+save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
 
@@
 ## morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
+```
 
-}
+Bapst et al 2016 Biol Letters
+-----------
+
+```{r}
+slug <- "Bapst2016"
+
+## Loading the data
+matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
+FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
+
+# Tidy the tree and the data
+tree <- makeNodeLabel(tree, method = "number", prefix = "n")
+tree$root.time <- max(tree.age(tree)[, 1])
+cleaned_data <- clean.data(matrix$matrix, tree)
+matrix$matrix <- cleaned_data$data
+tree <- cleaned_data$tree
+
+## Estimating the ancestral states (with a uncertainty threshold of 0.95)
+matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
+                                  uncertainty.threshold = 0.95)
+save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+
+## Add ancestral states to nodes
+load(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+row.names(matrix_nodes) <- tree$node.label
+matrix$matrix <- rbind(matrix$matrix, matrix_nodes)
+
+## Constructing the distance matrix
+matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
+save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+
+## Ordinating the matrix - no need to run this here, we will do this in the next script
+## load(paste0("../data/processed/distance_matrix.", slugs[slug] ,".Rda"))
+## morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
 ```
+
+Bapst et al 2016 Biol Letters
+-----------
+
+```{r}
+slug <- "Bapst2016"
+
+## Loading the data
+matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
+FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
+
+# Tidy the tree and the data
+tree <- makeNodeLabel(tree, method = "number", prefix = "n")
+tree$root.time <- max(tree.age(tree)[, 1])
+cleaned_data <- clean.data(matrix$matrix, tree)
+matrix$matrix <- cleaned_data$data
+tree <- cleaned_data$tree
+
+## Estimating the ancestral states (with a uncertainty threshold of 0.95)
+matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
+                                  uncertainty.threshold = 0.95)
+save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+
+## Add ancestral states to nodes
+load(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+row.names(matrix_nodes) <- tree$node.label
+matrix$matrix <- rbind(matrix$matrix, matrix_nodes)
+
+## Constructing the distance matrix
+matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
+save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+
+## Ordinating the matrix - no need to run this here, we will do this in the next script
+## load(paste0("../data/processed/distance_matrix.", slugs[slug] ,".Rda"))
+## morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
+```
+
+Bapst et al 2016 Biol Letters
+-----------
+
+```{r}
+slug <- "Bapst2016"
+
+## Loading the data
+matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
+FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
+
+# Tidy the tree and the data
+tree <- makeNodeLabel(tree, method = "number", prefix = "n")
+tree$root.time <- max(tree.age(tree)[, 1])
+cleaned_data <- clean.data(matrix$matrix, tree)
+matrix$matrix <- cleaned_data$data
+tree <- cleaned_data$tree
+
+## Estimating the ancestral states (with a uncertainty threshold of 0.95)
+matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
+                                  uncertainty.threshold = 0.95)
+save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+
+## Add ancestral states to nodes
+load(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+row.names(matrix_nodes) <- tree$node.label
+matrix$matrix <- rbind(matrix$matrix, matrix_nodes)
+
+## Constructing the distance matrix
+matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
+save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+
+## Ordinating the matrix - no need to run this here, we will do this in the next script
+## load(paste0("../data/processed/distance_matrix.", slugs[slug] ,".Rda"))
+## morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
+```
+Bapst et al 2016 Biol Letters
+-----------
+
+```{r}
+slug <- "Bapst2016"
+
+## Loading the data
+matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
+FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
+
+# Tidy the tree and the data
+tree <- makeNodeLabel(tree, method = "number", prefix = "n")
+tree$root.time <- max(tree.age(tree)[, 1])
+cleaned_data <- clean.data(matrix$matrix, tree)
+matrix$matrix <- cleaned_data$data
+tree <- cleaned_data$tree
+
+## Estimating the ancestral states (with a uncertainty threshold of 0.95)
+matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
+                                  uncertainty.threshold = 0.95)
+save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+
+## Add ancestral states to nodes
+load(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+row.names(matrix_nodes) <- tree$node.label
+matrix$matrix <- rbind(matrix$matrix, matrix_nodes)
+
+## Constructing the distance matrix
+matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
+save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+
+## Ordinating the matrix - no need to run this here, we will do this in the next script
+## load(paste0("../data/processed/distance_matrix.", slugs[slug] ,".Rda"))
+## morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
+```

### 2017-11-23 analyses/01-extract-data-for-analyses.Rmd
Add processing code for other datasets

@@
 
-Bapst et al 2016 Biol Letters
+Heath et al bears
 -----------
@@
 ```{r}
-slug <- "Bapst2016"
+slug <- "bears"
+
+## Loading the data
+matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
+FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+tree <- read.nexus(paste0("../data/trees/", slug, ".nex"))
+
+# Tidy the tree and the data
+tree <- makeNodeLabel(tree, method = "number", prefix = "n")
+tree$root.time <- max(tree.age(tree)[, 1])
+cleaned_data <- clean.data(matrix$matrix, tree)
+matrix$matrix <- cleaned_data$data
+tree <- cleaned_data$tree
+
+## Estimating the ancestral states (with a uncertainty threshold of 0.95)
+matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
+                                  uncertainty.threshold = 0.95)
+save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+
+## Add ancestral states to nodes
+load(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+row.names(matrix_nodes) <- tree$node.label
+matrix$matrix <- rbind(matrix$matrix, matrix_nodes)
+
+## Constructing the distance matrix
+matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
+save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+
+## Ordinating the matrix - no need to run this here, we will do this in the next script
+## load(paste0("../data/processed/distance_matrix.", slugs[slug] ,".Rda"))
+## morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
+```
+
+Brusatte et al 2014 Current Biol
+-----------
+
+```{r}
+slug <- "Brusatte2014"
 
@@
 ```
-
-Bapst et al 2016 Biol Letters
------------
-
-```{r}
-slug <- "Bapst2016"
-
-## Loading the data
-matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
-FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
-tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
-
-# Tidy the tree and the data
-tree <- makeNodeLabel(tree, method = "number", prefix = "n")
-tree$root.time <- max(tree.age(tree)[, 1])
-cleaned_data <- clean.data(matrix$matrix, tree)
-matrix$matrix <- cleaned_data$data
-tree <- cleaned_data$tree
-
-## Estimating the ancestral states (with a uncertainty threshold of 0.95)
-matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
-                                  uncertainty.threshold = 0.95)
-save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
-
-## Add ancestral states to nodes
-load(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
-row.names(matrix_nodes) <- tree$node.label
-matrix$matrix <- rbind(matrix$matrix, matrix_nodes)
-
-## Constructing the distance matrix
-matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
-save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
-
-## Ordinating the matrix - no need to run this here, we will do this in the next script
-## load(paste0("../data/processed/distance_matrix.", slugs[slug] ,".Rda"))
-## morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
-```
-
-Bapst et al 2016 Biol Letters
------------
-
-```{r}
-slug <- "Bapst2016"
-
-## Loading the data
-matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
-FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
-tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
-
-# Tidy the tree and the data
-tree <- makeNodeLabel(tree, method = "number", prefix = "n")
-tree$root.time <- max(tree.age(tree)[, 1])
-cleaned_data <- clean.data(matrix$matrix, tree)
-matrix$matrix <- cleaned_data$data
-tree <- cleaned_data$tree
-
-## Estimating the ancestral states (with a uncertainty threshold of 0.95)
-matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
-                                  uncertainty.threshold = 0.95)
-save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
-
-## Add ancestral states to nodes
-load(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
-row.names(matrix_nodes) <- tree$node.label
-matrix$matrix <- rbind(matrix$matrix, matrix_nodes)
-
-## Constructing the distance matrix
-matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
-save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
-
-## Ordinating the matrix - no need to run this here, we will do this in the next script
-## load(paste0("../data/processed/distance_matrix.", slugs[slug] ,".Rda"))
-## morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
-```
-Bapst et al 2016 Biol Letters
------------
-
-```{r}
-slug <- "Bapst2016"
-
-## Loading the data
-matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
-FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
-tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
-
-# Tidy the tree and the data
-tree <- makeNodeLabel(tree, method = "number", prefix = "n")
-tree$root.time <- max(tree.age(tree)[, 1])
-cleaned_data <- clean.data(matrix$matrix, tree)
-matrix$matrix <- cleaned_data$data
-tree <- cleaned_data$tree
-
-## Estimating the ancestral states (with a uncertainty threshold of 0.95)
-matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
-                                  uncertainty.threshold = 0.95)
-save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
-
-## Add ancestral states to nodes
-load(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
-row.names(matrix_nodes) <- tree$node.label
-matrix$matrix <- rbind(matrix$matrix, matrix_nodes)
-
-## Constructing the distance matrix
-matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
-save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
-
-## Ordinating the matrix - no need to run this here, we will do this in the next script
-## load(paste0("../data/processed/distance_matrix.", slugs[slug] ,".Rda"))
-## morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
-```

### 2017-11-24 functions/tree.age.fossil.R
tree.age.fossil function
To be updated with the proper dispRity v0.4.2 version

@@ -1,0 +1,73 @@
+#' @title Calculating the age of nodes and tips in a tree.
+#'
+#' @description Calculates the age of each node and tip in a tree give the height of the tree or some specified age.
+#'
+#' @param tree A \code{phylo} object.
+#' @param age The age of the tree. If missing the age is set to be the tree height.
+#' @param order Either "past" if the units express time since the present (e.g. million years ago), or "present" if the unit is expressed in time since the root.
+#' @param fossil.only \code{logical}, whether the tree contains only fossil taxa (\code{TRUE}) or not (\code{FALSE} - default). If the \code{tree} contains a \code{tree$root.time} element (for tree's root age), and that \code{order} is set to \code{"past"}, the ages in the tree are adjusted to be starting from the root.time, resulting in tips potentially older than age 0.
+#'
+#' @examples
+#' ## A dated random phylogeny with a root 50 units of time old.
+#' tree.age(rtree(10), age = 50)
+#' ## A random tree with the distance since the root.
+#' tree.age(rtree(10), order = 'present')
+#'
+#' @seealso \code{\link{slice.tree}}, \code{\link{time.subsets}}.
+#'
+#' @author Thomas Guillerme
+
+#Modified from [R-sig-phylo] nodes and taxa depth II - 21/06/2011 - Paolo Piras - ppiras(at)uniroma3.it
+
+tree.age.fossil <- function(tree, age, order = 'past'){
+
+#SANITYZING
+
+    #tree
+    check.class(tree, 'phylo', ' must be a phylo object.')
+
+    #age
+    if(missing(age)) {
+        #Using the tree height as age if age is missing
+        age <- max(dist.nodes(tree)[, Ntip(tree)+1])
+    }
+    check.class(age, 'numeric', ' must be a numerical value.')
+    check.length(age, '1', ' must a a single value.')
+
+    #order
+    check.method(order, c("past", "present"), "order argument")
+
+    ## Fossils only
+    check.class(fossil.only, "logical")
+
+    #CALCULATE THE EDGES AGE
+
+    if(age != 0) {
+        ages.table <- tree.age_scale(tree.age_table(tree), age)
+    } else {
+        ages.table <- tree.age_table(tree)
+    }
+
+    #Type
+    if(order != 'past') {
+        ages.table$ages <- round(ages.table$ages, digits = 7) # CHANGE PRECISION!
+    } else {
+        tree.height <- max(ages.table$ages)
[…]

### 2017-11-24 functions/get.bin.ages.fix.R
Quick fix functions that will be in dispRity 0.2.4

@@ -1,0 +1,74 @@
+#' @title Get time bins ages
+#'
+#' @description Gets time bins for a specific tree using stratigraphy
+#'
+#' @param tree A \code{phylo} object with a \code{$root.time} component
+#' @param what Which data to output. Can be \code{"Start"}, \code{"End"} (default), \code{"Range"} or \code{"Midpoint"}.
+#' @param type The type of stratigraphic frame. Can be \code{"Age"} (default), \code{"Eon"}, \code{"Epoch"}, \code{"Era"} or \code{"Period"}.
+#' @param ICS The reference year of the International Commission on Stratigraphy (default = \code{2015}).
+#' 
+#' @examples
+#' ## Loading the data
+#' data(BeckLee_tree)
+#' data(BeckLee_mat50)
+#' 
+#' ## Getting the stratigraphic data
+#' stratigraphy <- get.bin.ages(BeckLee_tree)
+#' 
+#' ## Making stratigraphic time subsets
+#' time.subsets(BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
+#'                 time = stratigraphy)
+#'
+#' @seealso \code{\link{time.subsets}}
+#' 
+#' @author Thomas Guillerme
+#' @import geoscale timescales
+#' @export
+
+get.bin.ages <- function(tree, what = "End", type = "Age", ICS = 2015) {    
+    ## Tree
+    # check.class(tree, "phylo")
+    if(is.null(tree$root.time)) {
+        stop("The tree must have a root time element (tree$root.time).")
+    }
+
+    # ## What
+    # check.method(what, c("Start", "End", "Range", "Midpoint"), msg = "'what' argument")
+
+    # ## Type
+    # check.method(type, c("Age", "Eon", "Epoch", "Era", "Period"), msg = "'type' argument")
+
+    # ## ICS
+    # check.class(ICS, "numeric")
+    # check.length(ICS, 1, errorif = FALSE, msg = " must be a year from 2008 to the present.")
+    ICS_available <- match(paste0("ICS", ICS), names(geoscale::timescales))
+    if(is.na(ICS_available)) {
+        stop(paste0("No stratigraphic data found for ", paste0("ICS", ICS), ".\nAvailable years are: ", paste(names(geoscale::timescales), collapse = ", "), "."))
+    } else {
+        ## Selecting the ICS
+        stratigraphy <- geoscale::timescales[[paste0("ICS", ICS)]]
+        ## Selecting the column of interest
[…]

### 2017-11-24 functions/tree.age.fossil.R

@@ -22,23 +22,49 @@
 tree.age.fossil <- function(tree, age, order = 'past'){
 
 #SANITYZING
+#FUNCTIONS FOR tree.age
 
-    #tree
-    check.class(tree, 'phylo', ' must be a phylo object.')
+#Extract the ages table from a tree
+#Calculating the tips and the elements age
+tree.age_table<-function(tree){
+    tree_length <- Ntip(tree)
+    
+    ages <- dist.nodes(tree)[tree_length + 1,]
+    tip.names <- tree$tip.label
+    if(is.null(tree$node.label)) {
+        nod.names <- c((tree_length + 1):length(dist.nodes(tree)[, 1]))
+    } else {
+        nod.names <- tree$node.label
+    }
+    elements <- c(tip.names, nod.names)
+    ages.table <- data.frame(ages = ages,elements = elements)
+    return(ages.table)
+}
 
-    #age
+
+#Scaling the ages from tree.age_table if scale != 1
+tree.age_scale <- function(ages.table, scale){
+    ages.table$ages <- ages.table$ages/max(ages.table$ages)
+    ages.table$ages <- ages.table$ages*scale
+    return(ages.table)
+}
+
+    # #tree
+    # check.class(tree, 'phylo', ' must be a phylo object.')
+
+    # #age
     if(missing(age)) {
         #Using the tree height as age if age is missing
         age <- max(dist.nodes(tree)[, Ntip(tree)+1])
     }
-    check.class(age, 'numeric', ' must be a numerical value.')
-    check.length(age, '1', ' must a a single value.')
+    # check.class(age, 'numeric', ' must be a numerical value.')
+    # check.length(age, '1', ' must a a single value.')
 
-    #order
-    check.method(order, c("past", "present"), "order argument")
+    # #order
+    # check.method(order, c("past", "present"), "order argument")
 
-    ## Fossils only
-    check.class(fossil.only, "logical")
+    # ## Fossils only
+    # check.class(fossil.only, "logical")
 
     #CALCULATE THE EDGES AGE
 

### 2017-11-24 analyses/01-extract-data-for-analyses.Rmd
Quick fixes for Wright's tree and FADLAD

@@
 ```{r}
+source("../functions/tree.age.fossil.R")
+
 slug <- "Wright2017"
@@
 matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
-FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+# FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+#TG: The FADLAD ages seems pretty weird, the data should be Ordovician/Devonian (-485/-360) and the range in these FADLADs is (-113/-10)!
+#TG: I've thus replaced it by a dummy FADLAD for now (just the taxa age) (see line ~169 after root.time)
 tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
@@
 # Multiply branches by 0.03517385
-tree$edge.length <- tree$edge.length * 0.03517385
+tree$edge.length <- tree$edge.length / 0.03517385 #TG: following the note above, it's divide not multiply (i.e. increase branch length because rate < 0)
 
@@
 tree$root.time <- 485.4
+
+# Dummy FADLAD
+tree_ages <- tree.age.fossil(tree)[1:Ntip(tree),]
+FADLAD <- as.data.frame(matrix(c(tree_ages$ages, tree_ages$ages), ncol = 2, byrow = FALSE, dimnames = list(as.character(tree_ages$elements), c("FAD", "LAD"))))
 

### 2017-11-24 analyses/02-time-slice-analyses-Wright2017.Rmd
Minor edits
Quick fixes for Wright's tree and FADLAD

@@
 
-## The latest version of dispRity on master branch
-## install_github("TGuillerme/dispRity", ref = "master")
+## The latest version of dispRity on release branch
+## install_github("TGuillerme/dispRity", ref = "release")
 library(dispRity)
@@
 source("../functions/simulations_fun.R")
+source("../functions/tree.age.fossil.R")
 ```
@@
 ## Loading the FADLAD
-FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+# FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+```
 
-## Loading the tree
-tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
-## Adding node labels and a root age (max tree age date)
-tree <- makeNodeLabel(tree, method = "number", prefix = "n")
-tree$root.time <- 485.4 # specific to Wright 2017 tree
+TG: There's a problem with the FADLAD (too recent! - see 01-extract-data-for-analysis.Rmd lines ~150).
+Also, I jus realised that the operations here are done twice, once in the extract-data file and once here.
+The clever way would be to do everything in the extract-data (e.g. the root time, rescaling etc...) so that only the time slice analysis is done here.
+We could export the data bundle (i.e. the `morphospace`, `tree` and `FADLAD`) in a `.Rda` format and simply load it here (so we're sure we have the correct root time etc.).
+
+```{r}
+#TG: Tree tinkering from 01-extract-data-for-analysis.Rmd
+
+# Replace zero branches with minimu branch length
+tree$edge.length[which(tree$edge.length == 0)] <- 
+  min(tree$edge.length[-which(tree$edge.length == 0)])
+
+# Multiply branches by 0.03517385
+tree$edge.length <- tree$edge.length / 0.03517385 #TG: following the note above, it's divide not multiply (i.e. increase branch length because rate < 0)
+
+# Add root time: 485.4
+tree$root.time <- 485.4
+
+# Dummy FADLAD
+tree_ages <- tree.age.fossil(tree)[1:Ntip(tree),]
+FADLAD <- as.data.frame(matrix(c(tree_ages$ages, tree_ages$ages), ncol = 2, byrow = FALSE, dimnames = list(as.character(tree_ages$elements), c("FAD", "LAD"))))
+
+# ## Loading the tree
+# tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
+# ## Adding node labels and a root age (max tree age date)
+# tree <- makeNodeLabel(tree, method = "number", prefix = "n")
+# tree$root.time <- 485.4 # specific to Wright 2017 tree
 

### 2017-11-24 analyses/02-time-slice-analyses.Rmd
updated function/arguments names for 0.4.2

@@
 > TG: for the following pipeline, I've reduced the number of rarefaction that highly increased the calculation time.
-We now only calculate the disparity for the non-rarefied subsamples and the subsamples rarefied to contain only 3 taxa (the lowest level of rarefaction).
+We now only calculate the disparity for the non-rarefied subsets and the subsets rarefied to contain only 3 taxa (the lowest level of rarefaction).
 We can use the full rarefaction if we want to plot rarefaction curves later on.
@@
 ## Set the list of available models
-models <- list("equalbins", "acctran", "deltran", "punctuated", "gradual")
+models <- list("equalbins", "acctran", "deltran", "random", "proximity")
 
@@
 ## Adding the legend
-legend(x = "bottomleft", legend = c("time bins", "acctran", "deltran", "punctuated", "gradual"),
+legend(x = "bottomleft", legend = c("time bins", "acctran", "deltran", "random", "proximity"),
   col = colors[1:5], lty = 1, bg = "white")

### 2017-11-24 functions/simulations_fun.R

@@ -23,7 +23,7 @@ get.bins <- function(tree, type){
 }
 
 ##---------------------------------------------------------------------------
-## Extract stratigraphy subsamples
+## Extract stratigraphy subsets
 ## morphospace is the cleaned morphospace
 ## tree is the phylogeny matched to the morphospace
 ## FADLAD is first and last occurrence data
@@ -31,14 +31,14 @@ get.bins <- function(tree, type){
 ## type is "Age" or "Epoch" - see get.bin.ages documentation
 ##--------------------------------------------------------------------------
 
-get.subsamples.stratigraphy <- function(morphospace, tree, FADLAD, inc.nodes, type){
-  time.subsamples(data = morphospace, tree = tree, 
+get.subsets.stratigraphy <- function(morphospace, tree, FADLAD, inc.nodes, type){
+  time.subsets(data = morphospace, tree = tree, 
                   method = "discrete", time = get.bins(tree, type)[[1]], 
                   FADLAD = FADLAD, inc.nodes = inc.nodes)
 }  
   
 ##---------------------------------------------------------------------------
-## Extract all types of subsample for both methods and four continuous models
+## Extract all types of subset for both methods and four continuous models
 ## morphospace is the cleaned morphospace
 ## tree is the phylogeny matched to the morphospace
 ## bins is a vector of bin times or number of bins
@@ -46,10 +46,10 @@ get.subsamples.stratigraphy <- function(morphospace, tree, FADLAD, inc.nodes, ty
 ## inc.nodes is logical for whether to estimate disparity using nodal values
 ##--------------------------------------------------------------------------
 
-get.subsamples <- function(morphospace, tree, bins, FADLAD, inc.nodes){
+get.subsets <- function(morphospace, tree, bins, FADLAD, inc.nodes){
   
   ## Subset samples by required number of time bins
-  subsamples.bins <- time.subsamples(data = morphospace, tree = tree, 
+  subsets.bins <- time.subsets(data = morphospace, tree = tree, 
   	                                 method = "discrete", time = bins, 
   	                                 FADLAD = FADLAD, inc.nodes = inc.nodes)
   
@@ -69,53 +69,53 @@ get.subsamples <- function(morphospace, tree, bins, FADLAD, inc.nodes){
   }
 
   ## Subset samples by required number of time slices with acctran model
-  subsamples.acctran <- time.subsamples(data = morphospace, tree = tree, 
[…]

### 2017-11-27 functions/simulations_fun.R
Updates to simulations fun

@@ -63,7 +63,7 @@ get.subsets <- function(morphospace, tree, bins, FADLAD, inc.nodes){
     ## Removing the last bin (that would be a negative slice)
     slices <- bins[-length(bins)]
     ## Getting the centre of the bins as slicing values
-    slices <- slices - (diff(bins)/2)
+    slices <- slices + (diff(bins)/2)
   } else {
     slices <- bins
   }
@@ -82,9 +82,9 @@ get.subsets <- function(morphospace, tree, bins, FADLAD, inc.nodes){
   
   ## Subset samples by required number of time slices with random model
   subsets.random <- time.subsets(data = morphospace, tree = tree, 
-  	                                       method = "continuous", time = slices, 
-  	                                       FADLAD = FADLAD, inc.nodes = inc.nodes,
-                                           model = "random")
+  	                                     method = "continuous", time = slices, 
+  	                                     FADLAD = FADLAD, inc.nodes = inc.nodes,
+                                         model = "random")
   
   ## Subset samples by required number of time slices with proximity model
   subsets.proximity <- time.subsets(data = morphospace, tree = tree, 
@@ -92,8 +92,20 @@ get.subsets <- function(morphospace, tree, bins, FADLAD, inc.nodes){
   	                                    FADLAD = FADLAD, inc.nodes = inc.nodes,
   	                                    model = "proximity")
 
-  return(list(subsets.bins, subsets.acctran, subsets.deltran, 
-  	          subsets.random, subsets.proximity))
+  # ## Subset samples by required number of time slices with proximity model
+  # subsets.punctuated <- time.subsets(data = morphospace, tree = tree, 
+  #                                       method = "continuous", time = slices, 
+  #                                       FADLAD = FADLAD, inc.nodes = inc.nodes,
+  #                                       model = "punctuated")
+
+  # ## Subset samples by required number of time slices with proximity model
+  # subsets.gradual <- time.subsets(data = morphospace, tree = tree, 
+  #                                       method = "continuous", time = slices, 
+  #                                       FADLAD = FADLAD, inc.nodes = inc.nodes,
+  #                                       model = "gradual")
+
+  return(list(subsets.bins, subsets.acctran, subsets.deltran, subsets.random,
+              subsets.proximity))#, subsets.punctuated, subsets.gradual))
 }
 
 ##--------------------------------------------------------------------------
[…]

### 2017-11-27 analyses/01-extract-data-for-analyses.Rmd
updated function/arguments names for 0.4.2
Remove FAD LAD - it isn't needed til the analyses
Remove FAD LADs and ordination code

--- 2017-11-23 analyses/01-extract-data-for-analyses.Rmd
@@
 
-Quick pipeline for getting data from discrete matrices.
+Pipeline for getting data from discrete matrices.
 ===============
@@
 matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
-
-## Loading the FADLAD
-FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
 
@@
 matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
-FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+
 tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
@@
 # Multiply branches by 0.03517385
-tree$edge.length <- tree$edge.length * 0.03517385
+tree$edge.length <- tree$edge.length / 0.03517385
 
@@
 
-## Ordinating the matrix - no need to run this here, we will do this in the next script
-## load(paste0("../data/processed/distance_matrix.", slug,".Rda"))
-## morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
 ```
@@
 matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
-FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
-# Note that these are the upper bound for FAD and lower bound for LAD
+
 tree <- read.nexus(paste0("../data/trees/", slug, ".nex"))
@@
 
-## Ordinating the matrix - no need to run this here, we will do this in the next script
-## load(paste0("../data/processed/distance_matrix.", slugs[slug] ,".Rda"))
-## morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
 ```
@@
 matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
-FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+
 tree <- read.nexus(paste0("../data/trees/", slug, ".nex"))

### 2017-11-27 analyses/02-time-slice-analyses.Rmd
Remove examining output section and save otuputs instead

@@
 ## Load functions using source
-# Doesn't always work... Rmd problem?
 source("../functions/simulations_fun.R")
@@
 To run `dispRity` analyses we need the **morphospace**, the **phylogeny** (with a root age and node labels) and the **first and last occurrence dates** (FADLAD).
-The tree and FADLAD are in the `data/` folder.
+The FADLAD is in the `data/` folder, the tree with node labels and cleaned of taxa not in the morphospace is in the `data/processed` folder.
 To construct the morphospace we need to use the `distance_matrix` files we created in `01-extract-data-for analyses.Rmd` and stored in `data/processed`.
@@
 
-## Loading the discrete morphological matrix
-## This is just to match the names with the tree
-matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
-
 ## Loading the FADLAD
@@
 ## Loading the tree
-tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
-## Adding node labels and a root age (max tree age date)
-tree <- makeNodeLabel(tree, method = "number", prefix = "n")
+## See "01-extract-data-for analyses.Rmd" for how we cleaned these
+tree <- read.nexus(paste0("../data/processed/tree.", slug, ".tre"))
+## Add root age (max tree age date)
 tree$root.time <- max(tree.age(tree)[, 1])
+```
 
-## Cleaning the tree to remove things that aren't in the matrix 
-cleaned_data <- clean.data(matrix$matrix, tree)
-matrix$matrix <- cleaned_data$matrix
-tree <- cleaned_data$tree
+Ordinating the matrix
+----------------
 
+Ordination is a classic NMDS (multidimensional scaling) including only the non-zero variance axes.
+
+```{r}
 ## Loading the distance matrices
@@
 morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
-
 ```
@@
 
-These analyses use a series of functions in `functions\`. These essentially wrap the existing `dispRity` functions so that running lots of models and extracting lots of data is not such a big issue.
-
-> TG: for the following pipeline, I've reduced the number of rarefaction that highly increased the calculation time.
-We now only calculate the disparity for the non-rarefied subsets and the subsets rarefied to contain only 3 taxa (the lowest level of rarefaction).
-We can use the full rarefaction if we want to plot rarefaction curves later on.
+These analyses use a series of functions in `functions\`.
+These essentially wrap the existing `dispRity` functions so that running lots of models and extracting lots of data is not such a big issue.
+The function `run.all.disparity.wrapper` will run all the scenarios above.
 
@@
 ## Run the function for Beck data
-slug <- "BeckLee"
+slug <- "Beck2014"
 
@@
                                 metric.name = "sum_var", rarefaction = FALSE)
-#readr::write_csv(out1, path = paste0("../outputs/age", "_", slug, ".csv"))
+
+## Save results to output folder
+save(out1, file = paste0("../outputs/age.", "Beck2014" ,".Rda"))
 
@@
                         rarefaction = FALSE)
-#readr::write_csv(out2, path = paste0("../outputs/epoch", "_", slug, ".csv"))
+
+## Save results to output folder
+save(out2, file = paste0("../outputs/epoch.", slug ,".Rda"))
 ```
-
-Note that you can use the previous output format (the data tables) as before by accessing them through `out$results$stratigraphy/duration/number`.
-Writing them as a csv will require some thinking since they don't have the same number of columns (though it's still doable I guess).
-Also, if you want to plot things through `ggplot`, you can use the values in these data tables!
-
-Examining the output 
---------------------
-
-The results as a table
-
-```{r, eval = TRUE, echo = FALSE}
-knitr::kable(out2$results$stratigraphy, caption = "Disparity per stratigraphic time: unequal time bins and non-equidistant time slices")
-knitr::kable(out2$results$duration, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the median stratigraphic epoch duration")
-knitr::kable(out2$results$number, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the number of stratigraphic epochs")
-```
-
-This is not really digest and we should find a better way to present that.
-Anyway, these boring tables are always good for the supplementaries.
-
-## Plotting the results `dispRity` style
-
-Here is an example of the results using `plot.dispRity`.
-The upper panel is with the CIs and the lower one is the same without the CIs.
-
-### The results for the Age type
-
-```{r, eval = TRUE, fig.width = 12, fig.height = 18, echo = TRUE}
-op <- par(bty = "n", mfrow = c(3,2))
-## Principal colors
-colors <- palette()[1:5][-1] #TG: that can be made fancier
-colors[5] <- "black"
-## Color combination (for the CI)
-colors_CI <- c("pink", "lightgreen", "lightblue", "lightcyan", "lightgray")
-## y limits
-ylim <- range(unlist(lapply(out1$results, function(X) return(c(min(X[,5], na.rm = TRUE), max(X[,5],
-  na.rm = TRUE))))))
-## Zooming
-ylim_zoom <- c(20, 35)
-
-## Plotting the different methods (with/without CIs)
-plot.results.dispRity(out1$object, method = 1, colors = colors, colors_CI = colors_CI,
-  main = "Stratigraphy", ylim = ylim, CI = TRUE, legend = TRUE)
-plot.results.dispRity(out1$object, method = 1, colors = colors, colors_CI = colors_CI,
-  main = "Stratigraphy (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)
-
-plot.results.dispRity(out1$object, method = 2, colors = colors, colors_CI = colors_CI,
-  main = "Duration", ylim = ylim, CI = TRUE)
-plot.results.dispRity(out1$object, method = 2, colors = colors, colors_CI = colors_CI,
-  main = "Duration (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)
-
-plot.results.dispRity(out1$object, method = 3, colors = colors, colors_CI = colors_CI,
-  main = "Number", ylim = ylim, CI = TRUE)
-plot.results.dispRity(out1$object, method = 3, colors = colors, colors_CI = colors_CI,
-  main = "Number (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)
-```
-
-### The results for the Epoch type
-
-```{r, eval = TRUE, fig.width = 12, fig.height = 18, echo = TRUE}
-op <- par(bty = "n", mfrow = c(3,2))
-## Principal colors
-colors <- palette()[1:5][-1] #TG: that can be made fancier
-colors[5] <- "black"
-## Color combination (for the CI)
-colors_CI <- c("pink", "lightgreen", "lightblue", "lightcyan", "lightgray")
-## y limits
-ylim <- range(unlist(lapply(out2$results, function(X) return(c(min(X[,5], na.rm = TRUE),
-  max(X[,5], na.rm = TRUE))))))
-
-## Zooming
-ylim_zoom <- c(20, 35)
-
-## Plotting the different methods (with/without CIS)
-plot.results.dispRity(out2$object, method = 1, colors = colors, colors_CI = colors_CI,
-  main = "Stratigraphy", ylim = ylim, CI = TRUE, legend = TRUE)
-plot.results.dispRity(out2$object, method = 1, colors = colors, colors_CI = colors_CI,
-  main = "Stratigraphy (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)
-
-plot.results.dispRity(out2$object, method = 2, colors = colors, colors_CI = colors_CI,
-  main = "Duration", ylim = ylim, CI = TRUE)
-plot.results.dispRity(out2$object, method = 2, colors = colors, colors_CI = colors_CI,
-  main = "Duration (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)
-
-plot.results.dispRity(out2$object, method = 3, colors = colors, colors_CI = colors_CI,
-  main = "Number", ylim = ylim, CI = TRUE)
-plot.results.dispRity(out2$object, method = 3, colors = colors, colors_CI = colors_CI,
-  main = "Number(no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)
-
-```
-
-## Plotting the results histogram style
-
-Another way for the plots would be kind of our histogram representation (you can certainly use `ggplot` here, below is just an example):
-
-### The results per method
-
-```{r, fig.width = 12, fig.height = 18}
-## Set the list of available models
-models <- list("equalbins", "acctran", "deltran", "random", "proximity")
-
-## Get the stratigraphic, duration and number data for the ages
-strat_data_ages <- lapply(models, make.class.histogram, out1$results, method = 1)
-durat_data_ages <- lapply(models, make.class.histogram, out1$results, method = 2)
-numbe_data_ages <- lapply(models, make.class.histogram, out1$results, method = 3)
-
-## Get the stratigraphic, duration and number data for the epochs
-strat_data_epoc <- lapply(models, make.class.histogram, out2$results, method = 1)
-durat_data_epoc <- lapply(models, make.class.histogram, out2$results, method = 2)
-numbe_data_epoc <- lapply(models, make.class.histogram, out2$results, method = 3)
-
-## The lower age and the lower disparity limit are arbitrarily fixed to 0 and 15 respectively.
-lower_age <- 0; lower_lim <- 15
-#TG: might want to change that for other datasets!
-
-## limits functions
-get.break <- function(x) return(x$breaks)
-get.count <- function(x) return(x$counts)
-## Ages limits
-xlim_ages <- c(max(unlist(c(sapply(strat_data_ages, get.break), sapply(durat_data_ages, get.break),
-  sapply(numbe_data_ages, get.break)))), lower_age)
-ylim_ages <- c(lower_lim, max(unlist(c(sapply(strat_data_ages, get.count),
-  sapply(durat_data_ages, get.count), sapply(numbe_data_ages, get.count)))))
-## Epoch limits
-xlim_epoc <- c(max(unlist(c(sapply(strat_data_epoc, get.break), sapply(durat_data_epoc, get.break),
-  sapply(numbe_data_epoc, get.break)))), lower_age)
-ylim_epoc <- c(lower_lim, max(unlist(c(sapply(strat_data_epoc, get.count),
-  sapply(durat_data_epoc, get.count), sapply(numbe_data_epoc, get.count)))))
-
-## Colors
-colors <- palette()[1:5]
-colors[1] <- "grey"
-
-## Graphics
-op <- par(bty = "n", mfrow = c(3,2))
-
-## Plot the stratigraphy - ages
-plot(strat_data_ages[[1]], xlim = xlim_ages, ylim = ylim_ages , xlab = "Time (Mya)",
-  ylab = "Sum of variance", col = colors[1], main = "Stratigraphy - ages")
-## Plot the slicing
-for(slice in 2:5) {
-  lines(strat_data_ages[[slice]]$breaks, strat_data_ages[[slice]]$counts, col = colors[slice])
-}
-
-## Adding the legend
-legend(x = "bottomleft", legend = c("time bins", "acctran", "deltran", "random", "proximity"),
-  col = colors[1:5], lty = 1, bg = "white")
-
-## Plot the stratigraphy - epoch
-plot(strat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
-  ylab = "Sum of variance", col = colors[1], main = "Stratigraphy - Epoch")
-## Plot the slicing
-for(slice in 2:5) {
-  lines(strat_data_epoc[[slice]]$breaks, strat_data_epoc[[slice]]$counts, col = colors[slice])
-}
-
-## Plot the duration - ages
-plot(durat_data_ages[[1]], xlim = xlim_ages, ylim = ylim_ages , xlab = "Time (Mya)",
-  ylab = "Sum of variance", col = colors[1], main = "Duration - ages")
-## Plot the slicing
-for(slice in 2:5) {
-  lines(durat_data_ages[[slice]]$breaks, durat_data_ages[[slice]]$counts, col = colors[slice])
-}
-
-## Plot the duration - epoch
-plot(durat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
-  ylab = "Sum of variance", col = colors[1], main = "Duration - Epoch")
-## Plot the slicing
-for(slice in 2:5) {
-  lines(durat_data_epoc[[slice]]$breaks, durat_data_epoc[[slice]]$counts, col = colors[slice])
-}
-
-## Plot the number - ages
-plot(numbe_data_ages[[1]], xlim = xlim_ages, ylim = ylim_ages , xlab = "Time (Mya)",
-  ylab = "Sum of variance", col = colors[1], main = "Number - ages")
-## Plot the slicing
-for(slice in 2:5) {
-  lines(numbe_data_ages[[slice]]$breaks, numbe_data_ages[[slice]]$counts, col = colors[slice])
-}
-
-## Plot the number - epoch
-plot(numbe_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
-  ylab = "Sum of variance", col = colors[1], main = "Number - Epoch")
-## Plot the slicing
-for(slice in 2:5) {
-  lines(numbe_data_epoc[[slice]]$breaks, numbe_data_epoc[[slice]]$counts, col = colors[slice])
-}
-```
-
-Some of the plots are a bit off set and everything but it gives us an idea
-
-### The results per model
-
-Something like that but with transparency:
-
-```{r, fig.width = 6, fig.height = 6}
-plot(strat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
-  ylab = "Sum of variance", col = colors[1], main = "Epoch")
-plot(durat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
-  ylab = "Sum of variance", col = colors[2], main = "", add = TRUE)
-plot(numbe_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
-  ylab = "Sum of variance", col = colors[3], main = "", add = TRUE)
-```
-
-## Comparing the disparity through time curves
-
-```{r}
-curve1 <- out2$object[[1]][[1]]
-curve2 <- out2$object[[1]][[2]]
-curve3 <- out2$object[[1]][[3]]
-curve4 <- out2$object[[1]][[4]]
-curve5 <- out2$object[[1]][[5]]
-
-```

### 2017-11-27 analyses/01-extract-data-for-analyses.Rmd
Remove non needed data etc.

@@
 The analyses require a NEXUS matrix of characters, a csv file of first and last occurrence dates (FADLAD) and a phylogeny.
+The matrix and phylogeny need to be cleaned and manipulated - we do this below.
+The FADLAD can be read in as is so doesn't get read in until we need it for analyses in later scripts.
 
@@
 ## Adding a root age (max tree age date)
+## Note that we need to redo this in the `02-time-slice-analyses` scripts
+## as write.nexus/tree won't save root times
 tree$root.time <- max(tree.age(tree)[, 1])
@@
 as.vector(na.omit(c(cleaned_data$dropped_tips, cleaned_data$dropped_rows)))
+
+## Saving the tree
+write.nexus(tree, file = paste0("../data/processed/tree.", slug ,".tre"))
 ```
@@
 
-Ordinating the matrix
-----------------
-
-Ordination is a classic NMDS (multidimensional scaling) including only the non-zero variance axes.
-
-```{r}
-## Loading the distance matrices
-load(paste0("../data/processed/distance_matrix.", slug ,".Rda"))
-
-## Ordinating the matrix
-morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
-```
-
 And that's it...
-we now have all the arguments for running the `dispRity` package properly: the space (`morphospace`), the tree (`tree`) and the FADLAD (`FADLAD`).
+we now have all the processed data for running the `dispRity` package properly: the distance matrices(that we will use to build the `morphospace`), and the cleaned tree (`tree`).
 Remember that for the `time.subsamples` function, you can use the `inc.nodes` logical option to include nodes or not!
@@
 
+## Removed taxa, i.e. those in the tree but not the matrix and vice versa
+as.vector(na.omit(c(cleaned_data$dropped_tips, cleaned_data$dropped_rows)))
+
+## Saving the tree
+write.nexus(tree, file = paste0("../data/processed/tree.", slug ,".tre"))
+
 ## Estimating the ancestral states (with a uncertainty threshold of 0.95)
@@
 save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
-
 ```

### 2017-11-27 analyses/01-extract-data-for-analyses.Rmd
Replace save and load with saveRDS and loadRDS, to prevent environment names issues

@@
 ## Saving the matrix
-save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+saveRDS(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 ```
@@
 ## Loading the matrix
-load(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 
@@
 ## Saving the matrix
-save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+saveRDS(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
 ```
@@
                                   uncertainty.threshold = 0.95)
-save(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+saveRDS(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 
 ## Add ancestral states to nodes
-load(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 row.names(matrix_nodes) <- tree$node.label
@@
 matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
-save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug,".Rda"))
+saveRDS(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug,".Rda"))
 
@@
 matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
-save(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+saveRDS(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
 ```

### 2017-11-27 analyses/02-time-slice-analyses.Rmd
Replace save and load with saveRDS and loadRDS, to prevent environment names issues

@@
 ## See "01-extract-data-for analyses.Rmd" for how we made these
-load(paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+readRDS(paste0("../data/processed/distance_matrix.", slug ,".Rda"))
 
@@
 ## Save results to output folder
-save(out1, file = paste0("../outputs/age.", "Beck2014" ,".Rda"))
+saveRDS(out1, file = paste0("../outputs/age.", "Beck2014" ,".Rda"))
 
@@
 ## Save results to output folder
-save(out2, file = paste0("../outputs/epoch.", slug ,".Rda"))
+saveRDS(out2, file = paste0("../outputs/epoch.", slug ,".Rda"))
 ```

### 2017-11-27 analyses/01-extract-data-for-analyses.Rmd
Tidy up code

@@
 ```{r}
-source("../functions/tree.age.fossil.R")
-
 slug <- "Wright2017"
 
-## Loading the data
+## Loading the data and tree
 matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
+tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
 
-tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
+# Replace zero branches with minimum branch length
+tree$edge.length[which(tree$edge.length == 0)] <- 
+  min(tree$edge.length[-which(tree$edge.length == 0)])
 
@@
 Note that to properly timescale the tree [e.g., in R] you must first divide the branch lengths by the corresponding clock rate, (= 0.03517385), and then set the root.time to 485.4]
-
-# Replace zero branches with minimu branch length
-tree$edge.length[which(tree$edge.length == 0)] <- 
-  min(tree$edge.length[-which(tree$edge.length == 0)])
-
-# Multiply branches by 0.03517385
+# Divide branches by 0.03517385
 tree$edge.length <- tree$edge.length / 0.03517385
@@
 
-## Loading the data
+## Loading the data and tree
 matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
+tree <- read.nexus(paste0("../data/trees/", slug, ".nex"))
 
-tree <- read.nexus(paste0("../data/trees/", slug, ".nex"))
+# Replace zero branches with minimum branch length
+tree$edge.length[which(tree$edge.length == 0)] <- 
+  min(tree$edge.length[-which(tree$edge.length == 0)])
 
@@
 
-## Loading the data
+## Loading the data and tree
 matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
+tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
 
-tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
+# Replace zero branches with minimum branch length
+tree$edge.length[which(tree$edge.length == 0)] <- 
+  min(tree$edge.length[-which(tree$edge.length == 0)])
 

### 2017-11-27 analyses/02-time-slice-analyses.Rmd
Add time slice analyses for all datasets

@@
 ```{r, eval = TRUE, warning = FALSE}
-## Run the function for Beck data
-slug <- "Beck2014"
 
@@
 ## Save results to output folder
-saveRDS(out1, file = paste0("../outputs/age.", "Beck2014" ,".Rda"))
+saveRDS(out1, file = paste0("../outputs/age.", slug ,".Rda"))
 
@@
 ```
+
+Running the analyses for other datasets
+===================================================
+Ideally we'd have written functions to do this, and/or a big old loop. But each dataset was slightly different so we've dealt with each individually.
+
+Wright 2017
+-----------
+```{r}
+## Choose a slug
+slug <- "Wright2017"
+
+## Loading the FADLAD
+FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+
+## Loading the tree
+## See "01-extract-data-for analyses.Rmd" for how we cleaned these
+tree <- read.nexus(paste0("../data/processed/tree.", slug, ".tre"))
+## Add root age as defined in Wright 2017
+tree$root.time <- 484.4
+
+## Loading the distance matrices
+## See "01-extract-data-for analyses.Rmd" for how we made these
+readRDS(paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+
+## Ordinating the matrix
+morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
+
+## Run the disparity analyses
+## Ages (rarefied to 3 taxa only)
+out1 <- run.all.disparity.wrapper(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
+                                bootstraps = 100, metric = c(sum, variances),
+                                metric.name = "sum_var", rarefaction = FALSE)
+
+## Save results to output folder
+saveRDS(out1, file = paste0("../outputs/age.", slug ,".Rda"))
+
+## Epochs (rarefied to 3 taxa only)
+out2 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
+                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var",
+                        rarefaction = FALSE)
+
+## Save results to output folder
+saveRDS(out2, file = paste0("../outputs/epoch.", slug ,".Rda"))
+```
+
+Bapst et al 2016 Biol Letters
+-----------
+
+```{r}
+slug <- "Bapst2016"
+
+## Loading the FADLAD
+FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+
+## Loading the tree
+## See "01-extract-data-for analyses.Rmd" for how we cleaned these
+tree <- read.nexus(paste0("../data/processed/tree.", slug, ".tre"))
+## Add root age (max tree age date)
+tree$root.time <- max(tree.age(tree)[, 1])
+
+## Loading the distance matrices
+## See "01-extract-data-for analyses.Rmd" for how we made these
+readRDS(paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+
+## Ordinating the matrix
+morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
+
+## Run the disparity analyses
+## Ages (rarefied to 3 taxa only)
+out1 <- run.all.disparity.wrapper(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
+                                bootstraps = 100, metric = c(sum, variances),
+                                metric.name = "sum_var", rarefaction = FALSE)
+
+## Save results to output folder
+saveRDS(out1, file = paste0("../outputs/age.", slug ,".Rda"))
+
+## Epochs (rarefied to 3 taxa only)
+out2 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
+                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var",
+                        rarefaction = FALSE)
+
+## Save results to output folder
+saveRDS(out2, file = paste0("../outputs/epoch.", slug ,".Rda"))
+```
+
+Heath et al bears
+-----------
+
+```{r}
+slug <- "bears"
+
+## Loading the FADLAD
+FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+
+## Loading the tree
+## See "01-extract-data-for analyses.Rmd" for how we cleaned these
+tree <- read.nexus(paste0("../data/processed/tree.", slug, ".tre"))
+## Add root age (max tree age date)
+tree$root.time <- max(tree.age(tree)[, 1])
+
+## Loading the distance matrices
+## See "01-extract-data-for analyses.Rmd" for how we made these
+readRDS(paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+
+## Ordinating the matrix
+morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
+
+## Run the disparity analyses
+## Ages (rarefied to 3 taxa only)
+out1 <- run.all.disparity.wrapper(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
+                                bootstraps = 100, metric = c(sum, variances),
+                                metric.name = "sum_var", rarefaction = FALSE)
+
+## Save results to output folder
+saveRDS(out1, file = paste0("../outputs/age.", slug ,".Rda"))
+
+## Epochs (rarefied to 3 taxa only)
+out2 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
+                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var",
+                        rarefaction = FALSE)
+
+## Save results to output folder
+saveRDS(out2, file = paste0("../outputs/epoch.", slug ,".Rda"))
+```
+
+Brusatte et al 2014 Current Biol
+-----------
+
+```{r}
+slug <- "Brusatte2014"
+
+## Loading the FADLAD
+FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+
+## Loading the tree
+## See "01-extract-data-for analyses.Rmd" for how we cleaned these
+tree <- read.nexus(paste0("../data/processed/tree.", slug, ".tre"))
+## Add root age (max tree age date)
+tree$root.time <- max(tree.age(tree)[, 1])
+
+## Loading the distance matrices
+## See "01-extract-data-for analyses.Rmd" for how we made these
+readRDS(paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+
+## Ordinating the matrix
+morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
+
+## Run the disparity analyses
+## Ages (rarefied to 3 taxa only)
+out1 <- run.all.disparity.wrapper(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
+                                bootstraps = 100, metric = c(sum, variances),
+                                metric.name = "sum_var", rarefaction = FALSE)
+
+## Save results to output folder
+saveRDS(out1, file = paste0("../outputs/age.", slug ,".Rda"))
+
+## Epochs (rarefied to 3 taxa only)
+out2 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
+                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var",
+                        rarefaction = FALSE)
+
+## Save results to output folder
+saveRDS(out2, file = paste0("../outputs/epoch.", slug ,".Rda"))
+```

### 2017-11-27 analyses/01-extract-data-for-analyses.Rmd
Add names to readRDS objects

@@
 ## Estimating the ancestral states (with a uncertainty threshold of 0.95)
+## This is slow, so on future runs we recommend using the saved data
 matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
@@
 ## Loading the matrix
-readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+## Not needed if running all the code at once
+## matrix_nodes <- readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 
@@
 ## Add ancestral states to nodes
-readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+## matrix_nodes <- readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 row.names(matrix_nodes) <- tree$node.label

### 2017-11-28 functions/simulations_fun.R
Updated pipelines analysis to v0.4.2

@@ -92,20 +92,20 @@ get.subsets <- function(morphospace, tree, bins, FADLAD, inc.nodes){
   	                                    FADLAD = FADLAD, inc.nodes = inc.nodes,
   	                                    model = "proximity")
 
-  # ## Subset samples by required number of time slices with proximity model
-  # subsets.punctuated <- time.subsets(data = morphospace, tree = tree, 
-  #                                       method = "continuous", time = slices, 
-  #                                       FADLAD = FADLAD, inc.nodes = inc.nodes,
-  #                                       model = "punctuated")
+  ## Subset samples by required number of time slices with proximity model
+  subsets.punctuated <- time.subsets(data = morphospace, tree = tree, 
+                                        method = "continuous", time = slices, 
+                                        FADLAD = FADLAD, inc.nodes = inc.nodes,
+                                        model = "punctuated")
 
-  # ## Subset samples by required number of time slices with proximity model
-  # subsets.gradual <- time.subsets(data = morphospace, tree = tree, 
-  #                                       method = "continuous", time = slices, 
-  #                                       FADLAD = FADLAD, inc.nodes = inc.nodes,
-  #                                       model = "gradual")
+  ## Subset samples by required number of time slices with proximity model
+  subsets.gradual <- time.subsets(data = morphospace, tree = tree, 
+                                        method = "continuous", time = slices, 
+                                        FADLAD = FADLAD, inc.nodes = inc.nodes,
+                                        model = "gradual")
 
   return(list(subsets.bins, subsets.acctran, subsets.deltran, subsets.random,
-              subsets.proximity))#, subsets.punctuated, subsets.gradual))
+              subsets.proximity, subsets.punctuated, subsets.gradual))
 }
 
 ##--------------------------------------------------------------------------
@@ -167,11 +167,11 @@ run.all.disparity <- function(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
                               bootstraps = bootstraps, metric, metric.name,
[…]

### 2017-11-28 analyses/tmp-time-slice-analyses.Rmd
renamed tmp for pre-merge

--- 2017-11-09 analyses/02-time-slice-analyses.Rmd
@@
 > TG: for the following pipeline, I've reduced the number of rarefaction that highly increased the calculation time.
-We now only calculate the disparity for the non-rarefied subsamples and the subsamples rarefied to contain only 3 taxa (the lowest level of rarefaction).
+We now only calculate the disparity for the non-rarefied subsets and the subsets rarefied to contain only 3 taxa (the lowest level of rarefaction).
 We can use the full rarefaction if we want to plot rarefaction curves later on.
-
-```{r, eval = FALSE, echo = FALSE}
-## Testing data: TO REMOVE
-#TG: it's just faster this way
-#TG: also, I've only tested it on "Epoch" since it take ages to run, should be pipelined all right though
-data(BeckLee_mat99)
-data(BeckLee_tree)
-data(BeckLee_ages)
-
-morphospace <- BeckLee_mat99
-tree <- BeckLee_tree
-FADLAD <- BeckLee_ages
-
-```
 
@@
 ## Principal colors
-colors <- palette()[1:5][-1] #TG: that can be made fancier
-colors[5] <- "black"
+colors <- palette()[1:7][-1] #TG: that can be made fancier
+colors[7] <- "black"
 ## Color combination (for the CI)
-colors_CI <- c("pink", "lightgreen", "lightblue", "lightcyan", "lightgray")
+colors_CI <- c("pink", "lightgreen", "lightblue", "lightcyan", "lightpink", "beige", "lightgray")
 ## y limits
@@
 ## Principal colors
-colors <- palette()[1:5][-1] #TG: that can be made fancier
-colors[5] <- "black"
+colors <- palette()[1:7][-1] #TG: that can be made fancier
+colors[7] <- "black"
 ## Color combination (for the CI)
-colors_CI <- c("pink", "lightgreen", "lightblue", "lightcyan", "lightgray")
+colors_CI <- c("pink", "lightgreen", "lightblue", "lightcyan", "lightpink", "beige", "lightgray")
 ## y limits
-ylim <- range(unlist(lapply(out2$results, function(X) return(c(min(X[,5], na.rm = TRUE),
-  max(X[,5], na.rm = TRUE))))))
+ylim <- range(unlist(lapply(out2$results, function(X) return(c(min(X[,5], na.rm = TRUE), max(X[,5],
+  na.rm = TRUE))))))
 
@@
 ## Set the list of available models
-models <- list("equalbins", "acctran", "deltran", "punctuated", "gradual")
+models <- list("equalbins", "acctran", "deltran", "random", "proximity", "punctuated", "gradual")
 
@@
 ## Colors
-colors <- palette()[1:5]
+colors <- palette()[1:7]
 colors[1] <- "grey"
@@
 ## Plot the slicing
-for(slice in 2:5) {
+for(slice in 2:7) {
   lines(strat_data_ages[[slice]]$breaks, strat_data_ages[[slice]]$counts, col = colors[slice])
@@
 ## Adding the legend
-legend(x = "bottomleft", legend = c("time bins", "acctran", "deltran", "punctuated", "gradual"),
-  col = colors[1:5], lty = 1, bg = "white")
+legend(x = "bottomleft", legend = c("time bins", "acctran", "deltran", "random", "proximity"),
+  col = colors[1:7], lty = 1, bg = "white")
 
@@
 ## Plot the slicing
-for(slice in 2:5) {
+for(slice in 2:7) {
   lines(strat_data_epoc[[slice]]$breaks, strat_data_epoc[[slice]]$counts, col = colors[slice])
@@
 ## Plot the slicing
-for(slice in 2:5) {
+for(slice in 2:7) {
   lines(durat_data_ages[[slice]]$breaks, durat_data_ages[[slice]]$counts, col = colors[slice])
@@
 ## Plot the slicing
-for(slice in 2:5) {
+for(slice in 2:7) {
   lines(durat_data_epoc[[slice]]$breaks, durat_data_epoc[[slice]]$counts, col = colors[slice])
@@
 ## Plot the slicing
-for(slice in 2:5) {
+for(slice in 2:7) {
   lines(numbe_data_ages[[slice]]$breaks, numbe_data_ages[[slice]]$counts, col = colors[slice])
@@
 ## Plot the slicing
-for(slice in 2:5) {
+for(slice in 2:7) {
   lines(numbe_data_epoc[[slice]]$breaks, numbe_data_epoc[[slice]]$counts, col = colors[slice])

### 2017-11-28 analyses/02-time-slice-analyses.Rmd

@@
 ## See "01-extract-data-for analyses.Rmd" for how we made these
-readRDS(paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+matrix_dist <- readRDS(paste0("../data/processed/distance_matrix.", slug ,".Rda"))
 

### 2017-11-28 analyses/03-examine-outputs.Rmd

--- 2017-11-22 analyses/02-time-slice-analyses-Wright2017.Rmd
@@
 ---
-title: "Time slice analyses"
+title: "Outputs from time slice analyses"
 author: "Natalie Cooper and Thomas Guillerme"
@@
 
-This code runs the analyses in the paper.
-Below we lay out the process in detail for one dataset, and then provide the code for the other datasets below this.
-Note that much of this uses functions in the `simulations.R` file.
+This code runs the output analyses in the paper, i.e. creates tables, figures and statistical tests.
 
@@
 
-Make sure you work with `dispRity` 0.4 and TG's version of `Claddis` (much faster than GL's).
-
 ```{r, eval = TRUE, message = FALSE}
-# The latest version of devtools and ape
 library(devtools)
 library(ape)
-
-## The latest version of Claddis (on development on TG branch)
-## install_github("TGuillerme/Claddis")
-library(Claddis)
-
-## The latest version of dispRity on master branch
-## install_github("TGuillerme/dispRity", ref = "master")
+library(tidyverse)
+library(viridis)
 library(dispRity) 
-
-## Load functions using source
-# Doesn't always work... Rmd problem?
-source("../functions/simulations_fun.R")
 ```
@@
 
-To run `dispRity` analyses we need the **morphospace**, the **phylogeny** (with a root age and node labels) and the **first and last occurrence dates** (FADLAD).
-The tree and FADLAD are in the `data/` folder.
-To construct the morphospace we need to use the `distance_matrix` files we created in `01-extract-data-for analyses.Rmd` and stored in `data/processed`.
+This script requires the outputs from `02-time-slice-analyses.Rmd`
 
 ```{r}
-## Choose a slug
-slug <- "Wright2017"
-
-## Loading the discrete morphological matrix
-## This is just to match the names with the tree
-matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
-
-## Loading the FADLAD
-FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
-
-## Loading the tree
-tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
-## Adding node labels and a root age (max tree age date)
-tree <- makeNodeLabel(tree, method = "number", prefix = "n")
-tree$root.time <- 485.4 # specific to Wright 2017 tree
-
-## Cleaning the tree to remove things that aren't in the matrix 
-cleaned_data <- clean.data(matrix$matrix, tree)
-matrix$matrix <- cleaned_data$matrix
-tree <- cleaned_data$tree
-
-## Loading the distance matrices
-## See "01-extract-data-for analyses.Rmd" for how we made these
-load(paste0("../data/processed/distance_matrix.", slug ,".Rda"))
-
-## Ordinating the matrix
-morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
-
+slug <- "Beck2014"
+out1 <- readRDS(out1, file = paste0("../outputs/age.", slug, ".Rda"))
+out2 <- readRDS(out2, file = paste0("../outputs/age.", slug, ".Rda"))
 ```
-
-Scenarios
---------------------------------
-
-We used the data to explore these different scenarios:
-
-1. Stratigraphy. This is the traditional method, where all the species within each stratigraphic period are included in the disparity calculation. This often leads to bins of unequal duration.
-
-2. Equally sized time bins. 
-    i. Where the *duration* of the bin is equal to the median duration of the stratigraphic period.
-    ii. Where the *number* of bins is equal to the number of stratigraphic periods.
-    iii. Where the *number* of bins is either 2, 5, 10, 15, or 20.
-
-3. Time slicing. 
-    i. Where the *interval* between slices is equal to the median duration of the stratigraphic period.
-    ii. Where the *number* of slices is equal to the number of stratigraphic periods.
-    iii. Where the *number* of slices is either 2, 5, 10, 15, or 20.
-
-Running the analyses and saving the outputs
---------------------------------------------
-
-These analyses use a series of functions in `functions\`. These essentially wrap the existing `dispRity` functions so that running lots of models and extracting lots of data is not such a big issue.
-
-> TG: for the following pipeline, I've reduced the number of rarefaction that highly increased the calculation time.
-We now only calculate the disparity for the non-rarefied subsamples and the subsamples rarefied to contain only 3 taxa (the lowest level of rarefaction).
-We can use the full rarefaction if we want to plot rarefaction curves later on.
-
-```{r, eval = TRUE, warning = FALSE}
-## Ages (rarefied to 3 taxa only)
-out1 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
-                                bootstraps = 100, metric = c(sum, variances),
-                                metric.name = "sum_var", rarefaction = FALSE)
-#readr::write_csv(out1, path = paste0("../outputs/age", "_", slug, ".csv"))
-
-## Epochs (rarefied to 3 taxa only)
-out2 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
-                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var",
-                        rarefaction = FALSE)
-#readr::write_csv(out2, path = paste0("../outputs/epoch", "_", slug, ".csv"))
-```
-
-Note that you can use the previous output format (the data tables) as before by accessing them through `out$results$stratigraphy/duration/number`.
-Writing them as a csv will require some thinking since they don't have the same number of columns (though it's still doable I guess).
-Also, if you want to plot things through `ggplot`, you can use the values in these data tables!
-
-<!-- TG: I've not run this analysis below -->
-<!-- We also ran the models using varying numbers of bins as follows. -->
-
-```{r, eval = FALSE, echo = FALSE}
-## Run the function for Beck data
-slug <- "BeckLee"
-
-for (bins in c(2, 5, 10, 15, 20)){
-  output <- run.all.disparity(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
-                              bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var")
-
-  readr::write_csv(output, path = paste0("../outputs/bins", "_", slug, ".csv"), append = TRUE)
-}
-
-```
-
-<!-- NC: Need to think more about the outputs here, currently they don't make much sense as there are no column headers -->
 
@@
 --------------------
+
+Outputs can be viewed using `out1$results$stratigraphy`, `out1$results$duration`, or `out1$results$number`.
 

### 2017-11-28 analyses/01-extract-data-for-analyses.Rmd
Change .tre to .nex for Brusatte data

@@
 ## The latest version of dispRity 
-install_github("TGuillerme/dispRity", ref = "release")
+install_github("TGuillerme/dispRity")
 library(dispRity) 
@@
 we now have all the processed data for running the `dispRity` package properly: the distance matrices(that we will use to build the `morphospace`), and the cleaned tree (`tree`).
-Remember that for the `time.subsamples` function, you can use the `inc.nodes` logical option to include nodes or not!
+Remember that for the `time.subsets` function, you can use the `inc.nodes` logical option to include nodes or not!
 
@@
 matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
-tree <- read.nexus(paste0("../data/trees/", slug, ".tre"))
+tree <- read.nexus(paste0("../data/trees/", slug, ".nex"))
 

### 2017-11-29 analyses/note-on-NA-distances.Rmd
Added a note for distance matrices with missing data

---
title: "Note on NAs in distance matrices"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Quick note on the NAs in distance matrices

```{r}
library(Claddis)
library(dispRity)
```

These NAs come from taxa having no overlapping data.
Because we're using a Gower distance, data is not imputed when missing (c.f. Generalised Euclidean Distance or other methods that replace NA by the mean).
The solution is to use a Generalised Euclidean Distance or to remove the problematic taxa.

## Using Generalised Euclidean Distances-ish

This solution is the easiest and don't generate any loss in data.
It's not exactly equivalent to the GED described in Graeme's `?MorphDistMatrix` (or `MorphDistMatrixFast` - same but faster) which replaces NAs by the mean values _during_ the distance matrix calculation.
With this trick, the proper distance type is used (i.e. GED, Gower, etc...) and the mean of these distances is used as a replacement value.
In other words, when the distance between two taxa is incalculable, this trick assume that the distance is average.

This "compactise" the morphospace by adding more "average" shapes in it.
Not a big problem when there are a couple of incomparable taxa, a bit more (much more?) when there are many.

The trick goes like (for example, on the Brusatte one):

```{r}
slug <- "Brusatte2014"

## Getting the distance matrix
matrix_dist <- readRDS(paste0("../data/processed/distance_matrix.", slug ,".Rda"))
any(is.na(matrix_dist))

## Replacing them by the mean
matrix_dist <- ifelse(!is.na(matrix_dist), matrix_dist, mean(matrix_dist, na.rm = TRUE))
any(is.na(matrix_dist))
dim(matrix_dist)
```

## Trimming the morphological matrix

This method consists in removing the taxa with non-overlap morphological character _prior_ to the distance analysis.
We can either go brutally using Graeme's `TrimMorphDistMatrix` function:

```{r}
## Getting the distance matrix
matrix_dist <- readRDS(paste0("../data/processed/distance_matrix.", slug ,".Rda"))

## Matrix dimensions before trimming
dim(matrix_dist)

## Trimming the distance matrix
trimming <- TrimMorphDistMatrix(matrix_dist)
matrix_dist <- trimming$dist.matrix
any(is.na(matrix_dist))

## Matrix dimensions after trimming
dim(matrix_dist) # -19 taxa
as.vector(trimming$removed.taxa)
```

The other solution is to remove the problematic taxa before the distance calculation using the `SafeTaxonomicReduction` function.
This should go with the following algorithm:

  1. Remove the _m1_ taxa with no overlapping discrete character resulting in a matrix of (_n_ - _m1_) rows
  2. Run ancestral states reconstruction resulting in a matrix of ((_n_ - _m1_) + (_n_ - _m1_ - 1)) rows (if tree is fully binary)
  3. Calculate pairwise distances of size ((_n_ - _m1_) + (_n_ - _m1_ - 1))

However, because we're using a really safe ancestral states reconstruction method (i.e. only keep reconstruction with a scaled likelihood > 0.95), step 3. is still likely to generate NAs.
We thus need to run a while loop until the step 3. generates no NAs.
This will go like:

  4. Return to step 1 and with _m1_ =  _m1_ + _m2_ with _m2_ being the taxa contributing to NAs in step 3.

This is likely to result in a matrix with really few taxa.

## Differences

In the end, although the GEDish method will generate shittier morphopsaces with increasing missing data than a proper trimming method (Graeme have shown that at least when talking about NAs in the morphological matrix; c.f. the distance matrix); I think it's not a huge problem compared with how the distances are calculated in general.
In fact, using the `MorphDistMatrix` completely ignores inapplicable data (and treats them as NAs), this probably "compactise" the morphospace as well since one could argue that an inapplicable data should create negative multidimensional axis (for example on Gavin's beak PCO duck axis, species with no beaks should be stored a neither end of this axis but probably perpendicularly to it).
But that's an entire new field of pain that I'm just slowly coming out from so let's skip it.

Long story short: method one is quick and easy; method two is more correct but reduces the amount of data (and is not totally correct either).

### 2017-11-29 analyses/01-extract-data-for-analyses.Rmd
Add trimming to matrix to remove taxa that do not overlap

@@
 
+One final issue is that to commute Gower distances we need all taxa to have some overlapping data.
+If not Gower distances will be NA (Gower distance does not impute missing data like some other methods).
+To avoid this problem we can remove the taxa with no overlaps as follows.
+This is commented out below because there are no non overlapping taxa in the Beck and Lee 2014 dataset.
+
+```{r}
+## Trimming the distance matrix
+## trim <- TrimMorphDistMatrix(matrix_dist)
+## matrix_dist <- trim$dist.matrix
+
+## Which taxa were removed?
+## as.vector(trim$removed.taxa)
+
+## Rewrite this to file for later use
+## saveRDS(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+```
+
 And that's it...
@@
 matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
+
+## Trim taxa with non overlapping data
+trim <- TrimMorphDistMatrix(matrix_dist)
+matrix_dist <- trim$dist.matrix
+
+## Which taxa were removed?
+as.vector(trim$removed.taxa)
+
+## Save
 saveRDS(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))

### 2017-11-29 analyses/01-extract-data-for-analyses.Rmd
Fix issues with non overlapping taxa and tree mismatch

--- 2017-11-28 analyses/01-extract-data-for-analyses.Rmd
@@
 
+One final issue is that to compute Gower distances we need all taxa to have some overlapping data.
+If not, Gower distances will be NA (Gower distance does not impute missing data like some other methods) and we can't estimate disparity.
+To avoid this problem we can remove the taxa with no overlaps using `TrimMorphDistMatrix` from `Claddis` on our distance matrix.
+Of course this then means we need to remove these from the tree as well.
+This then changes the node numbers which means we need to recalculate ancestral states.
+Because nothing is ever easy!
+Rather than going round in circles with this code, we just remove the species where this is a problem when we read in the trees (only an issue below for the Bapst and Brusatte datasets).
+Yes there's a more elegant way to do this, but time is tight and I'm pretty sure this project is trying to break us...
+
+If you needed to do this for your own data, this is how you can quickly work out which taxa to remove.
+
+```{r}
+## Trimming the distance matrix
+## trim <- TrimMorphDistMatrix(matrix_dist)
+## matrix_dist <- trim$dist.matrix
+
+## Which taxa were removed?
+## (remove_taxa <- as.vector(trim$removed.taxa))
+```
+
 And that's it...
@@
 
-# Replace zero branches with minimum branch length
+## Replace zero branches with minimum branch length
 tree$edge.length[which(tree$edge.length == 0)] <- 
   min(tree$edge.length[-which(tree$edge.length == 0)])
+
+## Remove taxa with non overlapping data (see note about Gower distances above)
+remove_Bapst <- c("Hagryphus_giganteus", "Atrociraptor_marshalli", "IGM100_1015_UndesDromaeosaurid", "Dromaeosaurus_albertensis", "Incisivosaurus_gauthieri", "Deinocheirus_mirificus", "Therizinosaurus_cheloniformis", "Anserimimus_planinychus", "Elmisaurus_rarus")
+
+tree <- drop.tip(tree, remove_Bapst)
 
@@
 
+## Remove taxa with non overlapping data (see note about Gower distances above)
+remove_Brusatte <- c("Shanag_ashile", "Atrociraptor_marshalli", "Proceratosaurus_bradleyi",     "Incisivosaurus_gauthieri", "Enigmosaurus", "Nanshiungosaurus_brevispinus", "Xixiasaurus", "Tsaagan_mangas", "Mirischia", "Pedopenna", "Suzhousaurus", "Juratyrant", "Vorona", "Bonapartenykus", "Teratophoneus", "Gobipteryx", "Songlingornis", "Liaoningornis_longidigitu", "Achillesaurus")
+
+tree <- drop.tip(tree, remove_Brusatte)
+
 # Tidy the tree and the data
@@
 as.vector(na.omit(c(cleaned_data$dropped_tips, cleaned_data$dropped_rows)))
+
+## Remove
 

### 2017-11-29 manuscript/tables/datasets.tex
Add draft table of dataset details

% Dataset descriptions table

\begin{tabular}[t]{l p{3cm} l l l l}

\hline
\textbf{Dataset} & \textbf{Group} & \textbf{$N_{taxa}$} & \textbf{$N_{characters}$} & \textbf{Age range} & \textbf{Reference} \\
\hline
Beck2014 & Eutherian mammals & 102 & 410 & 170 Ma - present & X \\
%-----------------------------------------------------------
Bapst2016 & theropods & 102 & 410 & 170 Ma - present & X \\
%-------------------------------------------------------
Brusatte2014 & theropods & 102 & 410 & 170 Ma - present & X \\
%-------------------------------------------------------
Heath2016 & bears & 102 & 410 & 170 Ma - present & X \\
%-------------------------------------------------------
Wright2017 & crinoids & 102 & 410 & 170 Ma - present & X \\
%-------------------------------------------------------
\hline
\end{tabular}

### 2017-11-29 manuscript/time-slice.tex
Edits to methods as far as disparity thru time section

@@
 \author{
-	Thomas Guillerme$^{1,2*}$ and Natalie Cooper$^{2}$
+	Thomas Guillerme$^{1*}$ and Natalie Cooper$^{2}$
 }
@@
 \affiliation{\noindent{\footnotesize
-	$^1$ Imperial\\ 
+	$^1$Thomas address\\ 
 	$^2$Department of Life Sciences, Natural History Museum, Cromwell Road, London, SW7 5BD, UK. natalie.cooper@nhm.ac.uk}\\
@@
 \label{overview-section}
-To test the different time subsampling methods, we first need to prepare the data for the disparity
+To test the different time subsampling methods, we followed the protocol below. 
+Note that all the code to reproduce these analyses (along with detailed instructions) is provided HERE.
 
@@
   \item Collate data.
-Disparity-through-time analysis requires a matrix containing morphological characters for each taxon, and information on the time range over which each taxon existed.
+Disparity-through-time analysis requires a matrix containing morphological characters for each taxon, and information on the first and last occurrence dates of each taxon.
 Time slicing methods also require a phylogeny.
@@
     \end{enumerate}
-  \item time subsample
+  \item Time subsampling
     \begin{enumerate}
@@
 \label{datasets}
-To test the different time slicing methods we selected X datasets. 
+To test the different time binning/slicing methods we selected five datasets (see Table \ref{datasets}). 
 Note that although disparity-through-time analyses are possible without a phylogeny, we only use examples with phylogenies here so that we can compare results of time slicing methods, which do require a phylogeny.
+Each dataset consists of a set of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
+More details on these datasets can be found in the appendix. All data are available HERE. And the processing code too.
 
-\paragraph{\textit{Mammal data - Halliday.}}
-We used the cladistic morphological matrices and the Total Evidence tip-dated trees from \citet[][103 taxa with 446 morphological characters;]{Slater2012MEE} and \citet[][102 taxa with 421 morphological characters]{beckancient2014}.
-%We chose these two datasets because they have a similar number of taxa and morphological characters.
-%\cite{Slater2012MEE} ranges from 310 million years ago (Ma; Late Carboniferous) to the present and focuses on the clade Mammaliaformes at the family-level and is called hereafter the Mammaliaformes dataset.
-%\cite{beckancient2014} ranges from 170 Ma (Middle Jurassic) to the present and focuses on Eutheria at the genus-level and is called hereafter the Eutheria dataset.
-%We used the first and last occurrences reported in \cite{Slater2012MEE} and \cite{beckancient2014} as the temporal range of each taxon in our analysis.
-%Both phylogenies are illustrated in figures \ref{fig:SlaterTree} and \ref{fig:BeckTree}.
-%Both trees contain few taxa compared to the overall species richness of living and fossil mammals \citep{bininda-emondsthe2007,archibald2011extinction}.
-%This is because Total Evidence trees need a lot of data, particularly morphological data for living taxa that can be hard to locate \citep[Chapter \ref{chap:TEM_paper} and \ref{chap:missing_mammals};][]{GuillermeCooperMissing,GuillermeCooper}.
-%Therefore, most Total Evidence studies to date contain one or two orders of magnitude fewer taxa than phylogenies based solely on molecular data (e.g.
-thousands of taxa in \citealt{bininda-emondsthe2007,meredithimpacts2011} \textit{vs.} hundreds in \citealt{ronquista2012,Slater2012MEE,beckancient2014}).
+  \begin{table}[!htbp]      
+    \caption[Datasets.]
+    {Dataset captions}
+    \input{tables/datasets.tex} 
+    \label{tab:data}  
+  \end{table}
 
@@
 
-\subsection{Building cladisto-spaces} % Maybe reduce detail here!
+\subsection{Building cladisto-spaces} 
 To explore disparity-through-time in our datasets we used a cladisto-space approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
@@
 the discrete morphological characters used to build a phylogenetic tree).
-Mathematically, a cladisto-space is an $n$ dimensional object that summarises the cladistic distances between the taxa present in a cladistic matrix (see details below).
+Mathematically, a cladisto-space is an $n$ dimensional object that summarises the cladistic distances between the taxa present in a cladistic matrix.
 Although empirically inter-taxon distances are the same in a morphospace or a cladisto-space \citep{foth2012different,hetherington2015cladistic}, we prefer the term cladisto-space to make it clear that this space is estimated using cladistic data and not morphometric data.
-%and because both objects have slightly different properties.
-%For example, because of its inherent combinatory properties, a cladisto-space is a finite theoretical object limited by the product of the number of character states, whereas a morphospace is an infinite theoretical object.
-%Thus a cladisto-space will be overloaded if the number of taxa is higher than the product of the number of character states, although this is rarely an issue with empirical data (our cladisto-spaces have maximal capacities of $1.9$$\times$$10^{181}$ taxa for the Mammaliaformes dataset, i.e.
-101 orders of magnitude more taxa than the number of particles in the universe; and $4.5$$\times$$10^{159}$ taxa for the Eutheria dataset).
 
@@
 To estimate the cladisto-spaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
-For each dataset separately, we calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e.
-the Euclidean distance between two taxa divided by the number of shared characters.
+We calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the Euclidean distance between two taxa divided by the number of shared characters. 
 This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e.
@@
 Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis} to remove pairs of taxa with no cladistic characters in common.
-%This led to us removing 11 taxa from the Mammaliaformes dataset but none from the Eutheria dataset.
-% NC: Add info on species we have to remove.
+This led to us removing nine taxa from the BAPST dataset, and 19 from the Brusatte dataset, but none from the other three datasets (see code for details of which species). % add citations
 
@@
 selecting only $m$ dimensions that represent up to 90\% of the variance in the distance matrix; \citealt{Brusatte12092008,toljagictriassic-jurassic2013}).
-% NC: Hmmm should we do this the usual way???
 
@@
 Second, because we include all $n$ dimensions in the analysis (see above), the products of ranges and variances will tend towards zero since the scores of the last dimension are usually really close to zero themselves.
-We therefore only use the sum of variances metric.
+We therefore use the sum of variances metric to estimate disparity here.
 

### 2017-11-29 manuscript/tables/datasets.tex
Add some details to data table

@@
 % Dataset descriptions table
+% ADD DETAILS!!!
 
-\begin{tabular}[t]{l p{3cm} l l l l}
+\begin{tabular}[t]{l p{3cm} l l l l p{6cm}}
 
 \hline
-\textbf{Dataset} & \textbf{Group} & \textbf{$N_{taxa}$} & \textbf{$N_{characters}$} & \textbf{Age range} & \textbf{Reference} \\
+\textbf{Dataset} & \textbf{Group} & \textbf{$N_{taxa}$} & \textbf{$N_{characters}$} & \textbf{Age range} & \textbf{Reference} & \textbf{Notes} \\
 \hline
-Beck2014 & Eutherian mammals & 102 & 410 & 170 Ma - present & X \\
+Beck2014 & Eutherian mammals & 102 & 410 & 170 Ma - present & X & Y \\
 %-----------------------------------------------------------
-Bapst2016 & theropods & 102 & 410 & 170 Ma - present & X \\
+Bapst2016 & theropods & 102 & 410 & 170 Ma - present & X & 2 + 9 taxa removed\\
 %-------------------------------------------------------
-Brusatte2014 & theropods & 102 & 410 & 170 Ma - present & X \\
+Brusatte2014 & theropods & 102 & 410 & 170 Ma - present & X & 12 + 19 taxa removed. Tree was randomly selected from all time-scaled trees\\
 %-------------------------------------------------------
-Heath2016 & bears & 102 & 410 & 170 Ma - present & X \\
+Heath2016 & bears & 102 & 410 & 170 Ma - present & X & 12 + 0 taxa removed\\
 %-------------------------------------------------------
-Wright2017 & crinoids & 102 & 410 & 170 Ma - present & X \\
+Wright2017 & crinoids & 102 & 410 & 170 Ma - present & X & Maximum clade credibility tree, time-scaled by dividing branches by 0.03517385 and setting root time to 485.4\\
 %-------------------------------------------------------

### 2017-11-29 manuscript/time-slice.tex
More edits to methods

@@
 
-We present this using data from Beck \& Lee and Halliday???
+We present this using five datasets collated from the literature.
 
@@
 \label{overview-section}
+% I'm unsure if this section is needed or not. The methods are long so it might help?
 To test the different time subsampling methods, we followed the protocol below. 
-Note that all the code to reproduce these analyses (along with detailed instructions) is provided HERE.
+Note that all the code needed to reproduce these analyses (along with detailed instructions) is provided HERE.
 
 \begin{enumerate}
-  \item Collate data.
+  \item Collating example datasets.
 Disparity-through-time analysis requires a matrix containing morphological characters for each taxon, and information on the first and last occurrence dates of each taxon.
 Time slicing methods also require a phylogeny.
-  \item Estimate ancestral character states at the nodes of the phylogeny. This step is required for the time slicing methods.
-  \item Build cladisto-spaces.  
-    \begin{enumerate}
-      \item Construct distance matrices.
-      \item ordiantion
-    \end{enumerate}
+  \item Processing example datasets.
+Various transformations of the datasets are required before the analyses can take place.
+We need to estimate ancestral character states at the nodes of the phylogeny for the time slicing methods, build cladisto-spaces via distance matrices and ordination, and remove taxa from the phylogeny that are missing from the matrix and vice versa.
   \item Time subsampling
-    \begin{enumerate}
-      \item stratigraphy
-      \item time bins
-        \begin{enumerate}
-          \item 5 million years
-          \item 10 million years
-          \item 15 million years
-        \end{enumerate}  
-      \item time slicing
-        \begin{enumerate}
-          \item punctuated
-          \item gradual
-          \item random
-        \end{enumerate}  
-    \end{enumerate}
   \item Estimate disparity for each time subsample
@@
 Each dataset consists of a set of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
-More details on these datasets can be found in the appendix. All data are available HERE. And the processing code too.
+More details on these datasets can be found in the appendix. All data are available HERE. And the processing code too. Probably also want to link to Dryad etc. HERE
 
@@
 Second, because we include all $n$ dimensions in the analysis (see above), the products of ranges and variances will tend towards zero since the scores of the last dimension are usually really close to zero themselves.
-We therefore use the sum of variances metric to estimate disparity here.
+We therefore use the sum of variances metric to estimate disparity.
 
-\subsection{Estimating disparity through time} 
+\subsection{Estimating disparity-through-time} 
 \label{time_slicing}
 
+To estimate disparity-through-time we first need to split the data into time subsamples.
+Here we explore the following scenarios.
+
+\begin{enumerate}
+  \item Stratigraphic time bins.
+This is the traditional method, where all the taxa within each stratigraphic period are included in the disparity calculation.
+This often leads to bins of unequal duration.
+Here we use stratigraphic ages and epochs.
+  \item Equally sized time bins.
+This is another commonly used method, where the time frame of interest is split into equally sized time bins, then all the taxa within each time bin are included in the disparity calculation.
+Here we use three different methods to determine the size of the time bins.
+    \begin{enumerate}
+      \item The *duration* of the bin is equal to the median duration of the stratigraphic period (age or epoch).
+      \item The *number* of bins is equal to the number of stratigraphic periods (ages or epochs).
+      \item The *number* of bins is either 2, 5, 10, 15, or 20.
+    \end{enumerate}
+  \item Time slicing.
+This method uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation.
+Here we use three different methods to determine the intervals between the time slices.
+    \begin{enumerate}
+      \item The *interval* between slices is equal to the median duration of the stratigraphic period (age or epoch).
+      \item The *number* of slices is equal to the number of stratigraphic periods (ages or epochs).
+      \item The *number* of slices is either 2, 5, 10, 15, or 20.
+    \end{enumerate}  
+\end{enumerate}
+
 % Might make a table with all the subsets???
+% Had a play - it's COMPLICATED!!!
+
+  \begin{table}[!htbp]      
+    \caption[Subsamples.]
+    {captions}
+    \input{tables/timesamples.tex} 
+    \label{tab:subsamples}  
+  \end{table}
 
@@
 \paragraph{Stratigraphic time bins.}
-We extracted the taxa found in each geological stage from each cladisto-space.
-These consisted of the Middle Jurassic to the present (XX subsamples) for dataset 1, the Middle Jurassic to the present (XX subsamples) for dataset 2 and the Middle Jurassic to the present (XX subsamples) X for dataset 3.
+We extracted the taxa found in each geological age or epoch from each cladisto-space.
 For each subsample, we estimated its disparity.
@@
 We then calculated the median disparity value for each subsample along with the 50\% and 95\% confidence intervals.
-We also recorded the number of taxa in each subsample as a proxy for taxonomic diversity.
+We also recorded the number of taxa in each subsample as a proxy for taxonomic diversity (Table \ref{tab:subsamples}).
 
@@
 We extracted the taxa found in each time bin for each cladisto-space.
-We varied the size of the time bins from 5 million year, 10 million years and 20 million years, to determine if the length of the time bin affected our results.
-There were XX subsamples for dataset 1, XX subsamples for dataset 2 and XX subsamples for dataset 3.
-For each subsample, we estimated its disparity.
-To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement by randomly resampling with replacement a new subsample of taxa from the observed taxa in the subsample 1000 times.
-We then calculated the median disparity value for each subsample along with the 50\% and 95\% confidence intervals.
-We also recorded the number of taxa in each subsample as a proxy for taxonomic diversity.
+We varied the size of the time bins so that either (i) the *duration* of the bin is equal to the median duration of the stratigraphic age or epoch, (ii) the *number* of bins is equal to the number of stratigraphic ages or epochs, (iii) or by using 2, 5, 10, 15, or 20 bins.
+For each subsample, we estimated its disparity and bootstrapped each disparity measurement as described above for stratigraphic time bins.
+We also recorded the number of taxa in each subsample as a proxy for taxonomic diversity (Table \ref{tab:subsamples}).
 
-\paragraph{Time slicing.}
+\paragraph{Time slicing.} % THOMAS TO UPDATE THIS BIT!!!
 The ``time-slicing'' approach considers subsets of taxa in the cladisto-space at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
-This results in even-sampling of the cladisto-space across time and permits us to define the underlying model of character evolution (punctuated or gradual).  
+This results in even-sampling of the cladisto-space across time and permits us to define the underlying model of character evolution (punctuated or gradual % and others ADD).  
 In practice, time-slicing considers the disparity of any element present in the phylogeny (branches, nodes and tips) at any point in time.
@@
 
-We applied our time-slicing approach separately to each cladisto-space, time-slicing the phylogeny every five million years.
-% Do we want to vary the time slice timings too?
-This resulted in XX subsamples for dataset 1, XX for dataset 2 and XX for dataset 3.
-For each subsample, we estimated its disparity assuming punctuated (ACCTRAN, DELTRAN and random) and gradual evolution as described above.
-To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement by randomly resampling with replacement a new subsample of taxa from the observed taxa in the subsample 1000 times.
-We then calculated the median disparity value for each subsample along with the 50\% and 95\% confidence intervals.
-We also recorded the number of phylogenetic elements (nodes and tips) in each subsample as a proxy for taxonomic diversity.
+We applied our time-slicing approach separately to each cladisto-space.
+We varied the intervals between time slices so that either (i) the *interval* between slices is equal to the median duration of the stratigraphic age or epoch; (ii) the *number* of slices is equal to the number of stratigraphic ages or epochs, or (iii) by setting the *number* of slices to 2, 5, 10, 15, or 20.
+For each subsample, we estimated its disparity assuming punctuated (ACCTRAN, DELTRAN and random) and gradual evolution as described above. % NEED TO ADD THE NEW MODELS.
+To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement as described above for the stratigraphic time bins. 
+We also recorded the number of phylogenetic elements (nodes and tips) in each subsample as a proxy for taxonomic diversity (Table \ref{tab:subsamples}).
 
@@
 %Testing the difference section
-
-\subsection{Rarefaction}
-
-Finally, disparity may be higher in subsamples with more phylogenetic elements simply because there are more taxa represented.
-To test whether this influenced our results, we repeated the analyses using rarefied disparities.
-In dataset 1, the minimum number of taxa in each stratigraphy subsample was eight, time bin XX and time slice was XX.
 
@@
 
-%\bibliographystyle{sysbio} 
-%\bibliography{time-refs}
+\bibliographystyle{sysbio} 
+\bibliography{time-refs}
 

### 2017-11-30 analyses/04-stats.Rmd
Add Rmd for all the stats and tables

---
title: "Outputs from time slice analyses"
author: "Natalie Cooper and Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 6
    fig_height: 6
---

This code runs the output analyses in the paper, i.e. creates tables, figures and statistical tests.

```{r}
set.seed(123)
```

Load packages and functions
-----------------------------------

```{r, eval = TRUE, message = FALSE}
library(tidyverse)
library(broom)
```

Loading the data
-----------------------------------

This script requires the outputs from `02-time-slice-analyses.Rmd`. The code below runs the analyses for the Beck and Lee 2014 dataset - chaneg the slug name to run the others. This of course can all be written into functions but we ran out of time so here is some suboptimal but working code!

```{r}
slug <- "Beck2014"
#slug <- "Brusatte2014"
#slug <- "Bapst2016"
#slug <- "bears"
#slug <- "Wright2017"
```

```{r}
out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
#out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
```

Outputs can be viewed using `out1$results$stratigraphy`, `out1$results$duration`, or `out1$results$number`.

Testing for differences among the models is a little tricky, as we are interested in how disparity-through-time is influenced by the choice of time subsampling method, rather than the disparities estimated by each method, especially as these will influenced by the number of taxa (and/or nodes) included in each subsample. We therefore present two different, simple ways of comparing the models: paired Wilcoxon tests and identification of disparity peaks.

Paired Wilcoxon tests
----------------------

First, we use paired Wilcoxon tests to compare (i) unequal sized time bins to unevenly spaced time slices where the bins are stratigraphic periods (ages or epochs) and slices occur at the midpoint of the stratigraphic period; (ii) equal time bins to equally spaced time slices where time bin size, or the interval between slices is defined by the average *duration* of the stratigraphic period (age or epoch); and (iii) equal time bins to equally spaced time slices where the number of time bins, or number of slices is defined by the average *number* of stratigraphic periods (ages or epochs) in the time frame of interest.

Due to the uneven spread of taxa across phylogenies, some time bins will contain one or no species, meaning that we cannot estimate disparity for that time bin. We first, therefore, remove the time bins, and corresponding time slices, without disparity estimates. We then perform paired Wilcoxon tests, so that bins and slices for the same time period are being compared. Significant results suggest that there is a systematic difference in disparity values at each time point, depending on whether bins or slices are used.

i) statigraphy

```{r}
strat <- out1$results$stratigraphy

# Work out how many disparity outputs per method there are
n_out <- 
  strat %>%
  filter(model == "acctran") %>%
  summarise(n = n())

# Get data into a format for wilcox tests
strat2 <- 
  strat %>% 
  # Add bin numbers so correct pairs can be compared  
  mutate(bin = rep(1:n_out$n, 7)) %>% 
  # Select only the required columns  
  select(bin, model, bs.median) %>%
  # Spread the models out by bin to make extracting NA bins easier
  spread(model, bs.median) %>%
  # Remove bins with NAs
  filter(complete.cases(.))

# Wilcox tests
w_acc <- wilcox.test(strat2$equalbins, strat2$acctran, paired = TRUE)
w_del <- wilcox.test(strat2$equalbins, strat2$deltran, paired = TRUE)
w_grad <- wilcox.test(strat2$equalbins, strat2$gradual, paired = TRUE)
w_prox <- wilcox.test(strat2$equalbins, strat2$proximity, paired = TRUE)
w_punc <- wilcox.test(strat2$equalbins, strat2$punctuated, paired = TRUE)
w_rand <- wilcox.test(strat2$equalbins, strat2$random, paired = TRUE)

# Save outputs
outputS <- rbind(tidy(w_acc), tidy(w_del), tidy(w_grad), tidy(w_prox), 
                tidy(w_punc), tidy(w_rand))
outputS$model <- c("acctran", "deltran", "gradual", "proximity", 
                        "punctuated", "random")
outputS$type <- rep("stratigraphy", 6)
outputS$period <- rep(strat$stratigraphy[1], 6)
```

ii) duration

```{r}
duration <- out1$results$duration

# Work out how many disparity outputs per method there are
n_out <- 
  duration %>%
  filter(model == "acctran") %>%
  summarise(n = n())

# Get data into a format for wilcox tests
duration2 <- 
  duration %>% 
  # Add bin numbers so correct pairs can be compared  
  mutate(bin = rep(1:n_out$n, 7)) %>% 
  # Select only the required columns  
  select(bin, model, bs.median) %>%
  # Spread the models out by bin to make extracting NA bins easier
  spread(model, bs.median) %>%
  # Remove bins with NAs
  filter(complete.cases(.))

# Wilcox tests
w_acc <- wilcox.test(duration2$equalbins, duration2$acctran, paired = TRUE)
w_del <- wilcox.test(duration2$equalbins, duration2$deltran, paired = TRUE)
w_grad <- wilcox.test(duration2$equalbins, duration2$gradual, paired = TRUE)
w_prox <- wilcox.test(duration2$equalbins, duration2$proximity, paired = TRUE)
w_punc <- wilcox.test(duration2$equalbins, duration2$punctuated, paired = TRUE)
w_rand <- wilcox.test(duration2$equalbins, duration2$random, paired = TRUE)

# Save outputs
outputD <- rbind(tidy(w_acc), tidy(w_del), tidy(w_grad), tidy(w_prox), 
                tidy(w_punc), tidy(w_rand))
outputD$model <- c("acctran", "deltran", "gradual", "proximity", 
                        "punctuated", "random")
outputD$type <- rep("duration", 6)
outputD$period <- rep(strat$stratigraphy[1], 6)
```

iii) number

```{r}
number <- out1$results$number

# Work out how many disparity outputs per method there are
n_out <- 
  number %>%
  filter(model == "acctran") %>%
  summarise(n = n())

# Get data into a format for wilcox tests
number2 <- 
  number %>% 
  # Add bin numbers so correct pairs can be compared  
  mutate(bin = rep(1:n_out$n, 7)) %>% 
  # Select only the required columns  
  select(bin, model, bs.median) %>%
  # Spread the models out by bin to make extracting NA bins easier
  spread(model, bs.median) %>%
  # Remove bins with NAs
  filter(complete.cases(.))

# Wilcox tests
w_acc <- wilcox.test(number2$equalbins, number2$acctran, paired = TRUE)
w_del <- wilcox.test(number2$equalbins, number2$deltran, paired = TRUE)
w_grad <- wilcox.test(number2$equalbins, number2$gradual, paired = TRUE)
w_prox <- wilcox.test(number2$equalbins, number2$proximity, paired = TRUE)
w_punc <- wilcox.test(number2$equalbins, number2$punctuated, paired = TRUE)
w_rand <- wilcox.test(number2$equalbins, number2$random, paired = TRUE)

# Save outputs
outputN <- rbind(tidy(w_acc), tidy(w_del), tidy(w_grad), tidy(w_prox), 
                tidy(w_punc), tidy(w_rand))
outputN$model <- c("acctran", "deltran", "gradual", "proximity", 
                        "punctuated", "random")
outputN$type <- rep("number", 6)
outputN$period <- rep(strat$stratigraphy[1], 6)
```

And finally save all the outputs...

```{r}
all_wilcox <- rbind(outputS, outputD, outputN)
all_wilcox <- select(all_wilcox, -c(method, alternative))
write.table(all_wilcox, file = paste0("../outputs/wilcox.", slug, ".csv"), quote = FALSE, 
          row.names = FALSE, col.names = FALSE, append = TRUE, sep = ",")
```

Repeat for epochs by uncommenting `out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))` and rerunning.

You can rerun for the other datasets by changing the slug name at the start.

Disparity peaks
----------------------

Set the slug

```{r}
slug <- "Beck2014"
#slug <- "Brusatte2014"
#slug <- "Bapst2016"
#slug <- "bears"
#slug <- "Wright2017"
```

Read in the data

```{r}
out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
out2 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
```

Extract the maximum bootstrapped median disparity values for each model, with details of the peak bin/slice times.

```{r, warning = FALSE}
# ages
strat <- out1$results$stratigraphy
duration <- out1$results$duration
number <- out1$results$number

strat <- 
  strat %>%
  group_by(model) %>%
  filter(bs.median == max(bs.median, na.rm = TRUE))

duration <-
  duration %>%
  group_by(model) %>%
  filter(bs.median == max(bs.median, na.rm = TRUE))

number <- 
  number %>%
  group_by(model) %>%
  filter(bs.median == max(bs.median, na.rm = TRUE))

# epochs
strat2 <- out2$results$stratigraphy
duration2 <- out2$results$duration
number2 <- out2$results$number

strat2 <- 
  strat2 %>%
  group_by(model) %>%
  filter(bs.median == max(bs.median, na.rm = TRUE))

duration2 <-
  duration2 %>%
  group_by(model) %>%
  filter(bs.median == max(bs.median, na.rm = TRUE))

number2 <- 
  number2 %>%
  group_by(model) %>%
  filter(bs.median == max(bs.median, na.rm = TRUE))

all_peaks <- rbind(strat, duration, number, strat2, duration2, number2)
write.csv(all_peaks, file = paste0("../outputs/peaks.", slug, ".csv"), quote = FALSE, 
          row.names = FALSE)

```
You can rerun for the other datasets by changing the slug name at the start.

Tables
----------

Finally, for the purposes of completeness here are the full results as tables.

Set the slug

```{r}
slug <- "Beck2014"
#slug <- "Brusatte2014"
#slug <- "Bapst2016"
#slug <- "bears"
#slug <- "Wright2017"
```
Read in the data

```{r}
out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
out2 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
```
Make the tables
```{r, eval = TRUE, echo = FALSE}
# Ages
knitr::kable(out1$results$stratigraphy, caption = "Disparity per stratigraphic age: unequal time bins and non-equidistant time slices")
knitr::kable(out1$results$duration, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the median stratigraphic age duration")
knitr::kable(out1$results$number, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the number of stratigraphic ages")

# Epochs
knitr::kable(out2$results$stratigraphy, caption = "Disparity per stratigraphic epoch: unequal time bins and non-equidistant time slices")
knitr::kable(out2$results$duration, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the median stratigraphic epoch duration")
knitr::kable(out2$results$number, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the number of stratigraphic epochs")
```

You can rerun for the other datasets by changing the slug name at the start.

### 2017-12-01 manuscript/time-slice.tex
title suggestion + affiliation

@@
 \title{Time for a rethink: disparity-through-time}
+%TG: Great title first part! How about changing the other part to 
+%Time for a rethink: different ways of slicing through the stratigraphic cake leads to different disparity through time results.
 
@@
 }
+%TG: authorship-wise it's up to you, in the end, since it's just the two of us, I think it really doesn't matter (i.e.
+it's still gonna be Cooper Guillerme or Guillerme Cooper), I'll hopefully will have a couple of first author papers coming up for 2018 so up to you if you want one as well.
 
@@
 \affiliation{\noindent{\footnotesize
-	$^1$Thomas address\\ 
+	$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia.\\
 	$^2$Department of Life Sciences, Natural History Museum, Cromwell Road, London, SW7 5BD, UK. natalie.cooper@nhm.ac.uk}\\

### 2017-12-03 manuscript/time-slice.tex
precisions in the models descriptions

@@
 The ``time-slicing'' approach considers subsets of taxa in the cladisto-space at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
-This results in even-sampling of the cladisto-space across time and permits us to define the underlying model of character evolution (punctuated or gradual % and others ADD).  
+This results in even-sampling of the cladisto-space across time and while permitting to use an underlying model of character evolution (punctuated or gradual).
+% and others ADD).
+%TG: I don't have any other ones in stock.
+But it could be easy to come up with new ones.
+They're easy to implement.
+
 In practice, time-slicing considers the disparity of any element present in the phylogeny (branches, nodes and tips) at any point in time.
-When the phylogenetic elements are nodes or tips, the eigenvector scores for the nodes (estimated using ancestral state reconstruction as described in section \ref{ace}) or tips are directly used for estimating disparity.
-When the phylogenetic elements are branches we chose the eigenvector score for the branch using one of two evolutionary models:
+When the phylogenetic elements are nodes or tips, the ordination scores for the nodes (estimated using ancestral state reconstruction as described in section \ref{ace}) or tips are directly used for estimating disparity.
+When the phylogenetic elements are branches we chose the ordination score for the branch using one of two evolutionary models:
+
+%TG: Not sure if it's the "egeinvector" score or the "eigen" score.
+I think there's a trick in naming them depending on how the algorithm does the transformation (soemthing with using eigenvalues or eigenvectors).
+I call it ordination score for now.
+
 \begin{enumerate}
+
     \item{\textbf{Punctuated evolution.}} 
-    This model selects the eigenvector score from either the ancestral node or the descendant node/tip of the branch regardless of the position of the slice along the branch. 
+    This model selects the ordination score from either the ancestral node or the descendant node/tip of the branch regardless of the position of the slice along the branch. 
     Similarly to the time interval approach, this reflects a model of punctuated evolution where changes in disparity occur either at the start or at the end of a branch over a relatively short time period and clades undergo long periods of stasis during their evolution \citep{Gould1977,Hunt20112007}.
@@
     \begin{enumerate}[(i)]
-      \item selecting the eigenvector score of the ancestral node of the branch (ACCTRAN).
-      \item selecting the eigenvector score of the descendant node/tip of the branch (DELTRAN).
-      \item randomly selecting either the eigenvector score of the ancestral node or the descendant node/tip of the branch (random).
+
+      \item The ``acctran'' model, always selecting the ordination score of the ancestral node of the branch.
+      \item The ``deltran'' model, always selecting the ordination score of the descendant node/tip of the branch.
+      \item The ``random'' model, selecting randomly between the ordination score of the ancestor or the descendant of the branch.
+      \item The ``punctuated'' model (probabilistic), selecting the ordination score from both the ancestor and the descendant with an equal probability
+          \begin{equation}
+          p(\text{ancestor}) = p(\text{descendant}) = 0.5
+          \end{equation}
+
     \end{enumerate}
-    Method (i) assumes that changes always occur early on the branch (accelerated transition, ACCTRAN) and (ii) assumes that changes always occur later (delayed transition, DELTRAN).
-    We prefer not to make either assumption so we report the results from (iii), although the ACCTRAN and DELTRAN results are available in the appendix \ref{chap:Appendix_STD} (Fig.
-\ref{Supp_disparity_all_Mammaliaformes}, \ref{Supp_disparity_all_Eutheria}, \ref{Supp_disparity_all_Mammaliaformes_rarefied}, \ref{Supp_disparity_all_Eutheria_rarefied}).
+
+    The two first models assume that changes always occur early (ACCelerated TRANsition) or late along the branches (DELayed TRANsition).
+    The third model makes neither assumption and simply selects randomly the data between the ancestor and the descendant.
+    These three first models do not allow the data to vary throughout eventual pseudo-replications in the analysis (the bootstraps).
+    For example, using the ``random'' model, once the the data of the ancestor has been randomly chosen, only this data will be used during the bootstraps procedure.
+    Conversely, the ``punctuated'' model allows both ancestor's and descendant's data to be used during the bootstrapping with an equal probability.
+    This results in the bootstrapped disparity at a specific slice to effectively incorporate both the data from the descendant and the ancestor.
+%TG: now that I'm mentioning that, I should maybe rename this model.
+Lets discuss that on monday.
+
     \item{\textbf{Gradual evolution.}}
-    This model also selects the eigenvector score from either the ancestral node or the descendant node/tip of the branch, but the choice depends on the distance between the sampling time point and the end of the branch.
-    If the sampling time point falls in the first half of the branch length the eigenvector score is taken from the ancestral node, conversely, if the sampling time point falls in the second half of the branch length the eigenvector score is taken from the descendant node/tip.
+    These model also selects the ordination score from either the ancestral node or the descendant node/tip of the branch, but the choice depends on the distance between the sampling time point and the end of the branch.
+    If the sampling time point falls in the first half of the branch length the ordination score is taken from the ancestral node, conversely, if the sampling time point falls in the second half of the branch length the ordination score is taken from the descendant node/tip.
     This reflects a model of gradual evolution where changes in disparity are gradual and cumulative along the branch.
-    Under this model, the gradual changes could be either directional or random, however, directional evolution has been empirically shown to be rare \citep[only 5\% of the time][]{Hunt20112007}.
-    We therefore considered that changes from a character state A to B were only dependent on the branch length.
+
+    \begin{enumerate}[(i)]
+    \setcounter{enumii}{4}
+    \item The ``proximity'' model, selecting the ordination of either the ancestor or the descendant if the slice occurs respectively before or after half the branch length.
+    \item The ``gradual'' model (probabilistic), selecting the ordination score from both the ancestor and the descendant with a probability equal to the length between the slice.
+          \begin{equation}
+              p(\text{ancestor}) = \frac{d(\text{ancestor},\text{slice})}{d(\text{ancestor},\text{descendant})}
+          \end{equation}
+          \begin{equation}
+              p(\text{descendant}) = 1 - p(\text{ancestor})
+          \end{equation}
+    \noindent Where $d(x,y)$ is the distance measured as branch length.
     \end{enumerate}
+
+    The ``proximity'' model is a variation of the ``random'' one described above were the selection of the data is not random but arbitrarily based on branch length.
+    However, similarly to the ``random'' one, this model doesn't allow variation.
+    The last ``gradual'' model is also a variation of the ``punctuated'' and allows variation in a similar way (but based on branch length).
+    For example, if slice occurs at the third quarter of a branch from A to B (75\% of the total branch length), after bootstrapping, the disparity results will be based on 25\% of the data from A and 75\% of the data from B.
+\end{enumerate}
+
+%TG: Justification on the fact that these models are not ACEs.
+It is important to notice that the time-slicing method is not an ancestral states estimation method \textit{per se}.
+This method does not estimate values along a branch applying a model (\textit{c.f.} methods described in \ref{ace}) but rather chooses between the two available information (the descendant or the ancestor) using the models described above.
+This allows this method to be used in post-ordination analysis were the data used in each time slice is data already present in the cladisto-space.
+In other words, this method does not require a re-ordination of the cladisto-space every time a slice goes through a branch, thus allowing the properties of the cladisto-space (e.g.
+distance between species, variance per axis, etc.) to remain constant.
 

### 2017-12-04 analyses/table-compiler.Rmd
Markdown to easily build the tables (example done with table 1)

---
title: "Table compiler"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Compiling the tables for the paper

Loading the packages

```{r}
library(xtable)
library(Claddis)
library(ape)
library(dispRity)
```

Loading the data:

```{r}
## Getting the data
slugs <- list("Beck2014", "Brusatte2014", "Bapst2016", "bears", "Wright2017")

## Matrices lists
matrices <- lapply(slugs, function(slug) ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))$matrix )

## Trees lists
trees <- lapply(slugs, function(slug) read.nexus(paste0("../data/processed/tree.", slug, ".tre")))
## Adding Wright's tree age
trees[[5]]$root.time <- 485.4
```

## Data set table

```{r}
## Column names
colnames <- c("Group", "n taxa", "m characters", "Age range", "Reference", "Notes")

## Dataset names
dataset <- unlist(slugs)
dataset[4] <- "Heath2016"

## Groups
group <- c("Eutherian mammals", "theropods", "theropods", "bears", "crinoids")

## Number of taxa
ntaxa <- unlist(lapply(matrices, nrow))

## Number of characters
mcharacters <- unlist(lapply(matrices, ncol))

## Age ranges
age_ranges <- lapply(trees, function(tree) rev(range(tree.age(tree)$age)))
age_ranges <- unlist(lapply(age_ranges, function(age) paste(age, collapse = " - ")))

## References and Notes (to be filled in LaTeX)
reference <- notes <- rep("", length(slugs))
notes[2] <- "2 + 9 taxa removed"
notes[3] <- "12 + 19 taxa removed. Tree was randomly selected from all time-scaled trees"
notes[4] <- "12 + 0 taxa removed"
notes[5] <-  "Maximum clade credibility tree, time-scaled by dividing branches by 0.03517385 and setting root time to 485.4"

## Formatting the table
table_out <- data.frame(group, ntaxa, mcharacters, age_ranges, reference, notes)
colnames(table_out) <- colnames
rownames(table_out) <- dataset

## Table caption
table_caption <- "Details of the datasets used in this study."

sink(file = "../manuscript/tables/datasets.tex")
xtable(table_out, caption = table_caption)
sink()
```

## Time subsets table

### 2017-12-04 manuscript/tables/datasets.tex
Markdown to easily build the tables (example done with table 1)

--- 2017-11-29 manuscript/tables/datasets.tex
@@
-% Dataset descriptions table
-
-\begin{tabular}[t]{l p{3cm} l l l l}
-
+% latex table generated in R 3.4.2 by xtable 1.8-2 package
+% Mon Dec  4 13:30:53 2017
+\begin{table}[ht]
+\centering
+\begin{tabular}{rlrrlll}
   \hline
-\textbf{Dataset} & \textbf{Group} & \textbf{$N_{taxa}$} & \textbf{$N_{characters}$} & \textbf{Age range} & \textbf{Reference} \\
+ & Group & n taxa & m characters & Age range & Reference & Notes \\ 
   \hline
-Beck2014 & Eutherian mammals & 102 & 410 & 170 Ma - present & X \\
-%-----------------------------------------------------------
-Bapst2016 & theropods & 102 & 410 & 170 Ma - present & X \\
-%-------------------------------------------------------
-Brusatte2014 & theropods & 102 & 410 & 170 Ma - present & X \\
-%-------------------------------------------------------
-Heath2016 & bears & 102 & 410 & 170 Ma - present & X \\
-%-------------------------------------------------------
-Wright2017 & crinoids & 102 & 410 & 170 Ma - present & X \\
-%-------------------------------------------------------
+Beck2014 & Eutherian mammals & 106 & 421 & 171.786 - 0 &  &  \\ 
+  Brusatte2014 & theropods & 152 & 853 & 102.535 - 0 &  & 2 + 9 taxa removed \\ 
+  Bapst2016 & theropods &  89 & 374 & 141.216 - 0 &  & 12 + 19 taxa removed. Tree was randomly selected from all time-scaled trees \\ 
+  Heath2016 & bears &  18 &  62 & 33.241 - 0 &  & 12 + 0 taxa removed \\ 
+  Wright2017 & crinoids &  42 &  87 & 485.4 - 372.2 &  & Maximum clade credibility tree, time-scaled by dividing branches by 0.03517385 and setting root time to 485.4 \\ 
    \hline
 \end{tabular}
+\caption{Details of the datasets used in this study.} 
+\end{table}

### 2017-12-04 analyses/04-stats.Rmd
Added extinction tests

@@
 library(broom)
+library(dispRity)
 ```
@@
 Make the tables
+
 ```{r, eval = TRUE, echo = FALSE}
@@
 You can rerun for the other datasets by changing the slug name at the start.
+
+Extinction effect
+----------
+
+Here we can test the effect of mass extinction on the different methods with the different models (for each dataset!).
+The way to test the extinction effect is a referential wilcoxon test.#' @param as.list \code{logical}, whether to output the results as a list for \code{\link{test.dispRity}} (\code{TRUE}) or not (\code{FALSE} - default).
+
+In other words, we compare the bins/slice just before the extinction event to the three ones just after (to account for an eventual lag effect).
+
+Set the slug:
+
+```{r}
+slug <- "Beck2014" ; extinction <- 66.0
+# slug <- "Brusatte2014"; extinction <- 66.0
+# slug <- "Bapst2016"; extinction <- 66.0
+# slug <- "bears"; extinction <- NA
+# slug <- "Wright2017"; extinction <- 443.0
+```
+
+Read in the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
+out2 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
+```
+
+Measure differences in disparity:
+
+```{r}
+## Get the extinction times (ages)
+ext_time_strat_age <- lapply(out1$objects$stratigraphy, extinction.subsets, extinction = extinction, lag = 3, as.list = TRUE)
+ext_time_durat_age <- lapply(out1$objects$duration, extinction.subsets, extinction = extinction, lag = 3, as.list = TRUE)
+ext_time_numbe_age <- lapply(out1$objects$number, extinction.subsets, extinction = extinction, lag = 3, as.list = TRUE)
+
+## Get the extinction times (epochs)
+ext_time_strat_epo <- lapply(out2$objects$stratigraphy, extinction.subsets, extinction = extinction, lag = 3, as.list = TRUE)
+ext_time_durat_epo <- lapply(out2$objects$duration, extinction.subsets, extinction = extinction, lag = 3, as.list = TRUE)
+ext_time_numbe_epo <- lapply(out2$objects$number, extinction.subsets, extinction = extinction, lag = 3, as.list = TRUE)
+
+## Test the effect of extinctions
+mapply.test <- function(data, times, test = wilcox.test, correction = "bonferroni") {
+  test <- try(output <- test.dispRity(data, test = test, comparisons = times, correction = correction), silent = TRUE)
+  if(class(test) == "try-error") {
+
+    ## Create an NA list
+    comparison_names <- unlist(lapply(times, function(time, data) paste(names(data$subsets)[time], collapse = " : "), data))
+    p_value <- statistic <- data.frame(rep(NA, 3), row.names = comparison_names)
+    colnames(statistic) <- "statistic: W"
+    colnames(p_value) <- "p.value"
+
+    return(list(statistic, p_value))
+  } else {
+    return(output)
+  }
+}
+
+## Applying the tests (ages)
+test_strat_age <- mapply(mapply.test, out1$objects$stratigraphy, ext_time_strat_age, SIMPLIFY = FALSE)
+test_durat_age <- mapply(mapply.test, out1$objects$duration, ext_time_durat_age, SIMPLIFY = FALSE)
+test_numbe_age <- mapply(mapply.test, out1$objects$number, ext_time_numbe_age, SIMPLIFY = FALSE)
+
+## Applying the tests (epochs)
+test_strat_epo <- mapply(mapply.test, out2$objects$stratigraphy, ext_time_strat_epo, SIMPLIFY = FALSE)
+test_durat_epo <- mapply(mapply.test, out2$objects$duration, ext_time_durat_epo, SIMPLIFY = FALSE)
+test_numbe_epo <- mapply(mapply.test, out2$objects$number, ext_time_numbe_epo, SIMPLIFY = FALSE)
+
+## Combining the tests into tables
+wrap.up.results <- function(results, type, method) {
+  results <- data.frame(matrix(unlist(results), ncol = 6, byrow = TRUE))
+  ## Adding the models
+  results <- cbind(c("bins", "acctran", "deltran", "random", "proximity", "punctuated", "gradual"),results)
+  ## Adding the type and method
+  results <- cbind(matrix(c(rep(type, 7), rep(method, 7)), ncol = 2), results)
+  ## Adding the column names
+  colnames(results) <- c("type", "method", "model", "e:1(W)", "e:2(W)", "e:3(W)", "e:1(p)", "e:2(p)", "e:3(p)")
+  return(results)
+}
+
+## Combining the results
+result <- wrap.up.results(test_strat_age, "age", "stratigraphy")
+result <- rbind(result, wrap.up.results(test_durat_age, "age", "duration"))
+result <- rbind(result, wrap.up.results(test_numbe_age, "age", "number"))
+result <- rbind(result, wrap.up.results(test_strat_epo, "epoch", "stratigraphy"))
+result <- rbind(result, wrap.up.results(test_durat_epo, "epoch", "duration"))
+result <- rbind(result, wrap.up.results(test_numbe_epo, "epoch", "number"))
+
+## Significant results
+result <- cbind(results, apply(result[,7:9], 2, function(col) ifelse(col < 0.05, TRUE, FALSE)))
+
+## Print the table
+knitr::kable(result)
+```

### 2017-12-04 manuscript/time-slice.tex
added couple of comments on the manuscript
Switch citation styles
Edits to methods

@@
 \section{Introduction}
+%TG: Plan idea
+%TG: §1: disparity-through-time really cool and well used
+%TG: §2: caveats with both disparity and through-time part
+%TG: §3: specific ways to look through-time and their problems
+%TG: §4: our kind of solution
+
 Disparity-through-time analyes are common in palaeontology.
@@
 However, our methods for analysing diversity-through-time have not changed all that much in the XX years they have been popular.
+%TG: small caveat on that after litterature search for the disparity review paper: disparity-through-time analysis have not changed much since Foote 1992, Foote and Gould 1992 and Briggs et al.
+1992 in the classic palaeobiology approach...
+%TG: However to not piss of people, there were several attempts to "rationalise" (Wills 2001), test (Ciampaglio 2001) and expand (Pavlinov 2011 and Gerber et al 2011) namely with univariate approaches and "dtt" analysis (Harmon et al 2003).
+%TG: One of the critisisms from one of the rude reviews was this idea that age of popular method has nothing to do with its performance, by nuancing this sentence we can basically say: yeah, some people have made great advances and proposed great ideas for CORRECTING ERRORS in the methods from the 90s but somehow everybody have ignored them (in palaeobiology).
 The way we perfrom these analyses may have profound effects on our conclusions.
@@
 
-We present this using five datasets collated from the literature.
+We present this using four datasets collated from the literature.
 
@@
 % I'm unsure if this section is needed or not. The methods are long so it might help?
-To test the different time subsampling methods, we followed the protocol below. 
-Note that all the code needed to reproduce these analyses (along with detailed instructions) is provided HERE.
+%TG: agree though we might just do a simple figure like fig1 in http://onlinelibrary.wiley.com/doi/10.1111/bij.12746/pdf (but with way less stuff!). I can get working on that.
+To test the different time subsampling methods, we followed the protocol below (Figure). 
+All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (https://github.com/nhcooper123/time-slice). % NC: might move this to the NHM github account
 
-\begin{enumerate}
-  \item Collating example datasets.
+%\begin{enumerate}
+ % \item Collating example datasets.
 Disparity-through-time analysis requires a matrix containing morphological characters for each taxon, and information on the first and last occurrence dates of each taxon.
 Time slicing methods also require a phylogeny.
-  \item Processing example datasets.
+  %\item Processing example datasets.
 Various transformations of the datasets are required before the analyses can take place.
 We need to estimate ancestral character states at the nodes of the phylogeny for the time slicing methods, build cladisto-spaces via distance matrices and ordination, and remove taxa from the phylogeny that are missing from the matrix and vice versa.
-  \item Time subsampling
-  \item Estimate disparity for each time subsample
-  \item Compare different methods
-\end{enumerate}
+  %\item Time subsampling
+  %\item Estimate disparity for each time subsample
+  %\item Compare different methods
+%\end{enumerate}
 
@@
 \label{datasets}
-To test the different time binning/slicing methods we selected five datasets (see Table \ref{datasets}). 
-Note that although disparity-through-time analyses are possible without a phylogeny, we only use examples with phylogenies here so that we can compare results of time slicing methods, which do require a phylogeny.
-Each dataset consists of a set of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
-More details on these datasets can be found in the appendix. All data are available HERE. And the processing code too. Probably also want to link to Dryad etc. HERE
+To test the different time binning/slicing methods we selected four datasets: a mammal dataset from \cite{beckancient2014}, two theropod datasets from \cite{brusatte2014gradual} and \cite{bapst2016topology}, and a crinoid dataset from \cite{wright2017bayesian}.
+Table \ref{datasets} and appendix ??? provide more details. % add appendix details
+%Note that although disparity-through-time analyses are possible without a phylogeny, we only use examples with phylogenies here so that we can compare results of time slicing methods, which do require a phylogeny.
+Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
+These datasets are freely available with their accompanying papers (Table \ref{datasets}), but for reproducibility purposes we also provide the data we used on the NHM Data Portal (CITE).
 
-  \begin{table}[!htbp]      
-    \caption[Datasets.]
-    {Dataset captions}
+  % \begin{table}[!htbp]      
+  %   \caption[Datasets.]
+  %   {Dataset captions}
     \input{tables/datasets.tex} 
     \label{tab:data}  
-  \end{table}
+  % \end{table}
+% remove notes, add strat ages, add refs to dat sources
 
@@
 \label{ace}
-For each dataset we used the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny, using the \texttt{rerootingMethod} function from the \texttt{R} package \texttt{phytools} version 0.4-45 \citep{phytools,R}.
-% Version still the same?
+For each dataset we used the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny, using the \texttt{rerootingMethod} function from the \texttt{R} package \texttt{phytools} version 0.6-44 \citep{phytools,R}.
 Where there were missing character data for a taxon we followed the method of \cite{Claddis} and treated missing data as any possible observed state for each character.
@@
 Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis} to remove pairs of taxa with no cladistic characters in common.
-This led to us removing nine taxa from the BAPST dataset, and 19 from the Brusatte dataset, but none from the other three datasets (see code for details of which species). % add citations
+This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see appendix for details of which species).
+% add appendix
 
@@
 
-\subsection{Estimating disparity-through-time} 
-\label{time_slicing}
+\subsection{Estimating disparity-through-time: time subsampling} 
+\label{time_subsamples}
 
@@
     \begin{enumerate}
-      \item The *duration* of the bin is equal to the median duration of the stratigraphic period (age or epoch).
-      \item The *number* of bins is equal to the number of stratigraphic periods (ages or epochs).
-      \item The *number* of bins is either 2, 5, 10, 15, or 20.
+      \item The \textit{duration} of the bin is equal to the median duration of the stratigraphic period (age or epoch).
+      \item The \textit{number} of bins is equal to the number of stratigraphic periods (ages or epochs).
+      \item The \textit{number} of bins is either 2, 5, 10, 15, or 20.
     \end{enumerate}
   \item Time slicing.
-This method uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation.
+We describe this in more detail below, but in brief, time slicing uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation.
 Here we use three different methods to determine the intervals between the time slices.
     \begin{enumerate}
-      \item The *interval* between slices is equal to the median duration of the stratigraphic period (age or epoch).
-      \item The *number* of slices is equal to the number of stratigraphic periods (ages or epochs).
-      \item The *number* of slices is either 2, 5, 10, 15, or 20.
-    \end{enumerate}  
+      \item The \textit{interval} between slices is equal to the median duration of the stratigraphic period (age or epoch).
+      \item The \textit{number} of slices is equal to the number of stratigraphic periods (ages or epochs).
+      \item The \textit{number} of slices is either 2, 5, 10, 15, or 20.
     \end{enumerate}
 
-% Might make a table with all the subsets???
-% Had a play - it's COMPLICATED!!!
+\end{enumerate}
 
-  \begin{table}[!htbp]      
-    \caption[Subsamples.]
-    {captions}
-    \input{tables/timesamples.tex} 
-    \label{tab:subsamples}  
-  \end{table}
-
-% Just tips or tips and nodes????
-\paragraph{Stratigraphic time bins.}
-We extracted the taxa found in each geological age or epoch from each cladisto-space.
-For each subsample, we estimated its disparity.
-To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement by randomly resampling with replacement a new subsample of taxa from the observed taxa in the subsample 1000 times.
-We then calculated the median disparity value for each subsample along with the 50\% and 95\% confidence intervals.
-We also recorded the number of taxa in each subsample as a proxy for taxonomic diversity (Table \ref{tab:subsamples}).
-
-\paragraph{Equally sized time bins.}
-We extracted the taxa found in each time bin for each cladisto-space.
-We varied the size of the time bins so that either (i) the *duration* of the bin is equal to the median duration of the stratigraphic age or epoch, (ii) the *number* of bins is equal to the number of stratigraphic ages or epochs, (iii) or by using 2, 5, 10, 15, or 20 bins.
-For each subsample, we estimated its disparity and bootstrapped each disparity measurement as described above for stratigraphic time bins.
-We also recorded the number of taxa in each subsample as a proxy for taxonomic diversity (Table \ref{tab:subsamples}).
-
-\paragraph{Time slicing.} % THOMAS TO UPDATE THIS BIT!!!
+\subsection{Estimating disparity-through-time: time slicing} 
+\label{time_slicing}
+% THOMAS TO UPDATE THIS BIT!!!
 The ``time-slicing'' approach considers subsets of taxa in the cladisto-space at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
@@
 
-\subsection{Comparing the results}
-%Testing the difference section
+\subsection{Estimating disparity-through-time: bootstrapping}
+Finally, to reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time subsampling method, by randomly resampling with replacement a new subsample of taxa from the observed taxa in the subsample 1000 times.
+We then calculated the median disparity value for each subsample along with the 50\% and 95\% confidence intervals.
+We also recorded the number of taxa (or taxa and nodes for time slicing methods) in each subsample as a proxy for taxonomic diversity (Table \ref{tab:subsamples}).
+
+\subsection{Comparing time subsampling methods}
+\label{testing}
+Testing for statistical differences among the time subsampling methods described above is a little tricky, as we need to compare similar units, and also to tackle questions important to the interpretation of disparity-through-time analyses.
+We therefore present three different, simple ways of comparing the time subsampling methods as follows.
+
+\paragraph{Systematic differences in disparity-through-time.} 
+To test whether using time bins or time slices resulted in significantly different disparity values at common time points, we used paired Wilcoxon tests to compare the following.
+
+\begin{enumerate}
+  \item Unequal time bins and unevenly spaced time slices, where the bins are stratigraphic periods (ages or epochs) and slices occur at the midpoint of the stratigraphic period.
+  \item Equal time bins and equally spaced time slices, where time bin size, or the interval between slices, is defined by the average \textit{duration} of the stratigraphic period (age or epoch).
+  \item Equal time bins and equally spaced time slices, where the number of time bins, or number of slices, is defined by the average \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest.
+\end{enumerate}
+
+Due to the uneven spread of taxa across phylogenies, some time bins will contain one or no species, meaning that we cannot estimate disparity for that time bin. 
+We first, therefore, remove the time bins, and corresponding time slices, without disparity estimates. 
+We then perform paired Wilcoxon tests, so that bins and slices for the same time period are being compared. 
+Significant results suggest that there is a systematic difference in disparity values at each time point, depending on whether bins or slices are used.
+
+\paragraph{Disparity peaks and troughs.}
+We are perhaps more interested in how the conclusions disparity-through-time analyses are influenced by the choice of time subsampling method, rather than the disparities estimated by each method \textit{per se}, especially as these will influenced by the number of taxa (and/or nodes) included in each subsample.
+Therefore, we also investigate where peaks and troughs of disparity occur in each of our datasets for the different time subsampling methods. 
+We calculated the maximum and minimum bootstrapped disparities for each dataset and for each time subsampling method, along with their associated confidence intervals.
+Significant shifts in disparity peaks or troughs suggest that the choice of time subsampling method will influence our conclusions about relative changes in the disparity of our example clades through time.
+
+\paragraph{Effects of mass extinction events.}
+Many analyses of disparity-through-time aim to demonstrate differences in disparity before and after mass extinction events [citations]. 
+Each of our four datasets span one mass extinction (Cretaceous-Paleogene 66 MYA; \cite{brusatte2014gradual,bapst2016topology,beckancient2014} and Ordovician- XX MYA \cite{wright2017bayesian}), so we use Wilcoxon tests to compare disparity in the time bin/slice prior to the approrpiate mass extinction, to that of the time bin/slice following the extinction event.
+Significant results suggest a significant effect of the mass extinction on disparity in the group.
+We then compare these results across the time subsampling methods to determine if our conclusions would change depending on the method used.
+We repeated these analyses using the two time bins/slices after the one immediately following the mass extinction event to account for any lag effects of the mass extinction on disparity.
 
@@
 
-\bibliographystyle{sysbio} 
+\bibliographystyle{palaeo} 
 \bibliography{time-refs}

### 2017-12-05 manuscript/time-slice.tex
Some comments added to results
updated models description

@@
     Similarly to the time interval approach, this reflects a model of punctuated evolution where changes in disparity occur either at the start or at the end of a branch over a relatively short time period and clades undergo long periods of stasis during their evolution \citep{Gould1977,Hunt20112007}.
-    We applied this model in three ways: 
+    We applied this model in four ways:
+
     \begin{enumerate}[(i)]
@@
       \item The ``random'' model, selecting randomly between the ordination score of the ancestor or the descendant of the branch.
+      \item The ``proximity'' model, selecting the ordination of either the ancestor or the descendant if the slice occurs respectively before or after half the branch length.
+
+    \end{enumerate}
+
+    The two first models assume that changes always occur early (ACCelerated TRANsition) or late along the branches (DELayed TRANsition).
+    The third model makes neither assumption and simply selects randomly the data between the ancestor and the descendant and the fourth base the selection of either the ancestor or the descendant on where the slice occurs along the branch.
+    The punctuated models only select either the ordination score from the ancestor and the descendant once in the whole disparity analysis.
+    For example, if using the ``random'' model, if the data of the ancestor has been randomly chosen, only this data will be used during the bootstraps procedure and for the disparity calculation.
+
+    \item{\textbf{Gradual evolution.}}
+    Contrary to the punctuated models, these following models do not select the ordination score of either the ancestor or the descendant but associate a probability to both.
+    This reflects a model of gradual evolution where changes in disparity are gradual and cumulative along the branch.
+
+    \begin{enumerate}[(i)]
+    \setcounter{enumii}{4}
       \item The ``punctuated'' model (probabilistic), selecting the ordination score from both the ancestor and the descendant with an equal probability
@@
 
-    \end{enumerate}
-
-    The two first models assume that changes always occur early (ACCelerated TRANsition) or late along the branches (DELayed TRANsition).
-    The third model makes neither assumption and simply selects randomly the data between the ancestor and the descendant.
-    These three first models do not allow the data to vary throughout eventual pseudo-replications in the analysis (the bootstraps).
-    For example, using the ``random'' model, once the the data of the ancestor has been randomly chosen, only this data will be used during the bootstraps procedure.
-    Conversely, the ``punctuated'' model allows both ancestor's and descendant's data to be used during the bootstrapping with an equal probability.
-    This results in the bootstrapped disparity at a specific slice to effectively incorporate both the data from the descendant and the ancestor.
-%TG: now that I'm mentioning that, I should maybe rename this model.
-Lets discuss that on monday.
-
-    \item{\textbf{Gradual evolution.}}
-    These model also selects the ordination score from either the ancestral node or the descendant node/tip of the branch, but the choice depends on the distance between the sampling time point and the end of the branch.
-    If the sampling time point falls in the first half of the branch length the ordination score is taken from the ancestral node, conversely, if the sampling time point falls in the second half of the branch length the ordination score is taken from the descendant node/tip.
-    This reflects a model of gradual evolution where changes in disparity are gradual and cumulative along the branch.
-
-    \begin{enumerate}[(i)]
-    \setcounter{enumii}{4}
-    \item The ``proximity'' model, selecting the ordination of either the ancestor or the descendant if the slice occurs respectively before or after half the branch length.
     \item The ``gradual'' model (probabilistic), selecting the ordination score from both the ancestor and the descendant with a probability equal to the length between the slice.
@@
 
-    The ``proximity'' model is a variation of the ``random'' one described above were the selection of the data is not random but arbitrarily based on branch length.
-    However, similarly to the ``random'' one, this model doesn't allow variation.
-    The last ``gradual'' model is also a variation of the ``punctuated'' and allows variation in a similar way (but based on branch length).
-    For example, if slice occurs at the third quarter of a branch from A to B (75\% of the total branch length), after bootstrapping, the disparity results will be based on 25\% of the data from A and 75\% of the data from B.
+    In these models, both the ancestor's and descendant's data will participate to the disparity calculation based on the following models:
+    For example, using the ``gradual.split'' model, if slice occurs at the third quarter of a branch from A to B (75\% of the total branch length), after bootstrapping, the disparity results will be based on 25\% of the data from A and 75\% of the data from B.
+    Because of the probabilistic nature of these models, there are only meaningful when calculating disparity from bootstrapped datasets.
 \end{enumerate}
@@
 We also recorded the number of phylogenetic elements (nodes and tips) in each subsample as a proxy for taxonomic diversity (Table \ref{tab:subsamples}).
+
+% Mention somewhere here that we focus on the proximity and gradual slice methods in the results.
 
@@
 
+\subsection{Estimating disparity-through-time}
+
+The full results tables can be found in appendix ??? % details.
+
+% Remind somewhere here that we focus on the proximity and gradual slice methods in the results. But full results are in appendix
+
+\subsection{Comparing time subsampling methods}
+% some demonstration of the differences in the binning - the ladder versus lines plot
+
+\paragraph{Systematic differences in disparity-through-time.} 
+% table? plots? Might be same as the plots above.
+
+\paragraph{Disparity peaks and troughs.}
+% Plots showing the differences
+
+\paragraph{Effects of mass extinction events.}
+% Big heat map table thing
+
 %Fig. or Figs
@@
 
+Benefits of time slices
+
+Further stuff to fix?
+
 \section{Conclusions}

### 2017-12-05 analyses/04-stats.Rmd
extinction csv saved

@@
 ## Significant results
-result <- cbind(results, apply(result[,7:9], 2, function(col) ifelse(col < 0.05, TRUE, FALSE)))
+result <- cbind(result, apply(result[,7:9], 2, function(col) ifelse(col < 0.05, TRUE, FALSE)))
 
@@
 knitr::kable(result)
+
+## Save the results
+write.csv(result, file = paste0("../outputs/extinction.", slug, ".csv"), quote = FALSE, 
+          row.names = FALSE)
+
 ```

### 2017-12-05 functions/curve.differences.R
removed old unused functions

@@ -1,148 +1,0 @@
-#' @title Curves differences
-#'
-#' @description Getting the differences between two curves
-#'
-#' @param x The \textit{x} axis coordinates
-#' @param y1 The \textit{y} coordinates for the first curve
-#' @param y2 The \textit{y} coordinates for the second curve (can be \code{NULL})
-# @param x2 Optional, the \textit{x} coordinates for the second curve (can be \code{NULL})
-#' 
-#' @examples
-#'
-#' @seealso
-#' 
-#' @author Thomas Guillerme
-#' @export
-
-absolute.curve.area <- function(x, y1, y2 = NULL) { # Add and x2 = NULL option if Xs don't match
-
-    ## Number of points
-    n <- length(x)
-
-    ## If y2 is null, get the area from the minimum y1 value
-    if(is.null(y2)) {
-        y2 <- rep(min(y1), n)
-    }
-
-    ## Detect any crossing points
-    crossing_points <- pracma::poly_crossings(rbind(x, y1), rbind(x, y2))
-
-    if(is.null(crossing_points) || length(crossing_points) == 0 ) {
-        ## Get the polygon coordinates
-        polygon_coords <- get.poly.coords(x, y1, y2)
-        return(abs(pracma::polyarea(polygon_coords[1,], polygon_coords[2,])))
-
-    } else {
-
-        ## Number of crossing points
-        n_crossing_points <- nrow(crossing_points)
-
-        ## Check if the two curves have the same starting final point
-        if(all(c(x[1], y1[1]) == c(x[1], y2[1])) || all(c(x[n], y1[n]) == c(x[n], y2[n]))) {
-            ## Check if the all the crossings are the starting point
-            if(all(c(x[1], y1[1]) == crossing_points) || all(c(x[n], y1[n]) == crossing_points)) {
-                ## Get the polygon coordinates
-                polygon_coords <- get.poly.coords(x, y1, y2)
-                return(abs(pracma::polyarea(polygon_coords[1,], polygon_coords[2,])))
-            } 
-        } else {
-
-            ## Remove first crossing point if necessary
-            if(all(c(x[1], y1[1]) == crossing_points[1,])) {
-                crossing_points <- crossing_points[-1,]
[…]

### 2017-12-05 functions/get.bin.ages.fix.R

@@ -1,74 +1,0 @@
-#' @title Get time bins ages
-#'
-#' @description Gets time bins for a specific tree using stratigraphy
-#'
-#' @param tree A \code{phylo} object with a \code{$root.time} component
-#' @param what Which data to output. Can be \code{"Start"}, \code{"End"} (default), \code{"Range"} or \code{"Midpoint"}.
-#' @param type The type of stratigraphic frame. Can be \code{"Age"} (default), \code{"Eon"}, \code{"Epoch"}, \code{"Era"} or \code{"Period"}.
-#' @param ICS The reference year of the International Commission on Stratigraphy (default = \code{2015}).
-#' 
-#' @examples
-#' ## Loading the data
-#' data(BeckLee_tree)
-#' data(BeckLee_mat50)
-#' 
-#' ## Getting the stratigraphic data
-#' stratigraphy <- get.bin.ages(BeckLee_tree)
-#' 
-#' ## Making stratigraphic time subsets
-#' time.subsets(BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
-#'                 time = stratigraphy)
-#'
-#' @seealso \code{\link{time.subsets}}
-#' 
-#' @author Thomas Guillerme
-#' @import geoscale timescales
-#' @export
-
-get.bin.ages <- function(tree, what = "End", type = "Age", ICS = 2015) {    
-    ## Tree
-    # check.class(tree, "phylo")
-    if(is.null(tree$root.time)) {
-        stop("The tree must have a root time element (tree$root.time).")
-    }
-
-    # ## What
-    # check.method(what, c("Start", "End", "Range", "Midpoint"), msg = "'what' argument")
-
-    # ## Type
-    # check.method(type, c("Age", "Eon", "Epoch", "Era", "Period"), msg = "'type' argument")
-
-    # ## ICS
-    # check.class(ICS, "numeric")
-    # check.length(ICS, 1, errorif = FALSE, msg = " must be a year from 2008 to the present.")
-    ICS_available <- match(paste0("ICS", ICS), names(geoscale::timescales))
-    if(is.na(ICS_available)) {
-        stop(paste0("No stratigraphic data found for ", paste0("ICS", ICS), ".\nAvailable years are: ", paste(names(geoscale::timescales), collapse = ", "), "."))
-    } else {
-        ## Selecting the ICS
-        stratigraphy <- geoscale::timescales[[paste0("ICS", ICS)]]
-        ## Selecting the column of interest
[…]

### 2017-12-05 functions/tree.age.fossil.R

@@ -1,99 +1,0 @@
-#' @title Calculating the age of nodes and tips in a tree.
-#'
-#' @description Calculates the age of each node and tip in a tree give the height of the tree or some specified age.
-#'
-#' @param tree A \code{phylo} object.
-#' @param age The age of the tree. If missing the age is set to be the tree height.
-#' @param order Either "past" if the units express time since the present (e.g. million years ago), or "present" if the unit is expressed in time since the root.
-#' @param fossil.only \code{logical}, whether the tree contains only fossil taxa (\code{TRUE}) or not (\code{FALSE} - default). If the \code{tree} contains a \code{tree$root.time} element (for tree's root age), and that \code{order} is set to \code{"past"}, the ages in the tree are adjusted to be starting from the root.time, resulting in tips potentially older than age 0.
-#'
-#' @examples
-#' ## A dated random phylogeny with a root 50 units of time old.
-#' tree.age(rtree(10), age = 50)
-#' ## A random tree with the distance since the root.
-#' tree.age(rtree(10), order = 'present')
-#'
-#' @seealso \code{\link{slice.tree}}, \code{\link{time.subsets}}.
-#'
-#' @author Thomas Guillerme
-
-#Modified from [R-sig-phylo] nodes and taxa depth II - 21/06/2011 - Paolo Piras - ppiras(at)uniroma3.it
-
-tree.age.fossil <- function(tree, age, order = 'past'){
-
-#SANITYZING
-#FUNCTIONS FOR tree.age
-
-#Extract the ages table from a tree
-#Calculating the tips and the elements age
-tree.age_table<-function(tree){
-    tree_length <- Ntip(tree)
-    
-    ages <- dist.nodes(tree)[tree_length + 1,]
-    tip.names <- tree$tip.label
-    if(is.null(tree$node.label)) {
-        nod.names <- c((tree_length + 1):length(dist.nodes(tree)[, 1]))
-    } else {
-        nod.names <- tree$node.label
-    }
-    elements <- c(tip.names, nod.names)
-    ages.table <- data.frame(ages = ages,elements = elements)
-    return(ages.table)
-}
-
-
-#Scaling the ages from tree.age_table if scale != 1
-tree.age_scale <- function(ages.table, scale){
-    ages.table$ages <- ages.table$ages/max(ages.table$ages)
-    ages.table$ages <- ages.table$ages*scale
-    return(ages.table)
-}
-
-    # #tree
[…]

### 2017-12-05 analyses/figures-compiler.Rmd
extinction fable plot

---
title: "Figures compiler"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Compiling the figures for the paper

Loading the code:

```{r}
source("../functions/figures_fun.R")
```

## Disparity peaks figures (maximum/minimum)

## Extinction effect figures

Initialising the slug and loading the data:

```{r}
## Slugs (choose one)
slug <- "Beck2014"
# slug <- "Brusatte2014"
# slug <- "Bapst2016"
# slug <- "Wright2017"

data <- read.csv(paste0("../outputs/extinction.", slug, ".csv"), stringsAsFactors = FALSE)
```

Single plots:

```{r, fig.width = 6, fig.height = 6}
colours <- c("grey", "white", "black") #Signif, non-signif, NA
plot.extinction(data, type = "age", xaxis = TRUE, yaxis = TRUE, xaxis2 = TRUE, main = slug, cols = colours)
```

Multiple plots:

```{r, fig.width = 12, fig.height = 4}
## Names of the datasets
slugs <- as.list(c("Beck2014", "Brusatte2014", "Bapst2016", "Wright2017"))

## Load all the data
data_list <- lapply(slugs, function(slug) read.csv(paste0("../outputs/extinction.", slug, ".csv"), stringsAsFactors = FALSE))

## Set up the colours (yeah, they're ugly for now)
colours <- c("lightblue", "yellow", "white") #Signif, non-signif, NA

## Plot all the results
multi.plot.extinction(data_list, type = c("age", "epoch"), data.names = unlist(slugs), cols = colours)
```

### 2017-12-05 functions/figures_fun.R

@@ -1,0 +1,324 @@
+#' @title Plotting TreeCmp results
+#'
+#' @description Simple wrapper for plotting the results from TreeCmp
+#'
+#' @param taxa_list a list of tree metrics from \code{read.TreeCmp} for all the scenario for a number of taxa
+#' @param metric a numeric of which metric to use (\code{1} = \code{MatchingCluster}, \code{2} = \code{R.F_Cluster}, \code{3} = \code{NodalSplitted}, \code{4} = \code{Triples})
+#' @param what which type of plot? Can be \code{"null"}, \code{"real"}, \code{"true"}, \code{"check"}. See details.
+#' @param ylim \code{ylim} parameter from \code{plot} (if missing is automatic)
+#' @param legend \code{logical}, whether to display the legend
+#' @param NTS \code{logical}, whether NTS was used
+#' @param ylab a \code{character} string to be passed to \code{boxplot(..., ylab)}
+#' @param xlab a \code{character} string to be passed to \code{boxplot(..., xlab)}
+#' @param axislab a \code{character} string to be passed to \code{axis(..., labels)}
+#' @param ... any additional argument to be passed to \code{boxplot}
+#'
+#' @details
+#' \code{"null"} = "maxi", "mini", "norm" tree vs. "rand"
+#' \code{"best"} = "maxi", "mini", "rand" tree vs. "norm"
+#' \code{"true"} = "norm", "rand" tree vs. "true"
+#' \code{"check"} = "norm", vs "rand"
+#' 
+#' @return
+#' A list of metrics for the comparisons.
+#' 
+#' @examples
+#' ##
+#' 
+#' @author Thomas Guillerme
+#' @export
+#' 
+
+
+plot.results.single <- function(taxa_list, metric, what, ylim, legend = FALSE, NTS = TRUE, ylab, xlab, axislab, ...) {
+    
+    ## Graphic parameters
+    ## Colours
+    cols <- c("red", "orange", "green3", "lightgreen", "blue", "lightblue")
+    
+    ## Y lab
+    if(missing(ylab)) {
+        if(metric == 1) ylab <- "Similarity (Matching Cluster)"
+        if(metric == 2) ylab <- "Similarity (Robinson-Foulds)"
+        if(metric == 3) ylab <- "Similarity (Nodal Split)"
+        if(metric == 4) ylab <- "Similarity (Triplets)"
+    }
+
+    ## Y lim
[…]

### 2017-12-05 analyses/04-stats.Rmd
added max and min tables

@@
 
-strat <- 
+strat_max <- 
   strat %>%
@@
 
-duration <-
+duration_max <-
   duration %>%
@@
 
-number <- 
+number_max <- 
   number %>%
@@
   filter(bs.median == max(bs.median, na.rm = TRUE))
+
+strat_min <- 
+  strat %>%
+  group_by(model) %>%
+  filter(bs.median == min(bs.median, na.rm = TRUE))
+
+duration_min <-
+  duration %>%
+  group_by(model) %>%
+  filter(bs.median == min(bs.median, na.rm = TRUE))
+
+number_min <- 
+  number %>%
+  group_by(model) %>%
+  filter(bs.median == min(bs.median, na.rm = TRUE))
 
@@
 
-strat2 <- 
+strat2_max <- 
   strat2 %>%
@@
 
-duration2 <-
+duration2_max <-
   duration2 %>%
@@
 
-number2 <- 
+number2_max <- 
   number2 %>%
@@
 
-all_peaks <- rbind(strat, duration, number, strat2, duration2, number2)
-write.csv(all_peaks, file = paste0("../outputs/peaks.", slug, ".csv"), quote = FALSE, 
+strat2_min <- 
+  strat2 %>%
+  group_by(model) %>%
+  filter(bs.median == min(bs.median, na.rm = TRUE))
+
+duration2_min <-
+  duration2 %>%
+  group_by(model) %>%
+  filter(bs.median == min(bs.median, na.rm = TRUE))
+
+number2_min <- 
+  number2 %>%
+  group_by(model) %>%
+  filter(bs.median == min(bs.median, na.rm = TRUE))
+
+all_peaks_max <- rbind(strat_max, duration_max, number_max, strat2_max, duration2_max, number2_max)
+write.csv(all_peaks_max, file = paste0("../outputs/peaks.max.", slug, ".csv"), quote = FALSE, 
           row.names = FALSE)
 
+all_peaks_min <- rbind(strat_min, duration_min, number_min, strat2_min, duration2_min, number2_min)
+write.csv(all_peaks_min, file = paste0("../outputs/peaks.min.", slug, ".csv"), quote = FALSE, 
+          row.names = FALSE)
+}
 ```

### 2017-12-05 analyses/figures-compiler.Rmd
Added individual plots for max/min plots

@@
 ```{r}
+library(ape)
+library(dispRity)
 source("../functions/figures_fun.R")
@@
 ## Disparity peaks figures (maximum/minimum)
+
+Initialising the slug and loading the data:
+
+Initialising the slug and loading the data:
+
+```{r}
+## Slugs (choose one)
+slug <- "Beck2014"
+# slug <- "Brusatte2014"
+# slug <- "Bapst2016"
+# slug <- "Wright2017"
+
+## Load the data
+data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"), stringsAsFactors = FALSE)
+
+## Get the relative age
+relative_time <- read.nexus(file = paste0("../data/processed/tree.", slug, ".tre"))
+if(slug == "Wright2017") {
+    relative_time <- 484.4
+} else {
+    relative_time <- max(tree.age(relative_time)$age)
+}
+```
+
+Single plots:
+
+```{r, fig.width = 12, fig.height = 6}
+op <- par(mfrow = c(1,2))
+plot.max.min(data_max, relative.time = relative_time, xaxis = TRUE, yaxis = TRUE, pch = 19, main = "Relative age at maximum disparity in Beck2014")
+plot.max.min(data_max, relative.time = relative_time, xaxis = TRUE, yaxis = TRUE, pch = 19, main = "Maximum disparity in Beck2014", what = "disparity", legend.coords = "bottomright")
+par(op)
+```
+
+Multiple plots:
+
+```{r, fig.width = 12, fig.height = 4}
+## Names of the datasets
+slugs <- as.list(c("Beck2014", "Brusatte2014", "Bapst2016", "Wright2017"))
+
+## Load all the data
+data_list <- lapply(slugs, function(slug) read.csv(paste0("../outputs/extinction.", slug, ".csv"), stringsAsFactors = FALSE))
+
+## Set up the colours (yeah, they're ugly for now)
+colours <- c("lightblue", "yellow", "white") #Signif, non-signif, NA
+
+## Plot all the results
+multi.plot.extinction(data_list, type = c("age", "epoch"), data.names = unlist(slugs), cols = colours)
+```
 

### 2017-12-05 functions/figures_fun.R

@@ -1,26 +1,19 @@
-#' @title Plotting TreeCmp results
+#' @title Plot max min
 #'
-#' @description Simple wrapper for plotting the results from TreeCmp
+#' @description  Plotting the maximum or minimum age/disparity
 #'
-#' @param taxa_list a list of tree metrics from \code{read.TreeCmp} for all the scenario for a number of taxa
-#' @param metric a numeric of which metric to use (\code{1} = \code{MatchingCluster}, \code{2} = \code{R.F_Cluster}, \code{3} = \code{NodalSplitted}, \code{4} = \code{Triples})
-#' @param what which type of plot? Can be \code{"null"}, \code{"real"}, \code{"true"}, \code{"check"}. See details.
-#' @param ylim \code{ylim} parameter from \code{plot} (if missing is automatic)
-#' @param legend \code{logical}, whether to display the legend
-#' @param NTS \code{logical}, whether NTS was used
-#' @param ylab a \code{character} string to be passed to \code{boxplot(..., ylab)}
-#' @param xlab a \code{character} string to be passed to \code{boxplot(..., xlab)}
-#' @param axislab a \code{character} string to be passed to \code{axis(..., labels)}
-#' @param ... any additional argument to be passed to \code{boxplot}
-#'
-#' @details
-#' \code{"null"} = "maxi", "mini", "norm" tree vs. "rand"
-#' \code{"best"} = "maxi", "mini", "rand" tree vs. "norm"
-#' \code{"true"} = "norm", "rand" tree vs. "true"
-#' \code{"check"} = "norm", vs "rand"
-#' 
-#' @return
-#' A list of metrics for the comparisons.
+#' @param data the max or minimum data
+#' @param cols the different colours for the models
+#' @param lty the different line types for the CIs
+#' @param pch the point type for the median
+#' @param relative.time the relative time for scaling
+#' @param what whether to have the age on the Y axis or the disparity
+#' @param adj the size adjustment between the point
+#' @param type "age" or "epoch"
+#' @param legend whether to plot the legend
+#' @param legend.coords where to plot the legend (can be left empty)
+#' @param xaxis whether to put the labels on the x axis
+#' @param yaxis whether to put the labels on the y axis
 #' 
 #' @examples
 #' ##
[…]

### 2017-12-06 analyses/figures-compiler.Rmd
Multiple plots updated

@@
 ## Load all the data
-data_list <- lapply(slugs, function(slug) read.csv(paste0("../outputs/extinction.", slug, ".csv"), stringsAsFactors = FALSE))
+data_list_max <- lapply(slugs, function(slug) read.csv(paste0("../outputs/peaks.max.", slug, ".csv"), stringsAsFactors = FALSE))
+data_list_min <- lapply(slugs, function(slug) read.csv(paste0("../outputs/peaks.min.", slug, ".csv"), stringsAsFactors = FALSE))
 
-## Set up the colours (yeah, they're ugly for now)
-colours <- c("lightblue", "yellow", "white") #Signif, non-signif, NA
+## Load the relative time
+relative_times <- lapply(slugs, function(slug) read.nexus(file = paste0("../data/processed/tree.", slug, ".tre")))
+relative_times <- lapply(relative_times, function(tree) max(tree.age(tree)$age))
+relative_times[[4]] <- 484.4
 
-## Plot all the results
-multi.plot.extinction(data_list, type = c("age", "epoch"), data.names = unlist(slugs), cols = colours)
+## Set up the colours (to be changed)
+colours <- palette()[1:7]
+
+## Plot all maximum ages
+multi.plot.max.min(data_list_max, what = c("age"), cols = colours, lty = c(1,2), pch = 19, adj = 0.1, relative.time = relative_times, legend = TRUE, names.data = unlist(slugs))
+
+## Plot all maximum disparity
+multi.plot.max.min(data_list_max, what = c("disparity"), cols = colours, lty = c(1,2), pch = 19, adj = 0.1, relative.time = relative_times, names.data = unlist(slugs))
+
+## Plot all minimum ages
+multi.plot.max.min(data_list_max, what = c("age"), cols = colours, lty = c(1,2), pch = 19, adj = 0.1, relative.time = relative_times, names.data = unlist(slugs))
+
+## Plot all minimum disparity
+multi.plot.max.min(data_list_max, what = c("disparity"), cols = colours, lty = c(1,2), pch = 19, adj = 0.1, relative.time = relative_times, names.data = unlist(slugs))
+
 ```

### 2017-12-06 functions/figures_fun.R

@@ -14,6 +14,7 @@
 #' @param legend.coords where to plot the legend (can be left empty)
 #' @param xaxis whether to put the labels on the x axis
 #' @param yaxis whether to put the labels on the y axis
+#' @param relative.disparity whether to plot the disparity relatively
 #' 
 #' @examples
 #' ##
@@ -22,7 +23,7 @@
 #' @export
 #' 
 
-plot.max.min <- function(data, cols = c("black", "red", "green3", "blue", "cyan", "magenta", "yellow"), lty = c(1,2), pch = 21, what = "age", adj = 0.1, type = "age", legend = TRUE, relative.time, xaxis = FALSE, yaxis = FALSE, legend.coords, ...) {
+plot.max.min <- function(data, cols = c("black", "red", "green3", "blue", "cyan", "magenta", "yellow"), lty = c(1,2), pch = 19, what = "age", adj = 0.1, type = "age", legend = TRUE, relative.time, xaxis = FALSE, yaxis = FALSE, legend.coords, relative.disparity = FALSE, ...) {
 
     ## select the data type
     if(type == "age") {
@@ -48,15 +49,25 @@ plot.max.min <- function(data, cols = c("black", "red", "green3", "blue", "cyan"
 
         ## Adjusting ages
         if(!missing(relative.time)) {
-            data_values <- lapply(data_values, function(X, relative.time) return(X/relative.time), relative.time)
-            ylim <- c(0,1)
+            if(relative.time != FALSE) {
+                data_values <- lapply(data_values, function(X, relative.time) return(X/relative.time), relative.time)
+                ylim <- c(0,1)
+            } else {
+                ylim <- range(unlist(data_values))
+            }
         } else {
             ylim <- range(unlist(data_values))
         }
 
     } else {
         data_values <- data[,4:8]
-        ylim <- range(as.vector(data_values))
+        if(relative.disparity) {
+            ## Scale disparity
+            data_values <- data_values/max(data_values)
+            ylim <- range(as.vector(data_values))
+        } else {
+            ylim <- range(as.vector(data_values))
+        }
     }
 
     ## Get the sorting
@@ -161,8 +172,89 @@ plot.max.min <- function(data, cols = c("black", "red", "green3", "blue", "cyan"
 }
 
 
-multi.plot.max.min <- function() {
-    return(NULL)
[…]

### 2017-12-06 analyses/note-interesting_results.Rmd
interesting notes on Brusatte's dataset

---
title: "Interesting results"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Some interesting results

Here's a good illustration of using stratigraphy number or duration on Brusatte's dataset.

```{r}
library(dispRity)
## Loading the data
slug <- "Brusatte2014"
out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
```

The red lines represent respectively (from left to right) the K-Pg (66 Mya) and the PETM (55.5 Mya).

```{r, fig.width = 12, fig.height = 12}

op <- par(mfrow = c(3,3))
## Stratigraphy
plot(out1$objects$stratigraphy[[1]], type = "p", main = "Brusatte2014 - bins/stratigraphy", ylim = c(5.5, 8.5))
abline(v = c(7.5, 9.5), col = "red")
plot(out1$objects$stratigraphy[[5]], main = "Brusatte2014 - proximity/stratigraphy", ylim = c(5.5, 8.5))
abline(v = c(6.5, 9.5), col = "red")
plot(out1$objects$stratigraphy[[7]], main = "Brusatte2014 - gradual.split/stratigraphy", ylim = c(5.5, 8.5))
abline(v = c(6.5, 9.5), col = "red")

## Duration
plot(out1$objects$duration[[1]], type = "p", main = "Brusatte2014 - bins/duration", ylim = c(5.5, 8.5))
abline(v = c(10.5, 13.5), col = "red")
plot(out1$objects$duration[[5]], main = "Brusatte2014 - proximity/duration", ylim = c(5.5, 8.5))
abline(v = c(10.5, 13.5), col = "red")
plot(out1$objects$duration[[7]], main = "Brusatte2014 - gradual.split/duration", ylim = c(5.5, 8.5))
abline(v = c(10.5, 13.5), col = "red")

## Stratigraphy
plot(out1$objects$number[[1]], type = "p", main = "Brusatte2014 - bins/number", ylim = c(5.5, 8.5))
abline(v = c(10.5, 13.5), col = "red")
plot(out1$objects$number[[5]], main = "Brusatte2014 - proximity/number", ylim = c(5.5, 8.5))
abline(v = c(6.5, 9.5), col = "red")
plot(out1$objects$number[[7]], main = "Brusatte2014 - gradual.split/number", ylim = c(5.5, 8.5))
abline(v = c(6.5, 9.5), col = "red")

par(op)
```

### 2017-12-06 functions/simulations_fun.R

@@ -93,19 +93,19 @@ get.subsets <- function(morphospace, tree, bins, FADLAD, inc.nodes){
   	                                    model = "proximity")
 
   ## Subset samples by required number of time slices with proximity model
-  subsets.punctuated <- time.subsets(data = morphospace, tree = tree, 
+  subsets.equal.split <- time.subsets(data = morphospace, tree = tree, 
                                         method = "continuous", time = slices, 
                                         FADLAD = FADLAD, inc.nodes = inc.nodes,
-                                        model = "punctuated")
+                                        model = "equal.split")
 
   ## Subset samples by required number of time slices with proximity model
-  subsets.gradual <- time.subsets(data = morphospace, tree = tree, 
+  subsets.gradual.split <- time.subsets(data = morphospace, tree = tree, 
                                         method = "continuous", time = slices, 
                                         FADLAD = FADLAD, inc.nodes = inc.nodes,
-                                        model = "gradual")
+                                        model = "gradual.split")
 
   return(list(subsets.bins, subsets.acctran, subsets.deltran, subsets.random,
-              subsets.proximity, subsets.punctuated, subsets.gradual))
+              subsets.proximity, subsets.equal.split, subsets.gradual.split))
 }
 
 ##--------------------------------------------------------------------------
@@ -195,7 +195,7 @@ run.all.disparity <- function(morphospace, tree, bins, FADLAD, inc.nodes = TRUE,
   
   ## Add model designations
   models <- c("equalbins", "acctran", "deltran", "random", "proximity",
-              "punctuated", "gradual")
+              "equal.split", "gradual.split")
   disparity.outputs <- mapply(cbind, disparity.object, "model" = models,
                               SIMPLIFY = FALSE)
   ## Flatten list to a dataframe
@@ -366,7 +366,7 @@ plot.results.dispRity <- function(slug.object, method, colors, colors_CI, main,
     ## Legend
     if(legend) {
       if(length(slug.object[[1]]) == 7) {
-        legend(x = "bottomright", legend = c("acctran", "deltran", "random", "proximity", "punctuated", "gradual", "time bins"), col = colors[1:7], lty = 1)
+        legend(x = "bottomright", legend = c("acctran", "deltran", "random", "proximity", "equal.split", "gradual.split", "time bins"), col = colors[1:7], lty = 1)
       } else {
         legend(x = "bottomright", legend = c("acctran", "deltran", "random", "proximity", "time bins"), col = colors[1:5], lty = 1)
       }

### 2017-12-06 analyses/04-stats.Rmd
interesting notes on Brusatte's dataset
updated results

@@
 This script requires the outputs from `02-time-slice-analyses.Rmd`.
-The code below runs the analyses for the Beck and Lee 2014 dataset - chaneg the slug name to run the others.
+The code below runs the analyses for the Beck and Lee 2014 dataset - change the slug name to run the others.
 This of course can all be written into functions but we ran out of time so here is some suboptimal but working code!
@@
           row.names = FALSE)
-}
 ```

### 2017-12-06 analyses/04-stats.Rmd
corrected peak ages

@@
 library(dispRity)
+source("../functions/stats_fun.R")
 ```
@@
 ```{r, warning = FALSE}
+
+## Getting the max/min for each age/epoch
+age_results_max <- do.call(rbind, lapply(out1$objects, select.max.min.method, fun = max))
+age_results_min <- do.call(rbind, lapply(out1$objects, select.max.min.method, fun = min))
+epoch_results_max <- do.call(rbind, lapply(out1$objects, select.max.min.method, fun = max))
+epoch_results_min <- do.call(rbind, lapply(out1$objects, select.max.min.method, fun = min))
+
+## Combining into the nice table formats
+age_results_max <- cbind(age_results_max, "stratigraphy" = rep("Age", nrow(age_results_max)),
+                         "bin_type" = unlist(strsplit(rownames(age_results_max), split = "\\..*")))
+age_results_min <- cbind(age_results_min, "stratigraphy" = rep("Age", nrow(age_results_min)),
+                         "bin_type" = unlist(strsplit(rownames(age_results_min), split = "\\..*")))
+epoch_results_max <- cbind(epoch_results_max, "stratigraphy" = rep("Epoch", nrow(epoch_results_max)),
+                        "bin_type" = unlist(strsplit(rownames(epoch_results_max), split = "\\..*")))
+epoch_results_min <- cbind(epoch_results_min, "stratigraphy" = rep("Epoch", nrow(epoch_results_min)),
+                        "bin_type" = unlist(strsplit(rownames(epoch_results_min), split = "\\..*")))
+
+## Combining the age and epoch results together
+all_peaks_max <- rbind(age_results_max, epoch_results_max)
+all_peaks_min <- rbind(age_results_min, epoch_results_min)
+
+write.csv(all_peaks_max, file = paste0("../outputs/peaks.max.", slug, ".csv"), quote = FALSE, 
+          row.names = FALSE)
+
+write.csv(all_peaks_min, file = paste0("../outputs/peaks.min.", slug, ".csv"), quote = FALSE, 
+          row.names = FALSE)
+```
+
+<!-- former version -->
+
+```{r, eval = FALSE, echo = FALSE}
 # ages

### 2017-12-06 functions/stats_fun.R

@@ -1,0 +1,38 @@
+#' @title Select max min methods
+#'
+#' @description Selects the maximum/minimum values for each model for each method
+#'
+#' @param method the objects for a bin type method (e.g. for stratigraphy, data$objects$stratigraphy)
+#' @param fun the selection function (e.g. max() or min())
+#' @param method.name optional, the method name
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+select.max.min.method <- function(method, fun = max, method.name) {
+
+  ## Selecting the max/min for one disparity object
+  select.max.min <- function(disparity, fun = max) {
+    ## Get the model
+    model <- disparity$call$subsets[length(disparity$call$subsets)]
+    model <- ifelse(model == "discrete", "equalbins", model)
+
+    ## Summarise the data
+    summary_data <- summary(disparity, round = 10)
+    return(cbind(summary_data[which(summary_data[,4] == fun(summary_data[,4], na.rm = TRUE)),], "model" = model))
+  }  
+
+  ## Apply the select.max.min function to the whole method
+  max.min_table <- do.call(rbind, lapply(method, select.max.min, fun = max))
+
+  ## Add the method name (optional)
+  if(!missing(method.name)) {
+      max.min_table <- cbind(max.min_table, "bin_type" = rep(method.name, nrow(max.min_table)))
+  }
+
+  return(max.min_table)
+}

### 2017-12-06 analyses/figures-compiler.Rmd
figures done

@@
 ## Plot all minimum ages
-multi.plot.max.min(data_list_max, what = c("age"), cols = colours, lty = c(1,2), pch = 19, adj = 0.1, relative.time = relative_times, names.data = unlist(slugs))
+multi.plot.max.min(data_list_min, what = c("age"), cols = colours, lty = c(1,2), pch = 19, adj = 0.1, relative.time = relative_times, names.data = unlist(slugs))
 
 ## Plot all minimum disparity
-multi.plot.max.min(data_list_max, what = c("disparity"), cols = colours, lty = c(1,2), pch = 19, adj = 0.1, relative.time = relative_times, names.data = unlist(slugs))
+multi.plot.max.min(data_list_min, what = c("disparity"), cols = colours, lty = c(1,2), pch = 19, adj = 0.1, relative.time = relative_times, names.data = unlist(slugs))
 

### 2017-12-06 functions/stats_fun.R

@@ -23,7 +23,7 @@ select.max.min.method <- function(method, fun = max, method.name) {
 
     ## Summarise the data
     summary_data <- summary(disparity, round = 10)
-    return(cbind(summary_data[which(summary_data[,4] == fun(summary_data[,4], na.rm = TRUE)),], "model" = model))
+    return(cbind(summary_data[which(summary_data[,4] == fun(summary_data[,4], na.rm = TRUE))[1],], "model" = model))
   }  
 
   ## Apply the select.max.min function to the whole method

### 2017-12-06 analyses/04-stats.Rmd
figures done
Tidy and loop

@@
 
-This code runs the output analyses in the paper, i.e. creates tables, figures and statistical tests.
+This code runs the output analyses in the paper, i.e. creates tables and performs statistical tests.
 
@@
 
-Loading the data
------------------------------------
-
-This script requires the outputs from `02-time-slice-analyses.Rmd`.
-The code below runs the analyses for the Beck and Lee 2014 dataset - change the slug name to run the others.
-This of course can all be written into functions but we ran out of time so here is some suboptimal but working code!
-
-```{r}
-slug <- "Beck2014"
-#slug <- "Brusatte2014"
-#slug <- "Bapst2016"
-#slug <- "bears"
-#slug <- "Wright2017"
-```
-
-```{r}
-out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
-#out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
-```
-
-Outputs can be viewed using `out1$results$stratigraphy`, `out1$results$duration`, or `out1$results$number`.
+This script requires the outputs from `02-time-slice-analyses.Rmd`. This of course can all be written into functions but we ran out of time so here is some suboptimal but working code!
 
@@
 
-i) statigraphy
+```{r}
+slugs <- c("Beck2014", "Brusatte2014", "Bapst2016", "Wright2017")
+```
 
 ```{r}
+# Loop through all four datasets
+for(i in 1:4) {
+  slug <- slugs[i]
+
+out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
+#out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
+
+# i) statigraphy
 strat <- out1$results$stratigraphy
@@
 w_del <- wilcox.test(strat2$equalbins, strat2$deltran, paired = TRUE)
-w_grad <- wilcox.test(strat2$equalbins, strat2$gradual, paired = TRUE)
+w_grad <- wilcox.test(strat2$equalbins, strat2$gradual.split, paired = TRUE)
 w_prox <- wilcox.test(strat2$equalbins, strat2$proximity, paired = TRUE)
-w_punc <- wilcox.test(strat2$equalbins, strat2$punctuated, paired = TRUE)
+w_punc <- wilcox.test(strat2$equalbins, strat2$equal.split, paired = TRUE)
 w_rand <- wilcox.test(strat2$equalbins, strat2$random, paired = TRUE)
@@
                 tidy(w_punc), tidy(w_rand))
-outputS$model <- c("acctran", "deltran", "gradual", "proximity", 
-                        "punctuated", "random")
+outputS$model <- c("acctran", "deltran", "gradual.split", "proximity", 
+                        "equal.split", "random")
 outputS$type <- rep("stratigraphy", 6)
 outputS$period <- rep(strat$stratigraphy[1], 6)
-```
 
-ii) duration
+## ii) duration
 
-```{r}
 duration <- out1$results$duration
@@
 w_del <- wilcox.test(duration2$equalbins, duration2$deltran, paired = TRUE)
-w_grad <- wilcox.test(duration2$equalbins, duration2$gradual, paired = TRUE)
+w_grad <- wilcox.test(duration2$equalbins, duration2$gradual.split, paired = TRUE)
 w_prox <- wilcox.test(duration2$equalbins, duration2$proximity, paired = TRUE)
-w_punc <- wilcox.test(duration2$equalbins, duration2$punctuated, paired = TRUE)
+w_punc <- wilcox.test(duration2$equalbins, duration2$equal.split, paired = TRUE)
 w_rand <- wilcox.test(duration2$equalbins, duration2$random, paired = TRUE)
@@
                 tidy(w_punc), tidy(w_rand))
-outputD$model <- c("acctran", "deltran", "gradual", "proximity", 
-                        "punctuated", "random")
+outputD$model <- c("acctran", "deltran", "gradual.split", "proximity", 
+                        "equal.split", "random")
 outputD$type <- rep("duration", 6)
 outputD$period <- rep(strat$stratigraphy[1], 6)
-```
 
-iii) number
+## iii) number
 
-```{r}
 number <- out1$results$number
@@
 w_del <- wilcox.test(number2$equalbins, number2$deltran, paired = TRUE)
-w_grad <- wilcox.test(number2$equalbins, number2$gradual, paired = TRUE)
+w_grad <- wilcox.test(number2$equalbins, number2$gradual.split, paired = TRUE)
 w_prox <- wilcox.test(number2$equalbins, number2$proximity, paired = TRUE)
-w_punc <- wilcox.test(number2$equalbins, number2$punctuated, paired = TRUE)
+w_punc <- wilcox.test(number2$equalbins, number2$equal.split, paired = TRUE)
 w_rand <- wilcox.test(number2$equalbins, number2$random, paired = TRUE)
@@
                 tidy(w_punc), tidy(w_rand))
-outputN$model <- c("acctran", "deltran", "gradual", "proximity", 
-                        "punctuated", "random")
+outputN$model <- c("acctran", "deltran", "gradual.split", "proximity", 
+                        "equal.split", "random")
 outputN$type <- rep("number", 6)
 outputN$period <- rep(strat$stratigraphy[1], 6)
-```
 
-And finally save all the outputs...
+## And finally save all the outputs...
 
-```{r}
 all_wilcox <- rbind(outputS, outputD, outputN)
@@
           row.names = FALSE, col.names = FALSE, append = TRUE, sep = ",")
+}
 ```
@@
 
-You can rerun for the other datasets by changing the slug name at the start.
-
 Disparity peaks
@@
 
-Set the slug
-
 ```{r}
-slug <- "Beck2014"
-#slug <- "Brusatte2014"
-#slug <- "Bapst2016"
-#slug <- "bears"
-#slug <- "Wright2017"
+slugs <- c("Beck2014", "Brusatte2014", "Bapst2016", "Wright2017")
 ```
 
-Read in the data
+```{r, warning = FALSE}
+# Loop through all four datasets
+for(i in 1:4) {
+  slug <- slugs[i]
 
-```{r}
+## Read in the data
 out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
 out2 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
-```
 
-Extract the maximum bootstrapped median disparity values for each model, with details of the peak bin/slice times.
-
-```{r, warning = FALSE}
+## Extract the maximum bootstrapped median disparity values for each model, with details of the ## peak bin/slice times.
 
@@
           row.names = FALSE)
+}
 ```
 
-<!-- former version -->
-
-```{r, eval = FALSE, echo = FALSE}
-# ages
-strat <- out1$results$stratigraphy
-duration <- out1$results$duration
-number <- out1$results$number
-
-strat_max <- 
-  strat %>%
-  group_by(model) %>%
-  filter(bs.median == max(bs.median, na.rm = TRUE))
-
-duration_max <-
-  duration %>%
-  group_by(model) %>%
-  filter(bs.median == max(bs.median, na.rm = TRUE))
-
-number_max <- 
-  number %>%
-  group_by(model) %>%
-  filter(bs.median == max(bs.median, na.rm = TRUE))
-
-strat_min <- 
-  strat %>%
-  group_by(model) %>%
-  filter(bs.median == min(bs.median, na.rm = TRUE))
-
-duration_min <-
-  duration %>%
-  group_by(model) %>%
-  filter(bs.median == min(bs.median, na.rm = TRUE))
-
-number_min <- 
-  number %>%
-  group_by(model) %>%
-  filter(bs.median == min(bs.median, na.rm = TRUE))
-
-# epochs
-strat2 <- out2$results$stratigraphy
-duration2 <- out2$results$duration
-number2 <- out2$results$number
-
-strat2_max <- 
-  strat2 %>%
-  group_by(model) %>%
-  filter(bs.median == max(bs.median, na.rm = TRUE))
-
-duration2_max <-
-  duration2 %>%
-  group_by(model) %>%
-  filter(bs.median == max(bs.median, na.rm = TRUE))
-
-number2_max <- 
-  number2 %>%
-  group_by(model) %>%
-  filter(bs.median == max(bs.median, na.rm = TRUE))
-
-strat2_min <- 
-  strat2 %>%
-  group_by(model) %>%
-  filter(bs.median == min(bs.median, na.rm = TRUE))
-
-duration2_min <-
-  duration2 %>%
-  group_by(model) %>%
-  filter(bs.median == min(bs.median, na.rm = TRUE))
-
-number2_min <- 
-  number2 %>%
-  group_by(model) %>%
-  filter(bs.median == min(bs.median, na.rm = TRUE))
-
-all_peaks_max <- rbind(strat_max, duration_max, number_max, strat2_max, duration2_max, number2_max)
-write.csv(all_peaks_max, file = paste0("../outputs/peaks.max.", slug, ".csv"), quote = FALSE, 
-          row.names = FALSE)
-
-all_peaks_min <- rbind(strat_min, duration_min, number_min, strat2_min, duration2_min, number2_min)
-write.csv(all_peaks_min, file = paste0("../outputs/peaks.min.", slug, ".csv"), quote = FALSE, 
-          row.names = FALSE)
-```
-You can rerun for the other datasets by changing the slug name at the start.
-
-Tables
-----------
-
-Finally, for the purposes of completeness here are the full results as tables.
-
-Set the slug
-
-```{r}
-slug <- "Beck2014"
-#slug <- "Brusatte2014"
-#slug <- "Bapst2016"
-#slug <- "bears"
-#slug <- "Wright2017"
-```
-Read in the data
-
-```{r}
-out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
-out2 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
-```
-Make the tables
-
-```{r, eval = TRUE, echo = FALSE}
-# Ages
-knitr::kable(out1$results$stratigraphy, caption = "Disparity per stratigraphic age: unequal time bins and non-equidistant time slices")
-knitr::kable(out1$results$duration, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the median stratigraphic age duration")
-knitr::kable(out1$results$number, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the number of stratigraphic ages")
-
-# Epochs
-knitr::kable(out2$results$stratigraphy, caption = "Disparity per stratigraphic epoch: unequal time bins and non-equidistant time slices")
-knitr::kable(out2$results$duration, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the median stratigraphic epoch duration")
-knitr::kable(out2$results$number, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the number of stratigraphic epochs")
-```
-
-You can rerun for the other datasets by changing the slug name at the start.
-
 Extinction effect
-----------
+------------------
 
@@
   ## Adding the models
-  results <- cbind(c("bins", "acctran", "deltran", "random", "proximity", "punctuated", "gradual"),results)
+  results <- cbind(c("bins", "acctran", "deltran", "random", "proximity", "equal.split", "gradual.split"),results)
   ## Adding the type and method
@@
 ```
+
+Tables
+----------
+
+Finally, for the purposes of completeness here are the full results as tables.
+
+Set the slug
+
+```{r}
+slug <- "Beck2014"
+#slug <- "Brusatte2014"
+#slug <- "Bapst2016"
+#slug <- "bears"
+#slug <- "Wright2017"
+```
+
+Read in the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
+out2 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
+```
+
+Make the tables
+
+```{r, eval = TRUE, echo = FALSE}
+# Ages
+knitr::kable(out1$results$stratigraphy, caption = "Disparity per stratigraphic age: unequal time bins and non-equidistant time slices")
+knitr::kable(out1$results$duration, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the median stratigraphic age duration")
+knitr::kable(out1$results$number, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the number of stratigraphic ages")
+
+# Epochs
+knitr::kable(out2$results$stratigraphy, caption = "Disparity per stratigraphic epoch: unequal time bins and non-equidistant time slices")
+knitr::kable(out2$results$duration, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the median stratigraphic epoch duration")
+knitr::kable(out2$results$number, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the number of stratigraphic epochs")
+```
+
+You can rerun for the other datasets by changing the slug name at the start.

### 2017-12-06 analyses/table-compiler.Rmd
Fixed details of table 1

@@
 ## Getting the data
-slugs <- list("Beck2014", "Brusatte2014", "Bapst2016", "bears", "Wright2017")
+slugs <- list("Beck2014", "Brusatte2014", "Bapst2016", "Wright2017")
 
@@
 ## Adding Wright's tree age
-trees[[5]]$root.time <- 485.4
+trees[[4]]$root.time <- 485.4
 ```
@@
 ## Column names
-colnames <- c("Group", "n taxa", "m characters", "Age range", "Reference", "Notes")
+colnames <- c("Group", "# taxa", "# characters", "Age range (MYA)", "Mass extinction (MYA)", "Reference(s)")
 
@@
 ## Groups
-group <- c("Eutherian mammals", "theropods", "theropods", "bears", "crinoids")
+group <- c("eutherian mammals", "theropods", "theropods", "crinoids")
 
@@
 
-## References and Notes (to be filled in LaTeX)
-reference <- notes <- rep("", length(slugs))
-notes[2] <- "2 + 9 taxa removed"
-notes[3] <- "12 + 19 taxa removed. Tree was randomly selected from all time-scaled trees"
-notes[4] <- "12 + 0 taxa removed"
-notes[5] <-  "Maximum clade credibility tree, time-scaled by dividing branches by 0.03517385 and setting root time to 485.4"
+## Mass extinction
+mass <- c("66", "66", "66", "443")
+
+## References (to be filled in LaTeX)
+reference <- rep("", length(slugs))
 
 ## Formatting the table
-table_out <- data.frame(group, ntaxa, mcharacters, age_ranges, reference, notes)
+table_out <- data.frame(group, ntaxa, mcharacters, age_ranges, mass, reference)
 colnames(table_out) <- colnames

### 2017-12-06 analyses/05-table-compiler.Rmd
Add new table compiler

--- 2017-12-06 analyses/table-compiler.Rmd
@@
     fig_width: 12
-    fig_height: 6
+    fig_height: 12
 ---
@@
 
-## Data set table
+## Data set table (Table 1)
 
@@
 ## Column names
-colnames <- c("Group", "# taxa", "# characters", "Age range (MYA)", "Mass extinction (MYA)", "Reference(s)")
+rownames <- c("Group", "# taxa", "# characters", "Age range (MYA)", "Mass extinction (MYA)", "Reference(s)")
 
@@
 dataset <- unlist(slugs)
-dataset[4] <- "Heath2016"
 
 ## Groups
-group <- c("eutherian mammals", "theropods", "theropods", "crinoids")
+group <- c("mammals", "theropods", "theropods", "crinoids")
 
@@
 ## Mass extinction
-mass <- c("66", "66", "66", "443")
+mass <- c("66 (K-Pg)", "66 (K-Pg)", "66 (K-Pg)", "443 (O-S)")
 
@@
 table_out <- data.frame(group, ntaxa, mcharacters, age_ranges, mass, reference)
-colnames(table_out) <- colnames
-rownames(table_out) <- dataset
+table_out <- t(table_out)
+colnames(table_out) <- dataset
+rownames(table_out) <- rownames
 

### 2017-12-06 analyses/03-statistical-analyses.Rmd
Remove bears

--- 2017-12-06 analyses/04-stats.Rmd
@@
 
-out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
-#out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
+#out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
+out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
 
@@
 # slug <- "Bapst2016"; extinction <- 66.0
-# slug <- "bears"; extinction <- NA
 # slug <- "Wright2017"; extinction <- 443.0
@@
 #slug <- "Bapst2016"
-#slug <- "bears"
 #slug <- "Wright2017"

### 2017-12-06 manuscript/tables/table_datasets.tex
Add latex formatted table

% latex table generated in R 3.4.2 by xtable 1.8-2 package
% Wed Dec  6 16:09:02 2017
\begin{table}[ht]
\centering
\begin{tabular}{lcccc}
  \hline
 & \textbf{Beck2014} & \textbf{Brusatte2014} & \textbf{Bapst2016} & \textbf{Wright2017} \\ 
  \hline
Group & mammals & theropods & theropods & crinoids \\ 
  \# taxa & 106 & 152 &  89 &  42 \\ 
  \# characters & 421 & 853 & 374 &  87 \\ 
  Age range (MYA) & 171.786 - 0 & 102.535 - 0 & 141.216 - 0 & 485.4 - 372.2 \\ 
  Mass extinction (MYA) & 66 (K-Pg) & 66 (K-Pg) & 66 (K-Pg) & 443 (O-S) \\ 
  Reference & \cite{beckancient2014} & \cite{brusatte2014gradual} & \cite{bapst2016topology} & \cite{wright2017bayesian} \\ 
  Data reference &  CHECK & \cite{dryad_84t75} & \cite{dryad_n2g80} &  \cite{dryad_6hb7j} \\ 
   \hline
\end{tabular}
\caption{Details of the datasets used in this study.} 
\end{table}

### 2017-12-06 manuscript/tables/raw_datasets.tex
Raw table output from table compiler

--- 2017-12-06 manuscript/tables/table_datasets.tex
@@
 \centering
-\begin{tabular}{lcccc}
+\begin{tabular}{rllll}
   \hline
- & \textbf{Beck2014} & \textbf{Brusatte2014} & \textbf{Bapst2016} & \textbf{Wright2017} \\ 
+ & Beck2014 & Brusatte2014 & Bapst2016 & Wright2017 \\ 
   \hline
@@
   Mass extinction (MYA) & 66 (K-Pg) & 66 (K-Pg) & 66 (K-Pg) & 443 (O-S) \\ 
-  Reference & \cite{beckancient2014} & \cite{brusatte2014gradual} & \cite{bapst2016topology} & \cite{wright2017bayesian} \\ 
-  Data reference &  CHECK & \cite{dryad_84t75} & \cite{dryad_n2g80} &  \cite{dryad_6hb7j} \\ 
+  Reference &  &  &  &  \\ 
+  Data reference &  &  &  &  \\ 
    \hline

### 2017-12-06 analyses/05-table-compiler.Rmd
Add data reference line

@@
 ```{r}
-## Column names
-rownames <- c("Group", "# taxa", "# characters", "Age range (MYA)", "Mass extinction (MYA)", "Reference(s)")
+## variable names
+rownames <- c("Group", "# taxa", "# characters", "Age range (MYA)", "Mass extinction (MYA)", "Reference", "Data reference")
 
@@
 reference <- rep("", length(slugs))
+dataref <- rep("", length(slugs))
 
 ## Formatting the table
-table_out <- data.frame(group, ntaxa, mcharacters, age_ranges, mass, reference)
+table_out <- data.frame(group, ntaxa, mcharacters, age_ranges, mass, reference, dataref)
 table_out <- t(table_out)
@@
 
-sink(file = "../manuscript/tables/datasets.tex")
+sink(file = "../manuscript/tables/raw_datasets.tex")
 xtable(table_out, caption = table_caption)

### 2017-12-06 analyses/04-figures.Rmd
disparity through time figures

---
title: "Figure 1: disparity through time - simplified and for appendix"
author: "Natalie Cooper"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

```{r, message = FALSE}
library(tidyverse)
```

Pick the slug

```{r}
slug <- "Brusatte2014"
```

Load the data

```{r}
out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))

strat <- out1$results$stratigraphy
duration <- out1$results$duration
number <- out1$results$number
```

Get the data into the correct format to plot

```{r}
# Stratigraphy
# Get ages for the bins, go for the midpoints
x <- str_split(strat$subset, " - ") 
y <- map(x, as.numeric)
z <- map(y, mean)

# Modify data - add midpoints, remove missing data, filter to just two models
strat2 <- 
  strat %>%
  mutate(time = unlist(z)) %>%
  filter(bs.median != 0 & !is.na(bs.median)) %>%
  filter(model == "equalbins" | model == "proximity")

# Duration
x <- str_split(duration$subset, " - ") 
y <- map(x, as.numeric)
z <- map(y, mean)

duration2 <- 
  duration %>%
  mutate(time = unlist(z)) %>%
  filter(bs.median != 0 & !is.na(bs.median)) %>%
  filter(model == "equalbins" | model == "proximity")

# Number
x <- str_split(number$subset, " - ") 
y <- map(x, as.numeric)
z <- map(y, mean)

number2 <- 
  number %>%
  mutate(time = unlist(z)) %>%
  filter(bs.median != 0 & !is.na(bs.median)) %>%
  filter(model == "equalbins" | model == "proximity")
```

Make plots in preparation for larger grid plot below

```{r}

brusatte_base <- 
  ggplot(strat2, aes(y = bs.median, x = time, colour = model)) +
  theme_classic(base_size = 16) + 
  xlab("Time (MYA)") + 
  ylab("Disparity") + 
  scale_y_continuous(breaks = seq(0, 10, 2), 
                     labels = seq(0, 10, 2)) +
  scale_x_reverse(limits = c(100, 0)) +
  expand_limits(x = 0, y = 3) +
  theme(legend.position = "none")

brusatte_s <- 
  brusatte_base +
  geom_point(data = strat2, size = 2, alpha = 0.8) +
  geom_errorbar(data = strat2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)

brusatte_d <- 
  brusatte_base +
  geom_point(data = duration2, size = 2, alpha = 0.8) +
  geom_errorbar(data = duration2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)

brusatte_n <- 
  brusatte_base +
  geom_point(data = number2, size = 2, alpha = 0.8) +
  geom_errorbar(data = number2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)

```

### 2017-12-06 analyses/04-figures.Rmd
disparity through time figures

@@
 library(tidyverse)
+library(gridExtra)
+```
+```{r}
+slug <- "Brusatte2014"
+```
+
+Load the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
+
+strat <- out1$results$stratigraphy
+duration <- out1$results$duration
+number <- out1$results$number
+```
+
+Get the data into the correct format to plot
+
+```{r}
+# Stratigraphy
+# Get ages for the bins, go for the midpoints
+x <- str_split(strat$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+# Modify data - add midpoints, remove missing data, filter to just two models
+strat2 <- 
+  strat %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity")
+
+# Duration
+x <- str_split(duration$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+duration2 <- 
+  duration %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity")
+
+# Number
+x <- str_split(number$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+number2 <- 
+  number %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity")
+```
+
+Make plots in preparation for larger grid plot below
+
+```{r}
+
+brusatte_base <- 
+  ggplot(strat2, aes(y = bs.median, x = time, colour = model)) +
+  theme_classic(base_size = 16) + 
+  xlab("Time (MYA)") + 
+  ylab("Disparity") + 
+  scale_y_continuous(breaks = seq(0, 10, 2), 
+                     labels = seq(0, 10, 2)) +
+  scale_x_reverse(limits = c(100, 0)) +
+  expand_limits(x = 0, y = 3) +
+  theme(legend.position = "none")
+
+brusatte_s <- 
+  brusatte_base +
+  geom_point(data = strat2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)
+
+brusatte_d <- 
+  brusatte_base +
+  geom_point(data = duration2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)
+
+brusatte_n <- 
+  brusatte_base +
+  geom_point(data = number2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)
+
 ```
@@
 ```
+
+```{r}
+slug <- "Bapst2016"
+```
+
+Load the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
+
+strat <- out1$results$stratigraphy
+duration <- out1$results$duration
+number <- out1$results$number
+```
+
+Get the data into the correct format to plot
+
+```{r}
+# Stratigraphy
+# Get ages for the bins, go for the midpoints
+x <- str_split(strat$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+# Modify data - add midpoints, remove missing data, filter to just two models
+strat2 <- 
+  strat %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity")
+
+# Duration
+x <- str_split(duration$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+duration2 <- 
+  duration %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity")
+
+# Number
+x <- str_split(number$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+number2 <- 
+  number %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity")
+```
+
+Make plots in preparation for larger grid plot below
+
+```{r}
+
+bapst_base <- 
+  ggplot(strat2, aes(y = bs.median, x = time, colour = model)) +
+  theme_classic(base_size = 16) + 
+  xlab("Time (MYA)") + 
+  ylab("Disparity") + 
+  scale_y_continuous(breaks = seq(0, 10, 2), 
+                     labels = seq(0, 10, 2)) +
+  scale_x_reverse(limits = c(100, 0)) +
+  expand_limits(x = 0, y = 4) +
+  theme(legend.position = "none")
+
+bapst_s <- 
+  bapst_base +
+  geom_point(data = strat2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)
+
+bapst_d <- 
+  bapst_base +
+  geom_point(data = duration2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)
+
+bapst_n <- 
+  bapst_base +
+  geom_point(data = number2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)
+
+```
+
+```{r}
+slug <- "Wright2017"
+```
+
+Load the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
+
+strat <- out1$results$stratigraphy
+duration <- out1$results$duration
+number <- out1$results$number
+```
+
+Get the data into the correct format to plot
+
+```{r}
+# Stratigraphy
+# Get ages for the bins, go for the midpoints
+x <- str_split(strat$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+# Modify data - add midpoints, remove missing data, filter to just two models
+strat2 <- 
+  strat %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity")
+
+# Duration
+x <- str_split(duration$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+duration2 <- 
+  duration %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity")
+
+# Number
+x <- str_split(number$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+number2 <- 
+  number %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity")
+```
+
+Make plots in preparation for larger grid plot below
+
+```{r}
+
+wright_base <- 
+  ggplot(strat2, aes(y = bs.median, x = time, colour = model)) +
+  theme_classic(base_size = 16) + 
+  xlab("Time (MYA)") + 
+  ylab("Disparity") + 
+  scale_y_continuous(breaks = seq(0, 3, 1), 
+                     labels = seq(0, 3, 1)) +
+  scale_x_reverse(limits = c(470, 370),
+                  breaks = c(seq(470, 370, by = -20))) +
+  expand_limits(y = c(1,3)) +
+  theme(legend.position = "none")
+
+wright_s <- 
+  wright_base +
+  geom_point(data = strat2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)
+
+wright_d <- 
+  wright_base +
+  geom_point(data = duration2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)
+
+wright_n <- 
+  wright_base +
+  geom_point(data = number2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)
+
+```
+
+Finally plot them all together
+
+```{r}
+
+all <- grid.arrange(brusatte_s, brusatte_d, brusatte_n,
+                    bapst_s, bapst_d, bapst_n,
+                    wright_s, wright_d, wright_n, ncol = 3)
+
+ggsave("", all, 
+       device = png(width = 600, height = 400))
+```

### 2017-12-07 manuscript/time-slice.tex
Change name of table
First draft intro

@@
 
-\runninghead{Time slicing}
-\keywords{??} 
+\runninghead{Time subsamples in disparity-through-time analyses}
+\keywords{time bin, time slice, disparity} 
 % up to 6
@@
 
-Disparity-through-time analyes are common in palaeontology.
-Examples
-They often provide us with exciting insights into mass extinctions, competitive replacements etc.
-However, our methods for analysing diversity-through-time have not changed all that much in the XX years they have been popular.
+Disparity-through-time analyes are common in palaeontology \citep{Foote01071994}.
+They reveal how the morphological diversity of clades has changed through time, and allow us to make inferences about the breadth of ecological niches extinct taxa occupied \citep{foote1997evolution}.
+Results from disparity-through-time studies also provide insights into the ecological impacts of mass extinctions, competitive replacements, and the evolution of morphological innovations \citep{Brusatte12092008,Foote29111996,friedmanexplosive2010}.
+Unfortunately, the way we perform these analyses may have profound effects on our conclusions.
+
+% However, our methods for analysing diversity-through-time have not changed all that much in the XX years they have been popular.
 %TG: small caveat on that after litterature search for the disparity review paper: disparity-through-time analysis have not changed much since Foote 1992, Foote and Gould 1992 and Briggs et al.
-1992 in the classic palaeobiology approach...
+1992 in the classic palaeobiology appro,ach...
 %TG: However to not piss of people, there were several attempts to "rationalise" (Wills 2001), test (Ciampaglio 2001) and expand (Pavlinov 2011 and Gerber et al 2011) namely with univariate approaches and "dtt" analysis (Harmon et al 2003).
 %TG: One of the critisisms from one of the rude reviews was this idea that age of popular method has nothing to do with its performance, by nuancing this sentence we can basically say: yeah, some people have made great advances and proposed great ideas for CORRECTING ERRORS in the methods from the 90s but somehow everybody have ignored them (in palaeobiology).
-The way we perfrom these analyses may have profound effects on our conclusions.
+% NC: OK so lets save ourselves some pain and not mention this at all and save it all for the bitching in the review paper!!!
 
-One aspect that is not always carefully considered, is how to divide taxa into subsets through time.
-The nature of disparity, i.e.
-a diversity metric, means it cannot be calculated using just one individual, so some way of subsetting taxa is required.
+Disparity-through-time analyes have two main analysis components: calculating disparity, and creating time subsamples of the data. 
+Here we focus on the latter.
+The nature of disparity (it is a diversity metric), means it cannot be calculated using just one individual, so some way of subsetting taxa is required.
+Changes in disparity-through-time are generally investigated by calculating the disparity of taxa present during specific time intervals or time bins \citep[e.g][]{cisneros2010,prentice2011,Hughes20082013,hopkinsdecoupling2013,bentonmodels2014,bensonfaunal2014}.
+These time bins are usually defined based on stratigraphy \citep[e.g.][]{cisneros2010,prentice2011,Hughes20082013,bentonmodels2014} but can also be arbitrarily chosen time bins of equal duration \citep{Butler2012,hopkinsdecoupling2013,bensonfaunal2014}.
+However, this approach has several limitations.
 
-Stratigraphy
-Time bins
-Time slicing (requires a phylogeny)
+First, time bins defined by stratigraphy are not of equal size, biasing higher disparity towards longer stratigraphic periods. 
+This can be dealt with using rarefaction, i.e. repeating the analysis using the smallest number of taxa present in any bin as the number of sampled taxa in each bin. 
+This can, however, lead to very large confidence intervals when there are stratigraphic periods with very few species.
+Other studies split large time bins so they are of roughly equal size, but this is quite an \textit{ad hoc} procedure that can introduce more bias depending on where bins are split.
+Second, all time binning approaches (whether bins are equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring \textit{between} intervals, rather than also allowing for gradual changes \textit{within} intervals \citep{Hunt21042015}.
+Finally, time bin analyses are often limited by the number of taxa in each bin.
+If there are insufficient taxa in a time bin, disparity cannot be calculated, so further analysis, e.g.
+correlations of disparity with climate proxies or other hypothesised drivers of morphological evolution, are not possible.
 
-Changes in disparity through time are generally investigated by calculating the disparity of taxa that occupy the cladisto-space during specific time intervals \citep[e.g][]{cisneros2010,prentice2011,Hughes20082013,hopkinsdecoupling2013,bentonmodels2014,bensonfaunal2014}.
-These time intervals are usually defined based on biostratigraphy \citep[e.g.][]{cisneros2010,prentice2011,Hughes20082013,bentonmodels2014} but can also be arbitrarily chosen time periods of equal duration \citep{Butler2012,hopkinsdecoupling2013,bensonfaunal2014}.
-However, this approach suffers from two main biases. 
-First, if biostratigraphy is used to determine the time intervals, disparity may be distorted towards higher differences between time intervals because biostratigraphical periods are geologically defined based on differences in the morphology of fossils found in the different strata.
-Second, this approach assumes that all characters evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring between intervals, rather than also allowing for gradual changes within intervals \citep{Hunt21042015}.
+To address these issues, we propose a ``time-slicing'' approach that takes advantage of the wealth of paleontological datasets which now have associated phylogenies. 
+Time slicing uses a phylogenetic tree and considers subsets of taxa at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
+This results in even-sampling across time and permits us to define the underlying model of character evolution (punctuated or gradual).  
+Time-slicing also includes any element present in the phylogeny (branches, nodes and tips) at the time slice in question as part of the disparity calculation.
+This allows us to measure disparity at time points where there are no terminal taxa, and increases the sample size at each time point, making downstream analyses of the drivers of disparity much more feasible.
 
-To address these issues, we used a ``time-slicing'' approach that considers subsets of taxa in the cladisto-space at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
-This results in even-sampling of the cladisto-space across time and permits us to define the underlying model of character evolution (punctuated or gradual).  
-In practice, time-slicing considers the disparity of any element present in the phylogeny (branches, nodes and tips) at any point in time.
-
-We present this using four datasets collated from the literature.
+Here we present the time slicing methods using four datasets taken from the literature.
+We calculate disparity-through-time for each dataset using a range of time binning and time slicing methods, and then compare these approaches with respect to the relative disparities calculated, but also with how the different approaches influence biological conclusions.
+We find that XXX
 
@@
   %   {Dataset captions}
-    \input{tables/datasets.tex} 
-    \label{tab:data}  
+    \input{tables/table_datasets.tex} 
+    \label{tab:datasets}  
   % \end{table}
-% remove notes, add strat ages, add refs to dat sources
 

### 2017-12-07 manuscript/tables/table_datasets.tex
Add ref to table

@@
   Reference & \cite{beckancient2014} & \cite{brusatte2014gradual} & \cite{bapst2016topology} & \cite{wright2017bayesian} \\ 
-  Data reference &  CHECK & \cite{dryad_84t75} & \cite{dryad_n2g80} &  \cite{dryad_6hb7j} \\ 
+  Data reference &  \cite{beckancient2014} & \cite{dryad_84t75} & \cite{dryad_n2g80} &  \cite{dryad_6hb7j} \\ 
    \hline

### 2017-12-07 manuscript/time-slice.tex
Minor edits

@@
 They reveal how the morphological diversity of clades has changed through time, and allow us to make inferences about the breadth of ecological niches extinct taxa occupied \citep{foote1997evolution}.
-Results from disparity-through-time studies also provide insights into the ecological impacts of mass extinctions, competitive replacements, and the evolution of morphological innovations \citep{Brusatte12092008,Foote29111996,friedmanexplosive2010}.
+Results from disparity-through-time studies also provide insights into the ecological impacts of mass extinctions, competitive replacements, and the drivers of morphological evolution \citep{Brusatte12092008,Foote29111996,friedmanexplosive2010}.
 Unfortunately, the way we perform these analyses may have profound effects on our conclusions.
@@
 This can, however, lead to very large confidence intervals when there are stratigraphic periods with very few species.
-Other studies split large time bins so they are of roughly equal size, but this is quite an \textit{ad hoc} procedure that can introduce more bias depending on where bins are split.
+Other studies split large time bins so they are of roughly equal size, but this is often an \textit{ad hoc} procedure that can introduce more bias depending on where bins are split.
 Second, all time binning approaches (whether bins are equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring \textit{between} intervals, rather than also allowing for gradual changes \textit{within} intervals \citep{Hunt21042015}.
 Finally, time bin analyses are often limited by the number of taxa in each bin.
-If there are insufficient taxa in a time bin, disparity cannot be calculated, so further analysis, e.g.
-correlations of disparity with climate proxies or other hypothesised drivers of morphological evolution, are not possible.
+If there are insufficient taxa in a time bin, disparity cannot be calculated, so further analyses, e.g.
+correlations of disparity with hypothesised drivers of morphological evolution, are not possible.
 
@@
 
-Here we present the time slicing methods using four datasets taken from the literature.
+Here we present our time slicing methods using four datasets taken from the literature.
 We calculate disparity-through-time for each dataset using a range of time binning and time slicing methods, and then compare these approaches with respect to the relative disparities calculated, but also with how the different approaches influence biological conclusions.
@@
     \input{tables/table_datasets.tex} 
-    \label{tab:datasets}  
+    \label{datasets}  
   % \end{table}
 
-\subsection{Estimating ancestral character states}
+\subsection{Preparing the data for disparity-through-time analysis}
 \label{ace}
+
+\paragraph{Estimating ancestral character states.}
 For each dataset we used the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny, using the \texttt{rerootingMethod} function from the \texttt{R} package \texttt{phytools} version 0.6-44 \citep{phytools,R}.
@@
 
-\subsection{Building cladisto-spaces} 
+\paragraph{Building cladisto-spaces.} 
 To explore disparity-through-time in our datasets we used a cladisto-space approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
@@
 
-\subsection{Calculating disparity}
+\subsection{Disparity-through-time analyses}
 \label{disparity_calc}
-Disparity can be estimated in many different ways \citep[e.g.][]{Wills1994,Ciampaglio2004,thorneresetting2011,hopkinsdecoupling2013,huang2015origins}, however most studies estimate disparity using four metrics: the sum and products of ranges and variances, each of which gives a slightly different estimate of how the data fits within the cladisto-space \citep{Foote01071994,Wills1994,brusatte50,Brusatte12092008,cisneros2010,thorneresetting2011,prentice2011,brusattedinosaur2012,toljagictriassic-jurassic2013,ruta2013,bentonmodels2014,bensonfaunal2014}.
+
+\paragraph{Calculating disparity.}
+Disparity can be calculated in many different ways \citep[e.g.][]{Wills1994,Ciampaglio2004,thorneresetting2011,hopkinsdecoupling2013,huang2015origins}, however most studies estimate disparity using four metrics: the sum and products of ranges and variances, each of which gives a slightly different estimate of how the data fits within the cladisto-space \citep{Foote01071994,Wills1994,brusatte50,Brusatte12092008,cisneros2010,thorneresetting2011,prentice2011,brusattedinosaur2012,toljagictriassic-jurassic2013,ruta2013,bentonmodels2014,bensonfaunal2014}.
 However, these metrics have limitations. 
@@
 Second, because we include all $n$ dimensions in the analysis (see above), the products of ranges and variances will tend towards zero since the scores of the last dimension are usually really close to zero themselves.
-We therefore use the sum of variances metric to estimate disparity.
+We therefore use the sum of variances metric to estimate disparity here.
+Note that there are still statistical issues with this metric, but for the purposes of comparison with previous work we want to use a standard metric for these analyses.
 
-\subsection{Estimating disparity-through-time: time subsampling} 
+\paragraph{Time subsampling} 
 \label{time_subsamples}
@@
 
-\subsection{Estimating disparity-through-time: time slicing} 
+\paragraph{Time slicing} 
 \label{time_slicing}

### 2017-12-07 functions/multiplot_fun.R
Function to easily plot multiple graphs

@@ -1,0 +1,45 @@
+# Modified multiplot function
+# Written by Scott Chamberlain 
+# https://gist.github.com/sckott/8444444
+
+multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL, 
+                      labs=list(), labpos=list(c(0.5,0.03), c(0.03,0.5))) {
+  require(grid)
+  
+  # Make a list from the ... arguments and plotlist
+  plots <- c(list(...), plotlist)
+  
+  numPlots = length(plots)
+  
+  # If layout is NULL, then use 'cols' to determine layout
+  if (is.null(layout)) {
+    # Make the panel
+    # ncol: Number of columns of plots
+    # nrow: Number of rows needed, calculated from # of cols
+    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
+                     ncol = cols, nrow = ceiling(numPlots/cols))
+  }
+  
+  if (numPlots==1) {
+    print(plots[[1]])
+    
+  } else {
+    # Set up the page
+    grid.newpage()
+    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
+    
+    # Make each plot, in the correct location
+    for (i in 1:numPlots) {
+      # Get the i,j matrix positions of the regions that contain this subplot
+      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
+      
+      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
+                                      layout.pos.col = matchidx$col))
+    }
+    
+    if(!length(labs) == 0){
+      grid.text(labs[1], x=labpos[[1]][1], y=labpos[[1]][2], gp=gpar(fontsize=16))
+      grid.text(labs[2], x=labpos[[2]][1], y=labpos[[2]][2], rot=90, gp=gpar(fontsize=16))
+    }
+  }
+}

### 2017-12-07 analyses/04-figures.Rmd
Tidying up dtt figure

--- 2017-12-06 analyses/04-figures.Rmd
@@
 library(tidyverse)
+library(gridExtra)
+source("../functions/multiplot_fun.R")
+```
+
+```{r}
+slug <- "Beck2014"
+```
+
+Load the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
+
+strat <- out1$results$stratigraphy
+duration <- out1$results$duration
+number <- out1$results$number
+```
+
+Get the data into the correct format to plot
+
+```{r}
+# Stratigraphy
+# Get ages for the bins, go for the midpoints
+x <- str_split(strat$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+# Modify data - add midpoints, remove missing data, filter to just two models,
+# make disparity relative by dividing by max values
+strat2 <- 
+  strat %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Duration
+x <- str_split(duration$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+duration2 <- 
+  duration %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Number
+x <- str_split(number$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+number2 <- 
+  number %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+```
+
+Make plots in preparation for larger grid plot below
+
+```{r}
+
+beck_base <- 
+  ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
+  theme_classic(base_size = 16) + 
+  scale_y_continuous(breaks = seq(0, 1, 0.2), 
+                     labels = seq(0, 1, 0.2)) +
+  scale_x_reverse(limits = c(100, 0)) +
+  expand_limits(x = 0, y = 0)
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        legend.position = "none")
+
+beck_s <- 
+  beck_base +
+  geom_point(data = strat2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+beck_d <- 
+  beck_base +
+  geom_point(data = duration2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+beck_n <- 
+  beck_base +
+  geom_point(data = number2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
 ```
@@
   filter(bs.median != 0 & !is.na(bs.median)) %>%
-  filter(model == "equalbins" | model == "proximity")
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
 
@@
   filter(bs.median != 0 & !is.na(bs.median)) %>%
-  filter(model == "equalbins" | model == "proximity")
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
 ```
@@
 brusatte_base <- 
-  ggplot(strat2, aes(y = bs.median, x = time, colour = model)) +
+  ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
   theme_classic(base_size = 16) + 
-  xlab("Time (MYA)") + 
-  ylab("Disparity") + 
-  scale_y_continuous(breaks = seq(0, 10, 2), 
-                     labels = seq(0, 10, 2)) +
+  scale_y_continuous(breaks = seq(0, 1, 0.2), 
+                     labels = seq(0, 1, 0.2)) +
   scale_x_reverse(limits = c(100, 0)) +
-  expand_limits(x = 0, y = 3) +
-  theme(legend.position = "none")
+  expand_limits(x = 0, y = 0)
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        legend.position = "none")
 
@@
   geom_point(data = strat2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = strat2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
 
@@
   geom_point(data = duration2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = duration2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
 
@@
   geom_point(data = number2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = number2, aes(ymin = `2.5%`, ymax = `97.5%`), width = 1, alpha = 0.8)
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
 
 ```
+
+```{r}
+slug <- "Bapst2016"
+```
+
+Load the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
+
+strat <- out1$results$stratigraphy
+duration <- out1$results$duration
+number <- out1$results$number
+```
+
+Get the data into the correct format to plot
+
+```{r}
+# Stratigraphy
+# Get ages for the bins, go for the midpoints
+x <- str_split(strat$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+# Modify data - add midpoints, remove missing data, filter to just two models
+strat2 <- 
+  strat %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Duration
+x <- str_split(duration$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+duration2 <- 
+  duration %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Number
+x <- str_split(number$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+number2 <- 
+  number %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+```
+
+Make plots in preparation for larger grid plot below
+
+```{r}
+
+bapst_base <- 
+  ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
+  theme_classic(base_size = 16) + 
+  scale_y_continuous(breaks = seq(0, 1, 0.2), 
+                     labels = seq(0, 1, 0.2)) +
+  scale_x_reverse(limits = c(100, 0)) +
+  expand_limits(x = 0, y = 0)
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        legend.position = "none")
+
+bapst_s <- 
+  bapst_base +
+  geom_point(data = strat2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+bapst_d <- 
+  bapst_base +
+  geom_point(data = duration2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+bapst_n <- 
+  bapst_base +
+  geom_point(data = number2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+```
+
+```{r}
+slug <- "Wright2017"
+```
+
+Load the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
+
+strat <- out1$results$stratigraphy
+duration <- out1$results$duration
+number <- out1$results$number
+```
+
+Get the data into the correct format to plot
+
+```{r}
+# Stratigraphy
+# Get ages for the bins, go for the midpoints
+x <- str_split(strat$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+# Modify data - add midpoints, remove missing data, filter to just two models
+strat2 <- 
+  strat %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Duration
+x <- str_split(duration$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+duration2 <- 
+  duration %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Number
+x <- str_split(number$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+number2 <- 
+  number %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+```
+
+Make plots in preparation for larger grid plot below
+
+```{r}
+
+wright_base <- 
+  ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
+  theme_classic(base_size = 16) + 
+  scale_y_continuous(breaks = seq(0, 1, 0.2), 
+                     labels = seq(0, 1, 0.2)) +
+  scale_x_reverse(limits = c(470, 370),
+                  breaks = c(seq(470, 370, by = -20))) +
+  expand_limits(x = 0, y = 0) +
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank())
+
+wright_s <- 
+  wright_base +
+  geom_point(data = strat2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  theme(legend.position = "none")
+
+wright_d <- 
+  wright_base +
+  geom_point(data = duration2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  theme(legend.position = "none")
+
+wright_n <- 
+  wright_base +
+  geom_point(data = number2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  theme(legend.position = "none")
+
+```
+
+Finally plot them all together
+
+```{r}
+# Using modified multiplot function
+# written by Scott Chamberlain
+all <-   
+  multiplot(beck_s, beck_d, beck_n,
+            brusatte_s, brusatte_d, brusatte_n,
+            bapst_s, bapst_d, bapst_n,
+            wright_s, wright_d, wright_n,
+            cols = 3, 
+            labs = list("Time (MYA)", 
+                        "Relative disparity"))
+
+ggsave("../manuscript/figures/figure-dtt.png", all, 
+       device = png(width = 600, height = 400))
+```

### 2017-12-07 functions/multiplot_fun.R
Reduce label size

@@ -38,8 +38,8 @@ multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL,
     }
     
     if(!length(labs) == 0){
-      grid.text(labs[1], x=labpos[[1]][1], y=labpos[[1]][2], gp=gpar(fontsize=16))
-      grid.text(labs[2], x=labpos[[2]][1], y=labpos[[2]][2], rot=90, gp=gpar(fontsize=16))
+      grid.text(labs[1], x=labpos[[1]][1], y=labpos[[1]][2], gp=gpar(fontsize=14))
+      grid.text(labs[2], x=labpos[[2]][1], y=labpos[[2]][2], rot=90, gp=gpar(fontsize=14))
     }
   }
 }

### 2017-12-07 analyses/04-figures.Rmd
Finished DTT figure

@@
 ---
-title: "Figure 1: disparity through time - simplified and for appendix"
+title: "Disparity through time - simplified and for appendix"
 author: "Natalie Cooper"
@@
 library(gridExtra)
+library(grid)
 source("../functions/multiplot_fun.R")
@@
   ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
-  theme_classic(base_size = 16) + 
-  scale_y_continuous(breaks = seq(0, 1, 0.2), 
-                     labels = seq(0, 1, 0.2)) +
-  scale_x_reverse(limits = c(100, 0)) +
-  expand_limits(x = 0, y = 0)
+  theme_classic(base_size = 12) + 
+  scale_y_continuous(breaks = seq(0, 1, 0.5), 
+                     labels = seq(0, 1, 0.5)) +
+  scale_x_reverse(limits = c(100, 0),
+                  breaks = c(seq(100, 0, by = -50))) +
+  expand_limits(x = 0, y = c(0,1)) +
   theme(axis.title.y = element_blank(),
@@
   geom_point(data = strat2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  ggtitle("Stratigraphy") +
+  theme(plot.title = element_text(size = 14, hjust = 0.5)) +
+  annotate("text", x = 75, y = 0.15, label = "Beck2014")
 
@@
   geom_point(data = duration2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  ggtitle("Duration") +
+  theme(plot.title = element_text(size = 14, hjust = 0.5))
 
@@
   geom_point(data = number2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  ggtitle("Number") +
+  theme(plot.title = element_text(size = 14, hjust = 0.5))
 
@@
   geom_point(data = strat2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  annotate("text", x = 70, y = 0.15, label = "Brusatte2014")
 
@@
 ```{r}
-
 bapst_base <- 
   ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
-  theme_classic(base_size = 16) + 
-  scale_y_continuous(breaks = seq(0, 1, 0.2), 
-                     labels = seq(0, 1, 0.2)) +
-  scale_x_reverse(limits = c(100, 0)) +
-  expand_limits(x = 0, y = 0)
+  theme_classic(base_size = 12) + 
+  scale_y_continuous(breaks = seq(0, 1, 0.5), 
+                     labels = seq(0, 1, 0.5)) +
+  scale_x_reverse(limits = c(100, 0),
+                  breaks = c(seq(100, 0, by = -50))) +
+  expand_limits(x = 0, y = c(0,1)) +
   theme(axis.title.y = element_blank(),
@@
   geom_point(data = strat2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  annotate("text", x = 75, y = 0.15, label = "Bapst2016")
 
@@
   ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
-  theme_classic(base_size = 16) + 
-  scale_y_continuous(breaks = seq(0, 1, 0.2), 
-                     labels = seq(0, 1, 0.2)) +
+  theme_classic(base_size = 12) + 
+  scale_y_continuous(breaks = seq(0, 1, 0.5), 
+                     labels = seq(0, 1, 0.5)) +
   scale_x_reverse(limits = c(470, 370),
-                  breaks = c(seq(470, 370, by = -20))) +
-  expand_limits(x = 0, y = 0) +
+                  breaks = c(seq(470, 370, by = -50))) +
+  expand_limits(x = 0, y = c(0,1)) +
   theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank())
+        axis.title.x = element_blank(),
+        legend.position = "none")
 
@@
   geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
-  theme(legend.position = "none")
+  annotate("text", x = 440, y = 0.15, label = "Wright2017")
 
@@
   geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
-  theme(legend.position = "none")
+  scale_x_reverse(limits = c(470, 370),
+                  breaks = c(seq(470, 370, by = -100)))
 
@@
   geom_point(data = number2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
-  theme(legend.position = "none")
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
 
@@
 # written by Scott Chamberlain
-all <-   
-  multiplot(beck_s, beck_d, beck_n,
-            brusatte_s, brusatte_d, brusatte_n,
-            bapst_s, bapst_d, bapst_n,
-            wright_s, wright_d, wright_n,
+
+pdf("../manuscript/figures/fig-dtt.pdf")  
+  multiplot(beck_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")),
+            brusatte_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
+            bapst_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")),
+            wright_s + theme(plot.margin=unit(c(5.5, 5.5, 12, 25),"points")), 
+            beck_d, brusatte_d, bapst_d, 
+            wright_d + theme(plot.margin=unit(c(5.5, 5.5, 12, 5.5),"points")),
+            beck_n,
+            brusatte_n,
+            bapst_n,
+            wright_n + theme(plot.margin=unit(c(5.5, 5.5, 12, 5.5),"points")),
             cols = 3, 
             labs = list("Time (MYA)", 
-                        "Relative disparity"))
-
-ggsave("../manuscript/figures/figure-dtt.png", all, 
-       device = png(width = 600, height = 400))
+                        "Relative disparity"),
+            labpos=list(c(0.5, 0.015), c(0.03, 0.5)))
+dev.off()
 ```

### 2017-12-07 manuscript/time-slice.tex
Small edits

@@
 \label{time_slicing}
-% THOMAS TO UPDATE THIS BIT!!!
 The ``time-slicing'' approach considers subsets of taxa in the cladisto-space at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
 This results in even-sampling of the cladisto-space across time and while permitting to use an underlying model of character evolution (punctuated or gradual).
-% and others ADD).
-%TG: I don't have any other ones in stock.
-But it could be easy to come up with new ones.
-They're easy to implement.
 
 In practice, time-slicing considers the disparity of any element present in the phylogeny (branches, nodes and tips) at any point in time.
-When the phylogenetic elements are nodes or tips, the ordination scores for the nodes (estimated using ancestral state reconstruction as described in section \ref{ace}) or tips are directly used for estimating disparity.
-When the phylogenetic elements are branches we chose the ordination score for the branch using one of two evolutionary models:
+When the phylogenetic elements are nodes or tips, the ordination scores for the nodes (estimated using ancestral state reconstruction as described above) or tips are directly used for estimating disparity.
+When the phylogenetic elements are branches we choose the ordination score for the branch using one of two evolutionary models:
 
@@
     This model selects the ordination score from either the ancestral node or the descendant node/tip of the branch regardless of the position of the slice along the branch. 
-    Similarly to the time interval approach, this reflects a model of punctuated evolution where changes in disparity occur either at the start or at the end of a branch over a relatively short time period and clades undergo long periods of stasis during their evolution \citep{Gould1977,Hunt20112007}.
-    We applied this model in four ways:
+    Similarly to the time bin approach, this reflects a model of punctuated evolution where changes in disparity occur either at the start or at the end of a branch over a relatively short time period and clades undergo long periods of stasis during their evolution \citep{Gould1977,Hunt20112007}.
+    We apply this model in four ways:
 
@@
       \item The ``deltran'' model, always selecting the ordination score of the descendant node/tip of the branch.
-      \item The ``random'' model, selecting randomly between the ordination score of the ancestor or the descendant of the branch.
+      \item The ``random'' model, randomly selecting the ordination score of either the ancestor or the descendant of the branch.
       \item The ``proximity'' model, selecting the ordination of either the ancestor or the descendant if the slice occurs respectively before or after half the branch length.
@@
 %TG: Justification on the fact that these models are not ACEs.
-It is important to notice that the time-slicing method is not an ancestral states estimation method \textit{per se}.
-This method does not estimate values along a branch applying a model (\textit{c.f.} methods described in \ref{ace}) but rather chooses between the two available information (the descendant or the ancestor) using the models described above.
+It is important to note that the time-slicing method is not an ancestral states estimation method \textit{per se}.
+This method does not estimate values along a branch applying a model (\textit{c.f.} methods described in \ref{ace}) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the models described above.
 This allows this method to be used in post-ordination analysis were the data used in each time slice is data already present in the cladisto-space.

### 2017-12-08 manuscript/time-slice.tex
grant up
Edits to end of time slice methods

@@
 To address these issues, we propose a ``time-slicing'' approach that takes advantage of the wealth of paleontological datasets which now have associated phylogenies. 
-Time slicing uses a phylogenetic tree and considers subsets of taxa at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
+Time-slicing uses a phylogenetic tree and considers subsets of taxa at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
 This results in even-sampling across time and permits us to define the underlying model of character evolution (punctuated or gradual).  
@@
 
-Here we present our time slicing methods using four datasets taken from the literature.
-We calculate disparity-through-time for each dataset using a range of time binning and time slicing methods, and then compare these approaches with respect to the relative disparities calculated, but also with how the different approaches influence biological conclusions.
+Here we present our time-slicing methods using four datasets taken from the literature.
+We calculate disparity-through-time for each dataset using a range of time binning and time-slicing methods, and then compare these approaches with respect to the relative disparities calculated, but also with how the different approaches influence biological conclusions.
 We find that XXX
@@
 \label{overview-section}
-% I'm unsure if this section is needed or not. The methods are long so it might help?
-%TG: agree though we might just do a simple figure like fig1 in http://onlinelibrary.wiley.com/doi/10.1111/bij.12746/pdf (but with way less stuff!). I can get working on that.
-To test the different time subsampling methods, we followed the protocol below (Figure). 
-All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (https://github.com/nhcooper123/time-slice). % NC: might move this to the NHM github account
+To test the different time subsampling methods, we followed the protocol below (Figure \ref{datasets}). 
+All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (https://github.com/nhcooper123/time-slice). 
+% NC: might move this to the NHM github account
 
-%\begin{enumerate}
- % \item Collating example datasets.
-Disparity-through-time analysis requires a matrix containing morphological characters for each taxon, and information on the first and last occurrence dates of each taxon.
-Time slicing methods also require a phylogeny.
-  %\item Processing example datasets.
-Various transformations of the datasets are required before the analyses can take place.
-We need to estimate ancestral character states at the nodes of the phylogeny for the time slicing methods, build cladisto-spaces via distance matrices and ordination, and remove taxa from the phylogeny that are missing from the matrix and vice versa.
-  %\item Time subsampling
-  %\item Estimate disparity for each time subsample
-  %\item Compare different methods
-%\end{enumerate}
+% Add overview figure
+
+  %\begin{figure}[!htbp]
+    %\centering
+    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/overview}
+    %\caption[caption title]
+    % {caption text}
+    %\label{overview}
+  %\end{figure}
 
@@
 Table \ref{datasets} and appendix ??? provide more details. % add appendix details
-%Note that although disparity-through-time analyses are possible without a phylogeny, we only use examples with phylogenies here so that we can compare results of time slicing methods, which do require a phylogeny.
 Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
@@
 
-  % \begin{table}[!htbp]      
-  %   \caption[Datasets.]
-  %   {Dataset captions}
+% Dataset details table
+  \begin{table}[!htbp]      
+    \caption[Datasets.]{caption}
     \input{tables/table_datasets.tex} 
     \label{datasets}  
-  % \end{table}
+  \end{table}
 
 \subsection{Preparing the data for disparity-through-time analysis}
-\label{ace}
 
 \paragraph{Estimating ancestral character states.}
+\label{ace}
 For each dataset we used the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny, using the \texttt{rerootingMethod} function from the \texttt{R} package \texttt{phytools} version 0.6-44 \citep{phytools,R}.
@@
 This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see appendix for details of which species).
-% add appendix
+% add appendix details
 
@@
 \subsection{Disparity-through-time analyses}
-\label{disparity_calc}
 
 \paragraph{Calculating disparity.}
+\label{disparity_calc}
 Disparity can be calculated in many different ways \citep[e.g.][]{Wills1994,Ciampaglio2004,thorneresetting2011,hopkinsdecoupling2013,huang2015origins}, however most studies estimate disparity using four metrics: the sum and products of ranges and variances, each of which gives a slightly different estimate of how the data fits within the cladisto-space \citep{Foote01071994,Wills1994,brusatte50,Brusatte12092008,cisneros2010,thorneresetting2011,prentice2011,brusattedinosaur2012,toljagictriassic-jurassic2013,ruta2013,bentonmodels2014,bensonfaunal2014}.
@@
 
-\paragraph{Time subsampling} 
+\paragraph{Time sub-sampling} 
 \label{time_subsamples}
@@
 To estimate disparity-through-time we first need to split the data into time subsamples.
-Here we explore the following scenarios.
+Here we use three time sub-sampling methods.
+
+\begin{enumerate}
+  \item Stratigraphic time bins.
+This is the traditional method, where all the taxa within each stratigraphic period are included in the disparity calculation.
+This often leads to bins of unequal duration.
+Here we use stratigraphic ages and epochs.
+  \item Equally sized time bins.
+This is another commonly used method, where the time frame of interest is split into equally sized time bins, then all the taxa within each time bin are included in the disparity calculation.
+  \item Time-slicing.
+We describe this in more detail below, but in brief, time-slicing uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation.
+\end{enumerate}
+
+\paragraph{Time-slicing} 
+\label{time_slicing}
+The ``time-slicing'' approach considers subsets of taxa in the cladisto-space at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
+This results in even-sampling of the cladisto-space across time and allows us to use different underlying models of character evolution (punctuated or gradual).
+
+In practice, time-slicing considers the disparity of any element present in the phylogeny (branches, nodes and tips) at any point in time.
+When the phylogenetic elements are nodes or tips, the ordination scores for the nodes (estimated using ancestral state reconstruction as described above) or tips are directly used for calculating disparity.
+When the phylogenetic elements are branches we choose the ordination score for the branch using one of two evolutionary models:
+
+%TG: Not sure if it's the "egeinvector" score or the "eigen" score.
+I think there's a trick in naming them depending on how the algorithm does the transformation (soemthing with using eigenvalues or eigenvectors).
+I call it ordination score for now.
+
+\begin{enumerate}
+
+    \item{\textbf{Punctuated evolution.}} 
+    This model selects the ordination score from either the ancestral node or the descendant node/tip of the branch regardless of the position of the slice along the branch. 
+    Similarly to the time bin approach, this reflects a model of punctuated evolution where changes in disparity occur either at the start or at the end of a branch over a relatively short time period, and clades undergo long periods of stasis during their evolution \citep{Gould1977,Hunt20112007}.
+    We apply this model in four ways:
+
+    \begin{enumerate}[(i)]
+
+      \item The ``acctran'' model, always selecting the ordination score of the ancestral node of the branch.
+      \item The ``deltran'' model, always selecting the ordination score of the descendant node/tip of the branch.
+      \item The ``random'' model, randomly selecting the ordination score of either the ancestor or the descendant of the branch.
+      \item The ``proximity'' model, selecting the ordination score of the ancestor if the slice occurs in the first half of the branch, and the descendant if the slice occurs in the second half of the branch.
+
+    \end{enumerate}
+
+    The two first models assume that changes always occur early (ACCelerated TRANsition) or late along the branches (DELayed TRANsition).
+    The third model makes neither assumption and simply selects data from the ancestor or the descendant at random, and the fourth bases the selection of either the ancestor or the descendant on where the slice occurs along the branch.
+    The punctuated models only select either the ordination score from the ancestor and the descendant once in the whole disparity analysis.
+    For example, if using the ``random'' model, if the data of the ancestor has been randomly chosen, only this data will be used during the bootstrapping (see below) and for the disparity calculation.
+
+    \item{\textbf{Gradual evolution.}}
+    Unlike the punctuated models, the following models do not select the ordination score of either the ancestor or the descendant but associate a probability to both.
+    This reflects a model of gradual evolution where changes in disparity are gradual and cumulative along the branch.
+
+    \begin{enumerate}[(i)]
+    \setcounter{enumii}{4}
+      \item The ``equal splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with an equal probability
+          \begin{equation}
+          p(\text{ancestor}) = p(\text{descendant}) = 0.5
+          \end{equation}
+
+    \item The ``gradual splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with a probability equal to the distance between the slice.
+    % NC: I'm unsure what you mean here: "the distance between the slice." Do you mean the branch length, the interval between slices, or the distance between the slice and the node?
+          \begin{equation}
+              p(\text{ancestor}) = \frac{d(\text{ancestor},\text{slice})}{d(\text{ancestor},\text{descendant})}
+          \end{equation}
+          \begin{equation}
+              p(\text{descendant}) = 1 - p(\text{ancestor})
+          \end{equation}
+    \noindent Where $d(x,y)$ is the distance measured as branch length.
+    \end{enumerate}
+
+    In these models, both the ordination scores of both the ancestor and descendant contribute to the disparity calculation based on the following models:
+    For example, using the ``gradual.split'' model, if slice occurs at the third quarter of a branch from A to B (75\% of the total branch length), after bootstrapping, the disparity results will be based on 25\% of the data from A and 75\% of the data from B.
+    Because of the probabilistic nature of these models, there are only meaningful when calculating disparity from bootstrapped datasets.
+\end{enumerate}
+
+%TG: Justification on the fact that these models are not ACEs.
+It is important to note that the time-slicing method is not an ancestral states estimation method \textit{per se}.
+This method does not estimate values along a branch applying a model (\textit{c.f.} methods described for reconstructing ancestral states in the preparing data for disparity-through-time section above) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the models described above.
+This allows the method to be used in post-ordination analysis where the data used in each time slice is data already present in the cladisto-space.
+In other words, this method does not require a re-ordination of the cladisto-space every time a slice goes through a branch, thus allowing the properties of the cladisto-space (e.g.
+distance between species, variance per axis, etc.) to remain constant.
 
@@
     \end{enumerate}
-  \item Time slicing.
-We describe this in more detail below, but in brief, time slicing uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation.
+  \item Time-slicing.
+We describe this in more detail below, but in brief, time-slicing uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation.
 Here we use three different methods to determine the intervals between the time slices.
@@
 
+Applying the models
+
+To compare our three approaches (stratigraphic time bins, equal time bins and time-slicing) we applied the methods as follows.
+
+\begin{enumerate}
+  \item Stratigraphic time bins. We calculated disparity for each stratigraphic 
+\begin{enumerate}
+  \item Unequal time bins and unevenly spaced time slices, where the bins are stratigraphic periods (ages or epochs) and slices occur at the midpoint of the stratigraphic period.
+  \item Equal time bins and equally spaced time slices, where time bin size, or the interval between slices, is defined by the average \textit{duration} of the stratigraphic period (age or epoch).
+  \item Equal time bins and equally spaced time slices, where the number of time bins, or number of slices, is defined by the average \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest.
 \end{enumerate}
 
-\paragraph{Time slicing} 
-\label{time_slicing}
-The ``time-slicing'' approach considers subsets of taxa in the cladisto-space at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
-This results in even-sampling of the cladisto-space across time and while permitting to use an underlying model of character evolution (punctuated or gradual).
-
-In practice, time-slicing considers the disparity of any element present in the phylogeny (branches, nodes and tips) at any point in time.
-When the phylogenetic elements are nodes or tips, the ordination scores for the nodes (estimated using ancestral state reconstruction as described above) or tips are directly used for estimating disparity.
-When the phylogenetic elements are branches we choose the ordination score for the branch using one of two evolutionary models:
-
-%TG: Not sure if it's the "egeinvector" score or the "eigen" score.
-I think there's a trick in naming them depending on how the algorithm does the transformation (soemthing with using eigenvalues or eigenvectors).
-I call it ordination score for now.
-
+  \item Equally sized time bins.
+This is another commonly used method, where the time frame of interest is split into equally sized time bins, then all the taxa within each time bin are included in the disparity calculation.
+Here we use three different methods to determine the size of the time bins.
     \begin{enumerate}
-
-    \item{\textbf{Punctuated evolution.}} 
-    This model selects the ordination score from either the ancestral node or the descendant node/tip of the branch regardless of the position of the slice along the branch. 
-    Similarly to the time bin approach, this reflects a model of punctuated evolution where changes in disparity occur either at the start or at the end of a branch over a relatively short time period and clades undergo long periods of stasis during their evolution \citep{Gould1977,Hunt20112007}.
-    We apply this model in four ways:
-
-    \begin{enumerate}[(i)]
-
-      \item The ``acctran'' model, always selecting the ordination score of the ancestral node of the branch.
-      \item The ``deltran'' model, always selecting the ordination score of the descendant node/tip of the branch.
-      \item The ``random'' model, randomly selecting the ordination score of either the ancestor or the descendant of the branch.
-      \item The ``proximity'' model, selecting the ordination of either the ancestor or the descendant if the slice occurs respectively before or after half the branch length.
-
+      \item The \textit{duration} of the bin is equal to the median duration of the stratigraphic period (age or epoch).
+      \item The \textit{number} of bins is equal to the number of stratigraphic periods (ages or epochs).
+      \item The \textit{number} of bins is either 2, 5, 10, 15, or 20.
     \end{enumerate}
-
-    The two first models assume that changes always occur early (ACCelerated TRANsition) or late along the branches (DELayed TRANsition).
-    The third model makes neither assumption and simply selects randomly the data between the ancestor and the descendant and the fourth base the selection of either the ancestor or the descendant on where the slice occurs along the branch.
-    The punctuated models only select either the ordination score from the ancestor and the descendant once in the whole disparity analysis.
-    For example, if using the ``random'' model, if the data of the ancestor has been randomly chosen, only this data will be used during the bootstraps procedure and for the disparity calculation.
-
-    \item{\textbf{Gradual evolution.}}
-    Contrary to the punctuated models, these following models do not select the ordination score of either the ancestor or the descendant but associate a probability to both.
-    This reflects a model of gradual evolution where changes in disparity are gradual and cumulative along the branch.
-
-    \begin{enumerate}[(i)]
-    \setcounter{enumii}{4}
-      \item The ``punctuated'' model (probabilistic), selecting the ordination score from both the ancestor and the descendant with an equal probability
-          \begin{equation}
-          p(\text{ancestor}) = p(\text{descendant}) = 0.5
-          \end{equation}
-
-    \item The ``gradual'' model (probabilistic), selecting the ordination score from both the ancestor and the descendant with a probability equal to the length between the slice.
-          \begin{equation}
-              p(\text{ancestor}) = \frac{d(\text{ancestor},\text{slice})}{d(\text{ancestor},\text{descendant})}
-          \end{equation}
-          \begin{equation}
-              p(\text{descendant}) = 1 - p(\text{ancestor})
-          \end{equation}
-    \noindent Where $d(x,y)$ is the distance measured as branch length.
+  \item Time-slicing.
+We describe this in more detail below, but in brief, time-slicing uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation.
+Here we use three different methods to determine the intervals between the time slices.
+    \begin{enumerate}
+      \item The \textit{interval} between slices is equal to the median duration of the stratigraphic period (age or epoch).
+      \item The \textit{number} of slices is equal to the number of stratigraphic periods (ages or epochs).
+      \item The \textit{number} of slices is either 2, 5, 10, 15, or 20.
     \end{enumerate}
-
-    In these models, both the ancestor's and descendant's data will participate to the disparity calculation based on the following models:
-    For example, using the ``gradual.split'' model, if slice occurs at the third quarter of a branch from A to B (75\% of the total branch length), after bootstrapping, the disparity results will be based on 25\% of the data from A and 75\% of the data from B.
-    Because of the probabilistic nature of these models, there are only meaningful when calculating disparity from bootstrapped datasets.
-\end{enumerate}
-
-%TG: Justification on the fact that these models are not ACEs.
-It is important to note that the time-slicing method is not an ancestral states estimation method \textit{per se}.
-This method does not estimate values along a branch applying a model (\textit{c.f.} methods described in \ref{ace}) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the models described above.
-This allows this method to be used in post-ordination analysis were the data used in each time slice is data already present in the cladisto-space.
-In other words, this method does not require a re-ordination of the cladisto-space every time a slice goes through a branch, thus allowing the properties of the cladisto-space (e.g.
-distance between species, variance per axis, etc.) to remain constant.
 
@@
 We then calculated the median disparity value for each subsample along with the 50\% and 95\% confidence intervals.
-We also recorded the number of taxa (or taxa and nodes for time slicing methods) in each subsample as a proxy for taxonomic diversity (Table \ref{tab:subsamples}).
+We also recorded the number of taxa (or taxa and nodes for time-slicing methods) in each subsample as a proxy for taxonomic diversity (Table \ref{tab:subsamples}).
 
@@
 \section{Acknowledgments}
-	We thank all the PalAss folk
+	We thank all the PalAss folk.
+  TG acknowledge support from the Australian Discovery Project Grant number DP170103227 awarded to Vera Weisbecker.
 

### 2017-12-08 manuscript/time-slice.tex
Edits to end of methods

@@
 
-\runninghead{Time subsamples in disparity-through-time analyses}
+\runninghead{Time sub-samples in disparity-through-time analyses}
 \keywords{time bin, time slice, disparity} 
@@
 
-Disparity-through-time analyes have two main analysis components: calculating disparity, and creating time subsamples of the data. 
+Disparity-through-time analyes have two main analysis components: calculating disparity, and creating time sub-samples of the data. 
 Here we focus on the latter.
@@
 \label{overview-section}
-To test the different time subsampling methods, we followed the protocol below (Figure \ref{datasets}). 
+To test the different time sub-sampling methods, we followed the protocol below (Figure \ref{datasets}). 
 All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (https://github.com/nhcooper123/time-slice). 
@@
 To test the different time binning/slicing methods we selected four datasets: a mammal dataset from \cite{beckancient2014}, two theropod datasets from \cite{brusatte2014gradual} and \cite{bapst2016topology}, and a crinoid dataset from \cite{wright2017bayesian}.
-Table \ref{datasets} and appendix ??? provide more details. % add appendix details
+Table \ref{table:datasets} and appendix ??? provide more details. % add appendix details
 Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
-These datasets are freely available with their accompanying papers (Table \ref{datasets}), but for reproducibility purposes we also provide the data we used on the NHM Data Portal (CITE).
+These datasets are freely available with their accompanying papers (Table \ref{table:datasets}), but for reproducibility purposes we also provide the data we used on the NHM Data Portal (CITE).
 
 % Dataset details table
-  \begin{table}[!htbp]      
-    \caption[Datasets.]{caption}
+\begin{landscape}
   \input{tables/table_datasets.tex} 
-    \label{datasets}  
-  \end{table}
+  \label{table:datasets}  
+\end{landscape}
 
@@
 the number of taxa in the matrix ($k$) minus the last two eigenvectors that are always null after applying the Cailliez correction.
-Contrary to previous studies \citep[e.g][]{brusatte50,cisneros2010,prentice2011,anderson2012using,Hughes20082013,bentonmodels2014}, we use all $n$ dimensions of our cladisto-spaces and not a subsample representing the majority of the variance in the distance matrix (e.g.
+Contrary to previous studies \citep[e.g][]{brusatte50,cisneros2010,prentice2011,anderson2012using,Hughes20082013,bentonmodels2014}, we use all $n$ dimensions of our cladisto-spaces and not a sub-sample representing the majority of the variance in the distance matrix (e.g.
 selecting only $m$ dimensions that represent up to 90\% of the variance in the distance matrix; \citealt{Brusatte12092008,toljagictriassic-jurassic2013}).
@@
 \paragraph{Time sub-sampling} 
-\label{time_subsamples}
+\label{time_sub-samples}
 
-To estimate disparity-through-time we first need to split the data into time subsamples.
+To estimate disparity-through-time we first need to split the data into time sub-samples.
 Here we use three time sub-sampling methods.
@@
 
-    In these models, both the ordination scores of both the ancestor and descendant contribute to the disparity calculation based on the following models:
-    For example, using the ``gradual.split'' model, if slice occurs at the third quarter of a branch from A to B (75\% of the total branch length), after bootstrapping, the disparity results will be based on 25\% of the data from A and 75\% of the data from B.
-    Because of the probabilistic nature of these models, there are only meaningful when calculating disparity from bootstrapped datasets.
+    In these models, the ordination scores of both the ancestor and descendant contribute to the disparity calculation.
+    For example, using the ``gradual.split'' model, if the slice occurs in the third quarter of a branch joining node A to node/tip B (75\% of the total branch length), after bootstrapping, the disparity results will be based on 25\% of the data from A and 75\% of the data from B.
+    Because of the probabilistic nature of these models, they are only meaningful when calculating disparity from bootstrapped datasets.
 \end{enumerate}
@@
 
+\subsection{Comparing time sub-sampling methods}
+
+To compare the time binning and time-slicing approaches we applied the methods as follows.
+
 \begin{enumerate}
-  \item Stratigraphic time bins.
-This is the traditional method, where all the taxa within each stratigraphic period are included in the disparity calculation.
-This often leads to bins of unequal duration.
-Here we use stratigraphic ages and epochs.
-  \item Equally sized time bins.
-This is another commonly used method, where the time frame of interest is split into equally sized time bins, then all the taxa within each time bin are included in the disparity calculation.
-Here we use three different methods to determine the size of the time bins.
+  \item Stratigraphy: time sub-samples defined by stratigraphic periods. 
   \begin{enumerate}
-      \item The \textit{duration} of the bin is equal to the median duration of the stratigraphic period (age or epoch).
-      \item The \textit{number} of bins is equal to the number of stratigraphic periods (ages or epochs).
-      \item The \textit{number} of bins is either 2, 5, 10, 15, or 20.
-    \end{enumerate}
-  \item Time-slicing.
-We describe this in more detail below, but in brief, time-slicing uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation.
-Here we use three different methods to determine the intervals between the time slices.
-    \begin{enumerate}
-      \item The \textit{interval} between slices is equal to the median duration of the stratigraphic period (age or epoch).
-      \item The \textit{number} of slices is equal to the number of stratigraphic periods (ages or epochs).
-      \item The \textit{number} of slices is either 2, 5, 10, 15, or 20.
+    \item Time bins (unequal). 
+    We calculated disparity for the taxa in each stratigraphic period (age or epoch). 
+    To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time bin by randomly resampling with replacement a new sub-sample of taxa from the observed taxa in the bin 1000 times.
+    We then calculated the median disparity value for each time bin along with the 50\% and 95\% confidence intervals.
+    \item Time-slices (unequal).
+    We calculated disparity using our time-slicing approach with slices occurring at the midpoint of each stratigraphic period (age or epoch), and using all six time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits).
+    To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement as described above for the stratigraphic time bins.
     \end{enumerate}
 
-Applying the models
-
-To compare our three approaches (stratigraphic time bins, equal time bins and time-slicing) we applied the methods as follows.
-
+  \item Duration: time sub-samples defined by the duration of stratigraphic periods. 
   \begin{enumerate}
-  \item Stratigraphic time bins. We calculated disparity for each stratigraphic 
-\begin{enumerate}
-  \item Unequal time bins and unevenly spaced time slices, where the bins are stratigraphic periods (ages or epochs) and slices occur at the midpoint of the stratigraphic period.
-  \item Equal time bins and equally spaced time slices, where time bin size, or the interval between slices, is defined by the average \textit{duration} of the stratigraphic period (age or epoch).
-  \item Equal time bins and equally spaced time slices, where the number of time bins, or number of slices, is defined by the average \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest.
+    \item Time bins (equal). 
+    We calculated disparity for the taxa in each time bin where time bin size was defined by the average \textit{duration} of the stratigraphic period (age or epoch), and bootstrapped the disparity values as described above.
+    \item Time-slices (unequal).
+    We calculated disparity using our time-slicing approach where the interval between slices, was defined by the average \textit{duration} of the stratigraphic period (age or epoch).
+    We used the six time-slicing methods and bootstrapped as described above.
   \end{enumerate}
 
-  \item Equally sized time bins.
-This is another commonly used method, where the time frame of interest is split into equally sized time bins, then all the taxa within each time bin are included in the disparity calculation.
-Here we use three different methods to determine the size of the time bins.
+  \item Number: time sub-samples defined by the number of stratigraphic periods. 
   \begin{enumerate}
-      \item The \textit{duration} of the bin is equal to the median duration of the stratigraphic period (age or epoch).
-      \item The \textit{number} of bins is equal to the number of stratigraphic periods (ages or epochs).
-      \item The \textit{number} of bins is either 2, 5, 10, 15, or 20.
-    \end{enumerate}
-  \item Time-slicing.
-We describe this in more detail below, but in brief, time-slicing uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation.
-Here we use three different methods to determine the intervals between the time slices.
-    \begin{enumerate}
-      \item The \textit{interval} between slices is equal to the median duration of the stratigraphic period (age or epoch).
-      \item The \textit{number} of slices is equal to the number of stratigraphic periods (ages or epochs).
-      \item The \textit{number} of slices is either 2, 5, 10, 15, or 20.
+    \item Time bins (equal). 
+    We calculated disparity for the taxa in each time bin where the number of time bins was defined by the average \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest, and bootstrapped the disparity values as described above.
+    \item Time-slices (unequal).
+    We calculated disparity using our time-slicing approach where the number of slices, was defined by the average \textit{number} of the stratigraphic periods (ages or epochs) in the time frame of interest..
+    We used the six time-slicing methods and bootstrapped as described above.
   \end{enumerate}
 
-We applied our time-slicing approach separately to each cladisto-space.
-We varied the intervals between time slices so that either (i) the *interval* between slices is equal to the median duration of the stratigraphic age or epoch; (ii) the *number* of slices is equal to the number of stratigraphic ages or epochs, or (iii) by setting the *number* of slices to 2, 5, 10, 15, or 20.
-For each subsample, we estimated its disparity assuming punctuated (ACCTRAN, DELTRAN and random) and gradual evolution as described above. % NEED TO ADD THE NEW MODELS.
-To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement as described above for the stratigraphic time bins. 
-We also recorded the number of phylogenetic elements (nodes and tips) in each subsample as a proxy for taxonomic diversity (Table \ref{tab:subsamples}).
+\end{enumerate}
+
+We also recorded the number of taxa (or taxa and nodes for time-slicing methods) in each sub-sample as a proxy for taxonomic diversity (Table \ref{tab:sub-samples}).
 
@@
 
-\subsection{Estimating disparity-through-time: bootstrapping}
-Finally, to reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time subsampling method, by randomly resampling with replacement a new subsample of taxa from the observed taxa in the subsample 1000 times.
-We then calculated the median disparity value for each subsample along with the 50\% and 95\% confidence intervals.
-We also recorded the number of taxa (or taxa and nodes for time-slicing methods) in each subsample as a proxy for taxonomic diversity (Table \ref{tab:subsamples}).
-
-\subsection{Comparing time subsampling methods}
+\subsection{Testing for differences in the time sub-sampling methods}
 \label{testing}
-Testing for statistical differences among the time subsampling methods described above is a little tricky, as we need to compare similar units, and also to tackle questions important to the interpretation of disparity-through-time analyses.
-We therefore present three different, simple ways of comparing the time subsampling methods as follows.
+Testing for statistical differences among the time sub-sampling methods described above is a little tricky, as we need to compare similar units, and also to tackle questions important to the interpretation of disparity-through-time analyses.
+We therefore present three different, simple ways of comparing the time sub-sampling methods as follows.
 
 \paragraph{Systematic differences in disparity-through-time.} 
-To test whether using time bins or time slices resulted in significantly different disparity values at common time points, we used paired Wilcoxon tests to compare the following.
-
-\begin{enumerate}
-  \item Unequal time bins and unevenly spaced time slices, where the bins are stratigraphic periods (ages or epochs) and slices occur at the midpoint of the stratigraphic period.
-  \item Equal time bins and equally spaced time slices, where time bin size, or the interval between slices, is defined by the average \textit{duration} of the stratigraphic period (age or epoch).
-  \item Equal time bins and equally spaced time slices, where the number of time bins, or number of slices, is defined by the average \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest.
-\end{enumerate}
+To test whether using time bins or time slices resulted in significantly different disparity values at common time points, we used paired Wilcoxon tests to compare the median bootstrapped disparities obtained in the stratigraphy (time sub-samples defined by stratigraphic periods), duration (time sub-samples defined by the duration of stratigraphic periods), and number (time sub-samples defined by the number of stratigraphic periods) analyses described above.
 
@@
 \paragraph{Disparity peaks and troughs.}
-We are perhaps more interested in how the conclusions disparity-through-time analyses are influenced by the choice of time subsampling method, rather than the disparities estimated by each method \textit{per se}, especially as these will influenced by the number of taxa (and/or nodes) included in each subsample.
-Therefore, we also investigate where peaks and troughs of disparity occur in each of our datasets for the different time subsampling methods. 
-We calculated the maximum and minimum bootstrapped disparities for each dataset and for each time subsampling method, along with their associated confidence intervals.
-Significant shifts in disparity peaks or troughs suggest that the choice of time subsampling method will influence our conclusions about relative changes in the disparity of our example clades through time.
+We are perhaps more interested in how the conclusions disparity-through-time analyses are influenced by the choice of time sub-sampling method, rather than the disparities estimated by each method \textit{per se}, especially as these will influenced by the number of taxa (and/or nodes) included in each sub-sample.
+Therefore, we also investigate where peaks and troughs of disparity occur in each of our datasets for the different time sub-sampling methods. 
+We calculated the maximum and minimum bootstrapped disparities for each dataset and for each time sub-sampling method, along with their associated confidence intervals.
+Significant shifts in disparity peaks or troughs suggest that the choice of time sub-sampling method will influence our conclusions about relative changes in the disparity of our example clades through time.
 
@@
 Many analyses of disparity-through-time aim to demonstrate differences in disparity before and after mass extinction events [citations]. 
-Each of our four datasets span one mass extinction (Cretaceous-Paleogene 66 MYA; \cite{brusatte2014gradual,bapst2016topology,beckancient2014} and Ordovician- XX MYA \cite{wright2017bayesian}), so we use Wilcoxon tests to compare disparity in the time bin/slice prior to the approrpiate mass extinction, to that of the time bin/slice following the extinction event.
+Each of our four datasets span one mass extinction (Cretaceous-Paleogene 66 MYA; \citealt{brusatte2014gradual,bapst2016topology,beckancient2014}; Ordovician-Silurian 455– 430 MYA; \citealt{wright2017bayesian}), so we use Wilcoxon tests to compare disparity in the time bin/slice prior to the approrpiate mass extinction, to that of the time bin/slice following the extinction event.
 Significant results suggest a significant effect of the mass extinction on disparity in the group.
-We then compare these results across the time subsampling methods to determine if our conclusions would change depending on the method used.
+We then compare these results across the time sub-sampling methods to determine if our conclusions would change depending on the method used.
 We repeated these analyses using the two time bins/slices after the one immediately following the mass extinction event to account for any lag effects of the mass extinction on disparity.
@@
 
-\subsection{Estimating disparity-through-time}
+\subsection{Disparity-through-time analyses}
 
@@
 
-\subsection{Comparing time subsampling methods}
+\subsection{Testing for differences in the time sub-sampling methods}
 % some demonstration of the differences in the binning - the ladder versus lines plot

### 2017-12-08 manuscript/time-slice.tex
Edits to acknowledgments

@@
   \item Stratigraphy: time sub-samples defined by stratigraphic periods. 
-  \begin{enumerate}
+  \begin{enumerate}[(i)]
     \item Time bins (unequal). 
@@
   \item Duration: time sub-samples defined by the duration of stratigraphic periods. 
-  \begin{enumerate}
+  \begin{enumerate}[(i)]
     \item Time bins (equal). 
@@
   \item Number: time sub-samples defined by the number of stratigraphic periods. 
-  \begin{enumerate}
+  \begin{enumerate}[(i)]
     \item Time bins (equal). 
@@
 \section{Acknowledgments}
-	We thank all the PalAss folk.
-  TG acknowledge support from the Australian Discovery Project Grant number DP170103227 awarded to Vera Weisbecker.
+	We thank Mark Sutton and Phillip Mannion for inviting us to contribute to the `Evolutionary Modelling' symposium at The Palaeontological Association Annual Meeting 2017.
+  TG acknowledges support from the Australian Discovery Project Grant number DP170103227 awarded to Vera Weisbecker.
 

### 2017-12-08 manuscript/time-slice.tex
Add results part 1

@@
 
-The full results tables can be found in appendix ??? % details.
+Disparity changes through time for each of the four datasets (Figure \ref{figure:dtt}, Figure in appendix, full tables in the appendix). 
+Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods.
+The six different time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits) show similar patterns, so we focus only of the results for one method with a punctuated model of evolution (specifically the `proximity' method), and one method with a gradual model of evolution (specifically the `gradual splits' method).
+Results for all six methods can be found in the appendix. % add details of appendix
 
-% Remind somewhere here that we focus on the proximity and gradual slice methods in the results. But full results are in appendix
+  \begin{figure}[!htbp]
+    \centering
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt.pdf}
+    \caption[Relative disparity through time for four example datasets.]
+    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
+    Pink points represent time binning methods, blue points are time slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
+    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods.
+    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
+    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
+    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
+    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
+    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).
+    Results for stratigraphic ages, and for other time-slicing models, are in the appendix.}
+    \label{figure:dtt}
+  \end{figure}
 
 \subsection{Testing for differences in the time sub-sampling methods}
-% some demonstration of the differences in the binning - the ladder versus lines plot
 
 \paragraph{Systematic differences in disparity-through-time.} 
-% table? plots? Might be same as the plots above.
+The results of the Wilcoxon tests show that...
 
@@
 \section{Acknowledgments}
-	We thank Mark Sutton and Phillip Mannion for inviting us to contribute to the `Evolutionary Modelling' symposium at The Palaeontological Association Annual Meeting 2017.
+	NC thanks Mark Sutton and Phillip Mannion for the invitation to contribute to the `Evolutionary Modelling' symposium at The Palaeontological Association Annual Meeting 2017.
   TG acknowledges support from the Australian Discovery Project Grant number DP170103227 awarded to Vera Weisbecker.
+  We thank Dave Bapst, Steve Brusatte, Graeme Lloyd, April Wright and Davey Wright for assistance in gathering data for the analyses.
 

### 2017-12-08 analyses/06-table-wilcox.Rmd
Add code to make wilcoxon tables for rresults and appendix

---
title: "Wilcoxon results tables"
author: "Natalie Cooper"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 12
---

# Compiling the tables for the paper

Loading the packages

```{r}
library(xtable)
library(Claddis)
library(ape)
library(dispRity)
library(tidyverse)
```

Loading the data:

```{r}
## Getting the data
slugs <- list("Beck2014", "Brusatte2014", "Bapst2016", "Wright2017")

## Wilcox results
wilk <- lapply(slugs, function(slug) read.csv(paste0("../outputs/wilcox.", slug, ".csv"), 
                                              header = FALSE))

```

Manipulate the data into the format we need.

```{r}
# Stick p values and W values together
# Repalce p values with *
p.values.function <- function(data){ 
  for (i in 1:length(data$V2)){
    if(data$V2[i] < 0.05 & data$V2[i] > 0.01){data$V2[i] <- "*"}
    if(data$V2[i] < 0.01 & data$V2[i] > 0.001){data$V2[i] <- "**"}
    if(data$V2[i] < 0.001){data$V2[i] <- "***"}
    if(data$V2[i] > 0.05){data$V2[i] <- ""}
  }
  data$V6 <- paste0(data$V1, data$V2)
return(data)
}

# Extract epoch only, and proximity and gradual split models
# Remove excess columns
# Add model column name
fix.wilcox <- function(data){
  data %>%
  filter(V5 == "Epoch") %>%
  select(-c(V1,V2, V5)) %>%
  filter(V3 == "proximity" | V3 == "gradual.split") %>%
  spread(V4, V6) %>%
  rename(model = V3)
}

wilk2 <- purrr::map(wilk, p.values.function)
wilk3 <- purrr::map_df(wilk2, fix.wilcox)
wilk3$dataset <- c(rep(unlist(slugs)[1], 2), rep(unlist(slugs)[2], 2), 
                   rep(unlist(slugs)[3], 2), rep(unlist(slugs)[4], 2))

```

## Create table for epochs

```{r}
## variable names
colnames <- c("Dataset", "Model", "Stratigraphy", "Duration", "Number")

## Formatting the table
table_out <- data.frame(wilk3$dataset, wilk3$model, wilk3$stratigraphy,
                        wilk3$duration, wilk3$number)
colnames(table_out) <- colnames

## Table caption
table_caption <- "Wilcoxon results"

sink(file = "../manuscript/tables/raw_wilcox_epoch.tex")
print(xtable(table_out, caption = table_caption), include.rownames = FALSE)
sink()
```

## Repeat for appendix tables with epoch and age and all models

Manipulate the data into the format we need.

```{r}
# Remove excess columns
# Add model column name
fix.wilcox.appendix <- function(data){
  data %>%
  select(-c(V1,V2)) %>%
  spread(V4, V6) %>%
  arrange(V5) %>%
  rename(model = V3) %>%
  rename(period = V5)
}

wilk2 <- purrr::map(wilk, p.values.function)
wilk3 <- purrr::map_df(wilk2, fix.wilcox.appendix)
wilk3$dataset <- c(rep(unlist(slugs)[1], 2), rep(unlist(slugs)[2], 2), 
                   rep(unlist(slugs)[3], 2), rep(unlist(slugs)[4], 2))

```

## Create table for all data

```{r}
## variable names
colnames <- c("Dataset", "Period", "Model", "Stratigraphy", "Duration", "Number")

## Formatting the table
table_out <- data.frame(wilk3$dataset, wilk3$period,
                        wilk3$model, wilk3$stratigraphy,
                        wilk3$duration, wilk3$number)
colnames(table_out) <- colnames

## Table caption
table_caption <- "Wilcoxon results for appendix"

sink(file = "../manuscript/tables/raw_wilcox_appendix.tex")
print(xtable(table_out, caption = table_caption), include.rownames = FALSE)
sink()
```

### 2017-12-08 analyses/04-figure-dtt.Rmd
Add appendix figures

--- 2017-12-07 analyses/04-figures.Rmd
@@
 
+```{r}
+slug <- "Beck2014"
+```
+
+Load the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
+
+strat <- out1$results$stratigraphy
+duration <- out1$results$duration
+number <- out1$results$number
+```
+
+Get the data into the correct format to plot
+
+```{r}
+# Stratigraphy
+# Get ages for the bins, go for the midpoints
+x <- str_split(strat$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+# Modify data - add midpoints, remove missing data, filter to just two models,
+# make disparity relative by dividing by max values
+strat2 <- 
+  strat %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Duration
+x <- str_split(duration$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+duration2 <- 
+  duration %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Number
+x <- str_split(number$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+number2 <- 
+  number %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+```
+
+Make plots in preparation for larger grid plot below
+
+```{r}
+
+beck_base <- 
+  ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
+  theme_classic(base_size = 12) + 
+  scale_y_continuous(breaks = seq(0, 1, 0.5), 
+                     labels = seq(0, 1, 0.5)) +
+  scale_x_reverse(limits = c(100, 0),
+                  breaks = c(seq(100, 0, by = -50))) +
+  expand_limits(x = 0, y = c(0,1)) +
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        legend.position = "none")
+
+beck_s <- 
+  beck_base +
+  geom_point(data = strat2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  ggtitle("Stratigraphy") +
+  theme(plot.title = element_text(size = 14, hjust = 0.5)) +
+  annotate("text", x = 75, y = 0.15, label = "Beck2014")
+
+beck_d <- 
+  beck_base +
+  geom_point(data = duration2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  ggtitle("Duration") +
+  theme(plot.title = element_text(size = 14, hjust = 0.5))
+
+beck_n <- 
+  beck_base +
+  geom_point(data = number2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  ggtitle("Number") +
+  theme(plot.title = element_text(size = 14, hjust = 0.5))
+
+```
+
+Pick the slug
+
+```{r}
+slug <- "Brusatte2014"
+```
+
+Load the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
+
+strat <- out1$results$stratigraphy
+duration <- out1$results$duration
+number <- out1$results$number
+```
+
+Get the data into the correct format to plot
+
+```{r}
+# Stratigraphy
+# Get ages for the bins, go for the midpoints
+x <- str_split(strat$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+# Modify data - add midpoints, remove missing data, filter to just two models
+strat2 <- 
+  strat %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Duration
+x <- str_split(duration$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+duration2 <- 
+  duration %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Number
+x <- str_split(number$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+number2 <- 
+  number %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+```
+
+Make plots in preparation for larger grid plot below
+
+```{r}
+
+brusatte_base <- 
+  ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
+  theme_classic(base_size = 12) + 
+  scale_y_continuous(breaks = seq(0, 1, 0.5), 
+                     labels = seq(0, 1, 0.5)) +
+  scale_x_reverse(limits = c(100, 0),
+                  breaks = c(seq(100, 0, by = -50))) +
+  expand_limits(x = 0, y = c(0,1)) +
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        legend.position = "none")
+
+brusatte_s <- 
+  brusatte_base +
+  geom_point(data = strat2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  annotate("text", x = 70, y = 0.15, label = "Brusatte2014")
+
+brusatte_d <- 
+  brusatte_base +
+  geom_point(data = duration2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+brusatte_n <- 
+  brusatte_base +
+  geom_point(data = number2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+```
+
+```{r}
+slug <- "Bapst2016"
+```
+
+Load the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
+
+strat <- out1$results$stratigraphy
+duration <- out1$results$duration
+number <- out1$results$number
+```
+
+Get the data into the correct format to plot
+
+```{r}
+# Stratigraphy
+# Get ages for the bins, go for the midpoints
+x <- str_split(strat$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+# Modify data - add midpoints, remove missing data, filter to just two models
+strat2 <- 
+  strat %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Duration
+x <- str_split(duration$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+duration2 <- 
+  duration %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Number
+x <- str_split(number$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+number2 <- 
+  number %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+```
+
+Make plots in preparation for larger grid plot below
+
+```{r}
+bapst_base <- 
+  ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
+  theme_classic(base_size = 12) + 
+  scale_y_continuous(breaks = seq(0, 1, 0.5), 
+                     labels = seq(0, 1, 0.5)) +
+  scale_x_reverse(limits = c(100, 0),
+                  breaks = c(seq(100, 0, by = -50))) +
+  expand_limits(x = 0, y = c(0,1)) +
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        legend.position = "none")
+
+bapst_s <- 
+  bapst_base +
+  geom_point(data = strat2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  annotate("text", x = 75, y = 0.15, label = "Bapst2016")
+
+bapst_d <- 
+  bapst_base +
+  geom_point(data = duration2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+bapst_n <- 
+  bapst_base +
+  geom_point(data = number2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+```
+
+```{r}
+slug <- "Wright2017"
+```
+
+Load the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
+
+strat <- out1$results$stratigraphy
+duration <- out1$results$duration
+number <- out1$results$number
+```
+
+Get the data into the correct format to plot
+
+```{r}
+# Stratigraphy
+# Get ages for the bins, go for the midpoints
+x <- str_split(strat$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+# Modify data - add midpoints, remove missing data, filter to just two models
+strat2 <- 
+  strat %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Duration
+x <- str_split(duration$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+duration2 <- 
+  duration %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Number
+x <- str_split(number$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+number2 <- 
+  number %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+```
+
+Make plots in preparation for larger grid plot below
+
+```{r}
+
+wright_base <- 
+  ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
+  theme_classic(base_size = 12) + 
+  scale_y_continuous(breaks = seq(0, 1, 0.5), 
+                     labels = seq(0, 1, 0.5)) +
+  scale_x_reverse(limits = c(470, 370),
+                  breaks = c(seq(470, 370, by = -50))) +
+  expand_limits(x = 0, y = c(0,1)) +
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        legend.position = "none")
+
+wright_s <- 
+  wright_base +
+  geom_point(data = strat2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  annotate("text", x = 440, y = 0.15, label = "Wright2017")
+
+wright_d <- 
+  wright_base +
+  geom_point(data = duration2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  scale_x_reverse(limits = c(470, 370),
+                  breaks = c(seq(470, 370, by = -100)))
+
+wright_n <- 
+  wright_base +
+  geom_point(data = number2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+```
+
+Finally plot them all together
+
+```{r}
+# Using modified multiplot function
+# written by Scott Chamberlain
+
+pdf("../manuscript/figures/fig-dtt-age.pdf")  
+  multiplot(beck_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")),
+            brusatte_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
+            bapst_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")),
+            wright_s + theme(plot.margin=unit(c(5.5, 5.5, 12, 25),"points")), 
+            beck_d, brusatte_d, bapst_d, 
+            wright_d + theme(plot.margin=unit(c(5.5, 5.5, 12, 5.5),"points")),
+            beck_n,
+            brusatte_n,
+            bapst_n,
+            wright_n + theme(plot.margin=unit(c(5.5, 5.5, 12, 5.5),"points")),
+            cols = 3, 
+            labs = list("Time (MYA)", 
+                        "Relative disparity"),
+            labpos=list(c(0.5, 0.015), c(0.03, 0.5)))
+dev.off()
+```
+
+Repeat for epochs
+-----------------
 ```{r}
@@
 
-pdf("../manuscript/figures/fig-dtt.pdf")  
+pdf("../manuscript/figures/fig-dtt-epoch.pdf")  
   multiplot(beck_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")),

### 2017-12-08 analyses/06-table-wilcox.Rmd
finish wilcox tables code

@@
 ---
-title: "Wilcoxon results tables"
+title: "Wilcoxon results tables - text and appendix"
 author: "Natalie Cooper"
@@
 
-# Extract epoch only, and proximity and gradual split models
+# Extract proximity and gradual split models
 # Remove excess columns
@@
   data %>%
-  filter(V5 == "Epoch") %>%
-  select(-c(V1,V2, V5)) %>%
+  select(-c(V1,V2)) %>%
   filter(V3 == "proximity" | V3 == "gradual.split") %>%
   spread(V4, V6) %>%
-  rename(model = V3)
+  arrange(V5) %>%
+  rename(model = V3) %>%
+  rename(period = V5)  
 }
@@
 wilk3 <- purrr::map_df(wilk2, fix.wilcox)
-wilk3$dataset <- c(rep(unlist(slugs)[1], 2), rep(unlist(slugs)[2], 2), 
-                   rep(unlist(slugs)[3], 2), rep(unlist(slugs)[4], 2))
+wilk3$dataset <- c(rep(unlist(slugs)[1], 4), rep(unlist(slugs)[2], 4), 
+                   rep(unlist(slugs)[3], 4), rep(unlist(slugs)[4], 4))
 
@@
 
-## Create table for epochs
+## Create table
 
@@
 ## variable names
-colnames <- c("Dataset", "Model", "Stratigraphy", "Duration", "Number")
+colnames <- c("Dataset", "Period", "Model", "Stratigraphy", "Duration", "Number")
 
 ## Formatting the table
-table_out <- data.frame(wilk3$dataset, wilk3$model, wilk3$stratigraphy,
+table_out <- data.frame(wilk3$dataset, wilk3$period, wilk3$model, wilk3$stratigraphy,
                         wilk3$duration, wilk3$number)
@@
 
-sink(file = "../manuscript/tables/raw_wilcox_epoch.tex")
+sink(file = "../manuscript/tables/raw_wilcox.tex")
 print(xtable(table_out, caption = table_caption), include.rownames = FALSE)
@@
 
-## Repeat for appendix tables with epoch and age and all models
+## Repeat for appendix tables with all models
 
@@
 wilk3 <- purrr::map_df(wilk2, fix.wilcox.appendix)
-wilk3$dataset <- c(rep(unlist(slugs)[1], 2), rep(unlist(slugs)[2], 2), 
-                   rep(unlist(slugs)[3], 2), rep(unlist(slugs)[4], 2))
+wilk3$dataset <- c(rep(unlist(slugs)[1], 12), rep(unlist(slugs)[2], 12), 
+                   rep(unlist(slugs)[3], 12), rep(unlist(slugs)[4], 12))
 

### 2017-12-08 manuscript/tables/table_wilcox.tex
Add Wilcoxon tables

% latex table generated in R 3.4.2 by xtable 1.8-2 package
% Fri Dec  8 17:24:09 2017
\begin{table}[!htbp]
\centering
\begin{tabular}{lllccc}
  \hline
\textbf{Dataset} & \textbf{Period} & \textbf{Model} & \textbf{Stratigraphy} & \textbf{Duration} & \textbf{Number} \\ 
  \hline
  Beck2014 & Age & gradual.split & 111 & 115*** & 65*** \\ 
  Beck2014 & Age & proximity & 105 & 83 & 68*** \\ 
  Beck2014 & Epoch & gradual.split & 21 & 39 & 43*** \\ 
  Beck2014 & Epoch & proximity & 21 & 36 & 32 \\ 
  Brusatte2014 & Age & gradual.split & 28*** & 61*** & 52*** \\ 
  Brusatte2014 & Age & proximity & 27*** & 31 & 28*** \\ 
  Brusatte2014 & Epoch & gradual.split & 3 & 6 & 6 \\ 
  Brusatte2014 & Epoch & proximity & 0 & 5*** & 5 \\ 
  Bapst2016 & Age & gradual.split & 93 & 153 & 165 \\ 
  Bapst2016 & Age & proximity & 57*** & 47 & 75*** \\ 
  Bapst2016 & Epoch & gradual.split & 4 & 6 & 12 \\ 
  Bapst2016 & Epoch & proximity & 2 & 0*** & 8 \\ 
  Wright2017 & Age & gradual.split & 152*** & 155 & 116 \\ 
  Wright2017 & Age & proximity & 160*** & 175*** & 101 \\ 
  Wright2017 & Epoch & gradual.split & 28 & 29 & 21 \\ 
  Wright2017 & Epoch & proximity & 23 & 28 & 18 \\ 
   \hline
\end{tabular}
\caption{Wilcoxon results} 
\end{table}

### 2017-12-08 manuscript/tables/table_wilcox_appendix.tex
Add Wilcoxon tables

% latex table generated in R 3.4.2 by xtable 1.8-2 package
% Fri Dec  8 17:06:33 2017
\begin{table}[!htbp]
\centering
\begin{tabular}{lllccc}
  \hline
\textbf{Dataset} & \textbf{Period} & \textbf{Model} & \textbf{Stratigraphy} & \textbf{Duration} & \textbf{Number} \\ 
  \hline
Beck2014 & Age & acctran & 39*** & 76 & 11 \\ 
  Beck2014 & Age & deltran & 188*** & 194*** & 171 \\ 
  Beck2014 & Age & equal.split & 91 & 119*** & 47 \\ 
  Beck2014 & Age & gradual.split & 111 & 115*** & 65*** \\ 
  Beck2014 & Age & proximity & 105 & 83 & 68*** \\ 
  Beck2014 & Age & random & 97 & 104*** & 45 \\ 
  Beck2014 & Epoch & acctran & 14 & 10 & 14 \\ 
  Beck2014 & Epoch & deltran & 21 & 45*** & 41*** \\ 
  Beck2014 & Epoch & equal.split & 21 & 40*** & 42*** \\ 
  Beck2014 & Epoch & gradual.split & 21 & 39 & 43*** \\ 
  Beck2014 & Epoch & proximity & 21 & 36 & 32 \\ 
  Beck2014 & Epoch & random & 21 & 37 & 45*** \\ 
  Brusatte2014 & Age & acctran & 27*** & 28 & 28*** \\ 
  Brusatte2014 & Age & deltran & 27*** & 29 & 31*** \\ 
  Brusatte2014 & Age & equal.split & 28*** & 58*** & 50*** \\ 
  Brusatte2014 & Age & gradual.split & 28*** & 61*** & 52*** \\ 
  Brusatte2014 & Age & proximity & 27*** & 31 & 28*** \\ 
  Brusatte2014 & Age & random & 27*** & 27 & 27*** \\ 
  Brusatte2014 & Epoch & acctran & 0 & 5*** & 5 \\ 
  Brusatte2014 & Epoch & deltran & 0 & 5*** & 5 \\ 
  Brusatte2014 & Epoch & equal.split & 3 & 6 & 6 \\ 
  Brusatte2014 & Epoch & gradual.split & 3 & 6 & 6 \\ 
  Brusatte2014 & Epoch & proximity & 0 & 5*** & 5 \\ 
  Brusatte2014 & Epoch & random & 0 & 5*** & 5 \\ 
  Bapst2016 & Age & acctran & 45*** & 47 & 72*** \\ 
  Bapst2016 & Age & deltran & 55*** & 46 & 78*** \\ 
  Bapst2016 & Age & equal.split & 93 & 147*** & 153 \\ 
  Bapst2016 & Age & gradual.split & 93 & 153 & 165 \\ 
  Bapst2016 & Age & proximity & 57*** & 47 & 75*** \\ 
  Bapst2016 & Age & random & 57*** & 48 & 81*** \\ 
  Bapst2016 & Epoch & acctran & 2 & 0*** & 8 \\ 
  Bapst2016 & Epoch & deltran & 2 & 0*** & 9 \\ 
  Bapst2016 & Epoch & equal.split & 4 & 6 & 13 \\ 
  Bapst2016 & Epoch & gradual.split & 4 & 6 & 12 \\ 
  Bapst2016 & Epoch & proximity & 2 & 0*** & 8 \\ 
  Bapst2016 & Epoch & random & 2 & 1*** & 8 \\ 
  Wright2017 & Age & acctran & 146*** & 146 & 84 \\ 
  Wright2017 & Age & deltran & 162*** & 138 & 101 \\ 
  Wright2017 & Age & equal.split & 151*** & 160 & 105 \\ 
  Wright2017 & Age & gradual.split & 152*** & 155 & 116 \\ 
  Wright2017 & Age & proximity & 160*** & 175*** & 101 \\ 
  Wright2017 & Age & random & 150*** & 147 & 111 \\ 
  Wright2017 & Epoch & acctran & 25 & 20 & 18 \\ 
  Wright2017 & Epoch & deltran & 27 & 26 & 25 \\ 
  Wright2017 & Epoch & equal.split & 29 & 30 & 25 \\ 
  Wright2017 & Epoch & gradual.split & 28 & 29 & 21 \\ 
  Wright2017 & Epoch & proximity & 23 & 28 & 18 \\ 
  Wright2017 & Epoch & random & 28 & 23 & 17 \\ 
   \hline
\end{tabular}
\caption{Wilcoxon results for appendix} 
\end{table}

### 2017-12-08 manuscript/time-slice.tex
Add appendix and modify results

@@
 
-% Mention somewhere here that we focus on the proximity and gradual slice methods in the results.
-
 \subsection{Testing for differences in the time sub-sampling methods}
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch.pdf}
     \caption[Relative disparity through time for four example datasets.]
@@
     Results for stratigraphic ages, and for other time-slicing models, are in the appendix.}
-    \label{figure:dtt}
+    \label{figure:dtt1}
   \end{figure}
@@
 \paragraph{Systematic differences in disparity-through-time.} 
-The results of the Wilcoxon tests show that...
+There is no overall significant systematic difference among the disparities calculated using time bins and those calculated using the time-slicing methods.
+(Table \ref{table:wilcox}, Appendix Table \ref{table:wilcox2}).
+Instead, the differences depend on the dataset and method in question.
+For example, the \cite{brusatte2014gradual}, \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, show significant differences when using bins versus time slices defined by stratigraphy, but the \cite{beckancient2014} dataset appears robust to these different models.
+Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} show different disparities when the number of bins or slices is the average number of stratigraphic periods, but this is not see in the \cite{wright2017bayesian} dataset.
+Note that for epochs, we find fewer significant differences simply because the smaller numbers of bins and slices being compared means we have low power to detect a significant difference.
+
+% Wilcoxon results table
+%\begin{landscape}
+  \input{tables/table_wilcox.tex} 
+  \label{table:wilcox}  
+%\end{landscape}
 
@@
 
+\section{APPENDIX}
+
+  \begin{figure}[!htbp]
+    \centering
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch-appendix.pdf}
+    \caption[Relative disparity through time for four example datasets.]
+    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
+    Pink points represent time binning methods, 
+    % blue points are time slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
+    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods.
+    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
+    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
+    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
+    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
+    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
+    \label{figure:dtt2}
+  \end{figure}
+
+  \begin{figure}[!htbp]
+    \centering
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-age-appendix.pdf}
+    \caption[Relative disparity through time for four example datasets.]
+    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
+    Pink points represent time binning methods, 
+    %blue points are time slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
+    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods.
+    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages. 
+    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages in the time frame of the dataset. 
+    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages in the time frame of the dataset. 
+    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
+    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
+    \label{figure:dtt3}
+  \end{figure}
+
+  % Wilcoxon results table
+%\begin{landscape}
+  \input{tables/table_wilcox_appendix.tex} 
+  \label{table:wilcox2}  
+%\end{landscape}
+
 \bibliographystyle{palaeo} 

### 2017-12-10 manuscript/time-slice.tex
Minor edits

@@
 All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (https://github.com/nhcooper123/time-slice).
+
+https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/details-of-specific-functions.html#time-slicing
+
 % NC: might move this to the NHM github account
@@
 Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
-These datasets are freely available with their accompanying papers (Table \ref{table:datasets}), but for reproducibility purposes we also provide the data we used on the NHM Data Portal (CITE).
+These datasets are freely available with their accompanying papers (Table \ref{table:datasets}), but for reproducibility purposes we also provide the data we used on GitHub (https://github.com/nhcooper123/time-slice/data).
 
@@
 
-%PCA figure
-	%\begin{figure}[!htbp]
-	%\centering
-	%\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/PCA_threeplot.png}
-	%\caption[Morphospace (principal components) plot of morphological diversity in tenrec and golden mole skulls.]
-	%	{Principal components plots of the morphospaces occupied by tenrecs (triangles, n=31 species) and golden moles (circles, n=12 species) for skulls in dorsal (top left), ventral (top right) and lateral (bottom left) views.
-Each point represents the average skull shape of an individual species.
-Axes are principal component 1 (PC1) and principal component 2 (PC2) of the average scores from principal components analyses of mean Procrustes shape coordinates for each species.}
-	%\label{fig:threePCA}
-	%\end{figure}
-
-% Results table
-	%\begin{table}[!htbp]			
-		%\caption[Comparing morphological diversity in tenrecs and golden moles.]
-		%{Morphological diversity in tenrecs compared to golden moles (12 species).
-N is the number of tenrec species: 31 species or 17 species including just five representatives of the \textit{Microgale} Genus.
-Morphological diversity of the Family is the mean Euclidean distance from each species to the Family centroid.
-Significant differences between the two Families (p$<$0.05) from two-tailed t-tests are highlighted in bold.}
-		%\input{tables/diversity.results} 
-		%\label{tab:diversity}  
-	%\end{table}
-
 \section{Discussion}

### 2017-12-11 manuscript/time-slice.tex
Added outline figure first try + answers to comments

@@
 % NC: OK so lets save ourselves some pain and not mention this at all and save it all for the bitching in the review paper!!!
+% TG: agree!
 
@@
 % NC: might move this to the NHM github account
+% TG: why that? It'll be just extra hassle no?
 
 % Add overview figure
+%TG: This is a first draft of the figure and its caption. Tell me what you think.
 
-  %\begin{figure}[!htbp]
-    %\centering
-    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/overview}
-    %\caption[caption title]
-    % {caption text}
-    %\label{overview}
-  %\end{figure}
+  \begin{figure}[!htbp]
+    \centering
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{outlinesvg.pdf}
+    \caption[caption title]
+    {Outline of the disparity-through-time pipeline applied to each data set: 1) we used ancestral characters estimation to infer nodal character states, 2) we measured the pairwise Gower distance between the tips and nodal characters, 3) we ordinated the distance matrix using a PCoA, 4) we split that matrix using stratigraphy bins, equal bins and the time slice method, 5) we measured disparity through time for each of these methods.}
+    \label{overview}
+  \end{figure}
 
@@
 
-    \item The ``gradual splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with a probability equal to the distance between the slice.
+    \item The ``gradual splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with a probability function of the distance between the nodes/tips and the slice.
     % NC: I'm unsure what you mean here: "the distance between the slice." Do you mean the branch length, the interval between slices, or the distance between the slice and the node?
+    %TG: I've clarified as above. Tell me if it's clearer. In my words: the distance between the slice and the node, like you illustrate in your presentation (nicely done by the way!)
           \begin{equation}

### 2017-12-11 analyses/09-figure-peaks.Rmd
Figures with peaks of disparity

---
title: "Disparity peaks and troughs"
author: "Natalie Cooper"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 12
---

```{r, message = FALSE}
library(tidyverse)
library(gridExtra)
library(grid)
source("../functions/multiplot_fun.R")
```

# Beck 2014

```{r}
slug <- "Beck2014"
```

Load the data

```{r}
data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
data_max$bin_type <- gsub("stratigraphy", "Stratigraphy", data_max$bin_type) 
data_max$bin_type <- gsub("duration", "Duration", data_max$bin_type) 
data_max$bin_type <- gsub("number", "Number", data_max$bin_type) 
data_max$bin_type <- as.factor(data_max$bin_type)
```

# Modify the data
```{r}
# Get ages for the bins, go for the midpoints
x <- str_split(data_max$subsets, " - ") 
y <- map(x, as.numeric)
z1 <- map(y, max)
z2 <- map(y, min)

# Modify data - add midpoints, remove missing data, filter to just three models
data_max2 <- 
  data_max %>%
  mutate(time1 = unlist(z1)) %>%
  mutate(time2 = unlist(z2)) %>%
  filter(bs.median != 0 & !is.na(bs.median)) %>%
  mutate(bin_type = relevel(data_max$bin_type, ref = 'Stratigraphy')) %>%
  filter(model == "equalbins" | model == "proximity" | model == "gradual.split")
data_max2_age <- filter(data_max2, stratigraphy == "Age")

data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
```

```{r}
beck_peaks <- 
  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
 # expand_limits(x = 0, y = c(0,1)) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.background = element_blank())

beck_peaks_age <- 
  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
 # expand_limits(x = 0, y = c(0,1)) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.background = element_blank())
```

# Brusatte 2014

```{r}
slug <- "Brusatte2014"
```

Load the data

```{r}
data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
```
```{r}
# Get ages for the bins, go for the midpoints
x <- str_split(data_max$subsets, " - ") 
y <- map(x, as.numeric)
z1 <- map(y, max)
z2 <- map(y, min)

# Modify data - add midpoints, remove missing data, filter to just three models
data_max2 <- 
  data_max %>%
  mutate(time1 = unlist(z1)) %>%
  mutate(time2 = unlist(z2)) %>%
  filter(bs.median != 0 & !is.na(bs.median)) %>%
  mutate(bin_type = relevel(data_max$bin_type, ref = 'stratigraphy')) %>%
  filter(model == "equalbins" | model == "proximity" | model == "gradual.split")

data_max2_age <- filter(data_max2, stratigraphy == "Age")

data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
```

```{r}
brusatte_peaks <- 
  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
  expand_limits(y = 0) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none", 
       panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.text = element_blank())

brusatte_peaks_age <- 
  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
  expand_limits(y = 0) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none", 
       panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.text = element_blank()) 
```

# Bapst 2016

```{r}
slug <- "Bapst2016"
```

Load the data

```{r}
data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
```
```{r}
# Get ages for the bins, go for the midpoints
x <- str_split(data_max$subsets, " - ") 
y <- map(x, as.numeric)
z1 <- map(y, max)
z2 <- map(y, min)

# Modify data - add midpoints, remove missing data, filter to just three models
data_max2 <- 
  data_max %>%
  mutate(time1 = unlist(z1)) %>%
  mutate(time2 = unlist(z2)) %>%
  filter(bs.median != 0 & !is.na(bs.median)) %>%
  mutate(bin_type = relevel(data_max$bin_type, ref = 'stratigraphy')) %>%
  filter(model == "equalbins" | model == "proximity" | model == "gradual.split")

data_max2_age <- filter(data_max2, stratigraphy == "Age")

data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
```

```{r}
bapst_peaks <- 
  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
 # expand_limits(x = 0, y = c(0,1)) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
       panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.text = element_blank())

bapst_peaks_age <- 
  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
 # expand_limits(x = 0, y = c(0,1)) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
       panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.text = element_blank())
```

# Wright 2017

```{r}
slug <- "Wright2017"
```

Load the data

```{r}
data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
```
```{r}
# Get ages for the bins, go for the midpoints
x <- str_split(data_max$subsets, " - ") 
y <- map(x, as.numeric)
z1 <- map(y, max)
z2 <- map(y, min)

# Modify data - add midpoints, remove missing data, filter to just three models
data_max2 <- 
  data_max %>%
  mutate(time1 = unlist(z1)) %>%
  mutate(time2 = unlist(z2)) %>%
  filter(bs.median != 0 & !is.na(bs.median)) %>%
  mutate(bin_type = relevel(data_max$bin_type, ref = 'stratigraphy')) %>%
  filter(model == "equalbins" | model == "proximity" | model == "gradual.split")

data_max2_age <- filter(data_max2, stratigraphy == "Age")

data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
```

```{r}
wright_peaks <- 
  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
 expand_limits(y = 480) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
       panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.text = element_blank())

wright_peaks_age <- 
  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
 expand_limits(y = 480) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
       panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.text = element_blank()) 
```

Finally plot them all together

```{r}
# Using modified multiplot function
# written by Scott Chamberlain

pdf("../manuscript/figures/fig-peaks-epoch.pdf")  
  multiplot(beck_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
            brusatte_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
            bapst_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
            wright_peaks + theme(plot.margin=unit(c(5.5, 5.5, 20, 25),"points")),
            cols = 1,
            labs = list("Time at disparity peak (MYA)", 
                        "Time bins or time slice method"),
            labpos = list(c(0.5, 0.015), c(0.03, 0.5)))

dev.off() 
```

# And for ages

```{r}

pdf("../manuscript/figures/fig-peaks-age.pdf")  
  multiplot(beck_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
            brusatte_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
            bapst_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
            wright_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 20, 25),"points")),
            cols = 1,
            labs = list("Time at disparity peak (MYA)", 
                        "Time bins or time slice method"),
            labpos = list(c(0.5, 0.015), c(0.03, 0.5)))

dev.off()

```

# Repeat for all models for appendix

# Beck 2014

```{r}
slug <- "Beck2014"
```

Load the data

```{r}
data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
data_max$bin_type <- gsub("stratigraphy", "Stratigraphy", data_max$bin_type) 
data_max$bin_type <- gsub("duration", "Duration", data_max$bin_type) 
data_max$bin_type <- gsub("number", "Number", data_max$bin_type) 
data_max$bin_type <- as.factor(data_max$bin_type)
```

# Modify the data
```{r}
# Get ages for the bins, go for the midpoints
x <- str_split(data_max$subsets, " - ") 
y <- map(x, as.numeric)
z1 <- map(y, max)
z2 <- map(y, min)

# Modify data - add midpoints, remove missing data, filter to just three models
data_max2 <- 
  data_max %>%
  mutate(time1 = unlist(z1)) %>%
  mutate(time2 = unlist(z2)) %>%
  filter(bs.median != 0 & !is.na(bs.median)) %>%
  mutate(bin_type = relevel(data_max$bin_type, ref = 'Stratigraphy')) %>%
  mutate(model = relevel(data_max$model, ref = 'equalbins')) #%>%
  #filter(model == "equalbins" | model == "proximity" | model == "gradual.split")

data_max2_age <- filter(data_max2, stratigraphy == "Age")

data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
```

```{r}
beck_peaks <- 
  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
 # expand_limits(x = 0, y = c(0,1)) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
       axis.text.x = element_blank(),
        #legend.position = "none",
        panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.background = element_blank())

beck_peaks_age <- 
  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
 # expand_limits(x = 0, y = c(0,1)) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        #legend.position = "none",
        panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.background = element_blank())
```

# Brusatte 2014

```{r}
slug <- "Brusatte2014"
```

Load the data

```{r}
data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
```
```{r}
# Get ages for the bins, go for the midpoints
x <- str_split(data_max$subsets, " - ") 
y <- map(x, as.numeric)
z1 <- map(y, max)
z2 <- map(y, min)

# Modify data - add midpoints, remove missing data, filter to just three models
data_max2 <- 
  data_max %>%
  mutate(time1 = unlist(z1)) %>%
  mutate(time2 = unlist(z2)) %>%
  filter(bs.median != 0 & !is.na(bs.median)) %>%
  mutate(bin_type = relevel(data_max$bin_type, ref = 'stratigraphy')) %>%
  mutate(model = relevel(data_max$model, ref = 'equalbins'))#%>%
  #filter(model == "equalbins" | model == "proximity" | model == "gradual.split")

data_max2_age <- filter(data_max2, stratigraphy == "Age")

data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
```

```{r}
brusatte_peaks <- 
  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
  expand_limits(y = 0) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
       axis.text.x = element_blank(), 
       legend.position = "none", 
       panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.text = element_blank())

brusatte_peaks_age <- 
  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
  expand_limits(y = 0) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none", 
       axis.text.x = element_blank(),
       panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.text = element_blank()) 
```

# Bapst 2016

```{r}
slug <- "Bapst2016"
```

Load the data

```{r}
data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
```
```{r}
# Get ages for the bins, go for the midpoints
x <- str_split(data_max$subsets, " - ") 
y <- map(x, as.numeric)
z1 <- map(y, max)
z2 <- map(y, min)

# Modify data - add midpoints, remove missing data, filter to just three models
data_max2 <- 
  data_max %>%
  mutate(time1 = unlist(z1)) %>%
  mutate(time2 = unlist(z2)) %>%
  filter(bs.median != 0 & !is.na(bs.median)) %>%
  mutate(bin_type = relevel(data_max$bin_type, ref = 'stratigraphy')) %>%
  mutate(model = relevel(data_max$model, ref = 'equalbins'))#%>%
  #filter(model == "equalbins" | model == "proximity" | model == "gradual.split")

data_max2_age <- filter(data_max2, stratigraphy == "Age")

data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
```

```{r}
bapst_peaks <- 
  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
 # scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
 # expand_limits(x = 0, y = c(0,1)) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
       axis.text.x = element_blank(),
       panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.text = element_blank())

bapst_peaks_age <- 
  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
 # expand_limits(x = 0, y = c(0,1)) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
       axis.text.x = element_blank(),
       panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.text = element_blank())
```

# Wright 2017

```{r}
slug <- "Wright2017"
```

Load the data

```{r}
data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
```
```{r}
# Get ages for the bins, go for the midpoints
x <- str_split(data_max$subsets, " - ") 
y <- map(x, as.numeric)
z1 <- map(y, max)
z2 <- map(y, min)

# Modify data - add midpoints, remove missing data, filter to just three models
data_max2 <- 
  data_max %>%
  mutate(time1 = unlist(z1)) %>%
  mutate(time2 = unlist(z2)) %>%
  filter(bs.median != 0 & !is.na(bs.median)) %>%
  mutate(bin_type = relevel(data_max$bin_type, ref = 'stratigraphy')) %>%
  mutate(model = relevel(data_max$model, ref = 'equalbins'))#%>%
  #filter(model == "equalbins" | model == "proximity" | model == "gradual.split")

data_max2_age <- filter(data_max2, stratigraphy == "Age")

data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
```

```{r}
wright_peaks <- 
  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
 expand_limits(y = 480) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
       axis.text.x = element_blank(),
       panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.text = element_blank())

wright_peaks_age <- 
  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
  theme_classic(base_size = 12) +
  geom_point() +
  geom_point(aes(y = time2, x = model, colour = model)) +
  facet_wrap(~bin_type, ncol = 3) +
  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
 expand_limits(y = 480) +
 theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
       axis.text.x = element_blank(),
       panel.spacing = unit(0, "lines"),
        panel.border = element_blank(),
        strip.text = element_blank()) 
```

Finally plot them all together

```{r}
# Using modified multiplot function
# written by Scott Chamberlain

pdf("../manuscript/figures/fig-peaks-epoch-appendix.pdf")  
  multiplot(beck_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
            brusatte_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
            bapst_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
            wright_peaks + theme(plot.margin=unit(c(5.5, 5.5, 20, 25),"points")),
            cols = 1,
            labs = list("Time at disparity peak (MYA)", 
                        "Time bins or time slice method"),
            labpos = list(c(0.5, 0.015), c(0.03, 0.5)))

dev.off() 
```

# And for ages

```{r}

pdf("../manuscript/figures/fig-peaks-age-appendix.pdf")  
  multiplot(beck_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
            brusatte_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
            bapst_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
            wright_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 20, 25),"points")),
            cols = 1,
            labs = list("Time at disparity peak (MYA)", 
                        "Time bins or time slice method"),
            labpos = list(c(0.5, 0.015), c(0.03, 0.5)))

dev.off()

```

### 2017-12-11 analyses/09-figure-peaks.Rmd
Switch X and Y labels

@@
 ---
-title: "Disparity peaks and troughs"
+title: "Disparity peaks"
 author: "Natalie Cooper"
@@
             cols = 1,
-            labs = list("Time at disparity peak (MYA)", 
-                        "Time bins or time slice method"),
+            labs = list("Time bins or time slice method",
+                        "Time at disparity peak (MYA)"),
             labpos = list(c(0.5, 0.015), c(0.03, 0.5)))

### 2017-12-11 manuscript/time-slice.tex
Add peaks figures etc

@@
 
-https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/details-of-specific-functions.html#time-slicing
+%https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/details-of-specific-functions.html#time-slicing$
 
@@
 % TG: why that? It'll be just extra hassle no?
+% NC: Nah it's just a group and makes it clearer here that I'm actually doing some programming work. Takes about 30s to transfer and just means you/I need to change our remotes which again takes 30s.
 
@@
 
+  \begin{figure}[!htbp]
+    \centering
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch.pdf}
+    \caption[Timing of peak disparity for four example datasets.]
+    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
+    Pink points represent time binning methods, green points are time slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
+    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
+    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
+    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
+    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
+    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (Table \ref{table:datasets}).
+    Results for stratigraphic ages, and for other time-slicing models, are in the appendix.}
+    \label{figure:peak1}
+  \end{figure}
+
 \paragraph{Effects of mass extinction events.}
@@
 
+\begin{figure}[!htbp]
+    \centering
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch-appendix.pdf}
+    \caption[Timing of peak disparity for four example datasets.]
+    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
+    %Pink points represent time binning methods, green points are time slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
+    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
+    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
+    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
+    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
+    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
+    \label{figure:peak2}
+  \end{figure}
+
+\begin{figure}[!htbp]
+    \centering
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-age-appendix.pdf}
+    \caption[Timing of peak disparity for four example datasets.]
+    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
+    %Pink points represent time binning methods, green points are time slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
+    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages. 
+    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages in the time frame of the dataset. 
+    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages in the time frame of the dataset. 
+    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
+    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
+    \label{figure:peak3}
+  \end{figure}
+
 \bibliographystyle{palaeo} 

### 2017-12-11 analyses/10-figure-extinction.Rmd
extinction figure (with ggplot palette)

---
title: "Extinction effect"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Compiling the figures for the paper

Loading the code:

```{r}
library(ape)
library(dispRity)
source("../functions/figures_fun.R")

## ggplot colors
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
```

## Extinction effect figures

Initialising the slug and loading the data:

```{r, eval = FALSE}
## Slugs (choose one)
slug <- "Beck2014"
# slug <- "Brusatte2014"
# slug <- "Bapst2016"
# slug <- "Wright2017"

data <- read.csv(paste0("../outputs/extinction.", slug, ".csv"), stringsAsFactors = FALSE)
```

Single plots:

```{r, fig.width = 6, fig.height = 6, eval = FALSE}
colours <- c("grey", "white", "black") #Signif, non-signif, NA
plot.extinction(data, type = "age", xaxis = TRUE, yaxis = TRUE, xaxis2 = TRUE, main = slug, cols = colours)
```

Multiple plots:

```{r, fig.width = 12, fig.height = 4}

## Names of the datasets
slugs <- as.list(c("Beck2014", "Brusatte2014", "Bapst2016", "Wright2017"))

## Load all the data
data_list <- lapply(slugs, function(slug) read.csv(paste0("../outputs/extinction.", slug, ".csv"), stringsAsFactors = FALSE))

## Set up the colours (yeah, they're ugly for now)
colours <- c(gg_color_hue(2), "white") #Signif, non-signif, NA

## Plot all the results
pdf("../manuscript/figures/fig-extinction-effect.pdf", width = 12, height = 4)  
multi.plot.extinction(data_list, type = c("age", "epoch"), data.names = unlist(slugs), cols = colours)
dev.off()
```

### 2017-12-12 manuscript/appendix.tex
New appendix

--- 2017-12-10 manuscript/time-slice.tex
@@
 
-\title{Time for a rethink: disparity-through-time}
-%TG: Great title first part! How about changing the other part to 
-%Time for a rethink: different ways of slicing through the stratigraphic cake leads to different disparity through time results.
+\title{Supporting Information from `Time for a rethink: time sub-sampling methods in disparity-through-time analyses'}
 
@@
 }
-%TG: authorship-wise it's up to you, in the end, since it's just the two of us, I think it really doesn't matter (i.e.
-it's still gonna be Cooper Guillerme or Guillerme Cooper), I'll hopefully will have a couple of first author papers coming up for 2018 so up to you if you want one as well.
 
@@
 
-\runninghead{Time sub-samples in disparity-through-time analyses}
-\keywords{time bin, time slice, disparity} 
+%\runninghead{Time sub-samples in disparity-through-time analyses}
+%\keywords{time bin, time slice, disparity} 
 % up to 6
@@
 
-\section{Abstract}
+%\section{Abstract}
 % 300 words max
@@
 
-\section{Introduction}
-%TG: Plan idea
-%TG: §1: disparity-through-time really cool and well used
-%TG: §2: caveats with both disparity and through-time part
-%TG: §3: specific ways to look through-time and their problems
-%TG: §4: our kind of solution
-
-Disparity-through-time analyes are common in palaeontology \citep{Foote01071994}.
-They reveal how the morphological diversity of clades has changed through time, and allow us to make inferences about the breadth of ecological niches extinct taxa occupied \citep{foote1997evolution}.
-Results from disparity-through-time studies also provide insights into the ecological impacts of mass extinctions, competitive replacements, and the drivers of morphological evolution \citep{Brusatte12092008,Foote29111996,friedmanexplosive2010}.
-Unfortunately, the way we perform these analyses may have profound effects on our conclusions.
-
-% However, our methods for analysing diversity-through-time have not changed all that much in the XX years they have been popular.
-%TG: small caveat on that after litterature search for the disparity review paper: disparity-through-time analysis have not changed much since Foote 1992, Foote and Gould 1992 and Briggs et al.
-1992 in the classic palaeobiology appro,ach...
-%TG: However to not piss of people, there were several attempts to "rationalise" (Wills 2001), test (Ciampaglio 2001) and expand (Pavlinov 2011 and Gerber et al 2011) namely with univariate approaches and "dtt" analysis (Harmon et al 2003).
-%TG: One of the critisisms from one of the rude reviews was this idea that age of popular method has nothing to do with its performance, by nuancing this sentence we can basically say: yeah, some people have made great advances and proposed great ideas for CORRECTING ERRORS in the methods from the 90s but somehow everybody have ignored them (in palaeobiology).
-% NC: OK so lets save ourselves some pain and not mention this at all and save it all for the bitching in the review paper!!!
-
-Disparity-through-time analyes have two main analysis components: calculating disparity, and creating time sub-samples of the data. 
-Here we focus on the latter.
-The nature of disparity (it is a diversity metric), means it cannot be calculated using just one individual, so some way of subsetting taxa is required.
-Changes in disparity-through-time are generally investigated by calculating the disparity of taxa present during specific time intervals or time bins \citep[e.g][]{cisneros2010,prentice2011,Hughes20082013,hopkinsdecoupling2013,bentonmodels2014,bensonfaunal2014}.
-These time bins are usually defined based on stratigraphy \citep[e.g.][]{cisneros2010,prentice2011,Hughes20082013,bentonmodels2014} but can also be arbitrarily chosen time bins of equal duration \citep{Butler2012,hopkinsdecoupling2013,bensonfaunal2014}.
-However, this approach has several limitations.
-
-First, time bins defined by stratigraphy are not of equal size, biasing higher disparity towards longer stratigraphic periods. 
-This can be dealt with using rarefaction, i.e. repeating the analysis using the smallest number of taxa present in any bin as the number of sampled taxa in each bin. 
-This can, however, lead to very large confidence intervals when there are stratigraphic periods with very few species.
-Other studies split large time bins so they are of roughly equal size, but this is often an \textit{ad hoc} procedure that can introduce more bias depending on where bins are split.
-Second, all time binning approaches (whether bins are equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring \textit{between} intervals, rather than also allowing for gradual changes \textit{within} intervals \citep{Hunt21042015}.
-Finally, time bin analyses are often limited by the number of taxa in each bin.
-If there are insufficient taxa in a time bin, disparity cannot be calculated, so further analyses, e.g.
-correlations of disparity with hypothesised drivers of morphological evolution, are not possible.
-
-To address these issues, we propose a ``time-slicing'' approach that takes advantage of the wealth of paleontological datasets which now have associated phylogenies. 
-Time-slicing uses a phylogenetic tree and considers subsets of taxa at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
-This results in even-sampling across time and permits us to define the underlying model of character evolution (punctuated or gradual).  
-Time-slicing also includes any element present in the phylogeny (branches, nodes and tips) at the time slice in question as part of the disparity calculation.
-This allows us to measure disparity at time points where there are no terminal taxa, and increases the sample size at each time point, making downstream analyses of the drivers of disparity much more feasible.
-
-Here we present our time-slicing methods using four datasets taken from the literature.
-We calculate disparity-through-time for each dataset using a range of time binning and time-slicing methods, and then compare these approaches with respect to the relative disparities calculated, but also with how the different approaches influence biological conclusions.
-We find that XXX
-
-\section{Materials and Methods}
-\subsection{Overview}
-\label{overview-section}
-To test the different time sub-sampling methods, we followed the protocol below (Figure \ref{datasets}). 
-All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (https://github.com/nhcooper123/time-slice).
-
-https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/details-of-specific-functions.html#time-slicing
-
-% NC: might move this to the NHM github account
-
-% Add overview figure
-
-  %\begin{figure}[!htbp]
-    %\centering
-    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/overview}
-    %\caption[caption title]
-    % {caption text}
-    %\label{overview}
-  %\end{figure}
-
-\subsection{Example datasets}
-\label{datasets}
-To test the different time binning/slicing methods we selected four datasets: a mammal dataset from \cite{beckancient2014}, two theropod datasets from \cite{brusatte2014gradual} and \cite{bapst2016topology}, and a crinoid dataset from \cite{wright2017bayesian}.
-Table \ref{table:datasets} and appendix ??? provide more details. % add appendix details
-Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
-These datasets are freely available with their accompanying papers (Table \ref{table:datasets}), but for reproducibility purposes we also provide the data we used on GitHub (https://github.com/nhcooper123/time-slice/data).
-
-% Dataset details table
-\begin{landscape}
-  \input{tables/table_datasets.tex} 
-  \label{table:datasets}  
-\end{landscape}
-
-\subsection{Preparing the data for disparity-through-time analysis}
-
-\paragraph{Estimating ancestral character states.}
-\label{ace}
-For each dataset we used the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny, using the \texttt{rerootingMethod} function from the \texttt{R} package \texttt{phytools} version 0.6-44 \citep{phytools,R}.
-Where there were missing character data for a taxon we followed the method of \cite{Claddis} and treated missing data as any possible observed state for each character.
-For example, if a character had two observed states ($0$ and $1$) across all taxa, we attributed the multi-state ``$0$\&$1$" value to the taxon with missing data, representing an equal probability of being either $0$ or $1$.
-This allows the ancestral node of a taxon with missing data to be estimated with no assumptions other than that the taxon has one of the observed character states.
-To prevent poor ancestral state reconstructions from biasing our results, especially when a lot of error is associated with the reconstruction, we only included ancestral state reconstructions with a scaled Likelihood $\geq$ $0.95$.
-Ancestral state reconstructions with scaled Likelihoods below this threshold were recoded as missing data (``?'').
-
-\paragraph{Building cladisto-spaces.} 
-To explore disparity-through-time in our datasets we used a cladisto-space approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
-This approach is similar to constructing a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010}, except a cladisto-space is an approximation of the morphospace based on cladistic data (i.e.
-the discrete morphological characters used to build a phylogenetic tree).
-Mathematically, a cladisto-space is an $n$ dimensional object that summarises the cladistic distances between the taxa present in a cladistic matrix.
-Although empirically inter-taxon distances are the same in a morphospace or a cladisto-space \citep{foth2012different,hetherington2015cladistic}, we prefer the term cladisto-space to make it clear that this space is estimated using cladistic data and not morphometric data.
-
-\paragraph{Constructing distance matrices.}
-To estimate the cladisto-spaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
-We calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the Euclidean distance between two taxa divided by the number of shared characters. 
-This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e.
-because some pairs of taxa share more characters in common than others, they are more likely to be similar).
-For cladistic matrices, using this corrected distance is preferable to the raw Euclidean distance because of its ability to deal with discrete or/and ordinated characters as well as with missing data \citep{anderson2012using}.
-However, the Gower distance cannot calculate distances when taxa have no overlapping data.
-Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis} to remove pairs of taxa with no cladistic characters in common.
-This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see appendix for details of which species).
-% add appendix details
-
-\paragraph{Ordination.}
-After constructing our distance matrices we transformed them using classical multidimensional scaling \citep[MDS;][]{torgerson1965multidimensional,GOWER01121966,cailliez1983analytical}.
-This method (also referred to as PCO; e.g. \citealt{Brusatte2015}; or PCoA; e.g. \citealt{paradisape:2004}) is an eigen decomposition of the distance matrix.
-Because we used Gower distances instead of raw Euclidean distances, negative eigenvalues can be calculated.
-To avoid this problem, we first transformed the distance matrices by applying the Cailliez correction \citep{cailliez1983analytical} which adds a constant $c^*$ to the values in a distance matrix (apart from the diagonal) so that all the Gower distances become Euclidean ($d_{Gower}+c^*=d_{Euclidean}$; \citealt{cailliez1983analytical}).
-We were then able to extract $n$ eigenvectors for each matrix (representing the $n$ dimensions of the cladisto-space) where $n$ is equal to $k-2$, i.e.
-the number of taxa in the matrix ($k$) minus the last two eigenvectors that are always null after applying the Cailliez correction.
-Contrary to previous studies \citep[e.g][]{brusatte50,cisneros2010,prentice2011,anderson2012using,Hughes20082013,bentonmodels2014}, we use all $n$ dimensions of our cladisto-spaces and not a sub-sample representing the majority of the variance in the distance matrix (e.g.
-selecting only $m$ dimensions that represent up to 90\% of the variance in the distance matrix; \citealt{Brusatte12092008,toljagictriassic-jurassic2013}).
-
-Note that our cladisto-spaces represent an ordination of all possible morphologies coded in each study through time.
-It is unlikely that all morphologies will co-occur at each time point, therefore, the disparity of the whole cladisto-space is expected to be greater than the disparity at any specific point in time.
-
-\subsection{Disparity-through-time analyses}
-
-\paragraph{Calculating disparity.}
-\label{disparity_calc}
-Disparity can be calculated in many different ways \citep[e.g.][]{Wills1994,Ciampaglio2004,thorneresetting2011,hopkinsdecoupling2013,huang2015origins}, however most studies estimate disparity using four metrics: the sum and products of ranges and variances, each of which gives a slightly different estimate of how the data fits within the cladisto-space \citep{Foote01071994,Wills1994,brusatte50,Brusatte12092008,cisneros2010,thorneresetting2011,prentice2011,brusattedinosaur2012,toljagictriassic-jurassic2013,ruta2013,bentonmodels2014,bensonfaunal2014}.
-However, these metrics have limitations. 
-First, the range metrics are affected by the uneven sampling of the fossil record \citep{Butler2012}.
-Second, because we include all $n$ dimensions in the analysis (see above), the products of ranges and variances will tend towards zero since the scores of the last dimension are usually really close to zero themselves.
-We therefore use the sum of variances metric to estimate disparity here.
-Note that there are still statistical issues with this metric, but for the purposes of comparison with previous work we want to use a standard metric for these analyses.
-
-\paragraph{Time sub-sampling} 
-\label{time_sub-samples}
-
-To estimate disparity-through-time we first need to split the data into time sub-samples.
-Here we use three time sub-sampling methods.
-
-\begin{enumerate}
-  \item Stratigraphic time bins.
-This is the traditional method, where all the taxa within each stratigraphic period are included in the disparity calculation.
-This often leads to bins of unequal duration.
-Here we use stratigraphic ages and epochs.
-  \item Equally sized time bins.
-This is another commonly used method, where the time frame of interest is split into equally sized time bins, then all the taxa within each time bin are included in the disparity calculation.
-  \item Time-slicing.
-We describe this in more detail below, but in brief, time-slicing uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation.
-\end{enumerate}
-
-\paragraph{Time-slicing} 
-\label{time_slicing}
-The ``time-slicing'' approach considers subsets of taxa in the cladisto-space at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
-This results in even-sampling of the cladisto-space across time and allows us to use different underlying models of character evolution (punctuated or gradual).
-
-In practice, time-slicing considers the disparity of any element present in the phylogeny (branches, nodes and tips) at any point in time.
-When the phylogenetic elements are nodes or tips, the ordination scores for the nodes (estimated using ancestral state reconstruction as described above) or tips are directly used for calculating disparity.
-When the phylogenetic elements are branches we choose the ordination score for the branch using one of two evolutionary models:
-
-%TG: Not sure if it's the "egeinvector" score or the "eigen" score.
-I think there's a trick in naming them depending on how the algorithm does the transformation (soemthing with using eigenvalues or eigenvectors).
-I call it ordination score for now.
-
-\begin{enumerate}
-
-    \item{\textbf{Punctuated evolution.}} 
-    This model selects the ordination score from either the ancestral node or the descendant node/tip of the branch regardless of the position of the slice along the branch. 
-    Similarly to the time bin approach, this reflects a model of punctuated evolution where changes in disparity occur either at the start or at the end of a branch over a relatively short time period, and clades undergo long periods of stasis during their evolution \citep{Gould1977,Hunt20112007}.
-    We apply this model in four ways:
-
-    \begin{enumerate}[(i)]
-
-      \item The ``acctran'' model, always selecting the ordination score of the ancestral node of the branch.
-      \item The ``deltran'' model, always selecting the ordination score of the descendant node/tip of the branch.
-      \item The ``random'' model, randomly selecting the ordination score of either the ancestor or the descendant of the branch.
-      \item The ``proximity'' model, selecting the ordination score of the ancestor if the slice occurs in the first half of the branch, and the descendant if the slice occurs in the second half of the branch.
-
-    \end{enumerate}
-
-    The two first models assume that changes always occur early (ACCelerated TRANsition) or late along the branches (DELayed TRANsition).
-    The third model makes neither assumption and simply selects data from the ancestor or the descendant at random, and the fourth bases the selection of either the ancestor or the descendant on where the slice occurs along the branch.
-    The punctuated models only select either the ordination score from the ancestor and the descendant once in the whole disparity analysis.
-    For example, if using the ``random'' model, if the data of the ancestor has been randomly chosen, only this data will be used during the bootstrapping (see below) and for the disparity calculation.
-
-    \item{\textbf{Gradual evolution.}}
-    Unlike the punctuated models, the following models do not select the ordination score of either the ancestor or the descendant but associate a probability to both.
-    This reflects a model of gradual evolution where changes in disparity are gradual and cumulative along the branch.
-
-    \begin{enumerate}[(i)]
-    \setcounter{enumii}{4}
-      \item The ``equal splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with an equal probability
-          \begin{equation}
-          p(\text{ancestor}) = p(\text{descendant}) = 0.5
-          \end{equation}
-
-    \item The ``gradual splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with a probability equal to the distance between the slice.
-    % NC: I'm unsure what you mean here: "the distance between the slice." Do you mean the branch length, the interval between slices, or the distance between the slice and the node?
-          \begin{equation}
-              p(\text{ancestor}) = \frac{d(\text{ancestor},\text{slice})}{d(\text{ancestor},\text{descendant})}
-          \end{equation}
-          \begin{equation}
-              p(\text{descendant}) = 1 - p(\text{ancestor})
-          \end{equation}
-    \noindent Where $d(x,y)$ is the distance measured as branch length.
-    \end{enumerate}
-
-    In these models, the ordination scores of both the ancestor and descendant contribute to the disparity calculation.
-    For example, using the ``gradual.split'' model, if the slice occurs in the third quarter of a branch joining node A to node/tip B (75\% of the total branch length), after bootstrapping, the disparity results will be based on 25\% of the data from A and 75\% of the data from B.
-    Because of the probabilistic nature of these models, they are only meaningful when calculating disparity from bootstrapped datasets.
-\end{enumerate}
-
-%TG: Justification on the fact that these models are not ACEs.
-It is important to note that the time-slicing method is not an ancestral states estimation method \textit{per se}.
-This method does not estimate values along a branch applying a model (\textit{c.f.} methods described for reconstructing ancestral states in the preparing data for disparity-through-time section above) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the models described above.
-This allows the method to be used in post-ordination analysis where the data used in each time slice is data already present in the cladisto-space.
-In other words, this method does not require a re-ordination of the cladisto-space every time a slice goes through a branch, thus allowing the properties of the cladisto-space (e.g.
-distance between species, variance per axis, etc.) to remain constant.
-
-\subsection{Comparing time sub-sampling methods}
-
-To compare the time binning and time-slicing approaches we applied the methods as follows.
-
-\begin{enumerate}
-  \item Stratigraphy: time sub-samples defined by stratigraphic periods. 
-  \begin{enumerate}[(i)]
-    \item Time bins (unequal). 
-    We calculated disparity for the taxa in each stratigraphic period (age or epoch). 
-    To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time bin by randomly resampling with replacement a new sub-sample of taxa from the observed taxa in the bin 1000 times.
-    We then calculated the median disparity value for each time bin along with the 50\% and 95\% confidence intervals.
-    \item Time-slices (unequal).
-    We calculated disparity using our time-slicing approach with slices occurring at the midpoint of each stratigraphic period (age or epoch), and using all six time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits).
-    To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement as described above for the stratigraphic time bins.
-    \end{enumerate}
-
-  \item Duration: time sub-samples defined by the duration of stratigraphic periods. 
-  \begin{enumerate}[(i)]
-    \item Time bins (equal). 
-    We calculated disparity for the taxa in each time bin where time bin size was defined by the average \textit{duration} of the stratigraphic period (age or epoch), and bootstrapped the disparity values as described above.
-    \item Time-slices (unequal).
-    We calculated disparity using our time-slicing approach where the interval between slices, was defined by the average \textit{duration} of the stratigraphic period (age or epoch).
-    We used the six time-slicing methods and bootstrapped as described above.
-  \end{enumerate}
-
-  \item Number: time sub-samples defined by the number of stratigraphic periods. 
-  \begin{enumerate}[(i)]
-    \item Time bins (equal). 
-    We calculated disparity for the taxa in each time bin where the number of time bins was defined by the average \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest, and bootstrapped the disparity values as described above.
-    \item Time-slices (unequal).
-    We calculated disparity using our time-slicing approach where the number of slices, was defined by the average \textit{number} of the stratigraphic periods (ages or epochs) in the time frame of interest..
-    We used the six time-slicing methods and bootstrapped as described above.
-  \end{enumerate}
-
-\end{enumerate}
-
-We also recorded the number of taxa (or taxa and nodes for time-slicing methods) in each sub-sample as a proxy for taxonomic diversity (Table \ref{tab:sub-samples}).
-
-\subsection{Testing for differences in the time sub-sampling methods}
-\label{testing}
-Testing for statistical differences among the time sub-sampling methods described above is a little tricky, as we need to compare similar units, and also to tackle questions important to the interpretation of disparity-through-time analyses.
-We therefore present three different, simple ways of comparing the time sub-sampling methods as follows.
-
-\paragraph{Systematic differences in disparity-through-time.} 
-To test whether using time bins or time slices resulted in significantly different disparity values at common time points, we used paired Wilcoxon tests to compare the median bootstrapped disparities obtained in the stratigraphy (time sub-samples defined by stratigraphic periods), duration (time sub-samples defined by the duration of stratigraphic periods), and number (time sub-samples defined by the number of stratigraphic periods) analyses described above.
-
-Due to the uneven spread of taxa across phylogenies, some time bins will contain one or no species, meaning that we cannot estimate disparity for that time bin. 
-We first, therefore, remove the time bins, and corresponding time slices, without disparity estimates. 
-We then perform paired Wilcoxon tests, so that bins and slices for the same time period are being compared. 
-Significant results suggest that there is a systematic difference in disparity values at each time point, depending on whether bins or slices are used.
-
-\paragraph{Disparity peaks and troughs.}
-We are perhaps more interested in how the conclusions disparity-through-time analyses are influenced by the choice of time sub-sampling method, rather than the disparities estimated by each method \textit{per se}, especially as these will influenced by the number of taxa (and/or nodes) included in each sub-sample.
-Therefore, we also investigate where peaks and troughs of disparity occur in each of our datasets for the different time sub-sampling methods. 
-We calculated the maximum and minimum bootstrapped disparities for each dataset and for each time sub-sampling method, along with their associated confidence intervals.
-Significant shifts in disparity peaks or troughs suggest that the choice of time sub-sampling method will influence our conclusions about relative changes in the disparity of our example clades through time.
-
-\paragraph{Effects of mass extinction events.}
-Many analyses of disparity-through-time aim to demonstrate differences in disparity before and after mass extinction events [citations]. 
-Each of our four datasets span one mass extinction (Cretaceous-Paleogene 66 MYA; \citealt{brusatte2014gradual,bapst2016topology,beckancient2014}; Ordovician-Silurian 455– 430 MYA; \citealt{wright2017bayesian}), so we use Wilcoxon tests to compare disparity in the time bin/slice prior to the approrpiate mass extinction, to that of the time bin/slice following the extinction event.
-Significant results suggest a significant effect of the mass extinction on disparity in the group.
-We then compare these results across the time sub-sampling methods to determine if our conclusions would change depending on the method used.
-We repeated these analyses using the two time bins/slices after the one immediately following the mass extinction event to account for any lag effects of the mass extinction on disparity.
-
-\section{Results}
-
-\subsection{Disparity-through-time analyses}
-
-Disparity changes through time for each of the four datasets (Figure \ref{figure:dtt}, Figure in appendix, full tables in the appendix). 
-Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods.
-The six different time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits) show similar patterns, so we focus only of the results for one method with a punctuated model of evolution (specifically the `proximity' method), and one method with a gradual model of evolution (specifically the `gradual splits' method).
-Results for all six methods can be found in the appendix. % add details of appendix
-
-  \begin{figure}[!htbp]
-    \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch.pdf}
-    \caption[Relative disparity through time for four example datasets.]
-    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Pink points represent time binning methods, blue points are time slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
-    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods.
-    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
-    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
-    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
-    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
-    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).
-    Results for stratigraphic ages, and for other time-slicing models, are in the appendix.}
-    \label{figure:dtt1}
-  \end{figure}
-
-\subsection{Testing for differences in the time sub-sampling methods}
-
-\paragraph{Systematic differences in disparity-through-time.} 
-There is no overall significant systematic difference among the disparities calculated using time bins and those calculated using the time-slicing methods.
-(Table \ref{table:wilcox}, Appendix Table \ref{table:wilcox2}).
-Instead, the differences depend on the dataset and method in question.
-For example, the \cite{brusatte2014gradual}, \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, show significant differences when using bins versus time slices defined by stratigraphy, but the \cite{beckancient2014} dataset appears robust to these different models.
-Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} show different disparities when the number of bins or slices is the average number of stratigraphic periods, but this is not see in the \cite{wright2017bayesian} dataset.
-Note that for epochs, we find fewer significant differences simply because the smaller numbers of bins and slices being compared means we have low power to detect a significant difference.
-
-% Wilcoxon results table
-%\begin{landscape}
-  \input{tables/table_wilcox.tex} 
-  \label{table:wilcox}  
-%\end{landscape}
-
-\paragraph{Disparity peaks and troughs.}
-% Plots showing the differences
-
-\paragraph{Effects of mass extinction events.}
-% Big heat map table thing
-
-%Fig. or Figs
-%Fig width 80mm single column
-%110mm - 2/3 width
-%166mm full page
-
-\section{Discussion}
-
-Benefits of time slices
-
-Further stuff to fix?
-
-\section{Conclusions}
-
-\section{Acknowledgments}
-	NC thanks Mark Sutton and Phillip Mannion for the invitation to contribute to the `Evolutionary Modelling' symposium at The Palaeontological Association Annual Meeting 2017.
-  TG acknowledges support from the Australian Discovery Project Grant number DP170103227 awarded to Vera Weisbecker.
-  We thank Dave Bapst, Steve Brusatte, Graeme Lloyd, April Wright and Davey Wright for assistance in gathering data for the analyses.
-
-\section{APPENDIX}
+\section{Additional tables and figures}
 

### 2017-12-12 manuscript/time-slice.tex
extinction figure (with ggplot palette)
New appendix

@@
 
-\title{Time for a rethink: disparity-through-time}
+\title{Time for a rethink: time sub-sampling methods in disparity-through-time analyses}
 %TG: Great title first part! How about changing the other part to 
@@
 
-% However, our methods for analysing diversity-through-time have not changed all that much in the XX years they have been popular.
-%TG: small caveat on that after litterature search for the disparity review paper: disparity-through-time analysis have not changed much since Foote 1992, Foote and Gould 1992 and Briggs et al.
-1992 in the classic palaeobiology appro,ach...
-%TG: However to not piss of people, there were several attempts to "rationalise" (Wills 2001), test (Ciampaglio 2001) and expand (Pavlinov 2011 and Gerber et al 2011) namely with univariate approaches and "dtt" analysis (Harmon et al 2003).
-%TG: One of the critisisms from one of the rude reviews was this idea that age of popular method has nothing to do with its performance, by nuancing this sentence we can basically say: yeah, some people have made great advances and proposed great ideas for CORRECTING ERRORS in the methods from the 90s but somehow everybody have ignored them (in palaeobiology).
-% NC: OK so lets save ourselves some pain and not mention this at all and save it all for the bitching in the review paper!!!
-% TG: agree!
-
 Disparity-through-time analyes have two main analysis components: calculating disparity, and creating time sub-samples of the data. 
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{outlinesvg.pdf}
-    \caption[caption title]
-    {Outline of the disparity-through-time pipeline applied to each data set: 1) we used ancestral characters estimation to infer nodal character states, 2) we measured the pairwise Gower distance between the tips and nodal characters, 3) we ordinated the distance matrix using a PCoA, 4) we split that matrix using stratigraphy bins, equal bins and the time slice method, 5) we measured disparity through time for each of these methods.}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/outlinesvg.pdf}
+    \caption[Outline of the disparity-through-time pipeline]
+    {1) We use ancestral character estimation to infer nodal character states; 2) we measure the pairwise Gower distance between the tip character states and nodal character states; 3) we ordinate the distance matrix using principal coordinates analysis (PCoA); 4) we time sub-sample the PCoA matrix using time bins defined by stratigraphic periods, equally sized time bins and time slices (using six methods to estimate ordination scores for branches); and finally 5) we measure disparity-through-time for each of these methods.}
     \label{overview}
@@
 
-%TG: Not sure if it's the "egeinvector" score or the "eigen" score.
-I think there's a trick in naming them depending on how the algorithm does the transformation (soemthing with using eigenvalues or eigenvectors).
-I call it ordination score for now.
-
 \begin{enumerate}
@@
 
-    \item The ``gradual splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with a probability function of the distance between the nodes/tips and the slice.
-    % NC: I'm unsure what you mean here: "the distance between the slice." Do you mean the branch length, the interval between slices, or the distance between the slice and the node?
-    %TG: I've clarified as above. Tell me if it's clearer. In my words: the distance between the slice and the node, like you illustrate in your presentation (nicely done by the way!)
+    \item The ``gradual splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with a probability function of the distance between the nodes/tip at the ends of the branch and the slice.
+
           \begin{equation}
@@
 
-%TG: Justification on the fact that these models are not ACEs.
 It is important to note that the time-slicing method is not an ancestral states estimation method \textit{per se}.
@@
 
-\paragraph{Disparity peaks and troughs.}
-We are perhaps more interested in how the conclusions disparity-through-time analyses are influenced by the choice of time sub-sampling method, rather than the disparities estimated by each method \textit{per se}, especially as these will influenced by the number of taxa (and/or nodes) included in each sub-sample.
+\paragraph{Disparity peaks.}
+We are perhaps more interested in how the conclusions disparity-through-time analyses are influenced by the choice of time sub-sampling method, rather than the disparities estimated by each method \textit{per se}, especially as these will be influenced by the number of taxa (and/or nodes) included in each sub-sample.
 Therefore, we also investigate where peaks and troughs of disparity occur in each of our datasets for the different time sub-sampling methods. 
-We calculated the maximum and minimum bootstrapped disparities for each dataset and for each time sub-sampling method, along with their associated confidence intervals.
-Significant shifts in disparity peaks or troughs suggest that the choice of time sub-sampling method will influence our conclusions about relative changes in the disparity of our example clades through time.
+We calculated the maximum bootstrapped disparities for each dataset and for each time sub-sampling method, along with their associated confidence intervals.
+Significant shifts in disparity peaks suggest that the choice of time sub-sampling method will influence our conclusions about relative changes in the disparity of our example clades through time.
 
@@
 
-Disparity changes through time for each of the four datasets (Figure \ref{figure:dtt}, Figure in appendix, full tables in the appendix). 
+Disparity changes through time for each of the four datasets (Figure \ref{figure:dtt}, Figure in appendix, full tables in the appendix). % add details of appendix
 Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods.
@@
 For example, the \cite{brusatte2014gradual}, \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, show significant differences when using bins versus time slices defined by stratigraphy, but the \cite{beckancient2014} dataset appears robust to these different models.
-Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} show different disparities when the number of bins or slices is the average number of stratigraphic periods, but this is not see in the \cite{wright2017bayesian} dataset.
+Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} show different disparities when the number of bins or slices is the average number of stratigraphic periods, but this is not seen in the \cite{wright2017bayesian} dataset.
 Note that for epochs, we find fewer significant differences simply because the smaller numbers of bins and slices being compared means we have low power to detect a significant difference.
@@
 
-\paragraph{Disparity peaks and troughs.}
+\paragraph{Disparity peaks}
+In the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets, disparity peaks occur much earlier when time slicing rather than time binning approaches are used (Figure \ref{figure:peak1}, Appendix figure).
+% appendix details
+This is also true for straigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, althought when using equal time bins the peaks are later than the time slicing methods, or very similar (Figure \ref{figure:peak1}, Appendix figure).
+Across the three time binning methods, the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets have similar disparity peaks whichever method is used, whereas in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Figure \ref{figure:peak1}, Appendix figure).
+
 % Plots showing the differences
-
   \begin{figure}[!htbp]
@@
 \paragraph{Effects of mass extinction events.}
+Mass extinction events influence disparity in all four datasets (Figure \ref{figure:extinction1}, appendix). 
+However, whether this change in disparity is significant or not depends on the dataset, and on the method used to create time sub-samples (Figure \ref{figure:extinction1}, appendix). % appendix!!!
+
 % Big heat map table thing
+  \begin{figure}[!htbp]
+    \centering
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch.pdf}
+    \caption[Extinction four example datasets.]
+    {XXX}
+    \label{figure:extinction1}
+  \end{figure}
 

### 2017-12-12 analyses/05-figure-dtt.Rmd
Reorder analysis files

--- 2017-12-08 analyses/04-figure-dtt.Rmd
@@
 
-# Modify data - add midpoints, remove missing data, filter to just two models,
+# Modify data - add midpoints, remove missing data, filter to just three models,
 # make disparity relative by dividing by max values
@@
 
-# Modify data - add midpoints, remove missing data, filter to just two models
+# Modify data - add midpoints, remove missing data, filter to just three models
 strat2 <- 

### 2017-12-12 analyses/figure-dtt-appendix.Rmd
Reorder analysis files

--- 2017-12-07 analyses/04-figures.Rmd
@@
 ---
-title: "Disparity through time - simplified and for appendix"
+title: "Disparity through time - for appendix"
 author: "Natalie Cooper"
@@
 
+```{r}
+slug <- "Beck2014"
+```
+
+Load the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
+
+strat <- out1$results$stratigraphy
+duration <- out1$results$duration
+number <- out1$results$number
+```
+
+Get the data into the correct format to plot
+
+```{r}
+# Stratigraphy
+# Get ages for the bins, go for the midpoints
+x <- str_split(strat$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+# Modify data - add midpoints, remove missing data, filter to just two models,
+# make disparity relative by dividing by max values
+strat2 <- 
+  strat %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Duration
+x <- str_split(duration$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+duration2 <- 
+  duration %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Number
+x <- str_split(number$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+number2 <- 
+  number %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+```
+
+Make plots in preparation for larger grid plot below
+
+```{r}
+
+beck_base <- 
+  ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
+  theme_classic(base_size = 12) + 
+  scale_y_continuous(breaks = seq(0, 1, 0.5), 
+                     labels = seq(0, 1, 0.5)) +
+  scale_x_reverse(limits = c(100, 0),
+                  breaks = c(seq(100, 0, by = -50))) +
+  expand_limits(x = 0, y = c(0,1)) +
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        legend.position = "none")
+
+beck_s <- 
+  beck_base +
+  geom_point(data = strat2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  ggtitle("Stratigraphy") +
+  theme(plot.title = element_text(size = 14, hjust = 0.5)) +
+  annotate("text", x = 75, y = 0.15, label = "Beck2014")
+
+beck_d <- 
+  beck_base +
+  geom_point(data = duration2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  ggtitle("Duration") +
+  theme(plot.title = element_text(size = 14, hjust = 0.5))
+
+beck_n <- 
+  beck_base +
+  geom_point(data = number2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  ggtitle("Number") +
+  theme(plot.title = element_text(size = 14, hjust = 0.5))
+
+```
+
+Pick the slug
+
+```{r}
+slug <- "Brusatte2014"
+```
+
+Load the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
+
+strat <- out1$results$stratigraphy
+duration <- out1$results$duration
+number <- out1$results$number
+```
+
+Get the data into the correct format to plot
+
+```{r}
+# Stratigraphy
+# Get ages for the bins, go for the midpoints
+x <- str_split(strat$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+# Modify data - add midpoints, remove missing data, filter to just two models
+strat2 <- 
+  strat %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Duration
+x <- str_split(duration$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+duration2 <- 
+  duration %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Number
+x <- str_split(number$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+number2 <- 
+  number %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+```
+
+Make plots in preparation for larger grid plot below
+
+```{r}
+
+brusatte_base <- 
+  ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
+  theme_classic(base_size = 12) + 
+  scale_y_continuous(breaks = seq(0, 1, 0.5), 
+                     labels = seq(0, 1, 0.5)) +
+  scale_x_reverse(limits = c(100, 0),
+                  breaks = c(seq(100, 0, by = -50))) +
+  expand_limits(x = 0, y = c(0,1)) +
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        legend.position = "none")
+
+brusatte_s <- 
+  brusatte_base +
+  geom_point(data = strat2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  annotate("text", x = 70, y = 0.15, label = "Brusatte2014")
+
+brusatte_d <- 
+  brusatte_base +
+  geom_point(data = duration2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+brusatte_n <- 
+  brusatte_base +
+  geom_point(data = number2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+```
+
+```{r}
+slug <- "Bapst2016"
+```
+
+Load the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
+
+strat <- out1$results$stratigraphy
+duration <- out1$results$duration
+number <- out1$results$number
+```
+
+Get the data into the correct format to plot
+
+```{r}
+# Stratigraphy
+# Get ages for the bins, go for the midpoints
+x <- str_split(strat$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+# Modify data - add midpoints, remove missing data, filter to just two models
+strat2 <- 
+  strat %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Duration
+x <- str_split(duration$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+duration2 <- 
+  duration %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Number
+x <- str_split(number$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+number2 <- 
+  number %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+```
+
+Make plots in preparation for larger grid plot below
+
+```{r}
+bapst_base <- 
+  ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
+  theme_classic(base_size = 12) + 
+  scale_y_continuous(breaks = seq(0, 1, 0.5), 
+                     labels = seq(0, 1, 0.5)) +
+  scale_x_reverse(limits = c(100, 0),
+                  breaks = c(seq(100, 0, by = -50))) +
+  expand_limits(x = 0, y = c(0,1)) +
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        legend.position = "none")
+
+bapst_s <- 
+  bapst_base +
+  geom_point(data = strat2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  annotate("text", x = 75, y = 0.15, label = "Bapst2016")
+
+bapst_d <- 
+  bapst_base +
+  geom_point(data = duration2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+bapst_n <- 
+  bapst_base +
+  geom_point(data = number2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+```
+
+```{r}
+slug <- "Wright2017"
+```
+
+Load the data
+
+```{r}
+out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
+
+strat <- out1$results$stratigraphy
+duration <- out1$results$duration
+number <- out1$results$number
+```
+
+Get the data into the correct format to plot
+
+```{r}
+# Stratigraphy
+# Get ages for the bins, go for the midpoints
+x <- str_split(strat$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+# Modify data - add midpoints, remove missing data, filter to just two models
+strat2 <- 
+  strat %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Duration
+x <- str_split(duration$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+duration2 <- 
+  duration %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+
+# Number
+x <- str_split(number$subset, " - ") 
+y <- map(x, as.numeric)
+z <- map(y, mean)
+
+number2 <- 
+  number %>%
+  mutate(time = unlist(z)) %>%
+  filter(bs.median != 0 & !is.na(bs.median)) %>%
+
+  mutate(relative.disparity = bs.median/max(bs.median)) %>%
+  mutate(relative.lwr = `2.5%`/max(bs.median)) %>%
+  mutate(relative.upr = `97.5%`/max(bs.median))
+```
+
+Make plots in preparation for larger grid plot below
+
+```{r}
+
+wright_base <- 
+  ggplot(strat2, aes(y = relative.disparity, x = time, colour = model)) +
+  theme_classic(base_size = 12) + 
+  scale_y_continuous(breaks = seq(0, 1, 0.5), 
+                     labels = seq(0, 1, 0.5)) +
+  scale_x_reverse(limits = c(470, 370),
+                  breaks = c(seq(470, 370, by = -50))) +
+  expand_limits(x = 0, y = c(0,1)) +
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        legend.position = "none")
+
+wright_s <- 
+  wright_base +
+  geom_point(data = strat2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  annotate("text", x = 440, y = 0.15, label = "Wright2017")
+
+wright_d <- 
+  wright_base +
+  geom_point(data = duration2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  scale_x_reverse(limits = c(470, 370),
+                  breaks = c(seq(470, 370, by = -100)))
+
+wright_n <- 
+  wright_base +
+  geom_point(data = number2, size = 2, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+
+```
+
+Finally plot them all together
+
+```{r}
+# Using modified multiplot function
+# written by Scott Chamberlain
+
+pdf("../manuscript/figures/fig-dtt-age-appendix.pdf")  
+  multiplot(beck_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")),
+            brusatte_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
+            bapst_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")),
+            wright_s + theme(plot.margin=unit(c(5.5, 5.5, 12, 25),"points")), 
+            beck_d, brusatte_d, bapst_d, 
+            wright_d + theme(plot.margin=unit(c(5.5, 5.5, 12, 5.5),"points")),
+            beck_n,
+            brusatte_n,
+            bapst_n,
+            wright_n + theme(plot.margin=unit(c(5.5, 5.5, 12, 5.5),"points")),
+            cols = 3, 
+            labs = list("Time (MYA)", 
+                        "Relative disparity"),
+            labpos=list(c(0.5, 0.015), c(0.03, 0.5)))
+dev.off()
+```
+
+Repeat for epochs
+-----------------
 ```{r}
@@
   filter(bs.median != 0 & !is.na(bs.median)) %>%
-  filter(model == "equalbins" | model == "proximity" | model == "gradual.split") %>%
+
   mutate(relative.disparity = bs.median/max(bs.median)) %>%
@@
 
-pdf("../manuscript/figures/fig-dtt.pdf")  
+pdf("../manuscript/figures/fig-dtt-epoch-appendix.pdf")  
   multiplot(beck_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")),

### 2017-12-12 manuscript/appendix.tex
Appendix edits

@@
 \usepackage{lineno} 	
-\usepackage{ms}    	
 \usepackage{hyperref}
@@
 \pagenumbering{arabic}
-\linenumbers
+
+\renewcommand{\thetable}{A\arabic{table}}
+\renewcommand{\thefigure}{A\arabic{figure}}
 
@@
 
-% lists - arabic in (1; (2); (3) .
-
-% Title page information
-
-\title{Supporting Information from `Time for a rethink: time sub-sampling methods in disparity-through-time analyses'}
-
-\author{
-	Thomas Guillerme$^{1*}$ and Natalie Cooper$^{2}$
-}
-
-\date{}
-
-\affiliation{\noindent{\footnotesize
-	$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia.\\
-	$^2$Department of Life Sciences, Natural History Museum, Cromwell Road, London, SW7 5BD, UK. natalie.cooper@nhm.ac.uk}\\
-	$^*$Corresponding author\\}
-
-\vfill
-
-%\runninghead{Time sub-samples in disparity-through-time analyses}
-%\keywords{time bin, time slice, disparity} 
-% up to 6
-
 \begin{document}
 
-\mstitlepage
-\parindent = 1.5em
-\addtolength{\parskip}{.3em}
+\noindent{\Large \bf Supporting Information from `Time for a rethink: time sub-sampling methods in disparity-through-time analyses'}
 
-%\section{Abstract}
-% 300 words max
+\raggedright
+\setlength{\parindent}{1cm}
+
+\section{Appendix S1: Extra details of datasets}
+
+Which species were removed? 
+Strap plots?
 
 \newpage
-\raggedright
-\doublespacing
-\setlength{\parindent}{1cm}
-
-\section{Additional tables and figures}
+\section{Appendix S2: Additional tables and figures}
 
@@
     For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
-    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
+    The four dataset names are on the first plot for each dataset (Table 1).}
     \label{figure:peak3}
@@
 
-\bibliographystyle{palaeo} 
-\bibliography{time-refs}
-
 \end{document}

### 2017-12-12 manuscript/time-slice.tex
Add appendix details

@@
 \titleformat*{\paragraph}{\small\itshape}
-
-% lists - arabic in (1; (2); (3) .
 
@@
 
-% NC: might move this to the NHM github account
-% TG: why that? It'll be just extra hassle no?
-% NC: Nah it's just a group and makes it clearer here that I'm actually doing some programming work. Takes about 30s to transfer and just means you/I need to change our remotes which again takes 30s.
-
-% Add overview figure
-%TG: This is a first draft of the figure and its caption. Tell me what you think.
-
+% overview figure
   \begin{figure}[!htbp]
@@
 To test the different time binning/slicing methods we selected four datasets: a mammal dataset from \cite{beckancient2014}, two theropod datasets from \cite{brusatte2014gradual} and \cite{bapst2016topology}, and a crinoid dataset from \cite{wright2017bayesian}.
-Table \ref{table:datasets} and appendix ??? provide more details. % add appendix details
+Table \ref{table:datasets} and the Supporting Information Appendix S1 provide more details. 
 Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
@@
 Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis} to remove pairs of taxa with no cladistic characters in common.
-This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see appendix for details of which species).
-% add appendix details
+This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see Supporting Information Appendix S1 for details of which species).
 
@@
 
-Disparity changes through time for each of the four datasets (Figure \ref{figure:dtt}, Figure in appendix, full tables in the appendix). % add details of appendix
+Disparity changes through time for each of the four datasets (Figure \ref{figure:dtt}, Supporting Information Appendix S2: Figure AX and Tables AX-X). % add details of appendix
 Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods.
 The six different time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits) show similar patterns, so we focus only of the results for one method with a punctuated model of evolution (specifically the `proximity' method), and one method with a gradual model of evolution (specifically the `gradual splits' method).
-Results for all six methods can be found in the appendix. % add details of appendix
+Results for all six methods can be found in Supporting Information Appendix S2: Figure AX. % add details of appendix
 
@@
     The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).
-    Results for stratigraphic ages, and for other time-slicing models, are in the appendix.}
+    Results for stratigraphic ages, and for other time-slicing models, are in the Supporting Information Appendix S2: Figure X} %appendix
     \label{figure:dtt1}
@@
 There is no overall significant systematic difference among the disparities calculated using time bins and those calculated using the time-slicing methods.
-(Table \ref{table:wilcox}, Appendix Table \ref{table:wilcox2}).
+(Table \ref{table:wilcox}, Supporting Information Appendix S2: Table X).
+% appendix
 Instead, the differences depend on the dataset and method in question.
@@
 \paragraph{Disparity peaks}
-In the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets, disparity peaks occur much earlier when time slicing rather than time binning approaches are used (Figure \ref{figure:peak1}, Appendix figure).
+In the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets, disparity peaks occur much earlier when time slicing rather than time binning approaches are used (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX).
 % appendix details
-This is also true for straigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, althought when using equal time bins the peaks are later than the time slicing methods, or very similar (Figure \ref{figure:peak1}, Appendix figure).
-Across the three time binning methods, the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets have similar disparity peaks whichever method is used, whereas in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Figure \ref{figure:peak1}, Appendix figure).
+This is also true for straigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, althought when using equal time bins the peaks are later than the time slicing methods, or very similar (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX).
+%appendix
+Across the three time binning methods, the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets have similar disparity peaks whichever method is used, whereas in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX).
+% appendix
 
@@
     The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (Table \ref{table:datasets}).
-    Results for stratigraphic ages, and for other time-slicing models, are in the appendix.}
+    Results for stratigraphic ages, and for other time-slicing models, are in the Supporting Information Appendix S2: Figure AX} %appendix
     \label{figure:peak1}
@@
 \paragraph{Effects of mass extinction events.}
-Mass extinction events influence disparity in all four datasets (Figure \ref{figure:extinction1}, appendix). 
-However, whether this change in disparity is significant or not depends on the dataset, and on the method used to create time sub-samples (Figure \ref{figure:extinction1}, appendix). % appendix!!!
+Mass extinction events influence disparity in all four datasets (Figure \ref{figure:extinction1}; Supporting Information Appendix S2: Figure AX). %appendix
+However, whether this change in disparity is significant or not depends on the dataset, and on the method used to create time sub-samples (Figure \ref{figure:extinction1}). % appendix!!!
 
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-extinction-effect.pdf}
     \caption[Extinction four example datasets.]
@@
 \section{Conclusions}
+The way you time sub-sample your taxa has important implications for the results of disparity-through-time analyses.
+
+\section{Data archiving statement}
+Data for this study are available from \cite{dryad_n2g80,dryad_6hb7j,dryad_84t75,beckancient2014} but for ease of reproducibility can also be found on GitHub (https://github.com/nhcooper123/time-slice/data).
+
+\section{Supporting information}
+Additional Supporting Information can be found in the online version of this article:
+Appendix S1: Extra details of datasets and Appendix S2: Additional tables and figures.
 
@@
 
-\section{APPENDIX}
-
-  \begin{figure}[!htbp]
-    \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch-appendix.pdf}
-    \caption[Relative disparity through time for four example datasets.]
-    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Pink points represent time binning methods, 
-    % blue points are time slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
-    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods.
-    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
-    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
-    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
-    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
-    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
-    \label{figure:dtt2}
-  \end{figure}
-
-  \begin{figure}[!htbp]
-    \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-age-appendix.pdf}
-    \caption[Relative disparity through time for four example datasets.]
-    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Pink points represent time binning methods, 
-    %blue points are time slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
-    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods.
-    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages. 
-    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages in the time frame of the dataset. 
-    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages in the time frame of the dataset. 
-    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
-    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
-    \label{figure:dtt3}
-  \end{figure}
-
-  % Wilcoxon results table
-%\begin{landscape}
-  \input{tables/table_wilcox_appendix.tex} 
-  \label{table:wilcox2}  
-%\end{landscape}
-
-\begin{figure}[!htbp]
-    \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch-appendix.pdf}
-    \caption[Timing of peak disparity for four example datasets.]
-    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    %Pink points represent time binning methods, green points are time slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
-    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
-    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
-    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
-    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
-    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
-    \label{figure:peak2}
-  \end{figure}
-
-\begin{figure}[!htbp]
-    \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-age-appendix.pdf}
-    \caption[Timing of peak disparity for four example datasets.]
-    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    %Pink points represent time binning methods, green points are time slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
-    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages. 
-    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages in the time frame of the dataset. 
-    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages in the time frame of the dataset. 
-    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
-    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
-    \label{figure:peak3}
-  \end{figure}
-
 \bibliographystyle{palaeo} 

### 2017-12-12 analyses/04-table-datasets.Rmd
Small edits

--- 2017-12-06 analyses/05-table-compiler.Rmd
@@
 ---
-title: "Table compiler"
+title: "Dataset table"
 author: "Thomas Guillerme"
@@
 
-# Compiling the tables for the paper
+# Compiling the dataset table
 

### 2017-12-13 analyses/03-statistical-analyses.Rmd
Fixed NA bug in Bapst

@@
 mapply.test <- function(data, times, test = wilcox.test, correction = "bonferroni") {
-  test <- try(output <- test.dispRity(data, test = test, comparisons = times, correction = correction), silent = TRUE)
-  if(class(test) == "try-error") {
+  test_run <- try(output <- test.dispRity(data, test = test, comparisons = times, correction = correction), silent = TRUE)
+  if(class(test_run) == "try-error") {
+
+    ## Try the test on less lags
+    subsets_to_test <- unique(unlist(times))
+    na_subsets <- summary(data)$obs[subsets_to_test]
+    if(any(is.na(na_subsets))) {
+      ## Remove the subsets with NA from the comparisons
+      subsets_to_remove <- unlist(lapply(lapply(times, function(X) match(X, subsets_to_test[which(is.na(na_subsets))])), function(X) return(any(!is.na(X)))))
+
+      ## Remove the subsets
+      times_no_na <- times[!subsets_to_remove]
+    } else {
+      subsets_to_remove <- rep(TRUE, length(times))
+    }
 
@@
     colnames(p_value) <- "p.value"
+
+    if(!all(subsets_to_remove)) {
+      output <- test.dispRity(data, test = test, comparisons = times_no_na, correction = correction)
+
+      ## Replacing the NAs
+      statistic[!subsets_to_remove,1] <- output[[1]][,1]
+      p_value[!subsets_to_remove,1] <- output[[2]][,1]
+    }
 

### 2017-12-14 manuscript/time-slice.tex
edits on the manuscript up until time-slicing.

@@
 
-Disparity-through-time analyes are common in palaeontology \citep{Foote01071994}.
+Disparity-through-time analyes are common in palaeontology since the 90s \citep{gould1991disparity,briggs1992morphological,Wills1994,Foote01071994}.
 They reveal how the morphological diversity of clades has changed through time, and allow us to make inferences about the breadth of ecological niches extinct taxa occupied \citep{foote1997evolution}.
@@
 
-Disparity-through-time analyes have two main analysis components: calculating disparity, and creating time sub-samples of the data. 
+Disparity-through-time analyes have two main analysis components: calculating disparity, and creating time sub-subsets of the data. 
 Here we focus on the latter.
-The nature of disparity (it is a diversity metric), means it cannot be calculated using just one individual, so some way of subsetting taxa is required.
+The nature of disparity (i.e. a diversity metric), means it cannot be calculated using a single individual, so some way of sampling many taxa in each sub-subset is required.
 Changes in disparity-through-time are generally investigated by calculating the disparity of taxa present during specific time intervals or time bins \citep[e.g][]{cisneros2010,prentice2011,Hughes20082013,hopkinsdecoupling2013,bentonmodels2014,bensonfaunal2014}.
@@
 First, time bins defined by stratigraphy are not of equal size, biasing higher disparity towards longer stratigraphic periods. 
-This can be dealt with using rarefaction, i.e. repeating the analysis using the smallest number of taxa present in any bin as the number of sampled taxa in each bin. 
+This can be dealt with using rarefaction methods, i.e.
+repeating the analysis while re-sampling the taxa to have the same amount of taxa in each bin (e.g.
+using bootstrapping with limited re-sampling methods).
 This can, however, lead to very large confidence intervals when there are stratigraphic periods with very few species.
 Other studies split large time bins so they are of roughly equal size, but this is often an \textit{ad hoc} procedure that can introduce more bias depending on where bins are split.
-Second, all time binning approaches (whether bins are equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring \textit{between} intervals, rather than also allowing for gradual changes \textit{within} intervals \citep{Hunt21042015}.
+Second, all time binning approaches (whether bins are equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring \textit{between} intervals, rather than also allowing for gradual changes \textit{within} intervals \citep[shown as being quiet common in the fossil record;][]{Hunt21042015}.
 Finally, time bin analyses are often limited by the number of taxa in each bin.
@@
 correlations of disparity with hypothesised drivers of morphological evolution, are not possible.
+%TG: Not sure where in this paragraph but it'll be good to add this point:
+%Also, in the case of looking at changes in disparity due to events specific in time (e.g.
+a mass extinction), time bins method leave us with a low temporal resolution: stratigraphic bins are limited to global events that defined the bins boundaries and time bins are limited by the duration of each bin (e.g.
+if time bins are every 20 Mya but an event occurs at time $t$ and $t+5$ Mya, it might not be possible to capture its effects).
 
@@
 Time-slicing also includes any element present in the phylogeny (branches, nodes and tips) at the time slice in question as part of the disparity calculation.
-This allows us to measure disparity at time points where there are no terminal taxa, and increases the sample size at each time point, making downstream analyses of the drivers of disparity much more feasible.
+This allows us to measure disparity at time points where there are no sampled terminal taxa, and increases the sample size at each time point, making downstream analyses of the drivers of disparity much more feasible.
 
@@
 To test the different time sub-sampling methods, we followed the protocol below (Figure \ref{datasets}). 
-All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (https://github.com/nhcooper123/time-slice).
-
+All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (\url{https://github.com/nhcooper123/time-slice}).
 %https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/details-of-specific-functions.html#time-slicing$
@@
 Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
-These datasets are freely available with their accompanying papers (Table \ref{table:datasets}), but for reproducibility purposes we also provide the data we used on GitHub (https://github.com/nhcooper123/time-slice/data).
+These datasets are freely available with their accompanying papers (Table \ref{table:datasets}), but for reproducibility purposes we also provide the data we used on GitHub (\url{https://github.com/nhcooper123/time-slice/data}).
 
@@
 \label{ace}
-For each dataset we used the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny, using the \texttt{rerootingMethod} function from the \texttt{R} package \texttt{phytools} version 0.6-44 \citep{phytools,R}.
-Where there were missing character data for a taxon we followed the method of \cite{Claddis} and treated missing data as any possible observed state for each character.
-For example, if a character had two observed states ($0$ and $1$) across all taxa, we attributed the multi-state ``$0$\&$1$" value to the taxon with missing data, representing an equal probability of being either $0$ or $1$.
-This allows the ancestral node of a taxon with missing data to be estimated with no assumptions other than that the taxon has one of the observed character states.
+For each dataset we estimated the ancestral character states at each node using the \texttt{AncStatesEstMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis,R}.
+This function uses the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny \citep[based on the \texttt{rerootingMethod} function in \texttt{phytools};][]{phytools}.
+% Inapplicable and missing character for any taxon were treated as ambiguous characters (i.e. any possible observed state for the character).
+% For example, if a character had two observed states ($0$ and $1$) across all taxa, we attributed the multi-state ``$0$\&$1$" value to the taxon with missing data, representing an equal probability of being either $0$ or $1$.
+% This allows the ancestral node of a taxon with missing data to be estimated with no assumptions other than that the taxon has one of the observed character states.
+%TG: These three sentences can be removed: it's pretty standard procedure.
+I'm not sure if we should leave that here.
+It sounded fancy during PhD time but now it's actually a bit lame to write it like that.
+Plus there is the inapplicable problem that Claddis do not deal with...
 To prevent poor ancestral state reconstructions from biasing our results, especially when a lot of error is associated with the reconstruction, we only included ancestral state reconstructions with a scaled Likelihood $\geq$ $0.95$.
 Ancestral state reconstructions with scaled Likelihoods below this threshold were recoded as missing data (``?'').
+This allowed our results to be less dependent on the quality (or the absence thereof) of the ancestral state reconstruction methods, especially in parts of the datasets where data was sparse.
 
@@
 To explore disparity-through-time in our datasets we used a cladisto-space approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
-This approach is similar to constructing a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010}, except a cladisto-space is an approximation of the morphospace based on cladistic data (i.e.
-the discrete morphological characters used to build a phylogenetic tree).
+This approach is similar to constructing a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010}, except a cladisto-space is an approximation of the morphospace based on discrete morphological data (i.e.
+the discrete morphological characters used to build a phylogenetic tree, often referred to as ``cladistic'' data).
+%TG: about that, I'm actually less and less convinced we should call it caldisto-space: first, semantically, cladistic data means data used to make clades (phylos) so it could be nucleotides as well.
+Second, during the PhD, we used cladisto-space to distinguish it from the morpho-space but at that time I was naively thinking the the morpho-space is the morphometric-space (which it often is!) but, again, semantically, the morpho-space can also be simply the morphology-space...
+However, I like the sound of cladisto-space and most people will understand it the way we do.
+I'm just thinking about pedantic-palaeo-reviewers...
 Mathematically, a cladisto-space is an $n$ dimensional object that summarises the cladistic distances between the taxa present in a cladistic matrix.
@@
 After constructing our distance matrices we transformed them using classical multidimensional scaling \citep[MDS;][]{torgerson1965multidimensional,GOWER01121966,cailliez1983analytical}.
-This method (also referred to as PCO; e.g. \citealt{Brusatte2015}; or PCoA; e.g. \citealt{paradisape:2004}) is an eigen decomposition of the distance matrix.
+This method (also referred to as PCO; e.g.
+\citealt{Brusatte2015}; or PCoA; e.g.
+\citealt{paradisape:2004}; but distinguished in \citealt{legendre2012numerical}) is an eigen decomposition of the distance matrix.
 Because we used Gower distances instead of raw Euclidean distances, negative eigenvalues can be calculated.
@@
 \subsection{Disparity-through-time analyses}
+
+The disparity-through-time analyses were performed using the \texttt{dispRity} \texttt{R} package \citep{dispRity.v02}. %TG: With some hope, the MEE paper might be out at the same time as this one!
 
@@
 This results in even-sampling of the cladisto-space across time and allows us to use different underlying models of character evolution (punctuated or gradual).
+
+%TG: Is it worth saying that Halliday and Goswami came up with a similar idea \citep{halliday2016eutherian} but that we came up with it first \cite{Guillerme2015} and we're the first to implement it \cite{dispRity.v02}?
 
@@
   We thank Dave Bapst, Steve Brusatte, Graeme Lloyd, April Wright and Davey Wright for assistance in gathering data for the analyses.
+  %TG: and actually: Dave Bapst, Graeme Lloyd and Davey Wrigth for discussions about the method?
 

### 2017-12-14 manuscript/time-slice.tex
Comments on results + discussion

@@
 However, the Gower distance cannot calculate distances when taxa have no overlapping data.
-Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis} to remove pairs of taxa with no cladistic characters in common.
+Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package to remove pairs of taxa with no cladistic characters in common.
 This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see Supporting Information Appendix S1 for details of which species).
@@
 \label{disparity_calc}
-Disparity can be calculated in many different ways \citep[e.g.][]{Wills1994,Ciampaglio2004,thorneresetting2011,hopkinsdecoupling2013,huang2015origins}, however most studies estimate disparity using four metrics: the sum and products of ranges and variances, each of which gives a slightly different estimate of how the data fits within the cladisto-space \citep{Foote01071994,Wills1994,brusatte50,Brusatte12092008,cisneros2010,thorneresetting2011,prentice2011,brusattedinosaur2012,toljagictriassic-jurassic2013,ruta2013,bentonmodels2014,bensonfaunal2014}.
+Disparity can be calculated in many different ways \citep[e.g.][]{Wills1994,Ciampaglio2004,thorneresetting2011,hopkinsdecoupling2013,huang2015origins}, however a majority of studies in palaeobiology estimate disparity using four metrics: the sum and products of ranges and variances, each of which gives a slightly different estimate of how the data fits within the cladisto-space \citep{Foote01071994,Wills1994,brusatte50,Brusatte12092008,cisneros2010,thorneresetting2011,prentice2011,brusattedinosaur2012,toljagictriassic-jurassic2013,ruta2013,bentonmodels2014,bensonfaunal2014}.
 However, these metrics have limitations. 
@@
 Second, because we include all $n$ dimensions in the analysis (see above), the products of ranges and variances will tend towards zero since the scores of the last dimension are usually really close to zero themselves.
-We therefore use the sum of variances metric to estimate disparity here.
-Note that there are still statistical issues with this metric, but for the purposes of comparison with previous work we want to use a standard metric for these analyses.
+We therefore use the sum of variances metric to estimate disparity here:
+
+\begin{equation}
+\text{disparity} = \sum\sigma^{2}{n_i}
+\end{equation}
+
+Where $\sigma^{2}{n_i}$ is the variance for the $n_i^{th}$ dimension ranging from $k$ to $k-2$ with $k$ being the number of taxa in the dataset.
+Note that there are still statistical issues with this metric (such as the co-variance between dimensions not being measured), but for the purposes of comparison with previous work we want to use a standard metric for these analyses.
 
@@
 
-    The two first models assume that changes always occur early (ACCelerated TRANsition) or late along the branches (DELayed TRANsition).
+    The two first models assume that changes always occur early (\textit{acc}elerated \textit{tran}sition) or late along the branches (\textit{del}ayed \textit{tran}sition).
     The third model makes neither assumption and simply selects data from the ancestor or the descendant at random, and the fourth bases the selection of either the ancestor or the descendant on where the slice occurs along the branch.
-    The punctuated models only select either the ordination score from the ancestor and the descendant once in the whole disparity analysis.
+    These punctuated models only select either the ordination score from the ancestor and the descendant once in the whole disparity analysis.
     For example, if using the ``random'' model, if the data of the ancestor has been randomly chosen, only this data will be used during the bootstrapping (see below) and for the disparity calculation.
@@
     To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time bin by randomly resampling with replacement a new sub-sample of taxa from the observed taxa in the bin 1000 times.
+    %TG: actually in the code the data is bootstrapped 100 times only.
+Should I change to 1000?
+To be fair, I think 1000 is too much (in terms of calculation and results) and 100 to few.
+A compromise could be 500.
     We then calculated the median disparity value for each time bin along with the 50\% and 95\% confidence intervals.
@@
   \end{enumerate}
+
+%TG: well done! Nice description of the method! I figured out it was pretty obvious in our (my) head but it's actually more painfull to write down.
 
@@
 Many analyses of disparity-through-time aim to demonstrate differences in disparity before and after mass extinction events [citations]. 
-Each of our four datasets span one mass extinction (Cretaceous-Paleogene 66 MYA; \citealt{brusatte2014gradual,bapst2016topology,beckancient2014}; Ordovician-Silurian 455– 430 MYA; \citealt{wright2017bayesian}), so we use Wilcoxon tests to compare disparity in the time bin/slice prior to the approrpiate mass extinction, to that of the time bin/slice following the extinction event.
-Significant results suggest a significant effect of the mass extinction on disparity in the group.
+Each of our four datasets span one mass extinction (Cretaceous-Paleogene 66 MYA; \citealt{brusatte2014gradual,bapst2016topology,beckancient2014}; Ordovician-Silurian 455– 430 MYA; \citealt{wright2017bayesian}), so we use Wilcoxon tests to compare disparity in the time bin/slice prior to the appropriate mass extinction, to that of the time bin/slice following the extinction event.
+Significant results suggest an effect of the mass extinction on disparity in the group.
 We then compare these results across the time sub-sampling methods to determine if our conclusions would change depending on the method used.
@@
 Disparity changes through time for each of the four datasets (Figure \ref{figure:dtt}, Supporting Information Appendix S2: Figure AX and Tables AX-X). % add details of appendix
-Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods.
+Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods (i.e.
+the data included in the time bins are only the nodes and the tips, the time-slice also contains the data along the branches).
 The six different time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits) show similar patterns, so we focus only of the results for one method with a punctuated model of evolution (specifically the `proximity' method), and one method with a gradual model of evolution (specifically the `gradual splits' method).
@@
 % appendix details
-This is also true for straigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, althought when using equal time bins the peaks are later than the time slicing methods, or very similar (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX).
+This is also true for stratigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, although when using equal time bins the peaks are later than the time slicing methods, or very similar (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX).
 %appendix
@@
 
+  %TG: Would it be possible to make the bins points into a thick line? Though that's just aesthetic...
+
 \paragraph{Effects of mass extinction events.}
@@
     \caption[Extinction four example datasets.]
-    {XXX}
+    {Results of the extinction effect in change in disparity.
+    Pink cells and green cells indicate respectively a significant or no significant change of disparity (i.e. an effect or no effect of the extinction on disparity).
+    White cells are due to not enough data for performing the test (less than 2 taxa sampled in that period).
+    \textit{e:1}, \textit{e:2}, and \textit{e:3} denote respectively the test between the last bin/slice before the extinction and the first, second and third bin/slice after it (to account for an eventual lag effect).}
     \label{figure:extinction1}
@@
 
+%TG: ideas for the discussion:
+  % Extinction test: in Brusatte's dataset, ther seems to be an effect of the model (gradual vs.
+punctuated) on recovering the effect of the extinction or not when using stratigraphy: not significant for bins and punctuated slices but signif for gradual slices.
+TG: kind of showing why it's cool to allow different models of evolution.
+  % peak test: the peak really depends on which time slicing method is used!
+This is probably expected for the disparity metric (different metrics give different results) but was never acknowledge to be the case for different methods of time subsampling!
+Kind of cool too!
+
 Benefits of time slices
 
+  - Allowing for more data to be used
+
+  - Allowing different models of evolution
+
+  - Avoiding biases from using stratigraphy (i.e. the stratigraphic bins are based on disparity in the fossil record)
+
 Further stuff to fix?
+
+  - Allowing proper reconstruction along the branches (basically slicing before the ordination).
+
+  - Testing the effect of slice/bin on properly simulated data set to test for systematic difference (or not)
+
+  - Testing the effect of slice/bin with different disparity metrics
 

### 2017-12-15 analyses/02-time-slice-analyses.Rmd
Change branch for dipRity install
Order slugs so it  matches later analyses and remove bears

@@
 Below we lay out the process in detail for one dataset, and then provide the code for the other datasets below this.
-Note that much of this uses functions in the `simulations.R` file.
+Note that much of this uses functions in the `simulations_fun.R` file.
 
@@
 
-## The latest version of dispRity on release branch #TG: master is under many changes!
-## install_github("TGuillerme/dispRity", ref = "release")
+## The latest version of dispRity
+## install_github("TGuillerme/dispRity", ref = "master")
 library(dispRity)
@@
 
-Wright 2017
+Brusatte et al 2014 Current Biol
 -----------
+
 ```{r}
-## Choose a slug
-slug <- "Wright2017"
+slug <- "Brusatte2014"
 
@@
 tree <- read.nexus(paste0("../data/processed/tree.", slug, ".tre"))
-## Add root age as defined in Wright 2017
-tree$root.time <- 484.4
+## Add root age (max tree age date)
+tree$root.time <- max(tree.age(tree)[, 1])
 
@@
 
-Heath et al bears
+Wright 2017
 -----------
-
 ```{r}
-slug <- "bears"
+## Choose a slug
+slug <- "Wright2017"
 
@@
 tree <- read.nexus(paste0("../data/processed/tree.", slug, ".tre"))
-## Add root age (max tree age date)
-tree$root.time <- max(tree.age(tree)[, 1])
+## Add root age as defined in Wright 2017
+tree$root.time <- 484.4
 
@@
 ```
-
-Brusatte et al 2014 Current Biol
------------
-
-```{r}
-slug <- "Brusatte2014"
-
-## Loading the FADLAD
-FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
-
-## Loading the tree
-## See "01-extract-data-for analyses.Rmd" for how we cleaned these
-tree <- read.nexus(paste0("../data/processed/tree.", slug, ".tre"))
-## Add root age (max tree age date)
-tree$root.time <- max(tree.age(tree)[, 1])
-
-## Loading the distance matrices
-## See "01-extract-data-for analyses.Rmd" for how we made these
-matrix_dist <- readRDS(paste0("../data/processed/distance_matrix.", slug ,".Rda"))
-
-## Ordinating the matrix
-morphospace <- cmdscale(matrix_dist, k = nrow(matrix_dist) - 2, add = TRUE)$points
-
-## Run the disparity analyses
-## Ages (rarefied to 3 taxa only)
-out1 <- run.all.disparity.wrapper(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
-                                bootstraps = 100, metric = c(sum, variances),
-                                metric.name = "sum_var", rarefaction = FALSE)
-
-## Save results to output folder
-saveRDS(out1, file = paste0("../outputs/age.", slug ,".Rda"))
-
-## Epochs (rarefied to 3 taxa only)
-out2 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
-                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var",
-                        rarefaction = FALSE)
-
-## Save results to output folder
-saveRDS(out2, file = paste0("../outputs/epoch.", slug ,".Rda"))
-```

### 2017-12-15 analyses/02-time-slice-analyses.Rmd
Corrected incorrect root age for Wright

@@
 ## Add root age as defined in Wright 2017
-tree$root.time <- 484.4
+tree$root.time <- 485.4
 

### 2017-12-15 manuscript/appendix.tex
Add tree details to appendix

@@
 
-Which species were removed? 
-Strap plots?
+\subsection{Beck2014 (Figure \ref{figure:beck})}
+The following taxa were removed because they were in the phylogeny but not the character matrix or vice versa: \textit{Montanalestes}, \textit{Lainodon}, \textit{Kharmerungulatum}, \textit{Alymlestes}.
+
+\subsection{Brusatte2014 (Figure \ref{figure:brusatte})} 
+We used one randomly selected time-scaled tree from Brusatte et al. (2014).
+Zero-length branches were replaced with the minimum branch length in the phylogeny.
+The following taxa were removed because they were in the phylogeny but not the character matrix or vice versa: \textit{Sinraptor dongi}, \textit{Hesperonychus elizabethae}, \textit{Pyroraptor olympius}, \textit{Limenavis patagonica}, \textit{Lithornis}, \textit{Crypturellus undulatus}, \textit{Gallus gallus}, \textit{Crax pauxi}, \textit{Anas platyrhynchus}, \textit{Chauna torquata}, \textit{Epidendrosaurus} and \textit{Kinnareemimus}.
+The following taxa were removed because they shared no characters in the morphological matrix: \textit{Shanag ashile}, \textit{Atrociraptor marshalli}, \textit{Proceratosaurus bradleyi}, \textit{Incisivosaurus gauthieri}, \textit{Enigmosaurus}, \textit{Nanshiungosaurus brevispinus}, \textit{Xixiasaurus}, \textit{Tsaagan mangas}, \textit{Mirischia}, \textit{Pedopenna}, \textit{Suzhousaurus}, \textit{Juratyrant}, \textit{Vorona}, \textit{Bonapartenykus}, \textit{Teratophoneus}, \textit{Gobipteryx}, \textit{Songlingornis}, \textit{Liaoningornis longidigitu} and \textit{Achillesaurus}.
+
+\subsection{Bapst2016 (Figure \ref{figure:bapst})} 
+We used the maximum clade credibility tree from Bapst et al. (2016).
+Zero-length branches were replaced with the minimum branch length in the phylogeny.
+The following taxa were removed because they were in the phylogeny but not the character matrix or vice versa: \textit{Mei long} and \textit{Mei lon}. 
+The following taxa were removed because they shared no characters in the morphological matrix: \textit{Hagryphus giganteus}, \textit{Atrociraptor marshalli}, \textit{IGM100 1015 UndesDromaeosaurid}, \textit{Dromaeosaurus albertensis}, \textit{Incisivosaurus gauthieri}, \textit{Deinocheirus mirificus}, \textit{Therizinosaurus cheloniformis},  \textit{Anserimimus planinychus} and \textit{Elmisaurus rarus}.
+
+\subsection{Wright2017 (Figure \ref{figure:wright})}
+We used the maximum clade credibility tree from Wright (2017). 
+To properly timescale the tree we followed the advice of Wright (2017) and divided the branch lengths by the corresponding clock rate (= 0.03517385)  and then set the root time to 485.4. 
+Zero-length branches were replaced with the minimum branch length in the phylogeny.
+No taxa were removed.
+
+\begin{figure}[!htbp]
+    \centering
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-tree-Beck2014-appendix.pdf}
+    \caption[Beck2014.]
+    {Phylogeny from Beck \& Lee (2014).}
+    \label{figure:beck}
+  \end{figure}
+
+\begin{figure}[!htbp]
+    \centering
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-tree-Brusatte2014-appendix.pdf}
+    \caption[Brusatte2014.]
+    {Phylogeny from Brusatte et al. (2014). This is one randomly selected tree from the time-scaled trees in the paper.}
+    \label{figure:brusatte}
+  \end{figure}
+
+\begin{figure}[!htbp]
+    \centering
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-tree-Bapst2016-appendix.pdf}
+    \caption[Bapst2016.]
+    {Phylogeny from Bapst et al. (2016). This is the maximum clade credibility tree.}
+    \label{figure:bapst}
+  \end{figure}
+
+\begin{figure}[!htbp]
+    \centering
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-tree-Wright2017-appendix.pdf}
+    \caption[Wright2017.]
+    {This is the maximum clade credibility tree from Wright (2017).}
+    \label{figure:wright}
+  \end{figure}
 

### 2017-12-15 analyses/09-appendix-trees-figures.Rmd
Add code to make strap trees

---
title: "Trees for the appendix"
author: "Natalie Cooper"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

Load the packages 
```{r}
library(strap)
```

Load the data

```{r}
## Choose a slug
slug <- "Beck2014"
#slug <- "Brusatte2014"
#slug <- "Bapst2016"
#slug <- "Wright2017"

## Loading the FADLAD
FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)

## Loading the tree
tree <- read.nexus(paste0("../data/processed/tree.", slug, ".tre"))
## Add root age (max tree age date)
## Remember that for Wright this is 485.4
tree$root.time <- max(tree.age(tree)[, 1])
```

Plot the trees and save as PDF
```{r}
pdf(paste0("../manuscript/figures/fig-tree-", slug, "-appendix.pdf"), width = 12, height = 12)  
geoscalePhylo(tree = tree, ages = FADLAD)
dev.off()
```

### 2017-12-15 analyses/10-appendix-dtt-figures.Rmd
Add legend

--- 2017-12-12 analyses/figure-dtt-appendix.Rmd
@@
 
-# Modify data - add midpoints, remove missing data, filter to just two models,
+# Modify data - add midpoints, remove missing data, 
 # make disparity relative by dividing by max values
@@
   filter(bs.median != 0 & !is.na(bs.median)) %>%
-
   mutate(relative.disparity = bs.median/max(bs.median)) %>%
@@
   theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        legend.position = "none")
+        axis.title.x = element_blank())
 
@@
   ggtitle("Stratigraphy") +
-  theme(plot.title = element_text(size = 14, hjust = 0.5)) +
+  theme(plot.title = element_text(size = 14, hjust = 0.5),
+        legend.position = "none") +
   annotate("text", x = 75, y = 0.15, label = "Beck2014")
@@
   ggtitle("Duration") +
-  theme(plot.title = element_text(size = 14, hjust = 0.5))
+  theme(plot.title = element_text(size = 14, hjust = 0.5),
+        legend.position = "none")
 
@@
   ggtitle("Number") +
-  theme(plot.title = element_text(size = 14, hjust = 0.5))
+  theme(plot.title = element_text(size = 14, hjust = 0.5),
+        legend.key.size = unit(0.2, "in"),
+        legend.text = element_text(size = 8),
+        legend.title=element_blank())
 
@@
 
-# Modify data - add midpoints, remove missing data, filter to just two models
+# Modify data - add midpoints, remove missing data
 strat2 <- 
@@
         axis.title.x = element_blank(),
-        legend.position = "none")
+        )
 

### 2017-12-15 analyses/11-appendix-figure-peaks.Rmd
Move appendix peak figures out to separate file

--- 2017-12-11 analyses/09-figure-peaks.Rmd
@@
 ```
-
-# Beck 2014
-
-```{r}
-slug <- "Beck2014"
-```
-
-Load the data
-
-```{r}
-data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
-data_max$bin_type <- gsub("stratigraphy", "Stratigraphy", data_max$bin_type) 
-data_max$bin_type <- gsub("duration", "Duration", data_max$bin_type) 
-data_max$bin_type <- gsub("number", "Number", data_max$bin_type) 
-data_max$bin_type <- as.factor(data_max$bin_type)
-```
-
-# Modify the data
-```{r}
-# Get ages for the bins, go for the midpoints
-x <- str_split(data_max$subsets, " - ") 
-y <- map(x, as.numeric)
-z1 <- map(y, max)
-z2 <- map(y, min)
-
-# Modify data - add midpoints, remove missing data, filter to just three models
-data_max2 <- 
-  data_max %>%
-  mutate(time1 = unlist(z1)) %>%
-  mutate(time2 = unlist(z2)) %>%
-  filter(bs.median != 0 & !is.na(bs.median)) %>%
-  mutate(bin_type = relevel(data_max$bin_type, ref = 'Stratigraphy')) %>%
-  filter(model == "equalbins" | model == "proximity" | model == "gradual.split")
-data_max2_age <- filter(data_max2, stratigraphy == "Age")
-
-data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
-```
-
-```{r}
-beck_peaks <- 
-  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- # expand_limits(x = 0, y = c(0,1)) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        legend.position = "none",
-        panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.background = element_blank())
-
-beck_peaks_age <- 
-  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- # expand_limits(x = 0, y = c(0,1)) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        legend.position = "none",
-        panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.background = element_blank())
-```
-
-# Brusatte 2014
-
-```{r}
-slug <- "Brusatte2014"
-```
-
-Load the data
-
-```{r}
-data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
-```
-```{r}
-# Get ages for the bins, go for the midpoints
-x <- str_split(data_max$subsets, " - ") 
-y <- map(x, as.numeric)
-z1 <- map(y, max)
-z2 <- map(y, min)
-
-# Modify data - add midpoints, remove missing data, filter to just three models
-data_max2 <- 
-  data_max %>%
-  mutate(time1 = unlist(z1)) %>%
-  mutate(time2 = unlist(z2)) %>%
-  filter(bs.median != 0 & !is.na(bs.median)) %>%
-  mutate(bin_type = relevel(data_max$bin_type, ref = 'stratigraphy')) %>%
-  filter(model == "equalbins" | model == "proximity" | model == "gradual.split")
-
-data_max2_age <- filter(data_max2, stratigraphy == "Age")
-
-data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
-```
-
-```{r}
-brusatte_peaks <- 
-  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
-  expand_limits(y = 0) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        legend.position = "none", 
-       panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.text = element_blank())
-
-brusatte_peaks_age <- 
-  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
-  expand_limits(y = 0) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        legend.position = "none", 
-       panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.text = element_blank()) 
-```
-
-# Bapst 2016
-
-```{r}
-slug <- "Bapst2016"
-```
-
-Load the data
-
-```{r}
-data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
-```
-```{r}
-# Get ages for the bins, go for the midpoints
-x <- str_split(data_max$subsets, " - ") 
-y <- map(x, as.numeric)
-z1 <- map(y, max)
-z2 <- map(y, min)
-
-# Modify data - add midpoints, remove missing data, filter to just three models
-data_max2 <- 
-  data_max %>%
-  mutate(time1 = unlist(z1)) %>%
-  mutate(time2 = unlist(z2)) %>%
-  filter(bs.median != 0 & !is.na(bs.median)) %>%
-  mutate(bin_type = relevel(data_max$bin_type, ref = 'stratigraphy')) %>%
-  filter(model == "equalbins" | model == "proximity" | model == "gradual.split")
-
-data_max2_age <- filter(data_max2, stratigraphy == "Age")
-
-data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
-```
-
-```{r}
-bapst_peaks <- 
-  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- # expand_limits(x = 0, y = c(0,1)) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        legend.position = "none",
-       panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.text = element_blank())
-
-bapst_peaks_age <- 
-  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- # expand_limits(x = 0, y = c(0,1)) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        legend.position = "none",
-       panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.text = element_blank())
-```
-
-# Wright 2017
-
-```{r}
-slug <- "Wright2017"
-```
-
-Load the data
-
-```{r}
-data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
-```
-```{r}
-# Get ages for the bins, go for the midpoints
-x <- str_split(data_max$subsets, " - ") 
-y <- map(x, as.numeric)
-z1 <- map(y, max)
-z2 <- map(y, min)
-
-# Modify data - add midpoints, remove missing data, filter to just three models
-data_max2 <- 
-  data_max %>%
-  mutate(time1 = unlist(z1)) %>%
-  mutate(time2 = unlist(z2)) %>%
-  filter(bs.median != 0 & !is.na(bs.median)) %>%
-  mutate(bin_type = relevel(data_max$bin_type, ref = 'stratigraphy')) %>%
-  filter(model == "equalbins" | model == "proximity" | model == "gradual.split")
-
-data_max2_age <- filter(data_max2, stratigraphy == "Age")
-
-data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
-```
-
-```{r}
-wright_peaks <- 
-  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- expand_limits(y = 480) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        legend.position = "none",
-       panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.text = element_blank())
-
-wright_peaks_age <- 
-  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- expand_limits(y = 480) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        legend.position = "none",
-       panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.text = element_blank()) 
-```
-
-Finally plot them all together
-
-```{r}
-# Using modified multiplot function
-# written by Scott Chamberlain
-
-pdf("../manuscript/figures/fig-peaks-epoch.pdf")  
-  multiplot(beck_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
-            brusatte_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
-            bapst_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
-            wright_peaks + theme(plot.margin=unit(c(5.5, 5.5, 20, 25),"points")),
-            cols = 1,
-            labs = list("Time bins or time slice method",
-                        "Time at disparity peak (MYA)"),
-            labpos = list(c(0.5, 0.015), c(0.03, 0.5)))
-
-dev.off() 
-```
-
-# And for ages
-
-```{r}
-
-pdf("../manuscript/figures/fig-peaks-age.pdf")  
-  multiplot(beck_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
-            brusatte_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
-            bapst_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
-            wright_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 20, 25),"points")),
-            cols = 1,
-            labs = list("Time bins or time slice method",
-                        "Time at disparity peak (MYA)"),
-            labpos = list(c(0.5, 0.015), c(0.03, 0.5)))
-
-dev.off()
-
-```
-
-# Repeat for all models for appendix
 
@@
         panel.border = element_blank(),
-        strip.background = element_blank())
+        strip.background = element_blank(),
+        strip.text = element_text(size = 14) )
 
@@
         panel.border = element_blank(),
-        strip.background = element_blank())
+        strip.background = element_blank(),
+        strip.text = element_text(size = 14))
 ```

### 2017-12-15 analyses/03-statistical-analyses.Rmd
Minor tidying of code

@@
 ---
-title: "Outputs from time slice analyses"
+title: "Statistical analyses of time slice outputs"
 author: "Natalie Cooper and Thomas Guillerme"
@@
 
-This code runs the output analyses in the paper, i.e. creates tables and performs statistical tests.
+This code runs the output analyses in the paper, i.e. performs statistical tests.
 
@@
 
-This script requires the outputs from `02-time-slice-analyses.Rmd`. This of course can all be written into functions but we ran out of time so here is some suboptimal but working code!
+This script requires the outputs from `02-time-slice-analyses.Rmd` and functions in `stats_fun.R`.
+This of course can all be written into functions but we ran out of time so here is some suboptimal but working code!
 
@@
 
-#out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
-out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
+out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
+#out1 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))
 
@@
                 tidy(w_punc), tidy(w_rand))
+# Add model names
 outputS$model <- c("acctran", "deltran", "gradual.split", "proximity", 
                         "equal.split", "random")
+# Add that this is the stratigraphy method
 outputS$type <- rep("stratigraphy", 6)
+# Add label for "Age" vs. "Epoch"
 outputS$period <- rep(strat$stratigraphy[1], 6)
@@
 
-Extinction effect
-------------------
+Mass extinction effects
+---------------------
 
-Here we can test the effect of mass extinction on the different methods with the different models (for each dataset!).
+Here we can test the effect of mass extinction on the different methods with the different models (for each dataset).
 The way to test the extinction effect is a referential wilcoxon test.#' @param as.list \code{logical}, whether to output the results as a list for \code{\link{test.dispRity}} (\code{TRUE}) or not (\code{FALSE} - default).
 
-In other words, we compare the bins/slice just before the extinction event to the three ones just after (to account for an eventual lag effect).
+In other words, we compare the bins/slice just before the extinction event to the three ones just after (to account for an possible lag effect).
 

### 2017-12-15 analyses/05-figure-dtt.Rmd
Minor tidying of code

@@
 ---
-title: "Disparity through time - simplified and for appendix"
+title: "Disparity through time figures"
 author: "Natalie Cooper"
@@
 ---
+
+Load packages
 
@@
 pdf("../manuscript/figures/fig-dtt-age.pdf")  
-  multiplot(beck_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")),
-            brusatte_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
-            bapst_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")),
-            wright_s + theme(plot.margin=unit(c(5.5, 5.5, 12, 25),"points")), 
+  multiplot(beck_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25), "points")),
+            brusatte_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25), "points")), 
+            bapst_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25), "points")),
+            wright_s + theme(plot.margin=unit(c(5.5, 5.5, 12, 25), "points")), 
             beck_d, brusatte_d, bapst_d, 
-            wright_d + theme(plot.margin=unit(c(5.5, 5.5, 12, 5.5),"points")),
+            wright_d + theme(plot.margin=unit(c(5.5, 5.5, 12, 5.5), "points")),
             beck_n,
@@
             bapst_n,
-            wright_n + theme(plot.margin=unit(c(5.5, 5.5, 12, 5.5),"points")),
+            wright_n + theme(plot.margin=unit(c(5.5, 5.5, 12, 5.5), "points")),
             cols = 3, 
@@
                         "Relative disparity"),
-            labpos=list(c(0.5, 0.015), c(0.03, 0.5)))
+            labpos = list(c(0.5, 0.015), c(0.03, 0.5)))
 dev.off()

### 2017-12-15 analyses/06-table-wilcox.Rmd
Minor tidying of code

@@
 # Stick p values and W values together
-# Repalce p values with *
+# Replace p values with *
 p.values.function <- function(data){ 
@@
 
+## Save
 sink(file = "../manuscript/tables/raw_wilcox.tex")
@@
 
+## Save
 sink(file = "../manuscript/tables/raw_wilcox_appendix.tex")
@@
 ```
+
+These tables are LaTeX format and can be added into documents after a small amount of manipulation.

### 2017-12-17 manuscript/time-slice.tex
Abstract ideas

@@
 % 300 words max
+
+Disparity through time analysis are really popular and routinely applied on palaeobiological datasets.
+
+Although many methods exist to measure disparity, all more or less assessed (metrics, tests, etc...), only one method exist for looking at disparity through time: pooling data into time bins.
+
+This may lead to some problems down the line as a methodological bottleneck (i.e. most disparity-through-time analysis are, in the end, based on pooling data into bins).
+
+Here we propose a novel method to complement the disparity-through-time tool-kit: time-slicing.
+
+This method allows to sample disparity through time at any fixed point in time rather than binning data between two fixed points in time.
+
+Additionally, this method allows to use all the phylogenetic data available (tips, nodes \textit{and} branches) and specify the implied model of evolution between each time-slice (punctuated or gradual).
+
+We test this method on four example datasets and compared how it performs at common disparity-through-time analysis: the effect of mass extinction and the time at the maximal and minimal disparity.
+
+We found that the sampling effect has a remarkable effect on our interpretation of disparity-through-time analysis.
+
+Using multiple methods for assessing disparity-through-time (rather than only one!), we can have a better picture of what is happening.
 

### 2017-12-18 analyses/01-extract-data-for-analyses.Rmd
Order slugs so it  matches later analyses and remove bears
Minor tidying of code
Corrected age bug in Brusatte and Bapst

@@
 
-## The latest version of Claddis (on development on my own branch)
+## The Thomas' Claddis development version
+## (contains faster AncStateEstMatrix and MorphDistMatrixFast functions)
 install_github("TGuillerme/Claddis")
@@
 ## Not needed if running all the code at once
-## matrix_nodes <- readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+matrix_nodes <- readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 
@@
 
+Brusatte et al 2014 Current Biol
+-----------
+
+```{r}
+slug <- "Brusatte2014"
+
+## Loading the data and tree
+matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
+tree <- read.nexus(paste0("../data/trees/", slug, ".nex"))
+
+# Replace zero branches with minimum branch length
+tree$edge.length[which(tree$edge.length == 0)] <- 
+  min(tree$edge.length[-which(tree$edge.length == 0)])
+
+## Remove taxa with non overlapping data (see note about Gower distances above)
+remove_Brusatte <- c("Shanag_ashile", "Atrociraptor_marshalli", "Proceratosaurus_bradleyi",     "Incisivosaurus_gauthieri", "Enigmosaurus", "Nanshiungosaurus_brevispinus", "Xixiasaurus", "Tsaagan_mangas", "Mirischia", "Pedopenna", "Suzhousaurus", "Juratyrant", "Vorona", "Bonapartenykus", "Teratophoneus", "Gobipteryx", "Songlingornis", "Liaoningornis_longidigitu", "Achillesaurus")
+
+tree <- drop.tip(tree, remove_Brusatte)
+
+# Tidy the tree and the data
+tree <- makeNodeLabel(tree, method = "number", prefix = "n")
+cleaned_data <- clean.data(matrix$matrix, tree)
+matrix$matrix <- cleaned_data$data
+tree <- cleaned_data$tree
+
+## Loading the FADLAD
+FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+
+## Remove the taxa not present in the tree but in the FADLAD
+absent_in_tree <- clean.data(FADLAD, tree)$dropped_rows
+if(!any(is.na(absent_in_tree))) {
+    FADLAD <- FADLAD[-match(absent_in_tree, rownames(FADLAD)),]
+}
+
+## Getting the tree ages
+tree_ages <- tree.age(tree)
+
+## Scaling with the latest FADLAD
+tree$root.time <- max(tree_ages[,1] + min(FADLAD[,2]))
+
+## Removed taxa, i.e. those in the tree but not the matrix and vice versa
+as.vector(na.omit(c(cleaned_data$dropped_tips, cleaned_data$dropped_rows)))
+
+## Saving the tree
+write.nexus(tree, file = paste0("../data/processed/tree.", slug ,".tre"))
+
+## Estimating the ancestral states (with a uncertainty threshold of 0.95)
+# matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
+#                                  uncertainty.threshold = 0.95)
+# saveRDS(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+
+## Add ancestral states to nodes
+matrix_nodes <- readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+row.names(matrix_nodes) <- tree$node.label
+matrix$matrix <- rbind(matrix$matrix, matrix_nodes)
+
+## Constructing the distance matrix
+matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
+
+## Trim taxa with non overlapping data
+trim <- TrimMorphDistMatrix(matrix_dist)
+matrix_dist <- trim$dist.matrix
+
+## Which taxa were removed?
+as.vector(trim$removed.taxa)
+
+## Save
+saveRDS(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+```
+
+Bapst et al 2016 Biol Letters
+-----------
+
+```{r}
+slug <- "Bapst2016"
+
+## Loading the data and tree
+matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
+tree <- read.nexus(paste0("../data/trees/", slug, ".nex"))
+
+## Replace zero branches with minimum branch length
+tree$edge.length[which(tree$edge.length == 0)] <- 
+  min(tree$edge.length[-which(tree$edge.length == 0)])
+
+## Remove taxa with non overlapping data (see note about Gower distances above)
+remove_Bapst <- c("Hagryphus_giganteus", "Atrociraptor_marshalli", "IGM100_1015_UndesDromaeosaurid", "Dromaeosaurus_albertensis", "Incisivosaurus_gauthieri", "Deinocheirus_mirificus", "Therizinosaurus_cheloniformis", "Anserimimus_planinychus", "Elmisaurus_rarus")
+
+tree <- drop.tip(tree, remove_Bapst)
+
+# Tidy the tree and the data
+tree <- makeNodeLabel(tree, method = "number", prefix = "n")
+cleaned_data <- clean.data(matrix$matrix, tree)
+matrix$matrix <- cleaned_data$data
+tree <- cleaned_data$tree
+
+## Loading the FADLAD
+FADLAD <- read.csv(paste0("../data/FADLAD/", slug, ".csv"), row.names = 1, header = TRUE)
+
+## Remove the taxa not present in the tree but in the FADLAD
+absent_in_tree <- clean.data(FADLAD, tree)$dropped_rows
+if(!any(is.na(absent_in_tree))) {
+    FADLAD <- FADLAD[-match(absent_in_tree, rownames(FADLAD)),]
+}
+
+## Getting the tree ages
+tree_ages <- tree.age(tree)
+
+## Scaling with the latest FADLAD
+tree$root.time <- max(tree_ages[,1] + min(FADLAD[,2]))
+
+## Removed taxa, i.e. those in the tree but not the matrix and vice versa
+as.vector(na.omit(c(cleaned_data$dropped_tips, cleaned_data$dropped_rows)))
+
+## Saving the tree
+write.nexus(tree, file = paste0("../data/processed/tree.", slug ,".tre"))
+
+## Estimating the ancestral states (with a uncertainty threshold of 0.95)
+# matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
+#                                   uncertainty.threshold = 0.95)
+# saveRDS(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+
+## Add ancestral states to nodes
+matrix_nodes <- readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+row.names(matrix_nodes) <- tree$node.label
+matrix$matrix <- rbind(matrix$matrix, matrix_nodes)
+
+## Constructing the distance matrix
+matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
+saveRDS(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
+```
+
 Wright 2017
@@
 ## Estimating the ancestral states (with a uncertainty threshold of 0.95)
-matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
-                                  uncertainty.threshold = 0.95)
-saveRDS(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+# matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
+#                                   uncertainty.threshold = 0.95)
+# saveRDS(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 
 ## Add ancestral states to nodes
-## matrix_nodes <- readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+matrix_nodes <- readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 row.names(matrix_nodes) <- tree$node.label
@@
 ```
-
-Bapst et al 2016 Biol Letters
------------
-
-```{r}
-slug <- "Bapst2016"
-
-## Loading the data and tree
-matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
-tree <- read.nexus(paste0("../data/trees/", slug, ".nex"))
-
-## Replace zero branches with minimum branch length
-tree$edge.length[which(tree$edge.length == 0)] <- 
-  min(tree$edge.length[-which(tree$edge.length == 0)])
-
-## Remove taxa with non overlapping data (see note about Gower distances above)
-remove_Bapst <- c("Hagryphus_giganteus", "Atrociraptor_marshalli", "IGM100_1015_UndesDromaeosaurid", "Dromaeosaurus_albertensis", "Incisivosaurus_gauthieri", "Deinocheirus_mirificus", "Therizinosaurus_cheloniformis", "Anserimimus_planinychus", "Elmisaurus_rarus")
-
-tree <- drop.tip(tree, remove_Bapst)
-
-# Tidy the tree and the data
-tree <- makeNodeLabel(tree, method = "number", prefix = "n")
-tree$root.time <- max(tree.age(tree)[, 1])
-cleaned_data <- clean.data(matrix$matrix, tree)
-matrix$matrix <- cleaned_data$data
-tree <- cleaned_data$tree
-
-## Removed taxa, i.e. those in the tree but not the matrix and vice versa
-as.vector(na.omit(c(cleaned_data$dropped_tips, cleaned_data$dropped_rows)))
-
-## Saving the tree
-write.nexus(tree, file = paste0("../data/processed/tree.", slug ,".tre"))
-
-## Estimating the ancestral states (with a uncertainty threshold of 0.95)
-matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
-                                  uncertainty.threshold = 0.95)
-saveRDS(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
-
-## Add ancestral states to nodes
-## matrix_nodes <- readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
-row.names(matrix_nodes) <- tree$node.label
-matrix$matrix <- rbind(matrix$matrix, matrix_nodes)
-
-## Constructing the distance matrix
-matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
-saveRDS(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
-```
-
-Heath et al bears
------------
-
-```{r}
-slug <- "bears"
-
-## Loading the data and tree
-matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
-tree <- read.nexus(paste0("../data/trees/", slug, ".nex"))
-
-# Replace zero branches with minimum branch length
-tree$edge.length[which(tree$edge.length == 0)] <- 
-  min(tree$edge.length[-which(tree$edge.length == 0)])
-
-# Tidy the tree and the data
-tree <- makeNodeLabel(tree, method = "number", prefix = "n")
-tree$root.time <- max(tree.age(tree)[, 1])
-cleaned_data <- clean.data(matrix$matrix, tree)
-matrix$matrix <- cleaned_data$data
-tree <- cleaned_data$tree
-
-## Removed taxa, i.e. those in the tree but not the matrix and vice versa
-as.vector(na.omit(c(cleaned_data$dropped_tips, cleaned_data$dropped_rows)))
-
-## Saving the tree
-write.nexus(tree, file = paste0("../data/processed/tree.", slug ,".tre"))
-
-## Estimating the ancestral states (with a uncertainty threshold of 0.95)
-matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
-                                  uncertainty.threshold = 0.95)
-saveRDS(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
-
-## Add ancestral states to nodes
-## matrix_nodes <- readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
-row.names(matrix_nodes) <- tree$node.label
-matrix$matrix <- rbind(matrix$matrix, matrix_nodes)
-
-## Constructing the distance matrix
-matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
-saveRDS(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
-```
-
-Brusatte et al 2014 Current Biol
------------
-
-```{r}
-slug <- "Brusatte2014"
-
-## Loading the data and tree
-matrix <- ReadMorphNexus(paste0("../data/matrices/", slug, ".nex"))
-tree <- read.nexus(paste0("../data/trees/", slug, ".nex"))
-
-# Replace zero branches with minimum branch length
-tree$edge.length[which(tree$edge.length == 0)] <- 
-  min(tree$edge.length[-which(tree$edge.length == 0)])
-
-## Remove taxa with non overlapping data (see note about Gower distances above)
-remove_Brusatte <- c("Shanag_ashile", "Atrociraptor_marshalli", "Proceratosaurus_bradleyi",     "Incisivosaurus_gauthieri", "Enigmosaurus", "Nanshiungosaurus_brevispinus", "Xixiasaurus", "Tsaagan_mangas", "Mirischia", "Pedopenna", "Suzhousaurus", "Juratyrant", "Vorona", "Bonapartenykus", "Teratophoneus", "Gobipteryx", "Songlingornis", "Liaoningornis_longidigitu", "Achillesaurus")
-
-tree <- drop.tip(tree, remove_Brusatte)
-
-# Tidy the tree and the data
-tree <- makeNodeLabel(tree, method = "number", prefix = "n")
-tree$root.time <- max(tree.age(tree)[, 1])
-cleaned_data <- clean.data(matrix$matrix, tree)
-matrix$matrix <- cleaned_data$data
-tree <- cleaned_data$tree
-
-## Removed taxa, i.e. those in the tree but not the matrix and vice versa
-as.vector(na.omit(c(cleaned_data$dropped_tips, cleaned_data$dropped_rows)))
-
-## Remove
-
-## Saving the tree
-write.nexus(tree, file = paste0("../data/processed/tree.", slug ,".tre"))
-
-## Estimating the ancestral states (with a uncertainty threshold of 0.95)
-matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
-                                  uncertainty.threshold = 0.95)
-saveRDS(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
-
-## Add ancestral states to nodes
-## matrix_nodes <- readRDS(paste0("../data/processed/ancestral_states.", slug ,".Rda"))
-row.names(matrix_nodes) <- tree$node.label
-matrix$matrix <- rbind(matrix$matrix, matrix_nodes)
-
-## Constructing the distance matrix
-matrix_dist <- MorphDistMatrixFast(matrix, distance = "Gower")
-
-## Trim taxa with non overlapping data
-trim <- TrimMorphDistMatrix(matrix_dist)
-matrix_dist <- trim$dist.matrix
-
-## Which taxa were removed?
-as.vector(trim$removed.taxa)
-
-## Save
-saveRDS(matrix_dist, file = paste0("../data/processed/distance_matrix.", slug ,".Rda"))
-```

### 2017-12-18 analyses/09-appendix-trees-figures.Rmd
Strap plots

@@
 Load the packages
+
 ```{r}
 library(strap)
+library(dispRity)
+library(ape)
 ```
@@
 slug <- "Beck2014"
-#slug <- "Brusatte2014"
-#slug <- "Bapst2016"
-#slug <- "Wright2017"
+# slug <- "Brusatte2014"
+# slug <- "Bapst2016" #ok
+# slug <- "Wright2017" #ok
+
+## Loading the tree
+tree <- read.nexus(paste0("../data/processed/tree.", slug, ".tre"))
 
@@
 
-## Loading the tree
-tree <- read.nexus(paste0("../data/processed/tree.", slug, ".tre"))
-## Add root age (max tree age date)
-## Remember that for Wright this is 485.4
-tree$root.time <- max(tree.age(tree)[, 1])
+## Remove the taxa not present in the tree but in the FADLAD
+absent_in_tree <- clean.data(FADLAD, tree)$dropped_rows
+if(!any(is.na(absent_in_tree))) {
+    FADLAD <- FADLAD[-match(absent_in_tree, rownames(FADLAD)),]
+}
+
+## Getting the tree ages
+tree_ages <- tree.age(tree)
+
+## Scaling with the latest FADLAD
+tree$root.time <- max(tree_ages[,1] + min(FADLAD[,2]))
+
+## Correct the root age for Wright2017
+tree$root.time <- ifelse(slug == "Wright2017", 485.4, tree$root.time)
+
+## Updating the tree age
+tree_ages <- tree.age(tree)
+
+## Updating the FADLAD (if missing tips)
+if(nrow(FADLAD) !=  Ntip(tree)) {
+    ## Selecting the missing taxa
+    missing_taxa <- which(is.na(match(tree_ages$elements[1:Ntip(tree)], rownames(FADLAD))))
+    ## Creating a FADLAD table for the missing taxa (FAD = LAD = tip age)
+    FADLAD_tmp <- data.frame(FAD = tree_ages$ages[missing_taxa], LAD = tree_ages$ages[missing_taxa],
+                             row.names = tree_ages$elements[missing_taxa])
+    ## Combining both.data frames
+    FADLAD <- rbind(FADLAD, FADLAD_tmp)
+}
 ```
@@
 Plot the trees and save as PDF
+
 ```{r}
 pdf(paste0("../manuscript/figures/fig-tree-", slug, "-appendix.pdf"), width = 12, height = 12) 
+## Adjust plot sizes (for offset labels with e.g.: x.lim = c(350, 485.4) and cex.tip = 0.5))
 geoscalePhylo(tree = tree, ages = FADLAD)

### 2017-12-18 analyses/02-time-slice-analyses.Rmd
Minor tidying of code
Reran full data

@@
     ii. Where the *number* of bins is equal to the number of stratigraphic periods.
-    iii. Where the *number* of bins is either 2, 5, 10, 15, or 20.
 
@@
     ii. Where the *number* of slices is equal to the number of stratigraphic periods.
-    iii. Where the *number* of slices is either 2, 5, 10, 15, or 20.
 
@@
 tree <- read.nexus(paste0("../data/processed/tree.", slug, ".tre"))
-## Add root age (max tree age date)
-tree$root.time <- max(tree.age(tree)[, 1])
+
+## Remove the taxa not present in the tree but in the FADLAD
+absent_in_tree <- clean.data(FADLAD, tree)$dropped_rows
+if(!any(is.na(absent_in_tree))) {
+    FADLAD <- FADLAD[-match(absent_in_tree, rownames(FADLAD)),]
+}
+
+## Getting the tree ages
+tree_ages <- tree.age(tree)
+
+## Scaling with the latest FADLAD
+tree$root.time <- max(tree_ages[,1] + min(FADLAD[,2]))
 

### 2017-12-18 analyses/12-appendix-disparity-tables.Rmd
added table compiler (96 pages of appendix!)

---
title: "Disparity tables for the appendix q"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Building the disparity tables for each dataset

```{r}
library(dispRity)
library(xtable)

## Function for only selecting the first occurrence of each level (for fancy tables)
level.selector <- function(column, data) {
    ## Selecting the data
    data_col <- data[, column]

    ## Selecting the levels
    all_levels <- levels(as.factor(data_col))

    ## Replacing them in the column
    for(one_level in 1:length(all_levels)) {
        ## Selecting the levels occurrences
        level_occurrences <- data_col %in% all_levels[one_level]
        ## Selecting the first of each
        first_level_occurrences <- sapply(seq(2:length(level_occurrences[-1])), function(X, test)
            return(ifelse((level_occurrences[X] == TRUE && level_occurrences[X-1] == FALSE)
                , TRUE, FALSE)), level_occurrences)

        ## Replacing the values in data_col
        data_col[which(first_level_occurrences)] <- paste0(data_col[which(first_level_occurrences)],
                                                           "@1st")
    }

    ## Remove the non first entries
    data_col[-grep("@1st", data_col)] <- ""
    data_col <- gsub("@1st", "", data_col)

    return(data_col)
}

## Rounding slices/bins values
round.slices <- function(slice, digits = 3) {
    options(warn = -1)
    ## Convert into numeric
    slice_value <- as.numeric(slice)
    options(warn = 0)

    if(is.na(slice_value)) {
        ## Is the value simply NA or a bin
        if(grep("-", slice) > 0) {
            ## Separating both values
            bin_values <- unlist(strsplit(slice, split = " - "))
            ## Rounding the values
            bin_values <- round(as.numeric(bin_values), digits = digits)
            ## Recombining them
            return(paste(bin_values, collapse = " - "))

        } else {
            ## Slice value is a real NA
            return(as.character(NA))
        }
    } else {
        ## Slice is a slice
        return(as.character(round(as.numeric(slice_value), digits = digits)))
    }
}

```

## Generate the raw results tables for each datasets

```{r, warn = FALSE}
slugs <- c("Beck2014", "Brusatte2014", "Bapst2016", "Wright2017")

for(one_slug in 1:length(slugs)) {

    slug <- slugs[one_slug]

    ## Loading the data
    out1 <- readRDS(file = paste0("../outputs/age.", slug, ".Rda"))
    out2 <- readRDS(file = paste0("../outputs/epoch.", slug, ".Rda"))

    ## bind all the data together
    results <- rbind(do.call(rbind, out1$results), do.call(rbind, out2$results))

    ## Re-order the columns
    results <- results[,c(10,11,9, 1:8)]
    results[,1:3] <- apply(results[,1:3], 2, as.character)

    ## Empty repetition cells (for aesthetics)
    ## Stratigraphy
    results[, 1] <- level.selector(1, results)
    ## Bin_type
    results[, 2] <- level.selector(2, results)
    ## Model
    results[, 3] <- level.selector(3, results)

    ## Correcting the first slice
    results[1, c(1,2,3)] <- c("Ages", "stratigraphy", "equalbins")

    ## Rounding the slices
    results[, 4] <- sapply(results[, 4], round.slices, digits = 2)

    ## Numbering the columns
    results[, 5:11] <- apply(results[, 5:11], 2, as.numeric)

    ## Table caption
    table_caption <- paste("Disparity measurement (sum of variance) for", slug, "dataset")

    ## Save the table (as longtable)
    write(print(xtable(results, caption = table_caption, digits = 2),
                tabular.environment = "longtable", include.rownames = FALSE, floating = FALSE),
          file = paste0("../manuscript/tables/disparity_measurement-", slug, ".tex"))
}
```

### 2017-12-18 analyses/08-figure-extinction.Rmd
Updated extinction effect (remove Brusatte and Bapst)

--- 2017-12-11 analyses/10-figure-extinction.Rmd
@@
 slug <- "Beck2014"
-# slug <- "Brusatte2014"
-# slug <- "Bapst2016"
 # slug <- "Wright2017"
@@
 ## Names of the datasets
-slugs <- as.list(c("Beck2014", "Brusatte2014", "Bapst2016", "Wright2017"))
+slugs <- as.list(c("Beck2014", "Wright2017"))
 
@@
 ## Plot all the results
-pdf("../manuscript/figures/fig-extinction-effect.pdf", width = 12, height = 4)  
+pdf("../manuscript/figures/fig-extinction-effect.pdf", width = 8, height = 4)  
 multi.plot.extinction(data_list, type = c("age", "epoch"), data.names = unlist(slugs), cols = colours)

### 2017-12-18 functions/figures_fun.R

@@ -352,11 +352,11 @@ multi.plot.extinction <- function(data, type, cols = c("blue", "orange", "white"
         par(mar = c(0, 6, 4, 0)) #c(bottom, left, top, right)
         plot.extinction(data[[1]], type = type[1], col = cols, yaxis = TRUE, xaxis2 = TRUE, main = data.names[[1]], ...)
 
-        ## Other plots (first row)
-        for(slug in 2:(length(data)-1)) {
-            par(mar = c(0, 0, 4, 0)) #c(bottom, left, top, right)
-            plot.extinction(data[[slug]], type = type[1], col = cols, main = data.names[[slug]], xaxis2 = TRUE, ...)
-        }
+        # ## Other plots (first row)
+        # for(slug in 2:(length(data)-1)) {
+        #     par(mar = c(0, 0, 4, 0)) #c(bottom, left, top, right)
+        #     plot.extinction(data[[slug]], type = type[1], col = cols, main = data.names[[slug]], xaxis2 = TRUE, ...)
+        # }
 
         ## Last plot (first row)
         par(mar = c(0, 0, 4, 3)) #c(bottom, left, top, right)
@@ -366,11 +366,11 @@ multi.plot.extinction <- function(data, type, cols = c("blue", "orange", "white"
         par(mar = c(4, 6, 0, 0)) #c(bottom, left, top, right)
         plot.extinction(data[[1]], type = type[2], col = cols, yaxis = TRUE, xaxis = TRUE, ...)
 
-        ## Other plots (second row)
-        for(slug in 2:(length(data)-1)) {
-            par(mar = c(4, 0, 0, 0)) #c(bottom, left, top, right)
-            plot.extinction(data[[slug]], type = type[2], col = cols, xaxis = TRUE, ...)
-        }
+        # ## Other plots (second row)
+        # for(slug in 2:(length(data)-1)) {
+        #     par(mar = c(4, 0, 0, 0)) #c(bottom, left, top, right)
+        #     plot.extinction(data[[slug]], type = type[2], col = cols, xaxis = TRUE, ...)
+        # }
 
         ## Last plot (first row)
         par(mar = c(4, 0, 0, 3)) #c(bottom, left, top, right)

### 2017-12-18 manuscript/time-slice.tex
Finish intro

@@
 \title{Time for a rethink: time sub-sampling methods in disparity-through-time analyses}
-%TG: Great title first part! How about changing the other part to 
-%Time for a rethink: different ways of slicing through the stratigraphic cake leads to different disparity through time results.
 
@@
 
-Disparity through time analysis are really popular and routinely applied on palaeobiological datasets.
+Disparity-through-time analyses are popular and routinely applied to palaeobiological datasets.
 
@@
 
-Additionally, this method allows to use all the phylogenetic data available (tips, nodes \textit{and} branches) and specify the implied model of evolution between each time-slice (punctuated or gradual).
+Additionally, this method allows to use all the phylogenetic data available (tips, nodes
+
+\textit{and} branches) and specify the implied model of evolution between each time-slice (punctuated or gradual).
 
@@
 \section{Introduction}
-%TG: Plan idea
-%TG: §1: disparity-through-time really cool and well used
-%TG: §2: caveats with both disparity and through-time part
-%TG: §3: specific ways to look through-time and their problems
-%TG: §4: our kind of solution
-
-Disparity-through-time analyes are common in palaeontology since the 90s \citep{gould1991disparity,briggs1992morphological,Wills1994,Foote01071994}.
+Disparity-through-time analyses are common in palaeontology \citep{gould1991disparity,briggs1992morphological,Wills1994,Foote01071994}.
 They reveal how the morphological diversity of clades has changed through time, and allow us to make inferences about the breadth of ecological niches extinct taxa occupied \citep{foote1997evolution}.
@@
 
-Disparity-through-time analyes have two main analysis components: calculating disparity, and creating time sub-subsets of the data. 
+Disparity-through-time analyses have two main analysis components: calculating disparity, and creating time sub-subsets of the data. 
 Here we focus on the latter.
-The nature of disparity (i.e. a diversity metric), means it cannot be calculated using a single individual, so some way of sampling many taxa in each sub-subset is required.
+The nature of disparity (i.e. it is a diversity metric), means it cannot be calculated using a single individual, so some way of sub-sampling taxa is required.
 Changes in disparity-through-time are generally investigated by calculating the disparity of taxa present during specific time intervals or time bins \citep[e.g][]{cisneros2010,prentice2011,Hughes20082013,hopkinsdecoupling2013,bentonmodels2014,bensonfaunal2014}.
-These time bins are usually defined based on stratigraphy \citep[e.g.][]{cisneros2010,prentice2011,Hughes20082013,bentonmodels2014} but can also be arbitrarily chosen time bins of equal duration \citep{Butler2012,hopkinsdecoupling2013,bensonfaunal2014}.
+These time bins are usually defined based on stratigraphy \citep[e.g.][]{cisneros2010,prentice2011,Hughes20082013,bentonmodels2014} but can also be arbitrarily chosen time bins of equal (or approximately equal) duration \citep{Butler2012,hopkinsdecoupling2013,bensonfaunal2014}.
 However, this approach has several limitations.
@@
 First, time bins defined by stratigraphy are not of equal size, biasing higher disparity towards longer stratigraphic periods. 
-This can be dealt with using rarefaction methods, i.e.
-repeating the analysis while re-sampling the taxa to have the same amount of taxa in each bin (e.g.
-using bootstrapping with limited re-sampling methods).
-This can, however, lead to very large confidence intervals when there are stratigraphic periods with very few species.
+This can be dealt with using rarefaction methods, i.e. repeating the analysis while re-sampling the taxa to have the same amount of taxa in each bin (e.g. using bootstrapping with limited resampling).
+This can, however, lead to large confidence intervals when there are stratigraphic periods with few species.
 Other studies split large time bins so they are of roughly equal size, but this is often an \textit{ad hoc} procedure that can introduce more bias depending on where bins are split.
-Second, all time binning approaches (whether bins are equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring \textit{between} intervals, rather than also allowing for gradual changes \textit{within} intervals \citep[shown as being quiet common in the fossil record;][]{Hunt21042015}.
+Second, all time binning approaches (whether bins are equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring \textit{between} intervals, rather than also allowing for gradual changes \textit{within} intervals \citep[fairly common in the fossil record;][]{Hunt21042015}.
+Third, when investigating changes in disparity due to events at a specific time point (e.g.
+a mass extinction), time bins may have not have high enough resolution to resolve changes at the event, for example if time bins are every 20 million years it may be hard to capture the effects of an event five million years into the bin.
 Finally, time bin analyses are often limited by the number of taxa in each bin.
@@
 correlations of disparity with hypothesised drivers of morphological evolution, are not possible.
-%TG: Not sure where in this paragraph but it'll be good to add this point:
-%Also, in the case of looking at changes in disparity due to events specific in time (e.g.
-a mass extinction), time bins method leave us with a low temporal resolution: stratigraphic bins are limited to global events that defined the bins boundaries and time bins are limited by the duration of each bin (e.g.
-if time bins are every 20 Mya but an event occurs at time $t$ and $t+5$ Mya, it might not be possible to capture its effects).
 
-To address these issues, we propose a ``time-slicing'' approach that takes advantage of the wealth of paleontological datasets which now have associated phylogenies. 
+To address these issues, we propose a ``time-slicing'' approach that takes advantage of the wealth of palaeontological datasets which now have associated phylogenies. 
 Time-slicing uses a phylogenetic tree and considers subsets of taxa at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
@@
 We calculate disparity-through-time for each dataset using a range of time binning and time-slicing methods, and then compare these approaches with respect to the relative disparities calculated, but also with how the different approaches influence biological conclusions.
-We find that XXX
+We find that the choice of time sub-sampling method can have profound effects on the conclusions of disparity-through-time analyses.
 

### 2017-12-20 analyses/01-extract-data-for-analyses.Rmd
Change dispRity version to release version

@@
 ## The latest version of dispRity 
-install_github("TGuillerme/dispRity")
+install_github("TGuillerme/dispRity@v0.5")
 library(dispRity) 
@@
 ## Estimating the ancestral states (with a uncertainty threshold of 0.95)
-# matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
-#                                  uncertainty.threshold = 0.95)
-# saveRDS(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
+                                 uncertainty.threshold = 0.95)
+saveRDS(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 
@@
 ## Estimating the ancestral states (with a uncertainty threshold of 0.95)
-# matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
-#                                   uncertainty.threshold = 0.95)
-# saveRDS(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
+matrix_nodes <- AncStateEstMatrix(matrix, tree, estimate.allchars = TRUE,
+                                   uncertainty.threshold = 0.95)
+saveRDS(matrix_nodes, file = paste0("../data/processed/ancestral_states.", slug ,".Rda"))
 

### 2017-12-20 analyses/04-table-datasets.Rmd
Minor tidying of code
Remove extinctions for Brusatte and Bapst data

@@
 ## Mass extinction
-mass <- c("66 (K-Pg)", "66 (K-Pg)", "66 (K-Pg)", "443 (O-S)")
+mass <- c("66 (K-Pg)", "NA", "NA", "443 (O-S)")
 
@@
 ## Table caption
-table_caption <- "Details of the datasets used in this study."
+table_caption <- "Details of the datasets used in this study.
+Note that although Brusatte2014 and Bapst2016 cross the Cretaceaous Palaeogene (K-Pg) mass extinction there are not enough taxa post K-Pg to investigate the effects of the mass extinction on disparity."
 
@@
 
-## Time subsets table
+This produces a nicely formatted LaTeX table. Some edits will need to be made before it is incorporated into a document but it contains all the basic information required.

### 2017-12-20 analyses/01-extract-data-for-analyses.Rmd
Add author info

@@
 title: "Extracting data for time slice analyses"
-author: "Natalie Cooper and Thomas Guillerme"
+author: "Thomas Guillerme and Natalie Cooper"
 date: "`r Sys.Date()`"

### 2017-12-20 analyses/12-appendix-disparity-tables.Rmd
Add author info

@@
 ---
-title: "Disparity tables for the appendix q"
+title: "Disparity tables for the Appendix"
 author: "Thomas Guillerme"

### 2017-12-20 manuscript/time-slice.tex
Minor edits to intro

@@
 First, time bins defined by stratigraphy are not of equal size, biasing higher disparity towards longer stratigraphic periods. 
-This can be dealt with using rarefaction methods, i.e. repeating the analysis while re-sampling the taxa to have the same amount of taxa in each bin (e.g. using bootstrapping with limited resampling).
+This can be dealt with using rarefaction methods, i.e. repeating the analysis while resampling the taxa to have the same number of taxa in each bin (e.g. using bootstrapping with limited resampling).
 This can, however, lead to large confidence intervals when there are stratigraphic periods with few species.
 Other studies split large time bins so they are of roughly equal size, but this is often an \textit{ad hoc} procedure that can introduce more bias depending on where bins are split.
-Second, all time binning approaches (whether bins are equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring \textit{between} intervals, rather than also allowing for gradual changes \textit{within} intervals \citep[fairly common in the fossil record;][]{Hunt21042015}.
+Second, all time binning approaches (whether bins are equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring \textit{between} intervals, rather than also allowing for gradual changes \textit{within} intervals \citep[a pattern that is fairly common in the fossil record;][]{Hunt21042015}.
 Third, when investigating changes in disparity due to events at a specific time point (e.g.
@@
 Here we present our time-slicing methods using four datasets taken from the literature.
-We calculate disparity-through-time for each dataset using a range of time binning and time-slicing methods, and then compare these approaches with respect to the relative disparities calculated, but also with how the different approaches influence biological conclusions.
+We calculate disparity-through-time for each dataset using a range of time binning and time-slicing methods, and then compare these approaches with respect to the relative disparities calculated, but also investigate how the different approaches influence biological conclusions.
 We find that the choice of time sub-sampling method can have profound effects on the conclusions of disparity-through-time analyses.
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch.pdf}
     \caption[Timing of peak disparity for four example datasets.]

### 2017-12-20 manuscript/tables/raw_datasets.tex
Edit to caption and extinctions

@@
-% latex table generated in R 3.4.2 by xtable 1.8-2 package
-% Wed Dec  6 16:09:02 2017
+% latex table generated in R 3.4.3 by xtable 1.8-2 package
+% Wed Dec 20 13:12:18 2017
 \begin{table}[ht]
@@
   Age range (MYA) & 171.786 - 0 & 102.535 - 0 & 141.216 - 0 & 485.4 - 372.2 \\ 
-  Mass extinction (MYA) & 66 (K-Pg) & 66 (K-Pg) & 66 (K-Pg) & 443 (O-S) \\ 
+  Mass extinction (MYA) & 66 (K-Pg) & NA & NA & 443 (O-S) \\ 
   Reference &  &  &  &  \\ 
@@
 \end{tabular}
-\caption{Details of the datasets used in this study.} 
+\caption{Details of the datasets used in this study.
+Note that although Brusatte2014 and Bapst2016 cross the Cretaceaous Palaeogene (K-Pg) mass extinction there are not enough taxa post K-Pg to investigate the effects of the mass extinction on disparity.}
 \end{table}

### 2017-12-20 manuscript/tables/table_datasets.tex
Edits to end of methods
Edit to caption and extinctions

--- 2017-12-06 manuscript/tables/table_datasets.tex
@@
-% latex table generated in R 3.4.2 by xtable 1.8-2 package
-% Wed Dec  6 16:09:02 2017
-\begin{table}[ht]
+% latex table generated in R 3.4.3 by xtable 1.8-2 package
+% Wed Dec 20 13:12:18 2017
+\begin{table}[!htbp]
 \centering
@@
   Age range (MYA) & 171.786 - 0 & 102.535 - 0 & 141.216 - 0 & 485.4 - 372.2 \\ 
-  Mass extinction (MYA) & 66 (K-Pg) & 66 (K-Pg) & 66 (K-Pg) & 443 (O-S) \\ 
+  Mass extinction (MYA) & 66 (K-Pg) & NA & NA & 443 (O-S) \\ 
   Reference & \cite{beckancient2014} & \cite{brusatte2014gradual} & \cite{bapst2016topology} & \cite{wright2017bayesian} \\ 
-  Data reference &  CHECK & \cite{dryad_84t75} & \cite{dryad_n2g80} &  \cite{dryad_6hb7j} \\ 
+  Data reference &  \cite{beckancient2014} & \cite{dryad_84t75} & \cite{dryad_n2g80} &  \cite{dryad_6hb7j} \\ 
    \hline
 \end{tabular}
-\caption{Details of the datasets used in this study.} 
+\caption{Details of the datasets used in this study.
+Note that although Brusatte2014 and Bapst2016 cross the Cretaceaous Palaeogene (K-Pg) mass extinction there are not enough taxa post K-Pg to investigate the effects of the mass extinction on disparity.}
+\label{table:datasets}  
 \end{table}

### 2017-12-20 analyses/04-table-datasets.Rmd
Add author info
Typo

@@
 title: "Dataset table"
-author: "Thomas Guillerme"
+author: "Thomas Guillerme and Natalie Cooper"
 date: "`r Sys.Date()`"
@@
 table_caption <- "Details of the datasets used in this study.
-Note that although Brusatte2014 and Bapst2016 cross the Cretaceaous Palaeogene (K-Pg) mass extinction there are not enough taxa post K-Pg to investigate the effects of the mass extinction on disparity."
+Note that although Brusatte2014 and Bapst2016 cross the Cretaceous Palaeogene (K-Pg) mass extinction there are not enough taxa post K-Pg to investigate the effects of the mass extinction on disparity."
 

### 2017-12-20 manuscript/tables/table_datasets.tex
Typo

@@
 \caption{Details of the datasets used in this study.
-Note that although Brusatte2014 and Bapst2016 cross the Cretaceaous Palaeogene (K-Pg) mass extinction there are not enough taxa post K-Pg to investigate the effects of the mass extinction on disparity.}
+Note that although Brusatte2014 and Bapst2016 cross the Cretaceous Palaeogene (K-Pg) mass extinction there are not enough taxa post K-Pg to investigate the effects of the mass extinction on disparity.}
 \label{table:datasets}  

### 2017-12-20 manuscript/time-slice.tex
Edits to end of methods

@@
 \runninghead{Time sub-samples in disparity-through-time analyses}
-\keywords{time bin, time slice, disparity} 
+\keywords{time bin, time-slice, disparity} 
 % up to 6
@@
 To address these issues, we propose a ``time-slicing'' approach that takes advantage of the wealth of palaeontological datasets which now have associated phylogenies. 
-Time-slicing uses a phylogenetic tree and considers subsets of taxa at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
+Time-slicing uses a phylogenetic tree and considers subsets of taxa at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time (a similar approach is outlined in \citep{halliday2016eutherian}).
 This results in even-sampling across time and permits us to define the underlying model of character evolution (punctuated or gradual).  
-Time-slicing also includes any element present in the phylogeny (branches, nodes and tips) at the time slice in question as part of the disparity calculation.
+Time-slicing also includes any element present in the phylogeny (branches, nodes and tips) at the time-slice in question as part of the disparity calculation.
 This allows us to measure disparity at time points where there are no sampled terminal taxa, and increases the sample size at each time point, making downstream analyses of the drivers of disparity much more feasible.
@@
 \label{overview-section}
-To test the different time sub-sampling methods, we followed the protocol below (Figure \ref{datasets}). 
-All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (\url{https://github.com/nhcooper123/time-slice}).
-%https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/details-of-specific-functions.html#time-slicing$
+To test the different time sub-sampling methods, we followed the protocol below (Figure \ref{fig:overview}). 
+All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (\url{https://github.com/nhcooper123/time-slice}.
+% NC: Add proper reference to this before publication.
 
@@
     \caption[Outline of the disparity-through-time pipeline]
-    {1) We use ancestral character estimation to infer nodal character states; 2) we measure the pairwise Gower distance between the tip character states and nodal character states; 3) we ordinate the distance matrix using principal coordinates analysis (PCoA); 4) we time sub-sample the PCoA matrix using time bins defined by stratigraphic periods, equally sized time bins and time slices (using six methods to estimate ordination scores for branches); and finally 5) we measure disparity-through-time for each of these methods.}
-    \label{overview}
+    {Outline of the disparity-through-time pipeline.
+1) We use ancestral character estimation to infer nodal character states; 2) we measure the pairwise Gower distance between the tip character states and nodal character states; 3) we ordinate the distance matrix using principal coordinates analysis (PCoA); 4) we time sub-sample the PCoA matrix using time bins defined by stratigraphic periods, equally sized time bins and time-slices (using six methods to estimate ordination scores for branches); and finally 5) we measure disparity-through-time for each of these methods.}
+    \label{fig:overview}
   \end{figure}
@@
   \input{tables/table_datasets.tex} 
-  \label{table:datasets}  
+  % Label in input or referencing doesn't work
 \end{landscape}
@@
 This function uses the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny \citep[based on the \texttt{rerootingMethod} function in \texttt{phytools};][]{phytools}.
-% Inapplicable and missing character for any taxon were treated as ambiguous characters (i.e. any possible observed state for the character).
-% For example, if a character had two observed states ($0$ and $1$) across all taxa, we attributed the multi-state ``$0$\&$1$" value to the taxon with missing data, representing an equal probability of being either $0$ or $1$.
-% This allows the ancestral node of a taxon with missing data to be estimated with no assumptions other than that the taxon has one of the observed character states.
-%TG: These three sentences can be removed: it's pretty standard procedure.
-I'm not sure if we should leave that here.
-It sounded fancy during PhD time but now it's actually a bit lame to write it like that.
-Plus there is the inapplicable problem that Claddis do not deal with...
+Inapplicable and missing characters for any taxon were treated as ambiguous characters (i.e. any possible observed state for the character).
+% NC: I've left this one bit to make it clear but removed the explanation (this was all just copied from your thesis)
 To prevent poor ancestral state reconstructions from biasing our results, especially when a lot of error is associated with the reconstruction, we only included ancestral state reconstructions with a scaled Likelihood $\geq$ $0.95$.
 Ancestral state reconstructions with scaled Likelihoods below this threshold were recoded as missing data (``?'').
-This allowed our results to be less dependent on the quality (or the absence thereof) of the ancestral state reconstruction methods, especially in parts of the datasets where data was sparse.
+This allowed our results to be less dependent on the quality (or the absence thereof) of the ancestral state reconstruction methods, especially in parts of the datasets where data were sparse.
 
@@
 I'm just thinking about pedantic-palaeo-reviewers...
+% NC: Good point. Let's leave it for now as I think this bit is nicely explained, but easy enough to edit if people don't like it.
 Mathematically, a cladisto-space is an $n$ dimensional object that summarises the cladistic distances between the taxa present in a cladistic matrix.
@@
 
-The disparity-through-time analyses were performed using the \texttt{dispRity} \texttt{R} package \citep{dispRity.v02}. %TG: With some hope, the MEE paper might be out at the same time as this one!
+Disparity-through-time analyses were performed using the \texttt{dispRity} \texttt{R} package \citep{dispRity.v02}. 
+% NC: Update reference here prior to publication
 
@@
 Where $\sigma^{2}{n_i}$ is the variance for the $n_i^{th}$ dimension ranging from $k$ to $k-2$ with $k$ being the number of taxa in the dataset.
-Note that there are still statistical issues with this metric (such as the co-variance between dimensions not being measured), but for the purposes of comparison with previous work we want to use a standard metric for these analyses.
+Note that there are still statistical issues with this metric (such as the co-variance between dimensions not being measured), but for the purposes of comparison with previous work we decided to use a standard metric for these analyses.
 
@@
 
-\paragraph{Time-slicing} 
+\paragraph{Time-slicing.} 
 \label{time_slicing}
@@
 %TG: Is it worth saying that Halliday and Goswami came up with a similar idea \citep{halliday2016eutherian} but that we came up with it first \cite{Guillerme2015} and we're the first to implement it \cite{dispRity.v02}?
+%NC: Yeah was going to ask about this - I've added a mention to the intro where I think it fits better, should we also cite the brusatte paper you said refers to it too?
 
@@
     In these models, the ordination scores of both the ancestor and descendant contribute to the disparity calculation.
-    For example, using the ``gradual.split'' model, if the slice occurs in the third quarter of a branch joining node A to node/tip B (75\% of the total branch length), after bootstrapping, the disparity results will be based on 25\% of the data from A and 75\% of the data from B.
+    For example, using the ``gradual splits'' model, if the slice occurs in the third quarter of a branch joining node A to node/tip B (75\% of the total branch length), after bootstrapping, the disparity results will be based on 25\% of the data from A and 75\% of the data from B.
     Because of the probabilistic nature of these models, they are only meaningful when calculating disparity from bootstrapped datasets.
@@
 It is important to note that the time-slicing method is not an ancestral states estimation method \textit{per se}.
-This method does not estimate values along a branch applying a model (\textit{c.f.} methods described for reconstructing ancestral states in the preparing data for disparity-through-time section above) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the models described above.
-This allows the method to be used in post-ordination analysis where the data used in each time slice is data already present in the cladisto-space.
+This method does not estimate values along a branch applying a model (\textit{c.f.} methods described for ancestral character estimation in the ``Preparing the data for disparity-through-time analysis'' section above) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the methods described above.
+This allows the method to be used in post-ordination analysis where the data used in each time-slice is data already present in the cladisto-space.
 In other words, this method does not require a re-ordination of the cladisto-space every time a slice goes through a branch, thus allowing the properties of the cladisto-space (e.g.
-distance between species, variance per axis, etc.) to remain constant.
+distance between species, variance of each axis, etc.) to remain constant.
 
@@
     We calculated disparity for the taxa in each stratigraphic period (age or epoch). 
-    To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time bin by randomly resampling with replacement a new sub-sample of taxa from the observed taxa in the bin 1000 times.
+    To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time bin by randomly resampling with replacement a new sub-sample of taxa from the observed taxa in the bin 100 times.
     %TG: actually in the code the data is bootstrapped 100 times only.
@@
 A compromise could be 500.
+    %NC: 100 is fine, I was just guessing!
     We then calculated the median disparity value for each time bin along with the 50\% and 95\% confidence intervals.
@@
     \item Time-slices (unequal).
-    We calculated disparity using our time-slicing approach where the number of slices, was defined by the average \textit{number} of the stratigraphic periods (ages or epochs) in the time frame of interest..
+    We calculated disparity using our time-slicing approach where the number of slices, was defined by the average \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest.
     We used the six time-slicing methods and bootstrapped as described above.
@@
 
-%TG: well done! Nice description of the method! I figured out it was pretty obvious in our (my) head but it's actually more painfull to write down.
+%TG: well done! Nice description of the method! I figured out it was pretty obvious in our (my) head but it's actually more painful to write down.
+% NC: I know! It took me ages to get it logically described!
 
@@
 
-We also recorded the number of taxa (or taxa and nodes for time-slicing methods) in each sub-sample as a proxy for taxonomic diversity (Table \ref{tab:sub-samples}).
+We also recorded the number of taxa (or taxa and nodes for time-slicing methods) in each sub-sample as a proxy for taxonomic diversity (Appendix ??? \ref{tab:sub-samples}). % appendix tables maybe?
 
@@
 \label{testing}
-Testing for statistical differences among the time sub-sampling methods described above is a little tricky, as we need to compare similar units, and also to tackle questions important to the interpretation of disparity-through-time analyses.
+Testing for statistical differences among the time sub-sampling methods described above is difficult, as we need to compare similar units, and also to tackle questions important to the interpretation of disparity-through-time analyses.
 We therefore present three different, simple ways of comparing the time sub-sampling methods as follows.
@@
 \paragraph{Systematic differences in disparity-through-time.} 
-To test whether using time bins or time slices resulted in significantly different disparity values at common time points, we used paired Wilcoxon tests to compare the median bootstrapped disparities obtained in the stratigraphy (time sub-samples defined by stratigraphic periods), duration (time sub-samples defined by the duration of stratigraphic periods), and number (time sub-samples defined by the number of stratigraphic periods) analyses described above.
+To test whether using time bins or time-slices resulted in significantly different disparity values at common time points, we used paired Wilcoxon tests to compare the median bootstrapped disparities obtained in the stratigraphy (time sub-samples defined by stratigraphic periods), duration (time sub-samples defined by the duration of stratigraphic periods), and number (time sub-samples defined by the number of stratigraphic periods) analyses described above.
 
 Due to the uneven spread of taxa across phylogenies, some time bins will contain one or no species, meaning that we cannot estimate disparity for that time bin. 
-We first, therefore, remove the time bins, and corresponding time slices, without disparity estimates. 
-We then perform paired Wilcoxon tests, so that bins and slices for the same time period are being compared. 
+We first, therefore, removed the time bins, and corresponding time-slices, without disparity estimates. 
+We then performed paired Wilcoxon tests, so that bins and slices for the same time period are being compared. 
 Significant results suggest that there is a systematic difference in disparity values at each time point, depending on whether bins or slices are used.
@@
 \paragraph{Disparity peaks.}
-We are perhaps more interested in how the conclusions disparity-through-time analyses are influenced by the choice of time sub-sampling method, rather than the disparities estimated by each method \textit{per se}, especially as these will be influenced by the number of taxa (and/or nodes) included in each sub-sample.
-Therefore, we also investigate where peaks and troughs of disparity occur in each of our datasets for the different time sub-sampling methods. 
+We are perhaps more interested in how the conclusions of disparity-through-time analyses are influenced by the choice of time sub-sampling method, rather than the disparities estimated by each method \textit{per se}, especially as these will be influenced by the number of taxa (and/or nodes) included in each sub-sample.
+Therefore, we also investigated where peaks of disparity occurred in each of our datasets for the different time sub-sampling methods. 
 We calculated the maximum bootstrapped disparities for each dataset and for each time sub-sampling method, along with their associated confidence intervals.
-Significant shifts in disparity peaks suggest that the choice of time sub-sampling method will influence our conclusions about relative changes in the disparity of our example clades through time.
+Significant shifts in disparity peaks suggest that the choice of time sub-sampling method will influence our conclusions about relative changes in the disparity of our groups through time.
 
 \paragraph{Effects of mass extinction events.}
-Many analyses of disparity-through-time aim to demonstrate differences in disparity before and after mass extinction events [citations]. 
-Each of our four datasets span one mass extinction (Cretaceous-Paleogene 66 MYA; \citealt{brusatte2014gradual,bapst2016topology,beckancient2014}; Ordovician-Silurian 455– 430 MYA; \citealt{wright2017bayesian}), so we use Wilcoxon tests to compare disparity in the time bin/slice prior to the appropriate mass extinction, to that of the time bin/slice following the extinction event.
+Many analyses of disparity-through-time aim to demonstrate differences in disparity before and after mass extinction events. 
+Two of our four datasets contain taxa before and immediately after a mass extinction (Cretaceous-Paleogene 66 MYA; \citealt{beckancient2014}; Ordovician-Silurian 455– 430 MYA; \citealt{wright2017bayesian}), so we used Wilcoxon tests to compare disparity in the time bin/slice prior to the appropriate mass extinction, to that of the time bin/slice following the extinction event.
 Significant results suggest an effect of the mass extinction on disparity in the group.
-We then compare these results across the time sub-sampling methods to determine if our conclusions would change depending on the method used.
+We then compare these results across the time sub-sampling methods to determine if our conclusions change depending on the method used.
 We repeated these analyses using the two time bins/slices after the one immediately following the mass extinction event to account for any lag effects of the mass extinction on disparity.
@@
 
-Disparity changes through time for each of the four datasets (Figure \ref{figure:dtt}, Supporting Information Appendix S2: Figure AX and Tables AX-X). % add details of appendix
+Disparity changes through time for each of the four datasets (F{}igure \ref{figure:dtt}, Supporting Information Appendix S2: Figure AX and Tables AX-X). % add details of appendix
 Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods (i.e.
@@
     {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Pink points represent time binning methods, blue points are time slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
+    Pink points represent time binning methods, blue points are time-slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time-slices with a gradual model of evolution (specifically the `gradual splits' method).
     Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods.
@@
 Instead, the differences depend on the dataset and method in question.
-For example, the \cite{brusatte2014gradual}, \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, show significant differences when using bins versus time slices defined by stratigraphy, but the \cite{beckancient2014} dataset appears robust to these different models.
+For example, the \cite{brusatte2014gradual}, \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, show significant differences when using bins versus time-slices defined by stratigraphy, but the \cite{beckancient2014} dataset appears robust to these different models.
 Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} show different disparities when the number of bins or slices is the average number of stratigraphic periods, but this is not seen in the \cite{wright2017bayesian} dataset.
@@
 \paragraph{Disparity peaks}
-In the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets, disparity peaks occur much earlier when time slicing rather than time binning approaches are used (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX).
+In the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets, disparity peaks occur much earlier when time-slicing rather than time binning approaches are used (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX).
 % appendix details
-This is also true for stratigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, although when using equal time bins the peaks are later than the time slicing methods, or very similar (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX).
+This is also true for stratigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, although when using equal time bins the peaks are later than the time-slicing methods, or very similar (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX).
 %appendix
@@
     {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Pink points represent time binning methods, green points are time slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
+    Pink points represent time binning methods, green points are time-slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time-slices with a gradual model of evolution (specifically the `gradual splits' method).
     Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
@@
 TG: kind of showing why it's cool to allow different models of evolution.
-  % peak test: the peak really depends on which time slicing method is used!
+  % peak test: the peak really depends on which time-slicing method is used!
 This is probably expected for the disparity metric (different metrics give different results) but was never acknowledge to be the case for different methods of time subsampling!
@@
 
-Benefits of time slices
+Benefits of time-slices
 

### 2017-12-20 manuscript/time-slice.tex
First draft abstract

@@
 
-Disparity-through-time analyses are popular and routinely applied to palaeobiological datasets.
-
-Although many methods exist to measure disparity, all more or less assessed (metrics, tests, etc...), only one method exist for looking at disparity through time: pooling data into time bins.
-
-This may lead to some problems down the line as a methodological bottleneck (i.e. most disparity-through-time analysis are, in the end, based on pooling data into bins).
-
-Here we propose a novel method to complement the disparity-through-time tool-kit: time-slicing.
-
-This method allows to sample disparity through time at any fixed point in time rather than binning data between two fixed points in time.
-
-Additionally, this method allows to use all the phylogenetic data available (tips, nodes
-
-\textit{and} branches) and specify the implied model of evolution between each time-slice (punctuated or gradual).
-
-We test this method on four example datasets and compared how it performs at common disparity-through-time analysis: the effect of mass extinction and the time at the maximal and minimal disparity.
-
-We found that the sampling effect has a remarkable effect on our interpretation of disparity-through-time analysis.
-
-Using multiple methods for assessing disparity-through-time (rather than only one!), we can have a better picture of what is happening.
+Disparity-through-time analyses can be used to determine how morphological diversity changes in response to mass extinctions, and to investigate the drivers of morphological change.
+These analyses are routinely applied to palaeobiological datasets, yet although there is much discussion about how to best calculate disparity, there has been little consideration of how taxa should be sub-sampled through time.
+Standard practice is to group taxa into discrete time bins, often based on stratigraphic periods. 
+However, this can introduce biases when bins are of unequal size, and implicitly assumes a punctuated model of evolution. 
+In addition, many time bins may have few or no taxa, meaning that disparity cannot be calculated for the bin and making it harder to complete downstream analyses of morphological change through time.
+Here we describe a different method to complement the disparity-through-time tool-kit: time-slicing.
+This method uses a time-calibrated phylogenetic tree to sample disparity-through-time at any fixed point in time rather than binning data.
+It uses all available data (tips, nodes and branches) to increase the power of the analyses, and specifies the implied model of evolution (punctuated or gradual).
+We test the time-slicing method on four example datasets and compare its performance in common disparity-through-time analyses.
+We find that the way you time sub-sample your taxa can change your interpretations of the results of disparity-through-time analyses.
+Using multiple methods for assessing disparity-through-time (rather than just time binning), we can have a better picture of what is happening.
 
@@
 Many analyses of disparity-through-time aim to demonstrate differences in disparity before and after mass extinction events. 
-Two of our four datasets contain taxa before and immediately after a mass extinction (Cretaceous-Paleogene 66 MYA; \citealt{beckancient2014}; Ordovician-Silurian 455– 430 MYA; \citealt{wright2017bayesian}), so we used Wilcoxon tests to compare disparity in the time bin/slice prior to the appropriate mass extinction, to that of the time bin/slice following the extinction event.
+Two of our four datasets contain taxa before and immediately after a mass extinction (Cretaceous-Paleogene 66 MYA; \citealt{beckancient2014}; Ordovician-Silurian 455-430 MYA; \citealt{wright2017bayesian}), so we used Wilcoxon tests to compare disparity in the time bin/slice prior to the appropriate mass extinction, to that of the time bin/slice following the extinction event.
 Significant results suggest an effect of the mass extinction on disparity in the group.
@@
 
-Disparity changes through time for each of the four datasets (F{}igure \ref{figure:dtt}, Supporting Information Appendix S2: Figure AX and Tables AX-X). % add details of appendix
-Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods (i.e.
-the data included in the time bins are only the nodes and the tips, the time-slice also contains the data along the branches).
-The six different time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits) show similar patterns, so we focus only of the results for one method with a punctuated model of evolution (specifically the `proximity' method), and one method with a gradual model of evolution (specifically the `gradual splits' method).
+Disparity changes through time for each of our four datasets (Figure \ref{figure:dtt1}, Supporting Information Appendix S2: Figure AX and Tables AX-X). % add details of appendix
+Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods.
+The six different time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits) show similar patterns, so we focus only on the results for one method with a punctuated model of evolution (specifically the `proximity' method), and one method with a gradual model of evolution (specifically the `gradual splits' method).
 Results for all six methods can be found in Supporting Information Appendix S2: Figure AX. % add details of appendix
@@
     \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch.pdf}
-    \caption[Relative disparity through time for four example datasets.]
-    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Pink points represent time binning methods, blue points are time-slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time-slices with a gradual model of evolution (specifically the `gradual splits' method).
+    \caption[Relative disparity-through-time.]
+    {Relative disparity-through-time. Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
+    Pink points represent time binning methods, blue points are time-slices with a punctuated model of evolution (`proximity' method), and green points are time-slices with a gradual model of evolution (`gradual splits' method).
     Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods.
-    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
-    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
-    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
+    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
+    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset.
+    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
     In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
-    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).
-    Results for stratigraphic ages, and for other time-slicing models, are in the Supporting Information Appendix S2: Figure X} %appendix
+    The four dataset names are on the first plot for each dataset (see Table \ref{table:datasets} for details).
+    Results for stratigraphic ages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figure X} %appendix
     \label{figure:dtt1}
@@
 \paragraph{Systematic differences in disparity-through-time.} 
-There is no overall significant systematic difference among the disparities calculated using time bins and those calculated using the time-slicing methods.
-(Table \ref{table:wilcox}, Supporting Information Appendix S2: Table X).
+There is no overall significant systematic difference among the disparities calculated using time bins and those calculated using the time-slicing methods (Table \ref{table:wilcox}, Supporting Information Appendix S2: Table X).
 % appendix
 Instead, the differences depend on the dataset and method in question.
-For example, the \cite{brusatte2014gradual}, \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, show significant differences when using bins versus time-slices defined by stratigraphy, but the \cite{beckancient2014} dataset appears robust to these different models.
-Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} show different disparities when the number of bins or slices is the average number of stratigraphic periods, but this is not seen in the \cite{wright2017bayesian} dataset.
+For example, the \cite{brusatte2014gradual}, \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, show significant differences when using bins versus time-slices defined by stratigraphy, but the \cite{beckancient2014} dataset appears robust to these different approaches.
+Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} datasets have different disparities when the number of bins or slices is the average number of stratigraphic periods, but this is not seen in the \cite{wright2017bayesian} dataset.
 Note that for epochs, we find fewer significant differences simply because the smaller numbers of bins and slices being compared means we have low power to detect a significant difference.
@@
   \input{tables/table_wilcox.tex}  
-  \label{table:wilcox}  
 %\end{landscape}
 
-\paragraph{Disparity peaks}
+\paragraph{Disparity peaks.}
 In the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets, disparity peaks occur much earlier when time-slicing rather than time binning approaches are used (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX).
@@
     \caption[Timing of peak disparity for four example datasets.]
-    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Pink points represent time binning methods, green points are time-slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time-slices with a gradual model of evolution (specifically the `gradual splits' method).
-    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
-    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
-    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
+    {Timing of peak disparity.
+    Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
+    Pink points represent time binning methods, green points are time-slices with a punctuated model of evolution (`proximity' method), and blue points are time-slices with a gradual model of evolution (`gradual splits' method).
+    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
+    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset.
+    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
     For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
-    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (Table \ref{table:datasets}).
-    Results for stratigraphic ages, and for other time-slicing models, are in the Supporting Information Appendix S2: Figure AX} %appendix
+    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (see Table \ref{table:datasets} for details).
+    Results for stratigraphic ages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figure AX} %appendix
     \label{figure:peak1}
@@
   %TG: Would it be possible to make the bins points into a thick line? Though that's just aesthetic...
+  % NC: Probably but I can't be arsed.
 
 \paragraph{Effects of mass extinction events.}
-Mass extinction events influence disparity in all four datasets (Figure \ref{figure:extinction1}; Supporting Information Appendix S2: Figure AX). %appendix
-However, whether this change in disparity is significant or not depends on the dataset, and on the method used to create time sub-samples (Figure \ref{figure:extinction1}). % appendix!!!
+Mass extinction events influence disparity in both the \cite{beckancient2014} and \cite{wright2017bayesian} datasets (Figure \ref{figure:extinction1}).
+However, whether this change in disparity is significant or not depends on the method used to create time sub-samples (Figure \ref{figure:extinction1}), and whether ages or epochs are used.
 
@@
     \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-extinction-effect.pdf}
-    \caption[Extinction four example datasets.]
-    {Results of the extinction effect in change in disparity.
-    Pink cells and green cells indicate respectively a significant or no significant change of disparity (i.e. an effect or no effect of the extinction on disparity).
-    White cells are due to not enough data for performing the test (less than 2 taxa sampled in that period).
-    \textit{e:1}, \textit{e:2}, and \textit{e:3} denote respectively the test between the last bin/slice before the extinction and the first, second and third bin/slice after it (to account for an eventual lag effect).}
+    \caption[Extinction example datasets.]
+    {Effects of mass extinction events on disparity.
+    Pink cells and blue cells indicate respectively a significant or non significant change in disparity before and after the mass extinction event (Cretaceous-Paleogene 66 MYA; \citealt{beckancient2014}; Ordovician-Silurian 455–430 MYA; \citealt{wright2017bayesian}).
+    \textit{e:1}, \textit{e:2}, and \textit{e:3} denote whether the comparison was between the time bin or time-slice immediately after the mass extinction (\textit{e:1}), or the second (\textit{e:2}) or third (\textit{e:3}) bin/slice after the mass extinction to account for any lag effect.
+    The top seven rows use stratigraphic ages and the bottom seven rows use stratigraphic epochs.
+    Labels on the left hand side indicate whether time bins (`bins') were used, or which of the six time-slicing methods was used.}
     \label{figure:extinction1}
@@
 \section{Discussion}
+Disparity-through-time analyses are strongly affected by the choice of time sub-sampling method used to divide the taxa. 
+These changes are idiosyncratic - it is not possible to predict \textit{a priori} which method will give which result.
 
-%TG: ideas for the discussion:
-  % Extinction test: in Brusatte's dataset, ther seems to be an effect of the model (gradual vs.
-punctuated) on recovering the effect of the extinction or not when using stratigraphy: not significant for bins and punctuated slices but signif for gradual slices.
-TG: kind of showing why it's cool to allow different models of evolution.
-  % peak test: the peak really depends on which time-slicing method is used!
-This is probably expected for the disparity metric (different metrics give different results) but was never acknowledge to be the case for different methods of time subsampling!
-Kind of cool too!
+Time-slicing has several advantages over time binning (using either equally or unequally sized bins) approaches.
+First, it allows us to use as much of the information available to us, in the form of phylogenetic relationships and ancestral taxa, as possible. 
+This increases our ability to investigate key biological questions, such as how do various drivers influence morphological diversity through time, and how do mass extinctions influence disparity, both by increasing the statistical power of analyses and through the availability of data at key time points in the history of clades of interest.
+Second, we are able to be more explicit about the mode of evolution in our clades; in time-slicing we can apply punctuated or gradual models of trait change rather than making an assumption of punctuated evolution.
+This may be important, as gradual change is a common pattern of trait evolution in the fossil record \citep{Hunt20112007}.
 
-Benefits of time-slices
-
-  - Allowing for more data to be used
-
-  - Allowing different models of evolution
-
-  - Avoiding biases from using stratigraphy (i.e. the stratigraphic bins are based on disparity in the fossil record)
-
-Further stuff to fix?
+As with any method, there are still features we could improve upon.
 
@@
 
-\section{Conclusions}
-The way you time sub-sample your taxa has important implications for the results of disparity-through-time analyses.
-
 \section{Data archiving statement}
-Data for this study are available from \cite{dryad_n2g80,dryad_6hb7j,dryad_84t75,beckancient2014} but for ease of reproducibility can also be found on GitHub (https://github.com/nhcooper123/time-slice/data).
+Data for this study are available from \cite{dryad_n2g80,dryad_6hb7j,dryad_84t75,beckancient2014} but for ease of reproducibility can also be found on GitHub (\url{https://github.com/nhcooper123/time-slice/data}).
 
@@
   TG acknowledges support from the Australian Discovery Project Grant number DP170103227 awarded to Vera Weisbecker.
-  We thank Dave Bapst, Steve Brusatte, Graeme Lloyd, April Wright and Davey Wright for assistance in gathering data for the analyses.
-  %TG: and actually: Dave Bapst, Graeme Lloyd and Davey Wrigth for discussions about the method?
+  We thank Dave Bapst, Steve Brusatte, Graeme Lloyd, April Wright and Davey Wright for assistance in gathering data for the analyses and/or discussions about the approach.
 

### 2017-12-20 analyses/06-table-wilcox.Rmd
Change table legend

@@
 ## variable names
-colnames <- c("Dataset", "Period", "Model", "Stratigraphy", "Duration", "Number")
+colnames <- c("Dataset", "Period", "Method", "Stratigraphy", "Duration", "Number")
 
@@
 ## Table caption
-table_caption <- "Wilcoxon results for appendix"
+table_caption <- "Results of paired Wilcoxon tests investigating whether disparities calculated using time bins are significantly different to those calculated using time-slices.
+Time-slices used either a  punctuated (‘proximity’ method) or gradual (‘gradual.split’ method) model of evolution.
+Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages or epochs.
+Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages or epochs in the time frame of the dataset.
+Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages or epochs in the time frame of the dataset.
+Results for other time-slicing methods are in the Supporting Information Appendix S2: Table A1"
 

### 2017-12-20 manuscript/time-slice-appendices.tex
Rename appendix to something more descriptive

--- 2017-12-15 manuscript/appendix.tex
@@
 \usepackage{setspace}
+\usepackage{longtable}
 
@@
 
-\noindent{\Large \bf Supporting Information from `Time for a rethink: time sub-sampling methods in disparity-through-time analyses'}
+\noindent{\Large \bf Supporting Information from ``Time for a rethink: time sub-sampling methods in disparity-through-time analyses''}
 
@@
 
+% Reset table and figure counters to 1
+\setcounter{figure}{0}  
+\setcounter{table}{0}
+
+% DTT figures
   \begin{figure}[!htbp]
@@
     \caption[Relative disparity through time for four example datasets.]
-    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Pink points represent time binning methods, 
-    % blue points are time slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
+    {Relative disparity-through-time for stratigraphic epochs. 
+    Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
     Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods.
-    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
-    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
-    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
+    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
+    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset.
+    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
     In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
-    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
+    The four dataset names are on the first plot for each dataset (see Table 1 for details).
+    Results for stratigraphic ages are shown in Figure A2.}
     \label{figure:dtt2}
@@
     \caption[Relative disparity through time for four example datasets.]
-    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Pink points represent time binning methods, 
-    %blue points are time slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
+    {Relative disparity-through-time for stratigraphic ages. 
+    Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
     Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods.
-    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages. 
-    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages in the time frame of the dataset. 
-    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages in the time frame of the dataset. 
+    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages. 
+    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages in the time frame of the dataset.
+    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages in the time frame of the dataset. 
     In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
-    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
+    The four dataset names are on the first plot for each dataset (see Table 1 for details).
+    Results for stratigraphic epochs are shown in Figure A1.}
     \label{figure:dtt3}
@@
 % Wilcoxon results table
-%\begin{landscape}
   \input{tables/table_wilcox_appendix.tex} 
   \label{table:wilcox2}
-%\end{landscape}
 
+% disparity peaks figures
 \begin{figure}[!htbp]
@@
     \caption[Timing of peak disparity for four example datasets.]
-    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    %Pink points represent time binning methods, green points are time slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
-    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
-    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
-    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
+    {Timing of peak disparity for stratigraphic epochs.
+    Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
+    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
+    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset.
+    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
     For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
-    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
+    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (see Table 1 for details).
+    Results for stratigraphic ages are in Figure A4.}
     \label{figure:peak2}
@@
     \caption[Timing of peak disparity for four example datasets.]
-    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    %Pink points represent time binning methods, green points are time slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
-    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages. 
-    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages in the time frame of the dataset. 
-    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages in the time frame of the dataset. 
+    {Timing of peak disparity for stratigraphic ages.
+    Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
+    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages. 
+    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages in the time frame of the dataset.
+    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages in the time frame of the dataset. 
     For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
-    The four dataset names are on the first plot for each dataset (Table 1).}
+    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (see Table 1 for details).
+    Results for stratigraphic epochs are in Figure A3.}
     \label{figure:peak3}
@@
 
+% Full results tables
+% These are so huge (100 pages!!!) I'm going to leave them out for now...
+
+%\begin{landscape}
+%  \input{tables/disparity_measurement-Beck2014.tex} 
+%\end{landscape}
+
+%\newpage
+%\begin{landscape}
+%  \input{tables/disparity_measurement-Brusatte2014.tex} 
+%\end{landscape}
+
+%\newpage
+%\begin{landscape}
+%  \input{tables/disparity_measurement-Bapst2016.tex} 
+%\end{landscape}
+
+%\newpage
+%\begin{landscape}
+ % \input{tables/disparity_measurement-Wright2017.tex}  
+%\end{landscape}
+
 \end{document}

### 2017-12-20 manuscript/time-slice-appendices.tex
Tidied up appendices

@@
 \usepackage{longtable}
+\usepackage{fancyhdr}
 
 \setcounter{secnumdepth}{0} 
-\raggedright 			
-\pagenumbering{arabic}
+%\raggedright 			
+%\pagenumbering{arabic}	
+\pagenumbering{gobble}
+
+\pagestyle{fancy}
+\rhead{\textit{Guillerme \& Cooper 2018}}
+\lhead{\textit{Supporting Information}}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrulewidth}{0pt}
+\setlength{\headsep}{0.3in}
 
@@
 
-\noindent{\Large \bf Supporting Information from ``Time for a rethink: time sub-sampling methods in disparity-through-time analyses''}
+\par{\centering{\Large \bf Supporting Information from ``Time for a rethink: time sub-sampling methods in disparity-through-time analyses''\par}}
 
-\raggedright
 \setlength{\parindent}{1cm}
@@
 Zero-length branches were replaced with the minimum branch length in the phylogeny.
-The following taxa were removed because they were in the phylogeny but not the character matrix or vice versa: \textit{Sinraptor dongi}, \textit{Hesperonychus elizabethae}, \textit{Pyroraptor olympius}, \textit{Limenavis patagonica}, \textit{Lithornis}, \textit{Crypturellus undulatus}, \textit{Gallus gallus}, \textit{Crax pauxi}, \textit{Anas platyrhynchus}, \textit{Chauna torquata}, \textit{Epidendrosaurus} and \textit{Kinnareemimus}.
+The following taxa were removed because they were in the phylogeny but not the character matrix or vice versa: \textit{Sinraptor dongi}, \textit{Hesperonychus elizabethae}, \textit{Pyroraptor olympius}, \textit{Limenavis patagonica}, \textit{Lithornis}, \textit{Crypturellus undulatus}, \textit{Gallus gallus}, \textit{Crax pauxi}, \textit{Anas}\\ \textit{platyrhynchus}, \textit{Chauna torquata}, \textit{Epidendrosaurus} and \textit{Kinnareemimus}.
 The following taxa were removed because they shared no characters in the morphological matrix: \textit{Shanag ashile}, \textit{Atrociraptor marshalli}, \textit{Proceratosaurus bradleyi}, \textit{Incisivosaurus gauthieri}, \textit{Enigmosaurus}, \textit{Nanshiungosaurus brevispinus}, \textit{Xixiasaurus}, \textit{Tsaagan mangas}, \textit{Mirischia}, \textit{Pedopenna}, \textit{Suzhousaurus}, \textit{Juratyrant}, \textit{Vorona}, \textit{Bonapartenykus}, \textit{Teratophoneus}, \textit{Gobipteryx}, \textit{Songlingornis}, \textit{Liaoningornis longidigitu} and \textit{Achillesaurus}.
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch-appendix.pdf}
+   % \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch-appendix.pdf}
     \caption[Timing of peak disparity for four example datasets.]
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-age-appendix.pdf}
+    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-age-appendix.pdf}
     \caption[Timing of peak disparity for four example datasets.]

### 2017-12-20 analyses/06-table-wilcox.Rmd
Fix wilcoxon table legends

@@
 ## Table caption
-table_caption <- "Wilcoxon results"
+table_caption <- "Results of paired Wilcoxon tests investigating whether disparities calculated using time bins are significantly different to those calculated using time-slices.
+Time-slices used either a  punctuated (‘proximity’ method) or gradual (‘gradual.split’ method) model of evolution.
+Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages or epochs.
+Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages or epochs in the time frame of the dataset.
+Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages or epochs in the time frame of the dataset.
+P-values were Bonferroni corrected.
+***p < 0.001.
+Results for other time-slicing methods are in the Supporting Information Appendix S2: Table A1."
 
@@
 table_caption <- "Results of paired Wilcoxon tests investigating whether disparities calculated using time bins are significantly different to those calculated using time-slices.
-Time-slices used either a  punctuated (‘proximity’ method) or gradual (‘gradual.split’ method) model of evolution.
+Time-slices used one of six methods.
 Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages or epochs.
@@
 Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages or epochs in the time frame of the dataset.
-Results for other time-slicing methods are in the Supporting Information Appendix S2: Table A1"
+.
+P-values were Bonferroni corrected.
+***p < 0.001."
 

### 2017-12-20 manuscript/time-slice.tex
Edits

@@
 However, this can introduce biases when bins are of unequal size, and implicitly assumes a punctuated model of evolution. 
-In addition, many time bins may have few or no taxa, meaning that disparity cannot be calculated for the bin and making it harder to complete downstream analyses of morphological change through time.
+In addition, many time bins may have few or no taxa, meaning that disparity cannot be calculated for the bin and making it harder to complete downstream analyses.
 Here we describe a different method to complement the disparity-through-time tool-kit: time-slicing.
-This method uses a time-calibrated phylogenetic tree to sample disparity-through-time at any fixed point in time rather than binning data.
-It uses all available data (tips, nodes and branches) to increase the power of the analyses, and specifies the implied model of evolution (punctuated or gradual).
+This method uses a time-calibrated phylogenetic tree to sample disparity-through-time at any fixed point in time rather than binning taxa.
+It uses all available data (tips, nodes and branches) to increase the power of the analyses, specifies the implied model of evolution (punctuated or gradual), and is implemented in \texttt{R}.
 We test the time-slicing method on four example datasets and compare its performance in common disparity-through-time analyses.
 We find that the way you time sub-sample your taxa can change your interpretations of the results of disparity-through-time analyses.
-Using multiple methods for assessing disparity-through-time (rather than just time binning), we can have a better picture of what is happening.
+We advise using multiple methods for time sub-sampling taxa, rather than just time binning, to gain a better understanding disparity-through-time.
 
@@
 
-We also recorded the number of taxa (or taxa and nodes for time-slicing methods) in each sub-sample as a proxy for taxonomic diversity (Appendix ??? \ref{tab:sub-samples}). % appendix tables maybe?
+We also recorded the number of taxa (or taxa and nodes for time-slicing methods) in each sub-sample as a proxy for taxonomic diversity. 
+%(Supporting Information Appendix S2: Tables A1-A4). 
+% Tables are so huge I'm going to leave them out for now...
 
@@
 We first, therefore, removed the time bins, and corresponding time-slices, without disparity estimates. 
-We then performed paired Wilcoxon tests, so that bins and slices for the same time period are being compared. 
+We then performed paired Wilcoxon tests with Bonferroni corrected p-values, so that bins and slices for the same time period are being compared. 
 Significant results suggest that there is a systematic difference in disparity values at each time point, depending on whether bins or slices are used.
@@
 Many analyses of disparity-through-time aim to demonstrate differences in disparity before and after mass extinction events. 
-Two of our four datasets contain taxa before and immediately after a mass extinction (Cretaceous-Paleogene 66 MYA; \citealt{beckancient2014}; Ordovician-Silurian 455-430 MYA; \citealt{wright2017bayesian}), so we used Wilcoxon tests to compare disparity in the time bin/slice prior to the appropriate mass extinction, to that of the time bin/slice following the extinction event.
+Two of our four datasets contain taxa before and immediately after a mass extinction (Cretaceous-Paleogene 66 MYA; \citealt{beckancient2014}; Ordovician-Silurian 455-430 MYA; \citealt{wright2017bayesian}), so we used Wilcoxon tests with Bonferroni corrected p-values to compare disparity in the time bin/slice prior to the appropriate mass extinction, to that of the time bin/slice following the extinction event.
 Significant results suggest an effect of the mass extinction on disparity in the group.
@@
 
-Disparity changes through time for each of our four datasets (Figure \ref{figure:dtt1}, Supporting Information Appendix S2: Figure AX and Tables AX-X). % add details of appendix
+Disparity changes through time for each of our four datasets (Figure \ref{figure:dtt1}, Supporting Information Appendix S2: Figures A1-A2). 
 Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods.
 The six different time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits) show similar patterns, so we focus only on the results for one method with a punctuated model of evolution (specifically the `proximity' method), and one method with a gradual model of evolution (specifically the `gradual splits' method).
-Results for all six methods can be found in Supporting Information Appendix S2: Figure AX. % add details of appendix
+Results for all six methods can be found in Supporting Information Appendix S2: Figures A1-A2.
 
@@
     The four dataset names are on the first plot for each dataset (see Table \ref{table:datasets} for details).
-    Results for stratigraphic ages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figure X} %appendix
+    Results for stratigraphic ages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A1-A2.}
     \label{figure:dtt1}
@@
 \paragraph{Systematic differences in disparity-through-time.} 
-There is no overall significant systematic difference among the disparities calculated using time bins and those calculated using the time-slicing methods (Table \ref{table:wilcox}, Supporting Information Appendix S2: Table X).
-% appendix
+There is no overall significant systematic difference among the disparities calculated using time bins and those calculated using the time-slicing methods (Table \ref{table:wilcox}, Supporting Information Appendix S2: Table A1).
 Instead, the differences depend on the dataset and method in question.
@@
 \paragraph{Disparity peaks.}
-In the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets, disparity peaks occur much earlier when time-slicing rather than time binning approaches are used (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX).
-% appendix details
-This is also true for stratigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, although when using equal time bins the peaks are later than the time-slicing methods, or very similar (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX).
-%appendix
-Across the three time binning methods, the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets have similar disparity peaks whichever method is used, whereas in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX).
-% appendix
+In the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets, disparity peaks occur much earlier when time-slicing rather than time binning approaches are used (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figures A3-A4).
+This is also true for stratigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, although when using equal time bins the peaks are later than the time-slicing methods, or very similar (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figures A3-A4).
+Across the three time binning methods, the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets have similar disparity peaks whichever method is used, whereas in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figures A3-A4).
 
@@
     The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (see Table \ref{table:datasets} for details).
-    Results for stratigraphic ages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figure AX} %appendix
+    Results for stratigraphic ages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A3-A4.}
     \label{figure:peak1}
@@
 However, whether this change in disparity is significant or not depends on the method used to create time sub-samples (Figure \ref{figure:extinction1}), and whether ages or epochs are used. 
+In general, for the \cite{beckancient2014} dataset, time binning tended to give more significant results than time-slicing methods, but this was not the case for the \cite{wright2017bayesian} dataset.
 
@@
     The top seven rows use stratigraphic ages and the bottom seven rows use stratigraphic epochs.
-    Labels on the left hand side indicate whether time bins (`bins') were used, or which of the six time-slicing methods was used.}
+    Labels on the left hand side indicate whether time bins (``bins'') were used, or which of the six time-slicing methods was used.}
     \label{figure:extinction1}
@@
 \section{Discussion}
-Disparity-through-time analyses are strongly affected by the choice of time sub-sampling method used to divide the taxa. 
+Disparity-through-time analyses are influenced by the choice of time sub-sampling method used to divide the taxa. 
 These changes are idiosyncratic - it is not possible to predict \textit{a priori} which method will give which result.
@@
 Data for this study are available from \cite{dryad_n2g80,dryad_6hb7j,dryad_84t75,beckancient2014} but for ease of reproducibility can also be found on GitHub (\url{https://github.com/nhcooper123/time-slice/data}).
+% add proper link to zenodo
 
@@
 \section{Acknowledgments}
-	NC thanks Mark Sutton and Phillip Mannion for the invitation to contribute to the `Evolutionary Modelling' symposium at The Palaeontological Association Annual Meeting 2017.
+	NC thanks Mark Sutton and Philip Mannion for the invitation to contribute to the `Evolutionary Modelling' symposium at The Palaeontological Association Annual Meeting 2017.
   TG acknowledges support from the Australian Discovery Project Grant number DP170103227 awarded to Vera Weisbecker.

### 2017-12-21 manuscript/time-slice.tex
First draft of discussion completed

@@
     {Effects of mass extinction events on disparity.
-    Pink cells and blue cells indicate respectively a significant or non significant change in disparity before and after the mass extinction event (Cretaceous-Paleogene 66 MYA; \citealt{beckancient2014}; Ordovician-Silurian 455–430 MYA; \citealt{wright2017bayesian}).
+    Pink cells and blue cells indicate respectively a significant or non significant change in disparity before and after the mass extinction event (Cretaceous-Paleogene 66 MYA; \citealt{beckancient2014}; Ordovician-Silurian 455-430 MYA; \citealt{wright2017bayesian}).
     \textit{e:1}, \textit{e:2}, and \textit{e:3} denote whether the comparison was between the time bin or time-slice immediately after the mass extinction (\textit{e:1}), or the second (\textit{e:2}) or third (\textit{e:3}) bin/slice after the mass extinction to account for any lag effect.
@@
 Disparity-through-time analyses are influenced by the choice of time sub-sampling method used to divide the taxa. 
-These changes are idiosyncratic - it is not possible to predict \textit{a priori} which method will give which result.
+While differences in the relative disparities calculated among time sub-sampling methods may not be of much biological importance, these changes can have important implications for the conclusions of downstream analyses.
+For example, using stratigraphic epochs as our reference time period, there are 21 potential methods for time sub-sampling our data (splitting by stratigraphy, number and duration, see methods, and using time bins or one of six time-slicing methods).
+Of these 21 methods, in 16 (76\%) we show that placental mammals \citep{beckancient2014} significantly increased in disparity in the time bin/slice immediately after the K-Pg mass extinction event, and in 20 (95\%) we show that crinoids \citep{wright2017bayesian} significantly decreased in disparity in the time bin/slice immediately after the O-S mass extinction event.
+Given the high congruence (76\% and 95\%) of these results, one could argue that time-sub-sampling methods are not important. 
+However, if we had chosen to investigate crinoid disparity only using time bins and splitting these so the number of time bins was equal to the number of epochs (number), we would have concluded that the O-S extinction had no effect on crinoid disparity.
+This highlights the sensitivity of disparity-through-time analyses to the choice of time sub-sampling method. 
+Fortunately this issue is easy to solve; either disparity-through-time analyses should use, and report results from, multiple time sub-sampling methods (as demonstrated here), or great care should be taken in determining the appropriate time sub-samples to answer the question of interest.
 
@@
 First, it allows us to use as much of the information available to us, in the form of phylogenetic relationships and ancestral taxa, as possible. 
-This increases our ability to investigate key biological questions, such as how do various drivers influence morphological diversity through time, and how do mass extinctions influence disparity, both by increasing the statistical power of analyses and through the availability of data at key time points in the history of clades of interest.
+This increases our ability to investigate key biological questions, such as how do various drivers influence morphological diversity through time, and how do mass extinctions influence disparity \citep{Brusatte12092008,Foote29111996,friedmanexplosive2010}, both by increasing the statistical power of analyses and through the availability of data at key time points in the history of our groups.
 Second, we are able to be more explicit about the mode of evolution in our clades; in time-slicing we can apply punctuated or gradual models of trait change rather than making an assumption of punctuated evolution.
 This may be important, as gradual change is a common pattern of trait evolution in the fossil record \citep{Hunt20112007}.
+Of course the method also has limitations.
+The main one of these is a practical one; it requires a time-calibrated phylogeny.
+Not all palaeontological datasets have phylogenies yet.
+Another limitation is that the method, like most phylogeny based methods, depends on ancestral state reconstructions.
+Care should be taken in interpreting these, as they are highly dependent on the data and models used to for the reconstructions \citep{Slateretal2012,doi:10.1080/10635150801910451}.
 
-As with any method, there are still features we could improve upon.
-
-  - Allowing proper reconstruction along the branches (basically slicing before the ordination).
-
-  - Testing the effect of slice/bin on properly simulated data set to test for systematic difference (or not)
-
-  - Testing the effect of slice/bin with different disparity metrics
+\section{Conclusions}
+The choice of time sub-sampling method can alter the conclusions we obtain from disparity-through-time studies. 
+Time-slicing methods, with explicit models of evolution, provide an alternative to traditional time binning approaches.
+Note that while we introduce the time-slicing methods here, and describe their advantages, we do not suggest time-slicing is necessarily the ``best'' method for time sub-sampling. 
+As with all methods, the choice of methodology should be appropriate for the question and data at hand.
+However, we do strongly recommend performing disparity-through-time analyses using a series of appropriate time sub-sampling methods, and comparing these, to ensure that results are not merely a consequence of the time sub-sampling method employed.
 

### 2017-12-21 analyses/01-extract-data-for-analyses.Rmd
Add comment about how long code takes to run

@@
 This code loads the data and extracts matrices and other files needed to run the analyses in the paper.
-Note that these analyses can take time, so we save the outputs to a `data/processed` folder and use these in later analyses so that the analyses can be replicated more quickly.
+Note that these analyses can take time (i.e.
+more than 24 hours!), so we save the outputs to a `data/processed` folder and use these in later analyses so that the analyses can be replicated more quickly.
 Below we lay out the process in detail for one dataset, and then provide the code for the other datasets below this.

### 2017-12-21 analyses/04-table-datasets.Rmd
Edits to root times for Bapst and Brusatte

@@
 age_ranges <- unlist(lapply(age_ranges, function(age) paste(age, collapse = " - ")))
+## Note that I've had to correct these for Bapst2016 (207.2) and Brusatte2014 (168.5) due to 
+## the rerooting that occurred when remvoing taxa
 
@@
 table_caption <- "Details of the datasets used in this study.
-Note that although Brusatte2014 and Bapst2016 cross the Cretaceous Palaeogene (K-Pg) mass extinction there are not enough taxa post K-Pg to investigate the effects of the mass extinction on disparity."
+Note that although Brusatte2014 and Bapst2016 cross the Cretaceous Palaeogene (K-Pg) mass extinction there are not enough taxa post K-Pg to investigate the effects of the mass extinction on disparity.
+Age ranges are root time to most recent tip taxon."
 

### 2017-12-21 manuscript/tables/table_datasets.tex
Edits to root times for Bapst and Brusatte

@@
   \# characters & 421 & 853 & 374 &  87 \\ 
-  Age range (MYA) & 171.786 - 0 & 102.535 - 0 & 141.216 - 0 & 485.4 - 372.2 \\ 
+  Age range (MYA) & 171.8 - 0 & 168.5 - 0 & 207.2 - 0 & 485.4 - 372.2 \\ 
   Mass extinction (MYA) & 66 (K-Pg) & NA & NA & 443 (O-S) \\ 
@@
 \caption{Details of the datasets used in this study.
-Note that although Brusatte2014 and Bapst2016 cross the Cretaceous Palaeogene (K-Pg) mass extinction there are not enough taxa post K-Pg to investigate the effects of the mass extinction on disparity.}
+Note that although Brusatte2014 and Bapst2016 cross the Cretaceous Palaeogene (K-Pg) mass extinction there are not enough taxa post K-Pg to investigate the effects of the mass extinction on disparity.
+Age ranges are root time to most recent tip taxon.}
 \label{table:datasets}  

### 2017-12-21 analyses/05-figure-dtt.Rmd
Edit to fit Bapst and Brusatte rescaled

@@
                      labels = seq(0, 1, 0.5)) +
-  scale_x_reverse(limits = c(100, 0),
-                  breaks = c(seq(100, 0, by = -50))) +
+  scale_x_reverse(limits = c(170, 70),
+                  breaks = c(seq(170, 70, by = -50))) +
   expand_limits(x = 0, y = c(0,1)) +
@@
   geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
-  annotate("text", x = 70, y = 0.15, label = "Brusatte2014")
+  annotate("text", x = 140, y = 0.15, label = "Brusatte2014")
 
@@
                      labels = seq(0, 1, 0.5)) +
-  scale_x_reverse(limits = c(100, 0),
-                  breaks = c(seq(100, 0, by = -50))) +
+  scale_x_reverse(limits = c(200, 50),
+                  breaks = c(seq(200, 50, by = -50))) +
   expand_limits(x = 0, y = c(0,1)) +
@@
   geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
-  annotate("text", x = 75, y = 0.15, label = "Bapst2016")
+  annotate("text", x = 120, y = 0.15, label = "Bapst2016")
 
@@
   geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
-  annotate("text", x = 75, y = 0.15, label = "Bapst2016")
+  annotate("text", x = 165, y = 0.15, label = "Bapst2016")
 

### 2017-12-21 analyses/10-appendix-figure-dtt.Rmd
Edit to fit Bapst and Brusatte rescaled

--- 2017-12-15 analyses/10-appendix-dtt-figures.Rmd
@@
   geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
-  annotate("text", x = 75, y = 0.15, label = "Bapst2016")
+  annotate("text", x = 125, y = 0.15, label = "Bapst2016")
 
@@
   theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
+        axis.title.x = element_blank()
         )

### 2017-12-21 analyses/07-figure-peaks.Rmd
Split out appendix figs

--- 2017-12-11 analyses/09-figure-peaks.Rmd
@@
 ```
-
-# Repeat for all models for appendix
-
-# Beck 2014
-
-```{r}
-slug <- "Beck2014"
-```
-
-Load the data
-
-```{r}
-data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
-data_max$bin_type <- gsub("stratigraphy", "Stratigraphy", data_max$bin_type) 
-data_max$bin_type <- gsub("duration", "Duration", data_max$bin_type) 
-data_max$bin_type <- gsub("number", "Number", data_max$bin_type) 
-data_max$bin_type <- as.factor(data_max$bin_type)
-```
-
-# Modify the data
-```{r}
-# Get ages for the bins, go for the midpoints
-x <- str_split(data_max$subsets, " - ") 
-y <- map(x, as.numeric)
-z1 <- map(y, max)
-z2 <- map(y, min)
-
-# Modify data - add midpoints, remove missing data, filter to just three models
-data_max2 <- 
-  data_max %>%
-  mutate(time1 = unlist(z1)) %>%
-  mutate(time2 = unlist(z2)) %>%
-  filter(bs.median != 0 & !is.na(bs.median)) %>%
-  mutate(bin_type = relevel(data_max$bin_type, ref = 'Stratigraphy')) %>%
-  mutate(model = relevel(data_max$model, ref = 'equalbins')) #%>%
-  #filter(model == "equalbins" | model == "proximity" | model == "gradual.split")
-
-data_max2_age <- filter(data_max2, stratigraphy == "Age")
-
-data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
-```
-
-```{r}
-beck_peaks <- 
-  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- # expand_limits(x = 0, y = c(0,1)) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-       axis.text.x = element_blank(),
-        #legend.position = "none",
-        panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.background = element_blank())
-
-beck_peaks_age <- 
-  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- # expand_limits(x = 0, y = c(0,1)) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        axis.text.x = element_blank(),
-        #legend.position = "none",
-        panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.background = element_blank())
-```
-
-# Brusatte 2014
-
-```{r}
-slug <- "Brusatte2014"
-```
-
-Load the data
-
-```{r}
-data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
-```
-```{r}
-# Get ages for the bins, go for the midpoints
-x <- str_split(data_max$subsets, " - ") 
-y <- map(x, as.numeric)
-z1 <- map(y, max)
-z2 <- map(y, min)
-
-# Modify data - add midpoints, remove missing data, filter to just three models
-data_max2 <- 
-  data_max %>%
-  mutate(time1 = unlist(z1)) %>%
-  mutate(time2 = unlist(z2)) %>%
-  filter(bs.median != 0 & !is.na(bs.median)) %>%
-  mutate(bin_type = relevel(data_max$bin_type, ref = 'stratigraphy')) %>%
-  mutate(model = relevel(data_max$model, ref = 'equalbins'))#%>%
-  #filter(model == "equalbins" | model == "proximity" | model == "gradual.split")
-
-data_max2_age <- filter(data_max2, stratigraphy == "Age")
-
-data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
-```
-
-```{r}
-brusatte_peaks <- 
-  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
-  expand_limits(y = 0) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-       axis.text.x = element_blank(), 
-       legend.position = "none", 
-       panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.text = element_blank())
-
-brusatte_peaks_age <- 
-  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
-  expand_limits(y = 0) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        legend.position = "none", 
-       axis.text.x = element_blank(),
-       panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.text = element_blank()) 
-```
-
-# Bapst 2016
-
-```{r}
-slug <- "Bapst2016"
-```
-
-Load the data
-
-```{r}
-data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
-```
-```{r}
-# Get ages for the bins, go for the midpoints
-x <- str_split(data_max$subsets, " - ") 
-y <- map(x, as.numeric)
-z1 <- map(y, max)
-z2 <- map(y, min)
-
-# Modify data - add midpoints, remove missing data, filter to just three models
-data_max2 <- 
-  data_max %>%
-  mutate(time1 = unlist(z1)) %>%
-  mutate(time2 = unlist(z2)) %>%
-  filter(bs.median != 0 & !is.na(bs.median)) %>%
-  mutate(bin_type = relevel(data_max$bin_type, ref = 'stratigraphy')) %>%
-  mutate(model = relevel(data_max$model, ref = 'equalbins'))#%>%
-  #filter(model == "equalbins" | model == "proximity" | model == "gradual.split")
-
-data_max2_age <- filter(data_max2, stratigraphy == "Age")
-
-data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
-```
-
-```{r}
-bapst_peaks <- 
-  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
- # scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- # expand_limits(x = 0, y = c(0,1)) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        legend.position = "none",
-       axis.text.x = element_blank(),
-       panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.text = element_blank())
-
-bapst_peaks_age <- 
-  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- # expand_limits(x = 0, y = c(0,1)) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        legend.position = "none",
-       axis.text.x = element_blank(),
-       panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.text = element_blank())
-```
-
-# Wright 2017
-
-```{r}
-slug <- "Wright2017"
-```
-
-Load the data
-
-```{r}
-data_max <- read.csv(paste0("../outputs/peaks.max.", slug, ".csv"))
-```
-```{r}
-# Get ages for the bins, go for the midpoints
-x <- str_split(data_max$subsets, " - ") 
-y <- map(x, as.numeric)
-z1 <- map(y, max)
-z2 <- map(y, min)
-
-# Modify data - add midpoints, remove missing data, filter to just three models
-data_max2 <- 
-  data_max %>%
-  mutate(time1 = unlist(z1)) %>%
-  mutate(time2 = unlist(z2)) %>%
-  filter(bs.median != 0 & !is.na(bs.median)) %>%
-  mutate(bin_type = relevel(data_max$bin_type, ref = 'stratigraphy')) %>%
-  mutate(model = relevel(data_max$model, ref = 'equalbins'))#%>%
-  #filter(model == "equalbins" | model == "proximity" | model == "gradual.split")
-
-data_max2_age <- filter(data_max2, stratigraphy == "Age")
-
-data_max2_epoch <- filter(data_max2, stratigraphy == "Epoch")
-```
-
-```{r}
-wright_peaks <- 
-  ggplot(data_max2_epoch, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- expand_limits(y = 480) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        legend.position = "none",
-       axis.text.x = element_blank(),
-       panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.text = element_blank())
-
-wright_peaks_age <- 
-  ggplot(data_max2_age, aes(y = time1, x = model, colour = model)) +
-  theme_classic(base_size = 12) +
-  geom_point() +
-  geom_point(aes(y = time2, x = model, colour = model)) +
-  facet_wrap(~bin_type, ncol = 3) +
-  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- expand_limits(y = 480) +
- theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank(),
-        legend.position = "none",
-       axis.text.x = element_blank(),
-       panel.spacing = unit(0, "lines"),
-        panel.border = element_blank(),
-        strip.text = element_blank()) 
-```
-
-Finally plot them all together
-
-```{r}
-# Using modified multiplot function
-# written by Scott Chamberlain
-
-pdf("../manuscript/figures/fig-peaks-epoch-appendix.pdf")  
-  multiplot(beck_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
-            brusatte_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
-            bapst_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
-            wright_peaks + theme(plot.margin=unit(c(5.5, 5.5, 20, 25),"points")),
-            cols = 1,
-            labs = list("Time bins or time slice method",
-                        "Time at disparity peak (MYA)"),
-            labpos = list(c(0.5, 0.015), c(0.03, 0.5)))
-
-dev.off() 
-```
-
-# And for ages
-
-```{r}
-
-pdf("../manuscript/figures/fig-peaks-age-appendix.pdf")  
-  multiplot(beck_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
-            brusatte_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
-            bapst_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
-            wright_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 20, 25),"points")),
-            cols = 1,
-            labs = list("Time bins or time slice method",
-                        "Time at disparity peak (MYA)"),
-            labpos = list(c(0.5, 0.015), c(0.03, 0.5)))
-
-dev.off()
-
-```

### 2017-12-21 manuscript/tables/table_wilcox.tex
Updated Wilcoxon results
Add changes to wilcoxon tables

@@
   \hline
-\textbf{Dataset} & \textbf{Period} & \textbf{Model} & \textbf{Stratigraphy} & \textbf{Duration} & \textbf{Number} \\ 
+\textbf{Dataset} & \textbf{Period} & \textbf{Method} & \textbf{Stratigraphy} & \textbf{Duration} & \textbf{Number} \\ 
   \hline
@@
 \end{tabular}
-\caption{Wilcoxon results} 
+\caption{Results of paired Wilcoxon tests investigating whether disparities calculated using time bins are significantly different to those calculated using time-slices.
+Time-slices used either a  punctuated (‘proximity’ method) or gradual (‘gradual.split’ method) model of evolution.
+Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages or epochs.
+Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages or epochs in the time frame of the dataset.
+Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages or epochs in the time frame of the dataset.
+P-values were Bonferroni corrected.
+***p < 0.001.
+Results for other time-slicing methods are in the Supporting Information Appendix S2: Table A1.}
+\label{table:wilcox} 
 \end{table}

### 2017-12-21 manuscript/tables/table_wilcox_appendix.tex
Add changes to wilcoxon tables

@@
 \end{tabular}
-\caption{Wilcoxon results for appendix} 
+\caption{Results of paired Wilcoxon tests investigating whether disparities calculated using time bins are significantly different to those calculated using time-slices.
+Time-slices used one of six methods.
+Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages or epochs.
+Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages or epochs in the time frame of the dataset.
+Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages or epochs in the time frame of the dataset.
+.
+P-values were Bonferroni corrected.
+***p < 0.001.}
 \end{table}

### 2017-12-22 manuscript/time-slice.tex
Final edits

@@
 \usepackage{graphicx} 	
-%\usepackage[nomarkers, nolists]{endfloat} 
+\usepackage[nomarkers, nolists, figuresonly]{endfloat} 
 \usepackage{pdflscape}	
@@
 \author{
-	Thomas Guillerme$^{1*}$ and Natalie Cooper$^{2}$
+	Thomas Guillerme$^{1}$ and Natalie Cooper$^{2*}$
 }
@@
 \label{overview-section}
-To test the different time sub-sampling methods, we followed the protocol below (Figure \ref{fig:overview}). 
+To test the different time sub-sampling methods, we followed the protocol below (Fig. \ref{fig:overview}). 
 All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (\url{https://github.com/nhcooper123/time-slice}.
@@
 % Dataset details table
-\begin{landscape}
-  \input{tables/table_datasets.tex} 
+%\begin{landscape}
+%\input{tables/table_datasets.tex} 
   % Label in input or referencing doesn't work
-\end{landscape}
+%\end{landscape}
 
@@
 
-      \item The ``acctran'' model, always selecting the ordination score of the ancestral node of the branch.
-      \item The ``deltran'' model, always selecting the ordination score of the descendant node/tip of the branch.
+      \item The ``acctran'' model, always selecting the ordination score of the descendant node/tip of the branch.
+      \item The ``deltran'' model, always selecting the ordination score of the ancestral node of the branch.
       \item The ``random'' model, randomly selecting the ordination score of either the ancestor or the descendant of the branch.
@@
 
-Disparity changes through time for each of our four datasets (Figure \ref{figure:dtt1}, Supporting Information Appendix S2: Figures A1-A2). 
+Disparity changes through time for each of our four datasets (Fig. \ref{figure:dtt1}, Supporting Information Appendix S2: Figs A1-A2). 
 Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods.
@@
 % Wilcoxon results table
-%\begin{landscape}
-  \input{tables/table_wilcox.tex}  
-%\end{landscape}
+%\input{tables/table_wilcox.tex}
 
 \paragraph{Disparity peaks.}
-In the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets, disparity peaks occur much earlier when time-slicing rather than time binning approaches are used (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figures A3-A4).
-This is also true for stratigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, although when using equal time bins the peaks are later than the time-slicing methods, or very similar (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figures A3-A4).
-Across the three time binning methods, the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets have similar disparity peaks whichever method is used, whereas in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figures A3-A4).
+In the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets, disparity peaks occur much earlier when time-slicing rather than time binning approaches are used (Fig.
+\ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
+This is also true for stratigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, although when using equal time bins the peaks are later than the time-slicing methods, or very similar (Fig.
+\ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
+Across the three time binning methods, the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets have similar disparity peaks whichever method is used, whereas in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Fig.
+\ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
 
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch.pdf}
     \caption[Timing of peak disparity for four example datasets.]
@@
 \paragraph{Effects of mass extinction events.}
-Mass extinction events influence disparity in both the \cite{beckancient2014} and \cite{wright2017bayesian} datasets (Figure \ref{figure:extinction1}).
-However, whether this change in disparity is significant or not depends on the method used to create time sub-samples (Figure \ref{figure:extinction1}), and whether ages or epochs are used. 
+Mass extinction events influence disparity in both the \cite{beckancient2014} and \cite{wright2017bayesian} datasets (Fig. \ref{figure:extinction1}).
+However, whether this change in disparity is significant or not depends on the method used to create time sub-samples (Fig. \ref{figure:extinction1}), and whether ages or epochs are used. 
 In general, for the \cite{beckancient2014} dataset, time binning tended to give more significant results than time-slicing methods, but this was not the case for the \cite{wright2017bayesian} dataset.
@@
 However, if we had chosen to investigate crinoid disparity only using time bins and splitting these so the number of time bins was equal to the number of epochs (number), we would have concluded that the O-S extinction had no effect on crinoid disparity.
-This highlights the sensitivity of disparity-through-time analyses to the choice of time sub-sampling method. 
+Likewise, the timing of peak disparity differs among methods. 
+This is particularly evident when comparing stratigraphic time bins to time slicing methods, where for most of our datasets we see a much later time to peak disparity. 
+This could have major implications for our understanding of how morphological diversity changes through time, for example in response to climate effects.
+These results highlight the sensitivity of disparity-through-time analyses to the choice of time sub-sampling method. 
 Fortunately this issue is easy to solve; either disparity-through-time analyses should use, and report results from, multiple time sub-sampling methods (as demonstrated here), or great care should be taken in determining the appropriate time sub-samples to answer the question of interest.
@@
 Another limitation is that the method, like most phylogeny based methods, depends on ancestral state reconstructions.
-Care should be taken in interpreting these, as they are highly dependent on the data and models used to for the reconstructions \citep{Slateretal2012,doi:10.1080/10635150801910451}.
+Care should be taken in interpreting these, as they are highly dependent on the data and models used for the reconstructions \citep{Slateretal2012,doi:10.1080/10635150801910451}.
 
@@
 Time-slicing methods, with explicit models of evolution, provide an alternative to traditional time binning approaches.
-Note that while we introduce the time-slicing methods here, and describe their advantages, we do not suggest time-slicing is necessarily the ``best'' method for time sub-sampling. 
+Note that while we introduce the time-slicing methods here, and describe their advantages, we do not suggest time-slicing is necessarily the ``best'' method for time sub-sampling in all cases. 
 As with all methods, the choice of methodology should be appropriate for the question and data at hand.
@@
 \section{Supporting information}
-Additional Supporting Information can be found in the online version of this article:
-Appendix S1: Extra details of datasets and Appendix S2: Additional tables and figures.
+Additional Supporting Information can be found in the online version of this article:\\
+Appendix S1: Extra details of datasets. \\
+Appendix S2: Additional tables and figures.
 
-\section{Acknowledgments}
+\paragraph{Acknowledgments.}
 	NC thanks Mark Sutton and Philip Mannion for the invitation to contribute to the `Evolutionary Modelling' symposium at The Palaeontological Association Annual Meeting 2017.
@@
 
+\section{References}
+
 \bibliographystyle{palaeo} 
@@
 
+\newpage
+\section{Tables and Figures}
+
+% Dataset details table
+\begin{landscape}
+\input{tables/table_datasets.tex} 
+\end{landscape}
+
+% Wilcoxon results table
+\input{tables/table_wilcox.tex}
+
 \end{document}

### 2018-01-28 manuscript/response_to_reviewers.tex
Reviewers comments

\documentclass[12pt,letterpaper]{article}

%Packages
\usepackage{pdflscape}
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{color}
\usepackage[usenames,dvipsnames]{xcolor}
\pagenumbering{arabic}

%Pagination style and stuff
%\linespread{2}

\raggedright
\setlength{\parindent}{0.5in}
\setcounter{secnumdepth}{0} 
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}

\setlength\parindent{0pt}

\begin{document}

\textbf{RE: PALA-12-17-4130-OA}\\
\bigskip
Dear Dr Smith,\\
\bigskip

We are very grateful to the four referees for their helpful and constructive comments, blablabal...

% ~~~~~~~~~~~~~~~~~~~~~~~~
%
% REVIEWER 1
%
% ~~~~~~~~~~~~~~~~~~~~~~~~

\section{Reviewer 1 (Steve Brusatte)}

\begin{enumerate}

\item{\textcolor{blue}{In the section on adding in reconstructed ancestors to the dataset before ordination, I suppose you should mention that this is similar to the approach that we outlined in our Paleobiology paper (Brusatte et al. 2011) and mention any key differences.}}

\item{\textcolor{blue}{[...]I wonder if the term ``cladisto-space'' is confusing, and is just yet another new term that may muddy the waters rather than clarify. If you want to keep using this term, that’s fine, but a sentence or two of explanation (about how previous authors have used discrete characters in a non-phylogenetic way) would be warranted.}}

%TG: let's just remove cladisto-space: I think (1) it's a nice word and (2) it's actually a good decoy to focus reviewers attention and can be changed easily. + Steve makes a point with adding to many jargon.

\item{\textcolor{blue}{Figure 3 would be easier to interpret if the dataset names were added to the graphic itself (like in Figure 2) rather than constrained to the figure caption.}}

\item{\textcolor{blue}{I may be misinterpreting Figure 3, but it seems like peak disparity values for the Brusatte et al. dataset are pretty much the same regardless of method used (second plot from top). This doesn’t jibe with the statement in the text (lines 435-437). Please clarify.}}

\end{enumerate}

% ~~~~~~~~~~~~~~~~~~~~~~~~
%
% REVIEWER 2
%
% ~~~~~~~~~~~~~~~~~~~~~~~~

\section{Reviewer 2}

\begin{enumerate}

\item{\textcolor{blue}{First, if I understand lines 278-299 (and the R annotation) correctly, the gradual evolution models are not in fact fundamentally different from the punctuated models. They are both assigning scores along the branch that are the same as either the ancestor or descendant—the difference is only that the selection is based on a probability in the ``gradual'' models, rather than a rule, as in the ``punctuated'' models. The concept of gradual change would be more closely modeled by assigning values along a branch that were in-between those assigned to the ancestor and descendent but based on how far along the branch any given point is. For example, let’s say the value of the ancestor along PCO1 is 0.04 and that of the descendent is 0.03. For a point that is halfway along the branch linking them, the value should be 0.035 (instead of assigning either 0.03 or 0.04 based on a probability). This, of course, makes assumptions about the course of evolution between the ancestor and descendant, but at least would be describing gradual change.}}

%TG: I can implement that pretty easily but I think that mathematically it's not cool (we can not add values post-ordination in the ordinated space). Two options: (1) implementing the suggestion but with heavy warnings; (2) writing down a clever response to this comment. In the dispRity package review, that was highlighted (I bet reviewer 2 is Graeme) and I responded something similar to option (2) so I'd rather go that way for consistency.

\item{\textcolor{blue}{Second, I do not see how Figure 3 supports the text (lines 435-447). Overall, these results show the exact opposite of what is stated in the text (lines 435-447): in the figure the time-binned-approach (pink) results show earlier disparity peaks than the time-slicing approaches (green and blue). Are the y-axes reversed from what they should be? Also the Brusatte et al 2014 (second row) shows very similar values, not different values (compare with lines 435-436). And the difference between the ages of the peaks in the Bapst et 2016 study (third row) is at most 6 million years...but the study itself is over 150 million years divided very coarsely in Figure 2.  On that scale, these are basically the "same time". Finally, whether the six different time-slicing methods show ``similar'' patterns (lines 412-413) depends on the question... it is hard to distinguish them in the relative disparity plots, but the timing of peak disparity varies widely (Figure A3) while the two shown in Figure 3 happen to be similar.}}

\item{\textcolor{blue}{Finally, I have spent some time thinking about by the statement ``all time binning approaches (whether equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model.'' (lines 70-71). I see what the authors are getting at—they are worried that doing this requires an assumption of constant evolution across the time bin. But I would argue that it is not necessary to interpret it this way. Instead it could simply represent an average across the interval sampled without the need to make any assumption about how change occurred within that interval. Plus, even in scenarios where there was enough change that at a course scale, it appeared that disparity was static, some of that volatility might be revealed if the time-scale were altered.}}

\item{\textcolor{blue}{But there is also a deeper issue here, in that the fossil record is at its heart a time-binned record: rates of sedimentation vary and sediment deposition is sporadic, such that local beds represent different amounts of time, or different ``packages'' of time, separated by gaps, and the fossils within them may be time-averaged to various degrees. [...] Therefore, even if you pick a ``point in time'' and count the number of branches/ranges that cross that point, some of those are being captured because the resolution of the end points only allows us to know that the taxon started and/or ended somewhere in the interval encompassing that point in time.}}

%TG: this is a fair point we should include in the discussion.

\item{\textcolor{blue}{As an aside, the Hunt et al 2015 citation does not clarify the matter (line 74-75). A case could be made that all time series are binned (if you consider that the samples are likely time-averaged and there are gaps between samples--although this is different from dividing up "everything" into adjacent boxes), and yet there are time series that are best characterized by gradual directional change (e.g., best fit by a biased random walk). In addition, Hunt et al found that directional change was actually more likely to be detected in series of long duration. Furthermore, homogenous rate patterns were more likely to be supported than heterogenous/punctuated change with time series of fewer samples, which includes cases where each ``step'' includes a longer interval of time over which shorter changes could be hidden.}}

%TG: not sure where the reviewer want us to go with that. Is that the same nuancing point as the comment before? In that case we can simply merge them and make a caveat paragraph.

\end{enumerate}

\subsection{Reviewer 2 - Minor comments}

\begin{enumerate}

\item{\textcolor{blue}{It is not immediately clear what is meant by ``defined by the average duration of the stratigraphic period'' (line 337, 341) vs ``defined by the average number of stratigraphic periods'' (line 348, 353). After thinking about it for awhile and then checking the R code, this is what appears to be going on. Let’s say we have a case study that spans 10 stratigraphic intervals over 40 million years. Thus for the stratigraphy approach, there are 10 sub-samples which vary in their duration. Let’s say the average duration of those intervals is 5 million years. For the duration approach, the 40 million year span is divided into 5-million-year bins, resulting in 8 sub-samples. For the number approach, the 40 million year span is divided into 10, each of which is 4 million years long. If this is the case, the use of the word ``average'' when describing the number approach is confusing. I am also having trouble reconciling this with the number of points in each panel in Figure 2, unless there is also a minimum sample size applied (for example in the Bapst 2016 study, there are only three disparity values for the time-binned series using the ``number'' approach, maybe because after dividing the time scale up this way, there were not enough taxa in the first and second time bins?). It may be helpful to include a graphic, even if in relegated to one of the appendices.}}

%TG: good idea, we (I) can make a figure designating the three different strategies.

\item{\textcolor{blue}{Line 158. Gower’s distance is not the Euclidean distance (= the square root of the squared distances) over the number of shared characters. For a set of unordered characters, it is the number of mismatched character states over the number of comparable/shared characters. For a set of ordered characters, it is sum of the differences in the values of the characters states for each pairwise comparison, over the number of comparable characters. This is an edit that needs to be made only to the text; because the authors used functions from the Claddis package (see github documentation), they would have calculated Gower’s distance correctly.}}

\item{\textcolor{blue}{Consider using ``stage'' instead of ``age'' when referring to the finer resolution time scale.}}

\item{\textcolor{blue}{Figure 2. It is very difficult to distinguish the green dots from the blue dots. For someone who is color-blind, it would also be difficult to distinguish the red dots from the green dots.}}

\end{enumerate}

% ~~~~~~~~~~~~~~~~~~~~~~~~
%
% REVIEWER 3
%
% ~~~~~~~~~~~~~~~~~~~~~~~~

%TG: actually not sure how to write these comments in this response document, they are only technical thingies, we can just state in the intro paragraph above that we dealt with all of them. I'll raise them as tickable issues on GitHub (yay!) for now.

% \section{Reviewer 3 (Sally Thomas)}

% \begin{enumerate}

% \item{\textcolor{blue}{Please submit your manuscript in Word format. Do not include the figures in the text file, but it is helpful if you do include a list of the figure captions at the end.}}

% \item{\textcolor{blue}{Could I suggest adding ‘sub-sampling’ to your key words?}}

% \end{enumerate}

### 2018-01-29 manuscript/time-slice.tex
Updated acknowledgements

@@
   TG acknowledges support from the Australian Discovery Project Grant number DP170103227 awarded to Vera Weisbecker.
-  We thank Dave Bapst, Steve Brusatte, Graeme Lloyd, April Wright and Davey Wright for assistance in gathering data for the analyses and/or discussions about the approach.
+  We thank Dave Bapst, Graeme Lloyd, April Wright and Davey Wright for assistance in gathering data for the analyses and/or discussions about the approach; and Steve Brusatte and one anonymous reviewer for helpful comments on the manuscript.
 

### 2018-01-30 manuscript/response_to_reviewers.tex
added todo
Updated response to review

@@
 
+%TG: TODO: fix ages in table 1
+
 % ~~~~~~~~~~~~~~~~~~~~~~~~
@@
 
-\item{\textcolor{blue}{[...]I wonder if the term ``cladisto-space'' is confusing, and is just yet another new term that may muddy the waters rather than clarify.
+We now explicitly refer to Brusatte et al. 2011 and mention the key difference:
+
+\textit{This approach is similar to \cite{brusatte2011phylogenetic} but uses model based estimations (rather than parsimony) allowing to control for ambiguous nodes (i.e. poorly estimated).} lines @@@
+
+\item{\textcolor{blue}{I’m not sure about the use of the term ‘cladisto-spaces’.
+To me this implies that the space itself is ordinated using phylogenetic information (e.g., branch lengths).
+Now, you are indeed including some phylogenetic information in your morphospaces, in terms of reconstructed ancestors and inferred states for branches.
+So sure, I think what you have could be called a ‘cladisto-space’.
+Same with most of the morphospace options produced by Claddis.
+But the ones that most of us have done in the past, simply using discrete characters to ordinate a morphospace, haven’t generally incorporated phylogeny.
+I’m talking about the references you cite in lines 141-143 here.
+These morphospaces been ordinated using discrete characters of the type that are used in phylogenetic analysis.
+But the characters are not phylogenetically transformed in any way.
+As far as the ordination analysis knows, they are simply discrete character statements.
+So I wonder if the term ‘cladisto-space’ is confusing, and is just yet another new term that may muddy the waters rather than clarify.
 If you want to keep using this term, that’s fine, but a sentence or two of explanation (about how previous authors have used discrete characters in a non-phylogenetic way) would be warranted.}}
@@
 + Steve makes a point with adding to many jargon.
+%TG: actually I had a go at trying to change it and this might require your writing skills to make it flow (+ I'm hangover). I've tried a paragraph (commented out) in the relevant section.
 
@@
 
+We've changed the ``stratigraphic age'' to ``stratigraphic stage'' throughout the manuscript.
+
 \item{\textcolor{blue}{Figure 2.

### 2018-01-30 manuscript/time-slice.tex
Updated response to review

@@
 % NC: I've left this one bit to make it clear but removed the explanation (this was all just copied from your thesis)
-To prevent poor ancestral state reconstructions from biasing our results, especially when a lot of error is associated with the reconstruction, we only included ancestral state reconstructions with a scaled Likelihood $\geq$ $0.95$.
-Ancestral state reconstructions with scaled Likelihoods below this threshold were recoded as missing data (``?'').
-This allowed our results to be less dependent on the quality (or the absence thereof) of the ancestral state reconstruction methods, especially in parts of the datasets where data were sparse.
+To prevent poor ancestral state estimations from biasing our results, especially when a lot of error is associated with the estimations, we only included ancestral state estimations with a scaled Likelihood $\geq$ $0.95$.
+Ancestral state estimations with scaled Likelihoods below this threshold were recoded as missing data (``?'').
+This allowed our results to be less dependent on the quality (or the absence thereof) of the ancestral state estimations methods, especially in parts of the datasets where data were sparse. 
+This approach is similar to \cite{brusatte2011phylogenetic} but uses model based estimations (rather than parsimony) allowing to control for ambiguous nodes (i.e. poorly estimated).
 
@@
 
+% To explore disparity-through-time in our datasets we used a discrete morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
+% This approach is a approximation of a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010} and is based on discrete morphological data (i.e.
+the discrete morphological characters used to build a phylogenetic tree, often referred to as ``cladistic'' data).
+% Although empirically inter-taxon distances are the same in a morphospace based on discrete or continuous data \citep{foth2012different,hetherington2015cladistic}, we like to highlight that this space is based on discrete morphological data and phylogeny (it could be referred to as a ``cladisto-space'').
+% Mathematically, a cladisto-space is an $n$ dimensional object that summarises the distances between the taxa present in a discrete morphological matrix.
+
 \paragraph{Constructing distance matrices.}
@@
 because some pairs of taxa share more characters in common than others, they are more likely to be similar).
-For cladistic matrices, using this corrected distance is preferable to the raw Euclidean distance because of its ability to deal with discrete or/and ordinated characters as well as with missing data \citep{anderson2012using}.
+For discrete morphological matrices, using this corrected distance is preferable to the raw Euclidean distance because of its ability to deal with discrete or/and ordinated characters as well as with missing data \citep{anderson2012using}.
 However, the Gower distance cannot calculate distances when taxa have no overlapping data.
@@
 This often leads to bins of unequal duration.
-Here we use stratigraphic ages and epochs.
+Here we use stratigraphic stages and epochs.
   \item Equally sized time bins.
@@
     The four dataset names are on the first plot for each dataset (see Table \ref{table:datasets} for details).
-    Results for stratigraphic ages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A1-A2.}
+    Results for stratigraphic stages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A1-A2.}
     \label{figure:dtt1}
@@
     The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (see Table \ref{table:datasets} for details).
-    Results for stratigraphic ages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A3-A4.}
+    Results for stratigraphic stages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A3-A4.}
     \label{figure:peak1}
@@
     \textit{e:1}, \textit{e:2}, and \textit{e:3} denote whether the comparison was between the time bin or time-slice immediately after the mass extinction (\textit{e:1}), or the second (\textit{e:2}) or third (\textit{e:3}) bin/slice after the mass extinction to account for any lag effect.
-    The top seven rows use stratigraphic ages and the bottom seven rows use stratigraphic epochs.
+    The top seven rows use stratigraphic stages and the bottom seven rows use stratigraphic epochs.
     Labels on the left hand side indicate whether time bins (``bins'') were used, or which of the six time-slicing methods was used.}
@@
 Not all palaeontological datasets have phylogenies yet.
-Another limitation is that the method, like most phylogeny based methods, depends on ancestral state reconstructions.
-Care should be taken in interpreting these, as they are highly dependent on the data and models used for the reconstructions \citep{Slateretal2012,doi:10.1080/10635150801910451}.
+Another limitation is that the method, like most phylogeny based methods, depends on ancestral state estimations.
+Care should be taken in interpreting these, as they are highly dependent on the data and models used for the estimations \citep{Slateretal2012,doi:10.1080/10635150801910451}.
 

### 2018-01-31 manuscript/response_to_reviewers.tex
changed cladisto-space to morphospace (sigh...)

@@
 
-\textit{This approach is similar to \cite{brusatte2011phylogenetic} but uses model based estimations (rather than parsimony) allowing to control for ambiguous nodes (i.e. poorly estimated).} lines @@@
+\textit{This approach is similar to Brusatte et al. (2011) but uses model based estimations (rather than parsimony) allowing to control for ambiguous nodes (i.e. poorly estimated).} lines @@@
 
@@
 
-%TG: let's just remove cladisto-space: I think (1) it's a nice word and (2) it's actually a good decoy to focus reviewers attention and can be changed easily.
-+ Steve makes a point with adding to many jargon.
-%TG: actually I had a go at trying to change it and this might require your writing skills to make it flow (+ I'm hangover). I've tried a paragraph (commented out) in the relevant section.
+We've change the term cladisto-space to morphospace throughout the manuscript.
+Although we prefer the sound and distinction from the ``cladisto-space'' term, we agree with this reviewer concern about adding a``new term that may muddy the waters rather than clarify''.
+We've emphasised the differences, however, with other uses of the word ``morphospace'' in an explanatory paragraph:
+
+\textit{To explore disparity-through-time in our datasets, we used a morphospace approach (CITATIONS).
+Morphospace can be obtained from any multidimensional morphological data set but can differ in the data used (e.g. discrete or continuous) or whether they include phylogenetic data or not.
+Although empirical morphospaces from discrete or continuous data have been shown to have similar properties (CITATIONS), we would like to highlight that our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis;} c.f.
+\textit{geometric morphometric data) and includes some phylogenetic information (see above).
+Mathematically, our morphospaces are $n$ dimensional objects that summarises the distances between discrete morphological characters of the taxa present and their ancestors.} lines @@@
+
+%TG: what do you think?
 

### 2018-01-31 manuscript/time-slice.tex
changed cladisto-space to morphospace (sigh...)

@@
 
-\paragraph{Building cladisto-spaces.} 
-To explore disparity-through-time in our datasets we used a cladisto-space approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
-This approach is similar to constructing a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010}, except a cladisto-space is an approximation of the morphospace based on discrete morphological data (i.e.
+\paragraph{Building morphospaces.} 
+% To explore disparity-through-time in our datasets we used a morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
+% This approach is similar to constructing a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010}, except a morphospace is an approximation of the morphospace based on discrete morphological data (i.e.
 the discrete morphological characters used to build a phylogenetic tree, often referred to as ``cladistic'' data).
-%TG: about that, I'm actually less and less convinced we should call it caldisto-space: first, semantically, cladistic data means data used to make clades (phylos) so it could be nucleotides as well.
-Second, during the PhD, we used cladisto-space to distinguish it from the morpho-space but at that time I was naively thinking the the morpho-space is the morphometric-space (which it often is!) but, again, semantically, the morpho-space can also be simply the morphology-space...
-However, I like the sound of cladisto-space and most people will understand it the way we do.
+% %TG: about that, I'm actually less and less convinced we should call it caldisto-space: first, semantically, cladistic data means data used to make clades (phylos) so it could be nucleotides as well.
+Second, during the PhD, we used morphospace to distinguish it from the morpho-space but at that time I was naively thinking the the morpho-space is the morphometric-space (which it often is!) but, again, semantically, the morpho-space can also be simply the morphology-space...
+However, I like the sound of morphospace and most people will understand it the way we do.
 I'm just thinking about pedantic-palaeo-reviewers...
-% NC: Good point. Let's leave it for now as I think this bit is nicely explained, but easy enough to edit if people don't like it.
-Mathematically, a cladisto-space is an $n$ dimensional object that summarises the cladistic distances between the taxa present in a cladistic matrix.
-Although empirically inter-taxon distances are the same in a morphospace or a cladisto-space \citep{foth2012different,hetherington2015cladistic}, we prefer the term cladisto-space to make it clear that this space is estimated using cladistic data and not morphometric data.
+% % NC: Good point. Let's leave it for now as I think this bit is nicely explained, but easy enough to edit if people don't like it.
+% Mathematically, a morphospace is an $n$ dimensional object that summarises the cladistic distances between the taxa present in a cladistic matrix.
+% Although empirically inter-taxon distances are the same in a morphospace or a morphospace \citep{foth2012different,hetherington2015cladistic}, we prefer the term morphospace to make it clear that this space is estimated using cladistic data and not morphometric data.
 
-% To explore disparity-through-time in our datasets we used a discrete morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
-% This approach is a approximation of a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010} and is based on discrete morphological data (i.e.
-the discrete morphological characters used to build a phylogenetic tree, often referred to as ``cladistic'' data).
-% Although empirically inter-taxon distances are the same in a morphospace based on discrete or continuous data \citep{foth2012different,hetherington2015cladistic}, we like to highlight that this space is based on discrete morphological data and phylogeny (it could be referred to as a ``cladisto-space'').
-% Mathematically, a cladisto-space is an $n$ dimensional object that summarises the distances between the taxa present in a discrete morphological matrix.
+To explore disparity-through-time in our datasets, we used a morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
+Morphospace can be obtained from any multidimensional morphological data set but can differ in the data used (e.g. discrete or continuous) or whether they include phylogenetic data or not.
+Although empirical morphospaces from discrete or continuous data have been shown to have similar properties \citep{foth2012different,hetherington2015cladistic}, we would like to highlight that our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis; \textit{c.f.} geometric morphometric data) and includes some phylogenetic information (see above).
+Mathematically, our morphospaces are $n$ dimensional objects that summarises the distances between discrete morphological characters of the taxa present and their ancestors.
 
 \paragraph{Constructing distance matrices.}
-To estimate the cladisto-spaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
+To estimate the morphospaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
 We calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the Euclidean distance between two taxa divided by the number of shared characters. 
@@
 To avoid this problem, we first transformed the distance matrices by applying the Cailliez correction \citep{cailliez1983analytical} which adds a constant $c^*$ to the values in a distance matrix (apart from the diagonal) so that all the Gower distances become Euclidean ($d_{Gower}+c^*=d_{Euclidean}$; \citealt{cailliez1983analytical}).
-We were then able to extract $n$ eigenvectors for each matrix (representing the $n$ dimensions of the cladisto-space) where $n$ is equal to $k-2$, i.e.
+We were then able to extract $n$ eigenvectors for each matrix (representing the $n$ dimensions of the morphospace) where $n$ is equal to $k-2$, i.e.
 the number of taxa in the matrix ($k$) minus the last two eigenvectors that are always null after applying the Cailliez correction.
-Contrary to previous studies \citep[e.g][]{brusatte50,cisneros2010,prentice2011,anderson2012using,Hughes20082013,bentonmodels2014}, we use all $n$ dimensions of our cladisto-spaces and not a sub-sample representing the majority of the variance in the distance matrix (e.g.
+Contrary to previous studies \citep[e.g][]{brusatte50,cisneros2010,prentice2011,anderson2012using,Hughes20082013,bentonmodels2014}, we use all $n$ dimensions of our morphospaces and not a sub-sample representing the majority of the variance in the distance matrix (e.g.
 selecting only $m$ dimensions that represent up to 90\% of the variance in the distance matrix; \citealt{Brusatte12092008,toljagictriassic-jurassic2013}).
 
-Note that our cladisto-spaces represent an ordination of all possible morphologies coded in each study through time.
-It is unlikely that all morphologies will co-occur at each time point, therefore, the disparity of the whole cladisto-space is expected to be greater than the disparity at any specific point in time.
+Note that our morphospaces represent an ordination of all possible morphologies coded in each study through time.
+It is unlikely that all morphologies will co-occur at each time point, therefore, the disparity of the whole morphospace is expected to be greater than the disparity at any specific point in time.
 
@@
 \label{disparity_calc}
-Disparity can be calculated in many different ways \citep[e.g.][]{Wills1994,Ciampaglio2004,thorneresetting2011,hopkinsdecoupling2013,huang2015origins}, however a majority of studies in palaeobiology estimate disparity using four metrics: the sum and products of ranges and variances, each of which gives a slightly different estimate of how the data fits within the cladisto-space \citep{Foote01071994,Wills1994,brusatte50,Brusatte12092008,cisneros2010,thorneresetting2011,prentice2011,brusattedinosaur2012,toljagictriassic-jurassic2013,ruta2013,bentonmodels2014,bensonfaunal2014}.
+Disparity can be calculated in many different ways \citep[e.g.][]{Wills1994,Ciampaglio2004,thorneresetting2011,hopkinsdecoupling2013,huang2015origins}, however a majority of studies in palaeobiology estimate disparity using four metrics: the sum and products of ranges and variances, each of which gives a slightly different estimate of how the data fits within the morphospace \citep{Foote01071994,Wills1994,brusatte50,Brusatte12092008,cisneros2010,thorneresetting2011,prentice2011,brusattedinosaur2012,toljagictriassic-jurassic2013,ruta2013,bentonmodels2014,bensonfaunal2014}.
 However, these metrics have limitations. 
@@
 \label{time_slicing}
-The ``time-slicing'' approach considers subsets of taxa in the cladisto-space at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
-This results in even-sampling of the cladisto-space across time and allows us to use different underlying models of character evolution (punctuated or gradual).
+The ``time-slicing'' approach considers subsets of taxa in the morphospace at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
+This results in even-sampling of the morphospace across time and allows us to use different underlying models of character evolution (punctuated or gradual).
 
@@
 This method does not estimate values along a branch applying a model (\textit{c.f.} methods described for ancestral character estimation in the ``Preparing the data for disparity-through-time analysis'' section above) but rather chooses between the two available pieces of information (the ordination score of the descendant or the ancestor) using the methods described above.
-This allows the method to be used in post-ordination analysis where the data used in each time-slice is data already present in the cladisto-space.
-In other words, this method does not require a re-ordination of the cladisto-space every time a slice goes through a branch, thus allowing the properties of the cladisto-space (e.g.
+This allows the method to be used in post-ordination analysis where the data used in each time-slice is data already present in the morphospace.
+In other words, this method does not require a re-ordination of the morphospace every time a slice goes through a branch, thus allowing the properties of the morphospace (e.g.
 distance between species, variance of each axis, etc.) to remain constant.

### 2018-01-31 manuscript/response_to_reviewers.tex
Fixed comments from Steve

@@
 
+%TG: OK, I've tried but miserably failed to do it.
+ggplot's logic is just completely different than base-plot's one and I don't understand it...
+I'm actually wondering if ggplot is something like Word vs.
+LaTeX: once you're use to one it's really hard to switch to the other.
+%TG: Also for these figures, I think we need to try to plot a line rather than two dots for the bins interval. Its actually not that straight forward to realise what's happening.
+
 \item{\textcolor{blue}{I may be misinterpreting Figure 3, but it seems like peak disparity values for the Brusatte et al.
@@
 Please clarify.}}
+
+We apologies for mismatching the datasets and the patterns described in the text.
+We've now correct and double check this as follows:
+
+\textit{In the Beck \& Lee (2014) and Bapst et al.
+(2016) datasets, disparity peaks occur much at much older ages when time-slicing rather than time binning approaches are used (Fig.
+3; Supporting Information Appendix S2: Figs A3-A4).
+This is also true for stratigraphic time bins in the Wright (2017) dataset, although when using equal time bins the peaks are later than the time-slicing methods, or very similar (Fig.
+3; Supporting Information Appendix S2: Figs A3-A4).
+Across the three time binning methods, the Brusatte et al.
+(2014) dataset have similar disparity peaks whichever method is used, the Wright (2017) dataset only had variation in peaks when using un-equal time bins (stratigraphy) whereas in the Bapst et al.
+(2016) and Beck \& Lee (2014) datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Fig.
+3; Supporting Information Appendix S2: Figs A3-A4).} lines @@@
+
+We've also added a note about discrepancies in the time-slicing models in Beck \& Lee (2014):
+
+\textit{Additionally, there seem to be low discrepancy within the time-slicing methods (gradual vs.
+proximity) expect in Beck \& Lee (2014) dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig.
+3; Supporting Information Appendix S2: Figs A3-A4).} lines @@@
 

### 2018-01-31 manuscript/time-slice.tex
Fixed comments from Steve

@@
 \paragraph{Disparity peaks.}
-In the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets, disparity peaks occur much earlier when time-slicing rather than time binning approaches are used (Fig.
+In the \cite{beckancient2014} and \cite{bapst2016topology} datasets, disparity peaks occur much at much older ages when time-slicing rather than time binning approaches are used (Fig.
 \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
-This is also true for stratigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, although when using equal time bins the peaks are later than the time-slicing methods, or very similar (Fig.
+This is also true for stratigraphic time bins in the \cite{wright2017bayesian} dataset, although when using equal time bins the peaks are later than the time-slicing methods, or very similar (Fig.
 \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
-Across the three time binning methods, the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets have similar disparity peaks whichever method is used, whereas in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Fig.
+Across the three time binning methods, the \cite{brusatte2014gradual} dataset have similar disparity peaks whichever method is used, the \cite{wright2017bayesian} dataset only had variation in peaks when using un-equal time bins (stratigraphy) whereas in the \cite{bapst2016topology} and \cite{beckancient2014} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Fig.
+\ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
+Additionally, there seem to be low discrepancy within the time-slicing methods (gradual \textit{vs.} proximity) expect in \cite{beckancient2014} dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig.
 \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).

### 2018-02-02 manuscript/response_to_reviewers.tex
Updated response

@@
 They are both assigning scores along the branch that are the same as either the ancestor or descendant—the difference is only that the selection is based on a probability in the ``gradual'' models, rather than a rule, as in the ``punctuated'' models.
-The concept of gradual change would be more closely modeled by assigning values along a branch that were in-between those assigned to the ancestor and descendent but based on how far along the branch any given point is.
+The concept of gradual change would be more closely modelled by assigning values along a branch that were in-between those assigned to the ancestor and descendent but based on how far along the branch any given point is.
 For example, let’s say the value of the ancestor along PCO1 is 0.04 and that of the descendent is 0.03.
@@
 This, of course, makes assumptions about the course of evolution between the ancestor and descendant, but at least would be describing gradual change.}}
+
+The reviewer does indeed understands the \texttt{equal.splits} and \texttt{gradual.splits} models correctly: they selects the values in the matrix from the edge's descendant or ancestor with a probability value for each.
+We believe however that the reviewer's proposition of making these models ``actually'' gradual (e.g.
+by selecting the mean value between two points) might lead to incorrect interpretation of the results.
+
+The rational for proposing the \texttt{equal.splits} and \texttt{gradual.splits} models rather than a pure gradual model is to not create any extra data points in the ordinated space.
+In fact, doing so pre-ordination can simply change the ordinated space geometry while doing in post-ordination leads to incorrect estimation of the space: (1) the number of elements changed hence changing parameters for future statistical test (e.g.
+degrees of freedom) and (2) the variance per axis will change as well.
+On the other hand, our methods allows to approach a pure gradual model without modifying the space structure and properties.
+In fact, when applying the \texttt{equal.split} model (or \texttt{gradual.splits}) on a decent number of bootstrap replicates (say 100), the ancestor's PCO1 value (0.04 in the reviewer's example) will be sampled 50\% of the time and the same for the descendant's value (0.03 - with the probability changing as a function of branch length for the \texttt{gradual.splits} model).
+The resulting metric, based on this value would be $0.5\times0.04 + 0.5\times0.03 = 0.035$ thus identical to what the reviewer suggests without modifying the multidimensional space.
+
+We added the following explanation in the text:
+
+\textit{For example, using the ``equal.split'' model between an ancestor and a descendant with PCO1 values of respectively $0.04$ and $0.03$, after a sufficient number of bootstrap replicates (e.g.
+100) the value along the branch will be close to $0.5\times0.04 + 0.5\times0.03 = 0.035$.
+By estimating this value rather than generating it (i.e.
+creating a new element midway along the branch that would be the average of the descendant and ancestor - $0.035$) we obtain the same results without modifying the morphospace properties.} lines @@@
 
@@
 % \end{enumerate}
+
+\end{document}

### 2018-02-02 manuscript/time-slice.tex
Updated response

@@
 distance between species, variance of each axis, etc.) to remain constant.
+For example, using the ``equal.split'' model between an ancestor and a descendant with PCO1 values of respectively $0.04$ and $0.03$, after a sufficient number of bootstrap replicates (e.g.
+100) the value along the branch will be close to $0.5\times0.04 + 0.5\times0.03 = 0.035$.
+By estimating this value rather than generating it (i.e.
+creating a new element midway along the branch that would be the average of the descendant and ancestor - $0.035$) we obtain the same results without modifying the morphospace properties.
 

### 2018-02-02 manuscript/WithinOrdinationACE.Rmd
pre/post ordination ACE

---
title: "Within ordination ACE"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

Let's consider a square of size one in four dimensions

```{r, fig.width = 6, fig.height = 6}
## square coordinates
square <- matrix(ncol = 4, nrow = 4, byrow = TRUE,
                  data = c(c(0,0,0,0), c(0,1,0,0), c(1,1,0,0), c(1,0,0,0)))
rownames(square) <- letters[1:4]
## The 2D square
par(bty = "n")
plot(square[,c(1:2)], xlab = "Dimension 1", ylab = "Dimension 2")
```

We can then ordinate this segment using an MDS

```{r, fig.width = 6, fig.height = 6}
## Ordination (reducing only one dimension)
ord_square <- cmdscale(dist(square), k = 3)

## Plotting the two first dimensions
par(bty = "n")
plot(ord_square[,c(1:2)], xlab = "Ordination 1", ylab = "Ordination 2")
```

We can then estimate the values on the edges of the square (the mean between the pairs of summits) either within the ordination (average ordination scores) or directly from the non-ordinated square.

## Adding data within the ordination

We will now add four points that will be on the centre of each edges of the ordinated square

```{r, fig.width = 6, fig.height = 6}
## Getting the average in 3 dimensions for each edges
ab <- apply(ord_square[1:2,], 2, mean) 
bc <- apply(ord_square[2:3,], 2, mean) 
cd <- apply(ord_square[3:4,], 2, mean) 
da <- apply(ord_square[c(4,1),], 2, mean) 
ord_square_edge <- rbind(ord_square, ab, bc, cd, da)

## Plotting the two first dimensions
par(bty = "n")
plot(ord_square_edge[,c(1:2)], xlab = "Ordination 1", ylab = "Ordination 2")
```

## Adding data before the ordination

Alternatively, we can add the data in a post or pre ordination way.

```{r, fig.width = 12, fig.height = 6}
par(bty = "n", mfrow = c(1,2))

## Adding the average for each square edge
ab <- apply(square[1:2,], 2, mean) 
bc <- apply(square[2:3,], 2, mean) 
cd <- apply(square[3:4,], 2, mean) 
da <- apply(square[c(4,1),], 2, mean) 
square_edge <- rbind(square, ab, bc, cd, da)

## The 2D square (with edges center)
plot(square_edge[,c(1:2)], xlab = "Dimension 1", ylab = "Dimension 2", main = "post ordination addition")

## Ordination (reducing only one dimension)
ord_square_edge2<- cmdscale(dist(square_edge), k = 3)

## Plotting the two first dimensions
plot(ord_square_edge2[,c(1:2)], xlab = "Ordination 1", ylab = "Ordination 2", main = "pre ordination addition")
```

In the case of simple structures, addition of elements pre or post ordination has no effect on the results.
However, for more complex structures (i.e. phylogenies), this has a bigger impact on the resulting ordination:

# A phylogenetic example

Here we'll simulate a four taxa ladder tree with equal branch length (to make node estimation post-ordination comparable to pre-ordination - i.e. no effect of branch length).
We generate a realistic multistate discrete matrix (`dispRity::sim.morpho`) and estimate ancestral states for each node (`Claddis::AncStateEstMatrix`).
We also calculate the variance per axis (expressed in a probability).

```{r}
library(Claddis)
library(dispRity)
library(ape)
set.seed(1)

## Simulating a 4 taxa tree
tree <- read.tree(text = "(((t1,t2),t3),t4);")
tree$edge.length <- rep(1, 6)
## Setting up the parameters
my_rates = c(rgamma, rate = 10, shape = 5)
## Mk matrix (4*14)
matrixMk <- sim.morpho(tree, characters = 14, states = c(0.8, 0.2), model = "ER", rates = my_rates, invariant = FALSE)

## Estimating the ancestral characters
ancestors <- AncStateEstMatrix(MakeMorphMatrix(matrixMk), tree)

## Ordinating the matrix (without the ancestors)
ordination <- cmdscale(dist(matrixMk), k = nrow(matrixMk)-2)
ord_var <- apply(ordination, 2, var)
cum_var <- ord_var/sum(ord_var)
```

We can then plot both the tree and the position of each tip in the ordinated space as follows:

```{r, fig.width = 12, fig.height = 6}
## Plotting the two first axis
par(bty = "n", mfrow = c(1,2))
plot(ordination, xlab = paste0("O1 (", round(cum_var[1], 3),")"),
                 ylab = paste0("O2 (", round(cum_var[2], 3),")"),
                 xlim = c(-2.2, 2.2), ylim = c(-2.2, 2.2))

## Adding the tips names
for(tip in 1:Ntip(tree)) {
    text(ordination[tip, 1], ordination[tip, 2], labels = rownames(ordination)[tip], adj = -1)
}

## Adding the tree
plot(tree)
nodelabels(c("root", "n3", "n1.2"))
```

Note here the values of the ordination on each axis sums up to 1 since the ordination creates only two dimensions (using the $k = n-2$ rule for generating the number of axis where $k$ are the number of dimensions and $n$ the number of elements in the matrix).

### Pre/post ordination nodes

We can then add the estimates of our nodes pre or post ordination.

 * The post ordination nodes coordinates are the average values of its descendants (since branch length is always equal to one).
 * The pre ordination nodes coordinates are obtained by adding the ancestral states estimates to the input discrete morphological matrix (`matrixMk`) and re-ordinating the resulting matrix.

```{r}
## Adding the nodes to the ordination (as averages values)
n1.2 <- apply(ordination[1:2,], 2, mean) 
n3 <- apply(rbind(n1.2, ordination[3,]), 2, mean) 
root <- apply(rbind(n3, ordination[4,]), 2, mean) 
post_ordination <- rbind(ordination, n1.2, n3, root)
## Recalculating variance
post_ord_var <- apply(post_ordination, 2, var)
cum_post_ord_var <- post_ord_var/sum(post_ord_var)

## Rerunning the ordination with the ancestors
pre_ordination <- cmdscale(dist(rbind(matrixMk, ancestors)), k = nrow(rbind(matrixMk, ancestors))-2)
rownames(pre_ordination)[5:7] <- c("n1.2", "n3", "root")
## Recalculating variance
pre_ord_var <- apply(pre_ordination, 2, var)
cum_pre_ord_var <- pre_ord_var/sum(pre_ord_var)
```

For both, we recalculated the the variance per axis.
Note that due to the difference in input matrix size, the `post_ordination` matrix has still the same number of dimensions as previously (2) but the `pre_ordination` matrix has now five dimensions (7-2).

```{r, fig.width = 12, fig.height = 6}
## Plotting the two spaces
par(bty = "n", mfrow = c(1,2))

## The ordination within one
plot(post_ordination[1:4,], xlab = paste0("O1 (", round(cum_post_ord_var[1], 3),")"),
                        ylab = paste0("O2 (", round(cum_post_ord_var[2], 3),")"),
                        xlim = c(-2.2, 2.2), ylim = c(-2.2, 2.2),
                        main = "nodes added post ordination")

## Adding the tips names
for(tip in 1:(Ntip(tree))) {
    text(post_ordination[tip, 1], post_ordination[tip, 2],
        labels = rownames(post_ordination)[tip], adj = -1)
}

## The ordination without one
plot(pre_ordination[1:4,], xlab = paste0("O1 (", round(cum_pre_ord_var[1], 3),")"),
                        ylab = paste0("O2 (", round(cum_pre_ord_var[2], 3),")"),
                        xlim = c(-2.2, 2.2), ylim = c(-2.2, 2.2),
                        main = "nodes added pre ordination")

## Adding the tips names
for(tip in 1:(Ntip(tree))) {
    text(pre_ordination[tip, 1], pre_ordination[tip, 2],
        labels = rownames(pre_ordination)[tip], adj = -1)
}
```

In this case, regardless of the nodes position and/or estimation, there is a big variation in the tips position in the ordination when adding the nodes post or pre-ordination.
Note also the change in the axis variance between the post-ordination plot and the previous one (without node reconstructions).

#### Nodes position

```{r, fig.width = 12, fig.height = 6}
## Plotting the two spaces
par(bty = "n", mfrow = c(1,2))

## The ordination within one
plot(post_ordination, xlab = paste0("O1 (", round(cum_post_ord_var[1], 3),")"),
                        ylab = paste0("O2 (", round(cum_post_ord_var[2], 3),")"),
                        xlim = c(-2.2, 2.2), ylim = c(-2.2, 2.2),
                        main = "nodes added post ordination")

## Adding the tips names
for(tip in 1:(Ntip(tree))) {
    text(post_ordination[tip, 1], post_ordination[tip, 2],
        labels = rownames(post_ordination)[tip], adj = -1)
}

## The ordination without one
plot(pre_ordination, xlab = paste0("O1 (", round(cum_pre_ord_var[1], 3),")"),
                        ylab = paste0("O2 (", round(cum_pre_ord_var[2], 3),")"),
                        xlim = c(-2.2, 2.2), ylim = c(-2.2, 2.2),
                        main = "nodes added pre ordination")

## Adding the tips names
for(tip in 1:(Ntip(tree))) {
    text(pre_ordination[tip, 1], pre_ordination[tip, 2],
        labels = rownames(pre_ordination)[tip], adj = -1)
}
```

### 2018-02-05 manuscript/response_to_reviewers.tex
added minor comments
Minor edits to TG comments

@@
 
-We are very grateful to the four referees for their helpful and constructive comments, blablabal...
+We are very grateful to the referees for their helpful and constructive comments. 
+Below we provide their comments in blue text with our responses.
 
@@
 
-\item{\textcolor{blue}{I’m not sure about the use of the term ‘cladisto-spaces’.
+\item{\textcolor{blue}{I'm not sure about the use of the term ‘cladisto-spaces’.
 To me this implies that the space itself is ordinated using phylogenetic information (e.g., branch lengths).
@@
 Same with most of the morphospace options produced by Claddis.
-But the ones that most of us have done in the past, simply using discrete characters to ordinate a morphospace, haven’t generally incorporated phylogeny.
+But the ones that most of us have done in the past, simply using discrete characters to ordinate a morphospace, haven't generally incorporated phylogeny.
 I’m talking about the references you cite in lines 141-143 here.
@@
 So I wonder if the term ‘cladisto-space’ is confusing, and is just yet another new term that may muddy the waters rather than clarify.
-If you want to keep using this term, that’s fine, but a sentence or two of explanation (about how previous authors have used discrete characters in a non-phylogenetic way) would be warranted.}}
+If you want to keep using this term, that's fine, but a sentence or two of explanation (about how previous authors have used discrete characters in a non-phylogenetic way) would be warranted.}}
 
 We've change the term cladisto-space to morphospace throughout the manuscript.
-Although we prefer the sound and distinction from the ``cladisto-space'' term, we agree with this reviewer concern about adding a``new term that may muddy the waters rather than clarify''.
+Although we prefer the ``cladisto-space'' term, we agree with this concern about adding a ``new term that may muddy the waters rather than clarify''.
 We've emphasised the differences, however, with other uses of the word ``morphospace'' in an explanatory paragraph:
@@
 \textit{To explore disparity-through-time in our datasets, we used a morphospace approach (CITATIONS).
-Morphospace can be obtained from any multidimensional morphological data set but can differ in the data used (e.g. discrete or continuous) or whether they include phylogenetic data or not.
-Although empirical morphospaces from discrete or continuous data have been shown to have similar properties (CITATIONS), we would like to highlight that our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis;} c.f.
+Morphospaces can be obtained from any multidimensional morphological data set but can differ in the data used (e.g.
+discrete or continuous), and whether they include phylogenetic data or not.
+% NC: do you mean cladistic data i.e.e data used to build phylogenies?
+Although empirical morphospaces from discrete or continuous data have been shown to have similar properties (CITATIONS), our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis;} c.f.
 \textit{geometric morphometric data) and includes some phylogenetic information (see above).
-Mathematically, our morphospaces are $n$ dimensional objects that summarises the distances between discrete morphological characters of the taxa present and their ancestors.} lines @@@
+Mathematically, our morphospaces are $n$ dimensional objects that summarise the distances between discrete morphological characters of the taxa present and their ancestors.} lines @@@
 
-%TG: what do you think?
+%TG: what do you think? % NC: sounds good, made a few minor edits above
 
@@
 LaTeX: once you're use to one it's really hard to switch to the other.
+% NC: Ha yeah the reason it's not like in Fig 2 is that it was a pain in the ass. I partly did all these in ggplot because I'm trying to get better at it. Will work on this.
 %TG: Also for these figures, I think we need to try to plot a line rather than two dots for the bins interval. Its actually not that straight forward to realise what's happening.
+% NC: No! The lines are misleading, technically must be points, as lines suggest there is data throughout rather than at discrete points. Bit of a bugbear of mine...
 
@@
 dataset are pretty much the same regardless of method used (second plot from top).
-This doesn’t jibe with the statement in the text (lines 435-437).
+This doesn't jibe with the statement in the text (lines 435-437).
 Please clarify.}}
+\label{fig3_feckup}
 
-We apologies for mismatching the datasets and the patterns described in the text.
-We've now correct and double check this as follows:
+We apologise for mismatching the datasets and the patterns described in the text.
+We've now corrected this error and double check this as follows:
 
@@
 Across the three time binning methods, the Brusatte et al.
-(2014) dataset have similar disparity peaks whichever method is used, the Wright (2017) dataset only had variation in peaks when using un-equal time bins (stratigraphy) whereas in the Bapst et al.
+(2014) dataset has similar disparity peaks whichever method is used, the Wright (2017) dataset only had variation in peaks when using unequal time bins (stratigraphy), whereas in the Bapst et al.
 (2016) and Beck \& Lee (2014) datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Fig.
@@
 
-\textit{Additionally, there seem to be low discrepancy within the time-slicing methods (gradual vs.
-proximity) expect in Beck \& Lee (2014) dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig.
+\textit{Additionally, there seems to be small discrepancies within the time-slicing methods (gradual vs.
+proximity) except in the Beck \& Lee (2014) dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig.
 3; Supporting Information Appendix S2: Figs A3-A4).} lines @@@
@@
 \item{\textcolor{blue}{First, if I understand lines 278-299 (and the R annotation) correctly, the gradual evolution models are not in fact fundamentally different from the punctuated models.
-They are both assigning scores along the branch that are the same as either the ancestor or descendant—the difference is only that the selection is based on a probability in the ``gradual'' models, rather than a rule, as in the ``punctuated'' models.
+They are both assigning scores along the branch that are the same as either the ancestor or descendant — the difference is only that the selection is based on a probability in the ``gradual'' models, rather than a rule, as in the ``punctuated'' models.
 The concept of gradual change would be more closely modelled by assigning values along a branch that were in-between those assigned to the ancestor and descendent but based on how far along the branch any given point is.
-For example, let’s say the value of the ancestor along PCO1 is 0.04 and that of the descendent is 0.03.
+For example, let's say the value of the ancestor along PCO1 is 0.04 and that of the descendent is 0.03.
 For a point that is halfway along the branch linking them, the value should be 0.035 (instead of assigning either 0.03 or 0.04 based on a probability).
@@
 
-The reviewer does indeed understands the \texttt{equal.splits} and \texttt{gradual.splits} models correctly: they selects the values in the matrix from the edge's descendant or ancestor with a probability value for each.
+The reviewer does indeed understand the \texttt{equal.splits} and \texttt{gradual.splits} models correctly: they select the values in the matrix from the edge's descendant or ancestor with a probability value for each.
 We believe however that the reviewer's proposition of making these models ``actually'' gradual (e.g.
@@
 The rational for proposing the \texttt{equal.splits} and \texttt{gradual.splits} models rather than a pure gradual model is to not create any extra data points in the ordinated space.
-In fact, doing so pre-ordination can simply change the ordinated space geometry while doing in post-ordination leads to incorrect estimation of the space: (1) the number of elements changed hence changing parameters for future statistical test (e.g.
+In fact, doing so pre-ordination can change the ordinated space geometry, while doing so post-ordination leads to the incorrect estimation of the space: (1) the number of elements changes hence changing parameters for future statistical tests (e.g.
 degrees of freedom) and (2) the variance per axis will change as well.
-On the other hand, our methods allows to approach a pure gradual model without modifying the space structure and properties.
+Our methods allow us to approach a pure gradual model without modifying the ordinated space's structure and properties.
 In fact, when applying the \texttt{equal.split} model (or \texttt{gradual.splits}) on a decent number of bootstrap replicates (say 100), the ancestor's PCO1 value (0.04 in the reviewer's example) will be sampled 50\% of the time and the same for the descendant's value (0.03 - with the probability changing as a function of branch length for the \texttt{gradual.splits} model).
-The resulting metric, based on this value would be $0.5\times0.04 + 0.5\times0.03 = 0.035$ thus identical to what the reviewer suggests without modifying the multidimensional space.
+The resulting metric, based on this value would be $0.5\times0.04 + 0.5\times0.03 = 0.035$ thus identical to what the reviewer suggests, without modifying the multidimensional space.
 
-We added the following explanation in the text:
+We realise we had not made this clear, thus we have added the following explanation in the text:
 
-\textit{For example, using the ``equal.split'' model between an ancestor and a descendant with PCO1 values of respectively $0.04$ and $0.03$, after a sufficient number of bootstrap replicates (e.g.
+\textit{For example, using the ``equal.splits'' model on an ancestor and a descendant with PCO1 values of respectively $0.04$ and $0.03$, after a sufficient number of bootstrap replicates (e.g.
 100) the value along the branch will be close to $0.5\times0.04 + 0.5\times0.03 = 0.035$.
 By estimating this value rather than generating it (i.e.
-creating a new element midway along the branch that would be the average of the descendant and ancestor - $0.035$) we obtain the same results without modifying the morphospace properties.} lines @@@
+creating a new element mid-way along the branch that would be the average of the descendant and ancestor - $0.035$) we obtain the same results without modifying the morphospace properties.} lines @@@
 
@@
 In the dispRity package review, that was highlighted (I bet reviewer 2 is Graeme) and I responded something similar to option (2) so I'd rather go that way for consistency.
+
+% NC: I like this response. I dunno if it is Graeme, they use US spellings...
 
@@
 
+We apologise for the confusion in the description of Figure 3 and have fixed it (see Reviewer 1, comment \ref{fig3_feckup}).
+We were rushing to get this submitted by the symposium deadline and this error slipped through the cracks.
+
 \item{\textcolor{blue}{Finally, I have spent some time thinking about by the statement ``all time binning approaches (whether equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model.'' (lines 70-71).
-I see what the authors are getting at—they are worried that doing this requires an assumption of constant evolution across the time bin.
+I see what the authors are getting at — they are worried that doing this requires an assumption of constant evolution across the time bin.
 But I would argue that it is not necessary to interpret it this way.
@@
 Plus, even in scenarios where there was enough change that at a course scale, it appeared that disparity was static, some of that volatility might be revealed if the time-scale were altered.}}
+
+%TG: tone done the statement? % NC: yeah, not sure what else we can do.
 
@@
 %TG: this is a fair point we should include in the discussion.
+%TG: though that is just a "practical" point for the vertebrate fossil record. Foraminiferas, for example, do not behave this way (or at a way finer scale).
 
@@
 %TG: not sure where the reviewer want us to go with that. Is that the same nuancing point as the comment before? In that case we can simply merge them and make a caveat paragraph. 
+% NC: Yeah I'm not sure either, but maybe needs a paragraph in the discussion? Currently the discussion is short (and written by me in about an hour!) so would be good to add something there.
 
@@
 
-We've changed the ``stratigraphic age'' to ``stratigraphic stage'' throughout the manuscript.
+We have changed ``stratigraphic age'' to ``stratigraphic stage'' throughout the manuscript.
 
@@
 For someone who is color-blind, it would also be difficult to distinguish the red dots from the green dots.}}
+
+%TG: I'll let you change the colours (I still don't want to touch ggplot!) how about the darkgrey/orange/blue palette?
+%Here are the hexadecimals:
+%colours_blind <- c("orange" = "#F65205", "blue" = "#3E9CBA", "darkgrey" = "#1C1C1C")
+%NC: The colours are colour blind friendly already, but I'll switch to using viridis.
+
+Interestingly the palettes in ggplot2 are chosen so that they are distinguishable by people that are colour blind.
+But it's no problem to change the colours.
+Hopefully the new colour scheme is easier to discern.
+We have used the viridis colour palettes which are designed for precisely this purpose (https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html), and are colour-blind friendly too.
 
@@
 
+% NC: For simplicity I suggest we put them here too, then it's easy for the editor to just tick them off rather than having to check. Also makes sure that we definitely do it all!
+
 % \section{Reviewer 3 (Sally Thomas)}

### 2018-02-05 analyses/05-figure-dtt.Rmd
Change colours to viridis

@@
 library(grid)
+library(viridis)
 source("../functions/multiplot_fun.R")
@@
         axis.title.x = element_blank(),
-        legend.position = "none")
+        legend.position = "none") +
+  scale_colour_viridis(discrete = TRUE)
 

### 2018-02-05 analyses/05-figure-dtt.Rmd
Remove yellow colour from scale and replace with green

@@
         legend.position = "none") +
-  scale_colour_viridis(discrete = TRUE)
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 
@@
         legend.position = "none") +
-  scale_colour_viridis(discrete = TRUE)
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8))
 
@@
 
-pdf("../manuscript/figures/fig-dtt-age.pdf")  
+pdf("../manuscript/figures/fig-dtt-age-revision.pdf")  
   multiplot(beck_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25), "points")),
@@
 
-pdf("../manuscript/figures/fig-dtt-epoch.pdf")  
+pdf("../manuscript/figures/fig-dtt-epoch-revision.pdf")  
   multiplot(beck_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")),

### 2018-02-05 analyses/05-figure-dtt.Rmd
Typo

@@
         legend.position = "none") +
-  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8))
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 

### 2018-02-05 analyses/05-figure-dtt.Rmd
Make error bars more transparent and wider

@@
   geom_point(data = strat2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 2, alpha = 0.3) +
   ggtitle("Stratigraphy") +
@@
   geom_point(data = duration2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 2, alpha = 0.3) +
   ggtitle("Duration") +
@@
   geom_point(data = number2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 2, alpha = 0.3) +
   ggtitle("Number") +
@@
   geom_point(data = strat2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 2, alpha = 0.3) +
   annotate("text", x = 140, y = 0.15, label = "Brusatte2014")
@@
   geom_point(data = duration2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 2, alpha = 0.3)
 
@@
   geom_point(data = number2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+  geom_errorbar(data = number2, aes(ymin = relative.lwr, ymax = relative.upr), width = 2, alpha = 0.3)
 
@@
   geom_point(data = strat2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 2, alpha = 0.3) +
   annotate("text", x = 120, y = 0.15, label = "Bapst2016")
@@
   geom_point(data = strat2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 2, alpha = 0.3) +
   annotate("text", x = 440, y = 0.15, label = "Wright2017")
@@
   geom_point(data = duration2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 2, alpha = 0.3) +
   scale_x_reverse(limits = c(470, 370),
@@
   geom_point(data = strat2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 2, alpha = 0.3) +
   annotate("text", x = 165, y = 0.15, label = "Bapst2016")

### 2018-02-05 manuscript/response_to_reviewers.tex
Add ggplot fix comment

@@
 
-%TG: OK, I've tried but miserably failed to do it.
-ggplot's logic is just completely different than base-plot's one and I don't understand it...
-I'm actually wondering if ggplot is something like Word vs.
-LaTeX: once you're use to one it's really hard to switch to the other.
-% NC: Ha yeah the reason it's not like in Fig 2 is that it was a pain in the ass. I partly did all these in ggplot because I'm trying to get better at it. Will work on this.
+We have done as suggested - we didn't do this the first time round because of issues with ggplot.
+
 %TG: Also for these figures, I think we need to try to plot a line rather than two dots for the bins interval. Its actually not that straight forward to realise what's happening.
@@
 
+A bigger issue is that a lot of the points lie on top of each other.
+We have played around with this a lot and there's no satisfactory solution.
+For now we have made the error bars more transparent so hopefully it is a bit easier to see which points lie where.
+
 \end{enumerate}

### 2018-02-05 analyses/07-figure-peaks.Rmd
Add viridis and annotations

@@
         panel.border = element_blank(),
-        strip.background = element_blank())
+        strip.background = element_blank()) +
+  annotate("text", x = 1, y = 0.15, label = "Beck2012")
 

### 2018-02-05 analyses/07-figure-peaks.Rmd
Add names of datasets to plots and increase facet label size

@@
 source("../functions/multiplot_fun.R")
+```
+
+```{r}
+# Make mini dataframe for annotations
+ann_text <- data.frame(model = "bins", time1 = 5, bin_type = "Stratigraphy")
 ```
@@
         panel.border = element_blank(),
-        strip.background = element_blank()) +
-  annotate("text", x = 1, y = 0.15, label = "Beck2012")
+        strip.background = element_blank(),
+        strip.text.x = element_text(size = 12)) +
+  geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Beck2012"), col = "black")
 
@@
         panel.border = element_blank(),
-        strip.background = element_blank())
+        strip.background = element_blank(),
+        strip.text.x = element_text(size = 12)) +
+ geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Beck2012"), col = "black")
 ```
@@
         panel.border = element_blank(),
-        strip.text = element_blank())
+        strip.text = element_blank(),
+       strip.text.x = element_text(size = 12)) +
+ geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Brusatte2014"), col = "black"))
 
@@
         panel.border = element_blank(),
-        strip.text = element_blank()) 
+        strip.text = element_blank(),
+       strip.text.x = element_text(size = 12)) +
+ geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Brusatte2014"), col = "black")) 
 ```
@@
         panel.border = element_blank(),
-        strip.text = element_blank())
+        strip.text = element_blank(),
+       strip.text.x = element_text(size = 12)) +
+ geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Bapst2016"), col = "black")
 
@@
         panel.border = element_blank(),
-        strip.text = element_blank())
+        strip.text = element_blank(),
+       strip.text.x = element_text(size = 12)) +
+ geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Bapst2016"), col = "black")
 ```
@@
         panel.border = element_blank(),
-        strip.text = element_blank())
+        strip.text = element_blank(),
+       strip.text.x = element_text(size = 12)) +
+ geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Wright2017"), col = "black")
 
@@
         panel.border = element_blank(),
-        strip.text = element_blank()) 
+        strip.text = element_blank(),
+       strip.text.x = element_text(size = 12)) +
+ geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Wright2017"), col = "black")  
 ```

### 2018-02-05 analyses/07-figure-peaks.Rmd
Change colours to viridis

--- 2018-02-05 analyses/07-figure-peaks.Rmd
@@
         panel.border = element_blank(),
-        strip.background = element_blank()) +
-  annotate("text", x = 1, y = 0.15, label = "Beck2012")
+        strip.background = element_blank(),
+        strip.text.x = element_text(size = 12)) +
+  geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Beck2012"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 
@@
         panel.border = element_blank(),
-        strip.background = element_blank())
+        strip.background = element_blank(),
+        strip.text.x = element_text(size = 12)) +
+ geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Beck2012"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 ```
@@
         panel.border = element_blank(),
-        strip.text = element_blank())
+        strip.text = element_blank(),
+       strip.text.x = element_text(size = 12)) +
+ geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Brusatte2014"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 
@@
         panel.border = element_blank(),
-        strip.text = element_blank()) 
+        strip.text = element_blank(),
+       strip.text.x = element_text(size = 12)) +
+ geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Brusatte2014"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 ```
@@
         panel.border = element_blank(),
-        strip.text = element_blank())
+        strip.text = element_blank(),
+       strip.text.x = element_text(size = 12)) +
+ geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Bapst2016"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 
@@
         panel.border = element_blank(),
-        strip.text = element_blank())
+        strip.text = element_blank(),
+       strip.text.x = element_text(size = 12)) +
+ geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Bapst2016"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 ```
@@
         panel.border = element_blank(),
-        strip.text = element_blank())
+        strip.text = element_blank(),
+       strip.text.x = element_text(size = 12)) +
+ geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Wright2017"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 
@@
         panel.border = element_blank(),
-        strip.text = element_blank()) 
+        strip.text = element_blank(),
+       strip.text.x = element_text(size = 12)) +
+ geom_text(data = ann_text, aes(x = 1, y = 5, 
+            label = "Wright2017"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8) 
 ```

### 2018-02-05 analyses/07-figure-peaks.Rmd
Change figure names

@@
 
-pdf("../manuscript/figures/fig-peaks-epoch.pdf")  
+pdf("../manuscript/figures/fig-peaks-epoch-revision.pdf")  
   multiplot(beck_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
@@
 
-pdf("../manuscript/figures/fig-peaks-age.pdf")  
+pdf("../manuscript/figures/fig-peaks-age-revision.pdf")  
   multiplot(beck_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 

### 2018-02-05 analyses/07-figure-peaks.Rmd
Fix label positioning

@@
         strip.text.x = element_text(size = 12)) +
-  geom_text(data = ann_text, aes(x = 1, y = 5, 
+  geom_text(data = ann_text, aes(x = 1, y = 10, 
             label = "Beck2012"), col = "black") +
@@
         strip.text.x = element_text(size = 12)) +
- geom_text(data = ann_text, aes(x = 1, y = 5, 
+ geom_text(data = ann_text, aes(x = 1, y = 10, 
             label = "Beck2012"), col = "black") +
@@
        strip.text.x = element_text(size = 12)) +
- geom_text(data = ann_text, aes(x = 1, y = 5, 
-            label = "Brusatte2014"), col = "black") +
+ geom_text(data = ann_text, aes(x = 1, y = 20, 
+            label = "   Brusatte2014"), col = "black") +
   scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
@@
        strip.text.x = element_text(size = 12)) +
- geom_text(data = ann_text, aes(x = 1, y = 5, 
+ geom_text(data = ann_text, aes(x = 1, y = 72, 
             label = "Bapst2016"), col = "black") +
@@
        strip.text.x = element_text(size = 12)) +
- geom_text(data = ann_text, aes(x = 1, y = 5, 
-            label = "Wright2017"), col = "black") +
+ geom_text(data = ann_text, aes(x = 1, y = 465, 
+            label = " Wright2017"), col = "black") +
   scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
@@
        strip.text.x = element_text(size = 12)) +
- geom_text(data = ann_text, aes(x = 1, y = 5, 
-            label = "Wright2017"), col = "black") +
+ geom_text(data = ann_text, aes(x = 1, y = 465, 
+            label = " Wright2017"), col = "black") +
   scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8) 

### 2018-02-05 analyses/07-figure-peaks.Rmd
Remove code for facet labels (they are removed in the last three datasets so cannot be modified)

@@
 ```
+
 ```{r}
@@
         panel.border = element_blank(),
-        strip.text = element_blank(),
-       strip.text.x = element_text(size = 12)) +
+        strip.text = element_blank()) +
  geom_text(data = ann_text, aes(x = 1, y = 20, 
@@
         panel.border = element_blank(),
-        strip.text = element_blank(),
-       strip.text.x = element_text(size = 12)) +
+        strip.text = element_blank()) +
  geom_text(data = ann_text, aes(x = 1, y = 72, 
@@
         panel.border = element_blank(),
-        strip.text = element_blank(),
-       strip.text.x = element_text(size = 12)) +
+        strip.text = element_blank()) +
  geom_text(data = ann_text, aes(x = 1, y = 465, 

### 2018-02-05 analyses/07-figure-peaks.Rmd
Fix to ensure correct annotations

@@
 source("../functions/multiplot_fun.R")
-```
-
-```{r}
-# Make mini dataframe for annotations
-ann_text <- data.frame(model = "bins", time1 = 5, bin_type = "Stratigraphy")
 ```
@@
 ```{r}
+# Make mini dataframe for annotations
+ann_text <- data.frame(model = "bins", time1 = 10, bin_type = "Stratigraphy")
+
 beck_peaks <- 
@@
 ```{r}
+# Make mini dataframe for annotations
+ann_text <- data.frame(model = "bins", time1 = 20, bin_type = "stratigraphy")
+
 brusatte_peaks <- 

### 2018-02-05 analyses/07-figure-peaks.Rmd
Edit alignment of Wright label

@@
   scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- expand_limits(y = 480) +
+ expand_limits(y = c(455, 480)) +
  theme(axis.title.y = element_blank(),
@@
         strip.text = element_blank()) +
- geom_text(data = ann_text, aes(x = 1, y = 465, 
+ geom_text(data = ann_text, aes(x = 1, y = 460, 
             label = " Wright2017"), col = "black") +

### 2018-02-05 manuscript/time-slice.tex
Add subsampling to keywords
Edit fig 2 legend

@@
 \runninghead{Time sub-samples in disparity-through-time analyses}
-\keywords{time bin, time-slice, disparity} 
+\keywords{time bin, time-slice, disparity, sub-sampling} 
 % up to 6
@@
     {Relative disparity-through-time. Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Pink points represent time binning methods, blue points are time-slices with a punctuated model of evolution (`proximity' method), and green points are time-slices with a gradual model of evolution (`gradual splits' method).
+    Green points represent time binning methods, purple points are time-slices with a punctuated model of evolution (`proximity' method), and blue points are time-slices with a gradual model of evolution (`gradual splits' method).
     Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods.

### 2018-02-05 manuscript/time-slice.tex
Change fig 2 and 3 to revision versions

@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch-revision.pdf}
     \caption[Relative disparity-through-time.]
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch-revision.pdf}
     \caption[Timing of peak disparity for four example datasets.]

### 2018-02-05 manuscript/time-slice.tex
Edits to fig 3 legend

@@
     Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Pink points represent time binning methods, green points are time-slices with a punctuated model of evolution (`proximity' method), and blue points are time-slices with a gradual model of evolution (`gradual splits' method).
+    Green points represent time binning methods, blue points are time-slices with a punctuated model of evolution (`proximity' method), and purple points are time-slices with a gradual model of evolution (`gradual splits' method).
     Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
@@
     For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
-    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (see Table \ref{table:datasets} for details).
+    The four dataset names are on the first plot for each dataset (see Table \ref{table:datasets} for details).
     Results for stratigraphic stages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A3-A4.}

### 2018-02-05 analyses/11-appendix-figure-peaks.Rmd
Move appendix peak figures out to separate file
Add edits from main text figure

@@
 ---
-title: "Disparity peaks"
+title: "Disparity peaks - for appendix"
 author: "Natalie Cooper"
@@
         strip.background = element_blank(),
-        strip.text = element_text(size = 14) )
+        strip.text = element_text(size = 14), 
+       strip.text.x = element_text(size = 12)) +
+  geom_text(data = ann_text, aes(x = 1, y = 10, 
+            label = "Beck2012"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 
@@
         strip.background = element_blank(),
-        strip.text = element_text(size = 14))
+        strip.text = element_text(size = 14),
+       strip.text.x = element_text(size = 12)) +
+  geom_text(data = ann_text, aes(x = 1, y = 10, 
+            label = "Beck2012"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 ```
@@
 ```{r}
+# Make mini dataframe for annotations
+ann_text <- data.frame(model = "bins", time1 = 10, bin_type = "stratigraphy")
+
 brusatte_peaks <- 
@@
         panel.border = element_blank(),
-        strip.text = element_blank())
+        strip.text = element_blank()) +
+ geom_text(data = ann_text, aes(x = 1, y = 20, 
+            label = "    Brusatte2014"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 
@@
         panel.border = element_blank(),
-        strip.text = element_blank()) 
+        strip.text = element_blank()) +
+ geom_text(data = ann_text, aes(x = 1, y = 20, 
+            label = "    Brusatte2014"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 ```
@@
         panel.border = element_blank(),
-        strip.text = element_blank())
+        strip.text = element_blank()) +
+ geom_text(data = ann_text, aes(x = 1, y = 72, 
+            label = "Bapst2016"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 
@@
         panel.border = element_blank(),
-        strip.text = element_blank())
+        strip.text = element_blank()) +
+ geom_text(data = ann_text, aes(x = 1, y = 72, 
+            label = "Bapst2016"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 ```
@@
   #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- expand_limits(y = 480) +
+ expand_limits(y = c(455, 480)) +
  theme(axis.title.y = element_blank(),
@@
         panel.border = element_blank(),
-        strip.text = element_blank())
+        strip.text = element_blank()) +
+ geom_text(data = ann_text, aes(x = 1, y = 460, 
+            label = " Wright2017"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 
@@
         panel.border = element_blank(),
-        strip.text = element_blank()) 
+        strip.text = element_blank()) +
+ geom_text(data = ann_text, aes(x = 1, y = 460, 
+            label = " Wright2017"), col = "black") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)   
 ```
@@
 
-pdf("../manuscript/figures/fig-peaks-epoch-appendix.pdf")  
+pdf("../manuscript/figures/fig-peaks-epoch-appendix-revision.pdf")  
   multiplot(beck_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 
@@
 
-pdf("../manuscript/figures/fig-peaks-age-appendix.pdf")  
+pdf("../manuscript/figures/fig-peaks-age-appendix-revision.pdf")  
   multiplot(beck_peaks_age + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 

### 2018-02-05 analyses/11-appendix-figure-peaks.Rmd
Extend colour scale and move labels one bar in

@@
        strip.text.x = element_text(size = 12)) +
-  geom_text(data = ann_text, aes(x = 1, y = 10, 
+  geom_text(data = ann_text, aes(x = 2, y = 10, 
             label = "Beck2012"), col = "black") +
-  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
+  scale_colour_viridis(discrete = TRUE, direction = -1)
 
@@
        strip.text.x = element_text(size = 12)) +
-  geom_text(data = ann_text, aes(x = 1, y = 10, 
+  geom_text(data = ann_text, aes(x = 2, y = 10, 
             label = "Beck2012"), col = "black") +
-  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
+  scale_colour_viridis(discrete = TRUE, direction = -1)
 ```
@@
         strip.text = element_blank()) +
- geom_text(data = ann_text, aes(x = 1, y = 20, 
+ geom_text(data = ann_text, aes(x = 2, y = 20, 
             label = "    Brusatte2014"), col = "black") +
-  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
+  scale_colour_viridis(discrete = TRUE, direction = -1)
 
@@
         strip.text = element_blank()) +
- geom_text(data = ann_text, aes(x = 1, y = 20, 
+ geom_text(data = ann_text, aes(x = 2, y = 20, 
             label = "    Brusatte2014"), col = "black") +
-  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
+  scale_colour_viridis(discrete = TRUE, direction = -1)
 ```
@@
         strip.text = element_blank()) +
- geom_text(data = ann_text, aes(x = 1, y = 72, 
+ geom_text(data = ann_text, aes(x = 2, y = 72, 
             label = "Bapst2016"), col = "black") +
-  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
+  scale_colour_viridis(discrete = TRUE, direction = -1)
 
@@
         strip.text = element_blank()) +
- geom_text(data = ann_text, aes(x = 1, y = 72, 
+ geom_text(data = ann_text, aes(x = 2, y = 72, 
             label = "Bapst2016"), col = "black") +
-  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
+  scale_colour_viridis(discrete = TRUE, direction = -1)
 ```
@@
         strip.text = element_blank()) +
- geom_text(data = ann_text, aes(x = 1, y = 460, 
+ geom_text(data = ann_text, aes(x = 2, y = 460, 
             label = " Wright2017"), col = "black") +
-  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
+  scale_colour_viridis(discrete = TRUE, direction = -1)
 
@@
         strip.text = element_blank()) +
- geom_text(data = ann_text, aes(x = 1, y = 460, 
+ geom_text(data = ann_text, aes(x = 2, y = 460, 
             label = " Wright2017"), col = "black") +
-  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)   
+  scale_colour_viridis(discrete = TRUE, direction = -1)   
 ```

### 2018-02-05 analyses/10-appendix-figure-dtt.Rmd
Add edits from main text plot - mostly viridis

@@
   theme(axis.title.y = element_blank(),
-        axis.title.x = element_blank())
+        axis.title.x = element_blank()) +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 
@@
         axis.title.x = element_blank(),
-        legend.position = "none")
+        legend.position = "none") +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 
@@
   geom_point(data = strat2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8) +
+  geom_errorbar(data = strat2, aes(ymin = relative.lwr, ymax = relative.upr), width = 2, alpha = 0.3) +
   annotate("text", x = 125, y = 0.15, label = "Bapst2016")

### 2018-02-05 analyses/10-appendix-figure-dtt.Rmd
Add edits to epoch appendix figure

@@
 
-pdf("../manuscript/figures/fig-dtt-age-appendix.pdf")  
+pdf("../manuscript/figures/fig-dtt-age-appendix-revision.pdf")  
   multiplot(beck_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")),
@@
         axis.title.x = element_blank()
-        )
+        ) +
+  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
 
@@
   geom_point(data = duration2, size = 2, alpha = 0.8) +
-  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 1, alpha = 0.8)
+  geom_errorbar(data = duration2, aes(ymin = relative.lwr, ymax = relative.upr), width = 3, alpha = 0.3)
 
@@
 
-pdf("../manuscript/figures/fig-dtt-epoch-appendix.pdf")  
+pdf("../manuscript/figures/fig-dtt-epoch-appendix-revision.pdf")  
   multiplot(beck_s + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")),

### 2018-02-05 analyses/10-appendix-figure-dtt.Rmd
Extend viridis scale

@@
         axis.title.x = element_blank()) +
-  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
+  scale_colour_viridis(discrete = TRUE, direction = -1)
 
@@
         legend.position = "none") +
-  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
+  scale_colour_viridis(discrete = TRUE, direction = -1)
 
@@
         ) +
-  scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)
+  scale_colour_viridis(discrete = TRUE, direction = -1)
 

### 2018-02-05 manuscript/response_to_reviewers.tex
Add comments about figure changes

@@
 We have done as suggested - we didn't do this the first time round because of issues with ggplot.
+We also made the same change in the appendix figures, and changed the colours in response to a comment by reviewer 2.
 
@@
 We have used the viridis colour palettes which are designed for precisely this purpose (https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html), and are colour-blind friendly too.
+We've done this for all figures in the text and in the appendices.
 
-A bigger issue is that a lot of the points lie on top of each other.
-We have played around with this a lot and there's no satisfactory solution.
+A bigger issue, and this may be what the reviewer was referring to, is that a lot of the points lie on top of each other.
+We have played around with this a lot (we tried jittering and dodging point positions, and various types of colour palettes and transparencies etc.) and there's no satisfactory solution.
 For now we have made the error bars more transparent so hopefully it is a bit easier to see which points lie where.

### 2018-02-05 manuscript/time-slice.tex
Figure edits and some minor technical edits

@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/outlinesvg.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure1-outlinesvg.pdf}
     \caption[Outline of the disparity-through-time pipeline]
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch-revision.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure2-fig-dtt-epoch-revision.pdf}
     \caption[Relative disparity-through-time.]
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch-revision.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure3-fig-peaks-epoch-revision.pdf}
     \caption[Timing of peak disparity for four example datasets.]
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-extinction-effect.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure4-fig-extinction-effect.pdf}
     \caption[Extinction example datasets.]
@@
 % add proper link to zenodo
+Supporting information can be found on Dryad at % Add Dryad link
 
@@
 Appendix S1: Extra details of datasets. \\
-Appendix S2: Additional tables and figures.
+Appendix S2: Additional figures. \\
+Appendix S3: Additional tables.
 

### 2018-02-05 manuscript/S1-time-slice-appendices.tex
New separately appendices for submission

--- 2017-12-20 manuscript/time-slice-appendices.tex
@@
 \rhead{\textit{Guillerme \& Cooper 2018}}
-\lhead{\textit{Supporting Information}}
+\lhead{\textit{Supporting Information: S1}}
 \renewcommand{\headrulewidth}{0pt}
@@
 
-\section{Appendix S1: Extra details of datasets}
+\section{Appendix S1: Additional details of datasets}
 
@@
 
-\newpage
-\section{Appendix S2: Additional tables and figures}
-
-% Reset table and figure counters to 1
-\setcounter{figure}{0}  
-\setcounter{table}{0}
-
-% DTT figures
-  \begin{figure}[!htbp]
-    \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch-appendix.pdf}
-    \caption[Relative disparity through time for four example datasets.]
-    {Relative disparity-through-time for stratigraphic epochs. 
-    Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods.
-    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
-    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset.
-    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
-    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
-    The four dataset names are on the first plot for each dataset (see Table 1 for details).
-    Results for stratigraphic ages are shown in Figure A2.}
-    \label{figure:dtt2}
-  \end{figure}
-
-  \begin{figure}[!htbp]
-    \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-age-appendix.pdf}
-    \caption[Relative disparity through time for four example datasets.]
-    {Relative disparity-through-time for stratigraphic ages. 
-    Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods.
-    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages. 
-    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages in the time frame of the dataset.
-    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages in the time frame of the dataset. 
-    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
-    The four dataset names are on the first plot for each dataset (see Table 1 for details).
-    Results for stratigraphic epochs are shown in Figure A1.}
-    \label{figure:dtt3}
-  \end{figure}
-
-% Wilcoxon results table
-  \input{tables/table_wilcox_appendix.tex} 
-  \label{table:wilcox2}
-
-% disparity peaks figures
-\begin{figure}[!htbp]
-    \centering
-   % \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch-appendix.pdf}
-    \caption[Timing of peak disparity for four example datasets.]
-    {Timing of peak disparity for stratigraphic epochs.
-    Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
-    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset.
-    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
-    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
-    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (see Table 1 for details).
-    Results for stratigraphic ages are in Figure A4.}
-    \label{figure:peak2}
-  \end{figure}
-
-\begin{figure}[!htbp]
-    \centering
-    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-age-appendix.pdf}
-    \caption[Timing of peak disparity for four example datasets.]
-    {Timing of peak disparity for stratigraphic ages.
-    Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages. 
-    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages in the time frame of the dataset.
-    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages in the time frame of the dataset. 
-    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
-    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (see Table 1 for details).
-    Results for stratigraphic epochs are in Figure A3.}
-    \label{figure:peak3}
-  \end{figure}
-
-% Full results tables
-% These are so huge (100 pages!!!) I'm going to leave them out for now...
-
-%\begin{landscape}
-%  \input{tables/disparity_measurement-Beck2014.tex} 
-%\end{landscape}
-
-%\newpage
-%\begin{landscape}
-%  \input{tables/disparity_measurement-Brusatte2014.tex} 
-%\end{landscape}
-
-%\newpage
-%\begin{landscape}
-%  \input{tables/disparity_measurement-Bapst2016.tex} 
-%\end{landscape}
-
-%\newpage
-%\begin{landscape}
- % \input{tables/disparity_measurement-Wright2017.tex}  
-%\end{landscape}
-
 \end{document}

### 2018-02-05 manuscript/S2-time-slice-additional-figures.tex
New separately appendices for submission

--- 2017-12-20 manuscript/time-slice-appendices.tex
@@
 \rhead{\textit{Guillerme \& Cooper 2018}}
-\lhead{\textit{Supporting Information}}
+\lhead{\textit{Supporting Information: S2}}
 \renewcommand{\headrulewidth}{0pt}
@@
 
-\section{Appendix S1: Extra details of datasets}
-
-\subsection{Beck2014 (Figure \ref{figure:beck})}
-The following taxa were removed because they were in the phylogeny but not the character matrix or vice versa: \textit{Montanalestes}, \textit{Lainodon}, \textit{Kharmerungulatum}, \textit{Alymlestes}.
-
-\subsection{Brusatte2014 (Figure \ref{figure:brusatte})} 
-We used one randomly selected time-scaled tree from Brusatte et al. (2014).
-Zero-length branches were replaced with the minimum branch length in the phylogeny.
-The following taxa were removed because they were in the phylogeny but not the character matrix or vice versa: \textit{Sinraptor dongi}, \textit{Hesperonychus elizabethae}, \textit{Pyroraptor olympius}, \textit{Limenavis patagonica}, \textit{Lithornis}, \textit{Crypturellus undulatus}, \textit{Gallus gallus}, \textit{Crax pauxi}, \textit{Anas}\\ \textit{platyrhynchus}, \textit{Chauna torquata}, \textit{Epidendrosaurus} and \textit{Kinnareemimus}.
-The following taxa were removed because they shared no characters in the morphological matrix: \textit{Shanag ashile}, \textit{Atrociraptor marshalli}, \textit{Proceratosaurus bradleyi}, \textit{Incisivosaurus gauthieri}, \textit{Enigmosaurus}, \textit{Nanshiungosaurus brevispinus}, \textit{Xixiasaurus}, \textit{Tsaagan mangas}, \textit{Mirischia}, \textit{Pedopenna}, \textit{Suzhousaurus}, \textit{Juratyrant}, \textit{Vorona}, \textit{Bonapartenykus}, \textit{Teratophoneus}, \textit{Gobipteryx}, \textit{Songlingornis}, \textit{Liaoningornis longidigitu} and \textit{Achillesaurus}.
-
-\subsection{Bapst2016 (Figure \ref{figure:bapst})} 
-We used the maximum clade credibility tree from Bapst et al. (2016).
-Zero-length branches were replaced with the minimum branch length in the phylogeny.
-The following taxa were removed because they were in the phylogeny but not the character matrix or vice versa: \textit{Mei long} and \textit{Mei lon}. 
-The following taxa were removed because they shared no characters in the morphological matrix: \textit{Hagryphus giganteus}, \textit{Atrociraptor marshalli}, \textit{IGM100 1015 UndesDromaeosaurid}, \textit{Dromaeosaurus albertensis}, \textit{Incisivosaurus gauthieri}, \textit{Deinocheirus mirificus}, \textit{Therizinosaurus cheloniformis},  \textit{Anserimimus planinychus} and \textit{Elmisaurus rarus}.
-
-\subsection{Wright2017 (Figure \ref{figure:wright})}
-We used the maximum clade credibility tree from Wright (2017). 
-To properly timescale the tree we followed the advice of Wright (2017) and divided the branch lengths by the corresponding clock rate (= 0.03517385)  and then set the root time to 485.4. 
-Zero-length branches were replaced with the minimum branch length in the phylogeny.
-No taxa were removed.
-
-\begin{figure}[!htbp]
-    \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-tree-Beck2014-appendix.pdf}
-    \caption[Beck2014.]
-    {Phylogeny from Beck \& Lee (2014).}
-    \label{figure:beck}
-  \end{figure}
-
-\begin{figure}[!htbp]
-    \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-tree-Brusatte2014-appendix.pdf}
-    \caption[Brusatte2014.]
-    {Phylogeny from Brusatte et al. (2014). This is one randomly selected tree from the time-scaled trees in the paper.}
-    \label{figure:brusatte}
-  \end{figure}
-
-\begin{figure}[!htbp]
-    \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-tree-Bapst2016-appendix.pdf}
-    \caption[Bapst2016.]
-    {Phylogeny from Bapst et al. (2016). This is the maximum clade credibility tree.}
-    \label{figure:bapst}
-  \end{figure}
-
-\begin{figure}[!htbp]
-    \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-tree-Wright2017-appendix.pdf}
-    \caption[Wright2017.]
-    {This is the maximum clade credibility tree from Wright (2017).}
-    \label{figure:wright}
-  \end{figure}
-
-\newpage
-\section{Appendix S2: Additional tables and figures}
+\section{Appendix S2: Additional figures}
 
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch-appendix.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch-appendix-revision.pdf}
     \caption[Relative disparity through time for four example datasets.]
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-age-appendix.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-age-appendix-revision.pdf}
     \caption[Relative disparity through time for four example datasets.]
@@
 
-% Wilcoxon results table
-  \input{tables/table_wilcox_appendix.tex} 
-  \label{table:wilcox2}
-
 % disparity peaks figures
@@
     \centering
-   % \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch-appendix.pdf}
+   \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch-appendix-revision.pdf}
     \caption[Timing of peak disparity for four example datasets.]
@@
     For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
-    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (see Table 1 for details).
+    The four dataset names are on the first plot for each dataset (see Table 1 for details).
     Results for stratigraphic ages are in Figure A4.}
@@
     \centering
-    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-age-appendix.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-age-appendix-revision.pdf}
     \caption[Timing of peak disparity for four example datasets.]
@@
     For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
-    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (see Table 1 for details).
+    The four dataset names are on the first plot for each dataset (see Table 1 for details).
     Results for stratigraphic epochs are in Figure A3.}
@@
 
-% Full results tables
-% These are so huge (100 pages!!!) I'm going to leave them out for now...
-
-%\begin{landscape}
-%  \input{tables/disparity_measurement-Beck2014.tex} 
-%\end{landscape}
-
-%\newpage
-%\begin{landscape}
-%  \input{tables/disparity_measurement-Brusatte2014.tex} 
-%\end{landscape}
-
-%\newpage
-%\begin{landscape}
-%  \input{tables/disparity_measurement-Bapst2016.tex} 
-%\end{landscape}
-
-%\newpage
-%\begin{landscape}
- % \input{tables/disparity_measurement-Wright2017.tex}  
-%\end{landscape}
-
 \end{document}

### 2018-02-05 manuscript/response_to_reviewers.tex
Minor changes

@@
 discrete or continuous), and whether they include phylogenetic data or not.
-% NC: do you mean cladistic data i.e.e data used to build phylogenies?
+% NC: do you mean cladistic data i.e.
+data used to build phylogenies?
 Although empirical morphospaces from discrete or continuous data have been shown to have similar properties (CITATIONS), our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis;} c.f.
@@
 %TG: Also for these figures, I think we need to try to plot a line rather than two dots for the bins interval. Its actually not that straight forward to realise what's happening.
-% NC: No! The lines are misleading, technically must be points, as lines suggest there is data throughout rather than at discrete points. Bit of a bugbear of mine...
 
@@
 After thinking about it for awhile and then checking the R code, this is what appears to be going on.
-Let’s say we have a case study that spans 10 stratigraphic intervals over 40 million years.
+Let's say we have a case study that spans 10 stratigraphic intervals over 40 million years.
 Thus for the stratigraphy approach, there are 10 sub-samples which vary in their duration.
@@
 
-%TG: actually not sure how to write these comments in this response document, they are only technical thingies, we can just state in the intro paragraph above that we dealt with all of them.
-I'll raise them as tickable issues on GitHub (yay!) for now.
+%NC: It says in the instructions to respond to these too!
 
-% NC: For simplicity I suggest we put them here too, then it's easy for the editor to just tick them off rather than having to check. Also makes sure that we definitely do it all!
+\section{Reviewer 3 (Sally Thomas)}
+\subsection{Technical comments}
 
-% \section{Reviewer 3 (Sally Thomas)}
+\begin{enumerate}
 
-% \begin{enumerate}
+\item{\textcolor{blue}{Please submit your manuscript in Word format. Do not include the figures in the text file, but it is helpful if you do include a list of the figure captions at the end.}}
 
-% \item{\textcolor{blue}{Please submit your manuscript in Word format. Do not include the figures in the text file, but it is helpful if you do include a list of the figure captions at the end.}}
+Done.
+% NC: Need to put the figure numbers onto the file names before submission.
 
-% \item{\textcolor{blue}{Could I suggest adding ‘sub-sampling’ to your key words?}}
+\item{\textcolor{blue}{Could I suggest adding ‘sub-sampling’ to your key words?
+We now recommend that as many key words as possible are also included in your title, as well as being well-used in the abstract.}}
 
-% \end{enumerate}
+Done.
+
+\item{\textcolor{blue}{Please upload your supporting information files to Dryad as separate files (with the table in an editable format).}}
+
+We have split the appendices into three files S1) additional details of the datasets; S2) additional figures; and S3) additional tables (now in .docx format rather than PDF).
+% NC: To upload but splitting is done
+
+\item{\textcolor{blue}{Please clarify what all the supplementary files are (there is a details field available for each file) and ensure that they are in a form that is commonly readable, or provide information on how to access them.}}
+
+Done. 
+% NC: To do when uploading
+
+\item{\textcolor{blue}{Please add the Dryad review link (even though Dryad instructs you not to) to the Data Archiving Statement.
+This makes it easier for the Editor and Referees, and the link will be updated later as required.}}
+
+Done.
+% NC: To do when uploaded
+
+\item{\textcolor{blue}{Where you refer to the data in the text please cite a reference in the form: CARPENTER, D. K., FALCON-LANG, H. J., BENTON, M. J. and GREY, M. 2015. Data
+from: Early Pennsylvanian (Langsettian) fish assemblages from the Joggins
+Formation, Canada, and their implications for palaeoecology and palaeogeography.
+Dryad Digital Repository. https://doi.org/10.5061/dryad.b0551.}}
+
+\end{enumerate}
+
+\subsection{References}
+
+\begin{enumerate}
+
+\item{\textcolor{blue}{Please replace the GitHub URL in the Materials \& Methods section with a citation of the data repository reference (as it is quoted on GitHub; Cooper \& Guillerme 2018).}}
+
+We have now archived the code on Zenodo, so provide the citation for this.
+% NC: To do
+
+\item{\textcolor{blue}{Please include all repository details and links in data references (e.g. Bapst et al. 2016a)(see sample Dryad reference above for general format).}}
+% NC: To do
+
+\item{\textcolor{blue}{Where you have two references with the same citation (e.g.
+Bapst et al.
+2016), please list them in the same order as they are cited in the text, irrespective of the second author name (currently Bapst et al.
+2016b and Brusatte et al.
+2014b are both cited before their respective a references; in each case the first reference cited should be a and be listed first).}}
+
+Sorry this is a bug in our reference manager software that we thought would be easier to fix at this stage. Now fixed!
+% NC: This should be fixable at the converting to Word stage
+
+\item{\textcolor{blue}{Guillerme (2016) reference requires some more information – is there a URL for the software that can be added? (Similar for any other software reference, e.g. Lloyd 2015.)}}
+
+We have fixed this.
+% NC: Fix at converting to Word stage, hopefully there will be a paper to cite by then...
+% NC: Actually, I think we have cited the claddis package not the paper - should I change that?
+
+\end{enumerate}
+
+\subsection{Figures}
+\textcolor{blue}{Please include the figure number in the file name.}
+
+Done.
+
+\subsection{Tables}
+\textcolor{blue}{Journal style uses a single phrase for table captions, with any additional notes in footnotes. Discussion of the data should be included in the text.}
+
+Done.
+
+% NC: Made new word versions of these
+
+We hope we have dealt with all these comments appropriately.
+
+Please let us know if you need any further information,
+
+Thomas Guillerme and Natalie Cooper
 

### 2018-02-06 analyses/07-figure-peaks.Rmd
Add line to join bins points

--- 2018-02-05 analyses/07-figure-peaks.Rmd
@@
   geom_point(aes(y = time2, x = model, colour = model)) +
+  geom_segment(aes(y = time1, x = model, yend = time2, xend = model, colour = model), size = 1) +
   facet_wrap(~bin_type, ncol = 3) +
@@
   geom_point(aes(y = time2, x = model, colour = model)) +
+  geom_segment(aes(y = time1, x = model, yend = time2, xend = model, colour = model), size = 1) +
   facet_wrap(~bin_type, ncol = 3) +
   scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- expand_limits(y = 480) +
+ expand_limits(y = c(455, 480)) +
  theme(axis.title.y = element_blank(),

### 2018-02-06 analyses/11-appendix-figure-peaks.Rmd
Add viridis library
Tidy and add lines to bins

@@
   mutate(bin_type = relevel(data_max$bin_type, ref = 'Stratigraphy')) %>%
-  mutate(model = relevel(data_max$model, ref = 'equalbins')) #%>%
-  #filter(model == "equalbins" | model == "proximity" | model == "gradual.split")
+  mutate(model = relevel(data_max$model, ref = 'equalbins'))
 
@@
   geom_point(aes(y = time2, x = model, colour = model)) +
+  geom_segment(aes(y = time1, x = model, yend = time2, xend = model, colour = model), size = 1) +
   facet_wrap(~bin_type, ncol = 3) +
-  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- # expand_limits(x = 0, y = c(0,1)) +
   theme(axis.title.y = element_blank(),
@@
   mutate(bin_type = relevel(data_max$bin_type, ref = 'stratigraphy')) %>%
-  mutate(model = relevel(data_max$model, ref = 'equalbins'))#%>%
-  #filter(model == "equalbins" | model == "proximity" | model == "gradual.split")
+  mutate(model = relevel(data_max$model, ref = 'equalbins'))
 
@@
   geom_point(aes(y = time2, x = model, colour = model)) +
+  geom_segment(aes(y = time1, x = model, yend = time2, xend = model, colour = model), size = 1) +
   facet_wrap(~bin_type, ncol = 3) +
-  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
   expand_limits(y = 0) +
@@
   geom_point(aes(y = time2, x = model, colour = model)) +
+  geom_segment(aes(y = time1, x = model, yend = time2, xend = model, colour = model), size = 1) +
   facet_wrap(~bin_type, ncol = 3) +
- # scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- # expand_limits(x = 0, y = c(0,1)) +
   theme(axis.title.y = element_blank(),
@@
   geom_point(aes(y = time2, x = model, colour = model)) +
+  geom_segment(aes(y = time1, x = model, yend = time2, xend = model, colour = model), size = 1) +
   facet_wrap(~bin_type, ncol = 3) +
-  #scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
   expand_limits(y = c(455, 480)) +

### 2018-02-06 analyses/07-figure-peaks.Rmd
Tidy code

@@
   filter(model == "equalbins" | model == "proximity" | model == "gradual.split")
+
 data_max2_age <- filter(data_max2, stratigraphy == "Age")
@@
   scale_x_discrete(labels = c("bins", "gradual", "proximity")) +
- # expand_limits(x = 0, y = c(0,1)) +
   theme(axis.title.y = element_blank(),

### 2018-02-06 analyses/02-time-slice-analyses.Rmd
Small comment about time the script takes to run
Change dispRity version to release version
Hide warnings in .Rmd

@@
 Note that much of this uses functions in the `simulations_fun.R` file.
+It also is also quite slow on some machines, but all outputs are available in the repo.
 
@@
 ## The latest version of dispRity
-## install_github("TGuillerme/dispRity", ref = "master")
+## install_github("TGuillerme/dispRity@v0.5")
 library(dispRity)
@@
 
-```{r}
+```{r, warning = FALSE}
 slug <- "Brusatte2014"
@@
 
-```{r}
+```{r, warning = FALSE}
 slug <- "Bapst2016"
@@
 -----------
-```{r}
+```{r, warning = FALSE}
 ## Choose a slug

### 2018-02-06 analyses/06-table-wilcox.Rmd
Edits to reshape the data - spread no longer working due to duplicate rows so I've gone back to reshape. This required some renames to match the rest of the code.

@@
   filter(V3 == "proximity" | V3 == "gradual.split") %>%
-  spread(V4, V6) %>%
+  reshape(idvar = c("V3","V5"), timevar = "V4", direction = "wide") %>%
   arrange(V5) %>%
   rename(model = V3) %>%
-  rename(period = V5)  
+  rename(period = V5) %>%
+  rename(stratigraphy = V6.stratigraphy) %>%
+  rename(duration = V6.duration) %>%
+  rename(number = V6.number) 
 }
@@
   select(-c(V1,V2)) %>%
-  spread(V4, V6) %>%
+  reshape(idvar = c("V3","V5"), timevar = "V4", direction = "wide") %>%
   arrange(V5) %>%
   rename(model = V3) %>%
-  rename(period = V5)
+  rename(period = V5) %>%
+  rename(stratigraphy = V6.stratigraphy) %>%
+  rename(duration = V6.duration) %>%
+  rename(number = V6.number) 
 }

### 2018-02-06 analyses/07-figure-peaks.Rmd
Correct output name for figure 3

@@
 
-pdf("../manuscript/figures/fig-peaks-epoch-revision.pdf")  
+pdf("../manuscript/figures/Figure3-fig-peaks-epoch-revision.pdf")  
   multiplot(beck_peaks + theme(plot.margin=unit(c(5.5, 5.5, 5.5, 25),"points")), 

### 2018-02-06 manuscript/S2-time-slice-additional-figures.tex
Edits to change ages to stages

@@
     The four dataset names are on the first plot for each dataset (see Table 1 for details).
-    Results for stratigraphic ages are shown in Figure A2.}
+    Results for stratigraphic stages are shown in Figure A2.}
     \label{figure:dtt2}
@@
     \caption[Relative disparity through time for four example datasets.]
-    {Relative disparity-through-time for stratigraphic ages. 
+    {Relative disparity-through-time for stratigraphic stages. 
     Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
     Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods.
-    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages. 
-    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages in the time frame of the dataset.
-    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages in the time frame of the dataset. 
+    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic stages. 
+    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic stages in the time frame of the dataset.
+    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic stages in the time frame of the dataset. 
     In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
@@
     Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
-    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
+    For time bins the points indicate the maximum and minimum ages of the time bin within which peak disparities appeared.
     The four dataset names are on the first plot for each dataset (see Table 1 for details).
-    Results for stratigraphic ages are in Figure A4.}
+    Results for stratigraphic stages are in Figure A4.}
     \label{figure:peak2}
@@
     \caption[Timing of peak disparity for four example datasets.]
-    {Timing of peak disparity for stratigraphic ages.
+    {Timing of peak disparity for stratigraphic stages.
     Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
-    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages. 
-    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages in the time frame of the dataset.
-    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages in the time frame of the dataset. 
-    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
+    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic stages. 
+    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic stages in the time frame of the dataset.
+    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic stages in the time frame of the dataset. 
+    For time bins the points indicate the maximum and minimum ages of the time bin within which peak disparities appeared.
     The four dataset names are on the first plot for each dataset (see Table 1 for details).

### 2018-02-06 manuscript/time-slice.tex
Edit figure legend now we have lines
Edits to change ages to stages

@@
     Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
-    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
+    For time bins the points indicate the maximum and minimum ages of the time bin within which peak disparities appeared.
     The four dataset names are on the first plot for each dataset (see Table \ref{table:datasets} for details).
@@
 
-  %TG: Would it be possible to make the bins points into a thick line? Though that's just aesthetic...
-  % NC: Probably but I can't be arsed.
-
 \paragraph{Effects of mass extinction events.}
 Mass extinction events influence disparity in both the \cite{beckancient2014} and \cite{wright2017bayesian} datasets (Fig. \ref{figure:extinction1}).
-However, whether this change in disparity is significant or not depends on the method used to create time sub-samples (Fig. \ref{figure:extinction1}), and whether ages or epochs are used. 
+However, whether this change in disparity is significant or not depends on the method used to create time sub-samples (Fig. \ref{figure:extinction1}), and whether stages or epochs are used. 
 In general, for the \cite{beckancient2014} dataset, time binning tended to give more significant results than time-slicing methods, but this was not the case for the \cite{wright2017bayesian} dataset.

### 2018-02-06 manuscript/time-slice.tex
Edits for final version - removing figures and tables
Remove figures

@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure1-outlinesvg.pdf}
+    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure1-outlinesvg.pdf}
     \caption[Outline of the disparity-through-time pipeline]
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure2-fig-dtt-epoch-revision.pdf}
+    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure2-fig-dtt-epoch-revision.pdf}
     \caption[Relative disparity-through-time.]
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure3-fig-peaks-epoch-revision.pdf}
+    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure3-fig-peaks-epoch-revision.pdf}
     \caption[Timing of peak disparity for four example datasets.]
@@
     \centering
-    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure4-fig-extinction-effect.pdf}
+    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure4-fig-extinction-effect.pdf}
     \caption[Extinction example datasets.]
@@
 \section{Supporting information}
-Additional Supporting Information can be found in the online version of this article:\\
+Additional Supporting information can be found on Dryad at:\\
 Appendix S1: Extra details of datasets. \\
@@
 \newpage
-\section{Tables and Figures}
+\section{Figure legends}
 
 % Dataset details table
-\begin{landscape}
-\input{tables/table_datasets.tex} 
-\end{landscape}
+%\begin{landscape}
+%\input{tables/table_datasets.tex} 
+%\end{landscape}
 
 % Wilcoxon results table
-\input{tables/table_wilcox.tex}
+%\input{tables/table_wilcox.tex}
 

### 2018-02-06 analyses/07-figure-peaks.Rmd
Fix typo - Beck 2014 not 2012

@@
   geom_text(data = ann_text, aes(x = 1, y = 10, 
-            label = "Beck2012"), col = "black") +
+            label = "Beck2014"), col = "black") +
   scale_colour_viridis(discrete = TRUE, direction = -1, end = 0.8)

### 2018-02-06 analyses/11-appendix-figure-peaks.Rmd
Correct Beck 2014 typo

@@
   geom_text(data = ann_text, aes(x = 2, y = 10, 
-            label = "Beck2012"), col = "black") +
+            label = "Beck2014"), col = "black") +
   scale_colour_viridis(discrete = TRUE, direction = -1)

### 2018-02-06 manuscript/response_to_reviewers.tex
Updated response to include stuff I fixed today

@@
 %TG: TODO: fix ages in table 1
+% NC: What are these? I converted this to word already.
 
@@
 We also made the same change in the appendix figures, and changed the colours in response to a comment by reviewer 2.
-
-%TG: Also for these figures, I think we need to try to plot a line rather than two dots for the bins interval. Its actually not that straight forward to realise what's happening.
 
@@
 
-% NC: I like this response. I dunno if it is Graeme, they use US spellings...
+% NC: I like this response. I dunno if it is Graeme, they use US spellings... Maybe Bapst?
 
@@
 \item{\textcolor{blue}{Line 158.
-Gower’s distance is not the Euclidean distance (= the square root of the squared distances) over the number of shared characters.
+Gower's distance is not the Euclidean distance (= the square root of the squared distances) over the number of shared characters.
 For a set of unordered characters, it is the number of mismatched character states over the number of comparable/shared characters.
 For a set of ordered characters, it is sum of the differences in the values of the characters states for each pairwise comparison, over the number of comparable characters.
-This is an edit that needs to be made only to the text; because the authors used functions from the Claddis package (see github documentation), they would have calculated Gower’s distance correctly.}}
+This is an edit that needs to be made only to the text; because the authors used functions from the Claddis package (see github documentation), they would have calculated Gower's distance correctly.}}
 
@@
 Done.
-% NC: Need to put the figure numbers onto the file names before submission.
 
@@
 We have split the appendices into three files S1) additional details of the datasets; S2) additional figures; and S3) additional tables (now in .docx format rather than PDF).
-% NC: To upload but splitting is done
+% NC: Just need to sort S2 when we have the new figure. I've added them as is to Dryad to get the reference link.
 
@@
 Done.
-% NC: To do when uploading
 
@@
 Dryad Digital Repository. https://doi.org/10.5061/dryad.b0551.}}
+
+Done.
 
@@
 \item{\textcolor{blue}{Please include all repository details and links in data references (e.g. Bapst et al. 2016a)(see sample Dryad reference above for general format).}}
-% NC: To do
+
+Done.
 
@@
 
-Sorry this is a bug in our reference manager software that we thought would be easier to fix at this stage. Now fixed!
-% NC: This should be fixable at the converting to Word stage
+Sorry this is a bug in our reference manager software. Now fixed!
 
@@
 % NC: Fix at converting to Word stage, hopefully there will be a paper to cite by then...
-% NC: Actually, I think we have cited the claddis package not the paper - should I change that?
+for now I've used the v 0.5 zenodo citation.
+I've changed the citation of Claddis to Graeme's Biol Linn soc paper.
 
@@
 
-Done.
-
-% NC: Made new word versions of these
+Done, and converted to docx.
 

### 2018-02-07 manuscript/response_to_reviewers.tex
respond to couple of reviewer 2 major comments
also changes time-slices (non equal) to time-slices
(equidistant/nonequidistant).

@@
 
-%TG: tone done the statement? % NC: yeah, not sure what else we can do.
+We've tone down this statement by included the fact that bins can simply represent averages (as suggested by the reviewer).
+
+\textit{Second, the time binning approaches (whether bins are equally sized or not) favour punctuated equilibrium modes of evolution.
+Whether the disparity represents an average across the interval (with no interpretation on if or how it varied within the time bin) or it is effectively postulated to be constant, when analysing the changes in disparity through time, this method will only allow changes in disparity (in average or constant) to occur only between intervals rather than also allowing for gradual changes within intervals (a pattern that is fairly common in the fossil record, CITATION).} lines @@@
+
+%TG: tone done the statement? % NC: yeah, not sure what else we can do. %TG: Does that kinda work?
 
@@
 %TG: though that is just a "practical" point for the vertebrate fossil record. Foraminiferas, for example, do not behave this way (or at a way finer scale).
+
+We've add the following caveat to the discussion:
+
+\textit{Finally, the difference between the time-binning and time-slicing methods could also simply be favoured towards time binning due to the nature of the fossil record itself.
+The rates of sedimentation varies in time a space and can be really sporadic, influencing the groups found within and their temporal distribution (e.g. vertebrates, c.f. foraminifera).
+In this case, different beds could represent different ``packages'' of fauna through time separated by gaps thus effectively binning the data rather than slicing.
+For example, slicing through such strata will yield similar results whether where in time does the slice occur.
+It is important to note however, that, as specified above, the time slicing method also includes ancestral estimations data (whether through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.} lines @@@
 

### 2018-02-07 manuscript/time-slice.tex
respond to couple of reviewer 2 major comments
also changes time-slices (non equal) to time-slices
(equidistant/nonequidistant).

@@
 Other studies split large time bins so they are of roughly equal size, but this is often an \textit{ad hoc} procedure that can introduce more bias depending on where bins are split.
-Second, all time binning approaches (whether bins are equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring \textit{between} intervals, rather than also allowing for gradual changes \textit{within} intervals \citep[a pattern that is fairly common in the fossil record;][]{Hunt21042015}.
+Second, the time binning approaches (whether bins are equally sized or not) favour punctuated equilibrium modes of evolution.
+Whether the disparity represents an average across the interval (with no interpretation on if or how it varied within the time bin) or it is \textit{effectively} postulated to be constant, when analysing the changes in disparity through time, this method will only allow changes in disparity (in average or constant) to occur only \textit{between} intervals rather than also allowing for gradual changes \textit{within} intervals \citep[a pattern that is fairly common in the fossil record;][]{Hunt21042015}.
 Third, when investigating changes in disparity due to events at a specific time point (e.g.
@@
     We then calculated the median disparity value for each time bin along with the 50\% and 95\% confidence intervals.
-    \item Time-slices (unequal).
+    \item Time-slices (non-equidistant).
     We calculated disparity using our time-slicing approach with slices occurring at the midpoint of each stratigraphic period (age or epoch), and using all six time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits).
@@
     \item Time bins (equal). 
-    We calculated disparity for the taxa in each time bin where time bin size was defined by the average \textit{duration} of the stratigraphic period (age or epoch), and bootstrapped the disparity values as described above.
-    \item Time-slices (unequal).
-    We calculated disparity using our time-slicing approach where the interval between slices, was defined by the average \textit{duration} of the stratigraphic period (age or epoch).
+    We calculated disparity for the taxa in each time bin where time bin size was defined by the mean \textit{duration} of the stratigraphic period (age or epoch), and bootstrapped the disparity values as described above.
+    \item Time-slices (equidistant).
+    We calculated disparity using our time-slicing approach where the interval between slices, was defined by the mean \textit{duration} of the stratigraphic period (age or epoch).
     We used the six time-slicing methods and bootstrapped as described above.
@@
     \item Time bins (equal). 
-    We calculated disparity for the taxa in each time bin where the number of time bins was defined by the average \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest, and bootstrapped the disparity values as described above.
-    \item Time-slices (unequal).
-    We calculated disparity using our time-slicing approach where the number of slices, was defined by the average \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest.
+    We calculated disparity for the taxa in each time bin where the number of time bins was defined by the \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest, and bootstrapped the disparity values as described above.
+    \item Time-slices (equidistant).
+    We calculated disparity using our time-slicing approach where the number of slices, was defined by the \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest.
     We used the six time-slicing methods and bootstrapped as described above.
@@
 This may be important, as gradual change is a common pattern of trait evolution in the fossil record \citep{Hunt20112007}.
+
 Of course the method also has limitations.
-The main one of these is a practical one; it requires a time-calibrated phylogeny.
-Not all palaeontological datasets have phylogenies yet.
-Another limitation is that the method, like most phylogeny based methods, depends on ancestral state estimations.
+The main one of these is a practical one; it requires a time-calibrated phylogeny which are not available for all palaeontological datasets.
+Furthermore, like most phylogeny based methods, time-slicing depends on ancestral state estimations.
 Care should be taken in interpreting these, as they are highly dependent on the data and models used for the estimations \citep{Slateretal2012,doi:10.1080/10635150801910451}.
+Finally, the difference between the time-binning and time-slicing methods could also simply be favoured towards time binning due to the nature of the fossil record itself.
+The rates of sedimentation varies in time a space and can be really sporadic, influencing the groups found within and their temporal distribution (e.g. vertebrates, c.f. foraminifera).
+In this case, different beds could represent different ``packages'' of fauna through time separated by gaps thus effectively binning the data rather than slicing.
+For example, slicing through such strata will yield similar results whether where in time does the slice occur.
+It is important to note however, that, as specified above, the time slicing method also includes ancestral estimations data (whether through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
 

### 2018-02-07 manuscript/response_to_reviewers.tex
Updated comments

@@
 % NC: What are these? I converted this to word already.
+%TG: In Table 1, ages for Brusatte and Bapst should be respectively 168.5 - 66 and 207.2 - 66  (they are currently set to 0).
 
@@
 Mathematically, our morphospaces are $n$ dimensional objects that summarise the distances between discrete morphological characters of the taxa present and their ancestors.} lines @@@
-
-%TG: what do you think? % NC: sounds good, made a few minor edits above
 
@@
 
-%TG: I can implement that pretty easily but I think that mathematically it's not cool (we can not add values post-ordination in the ordinated space).
-Two options: (1) implementing the suggestion but with heavy warnings; (2) writing down a clever response to this comment.
-In the dispRity package review, that was highlighted (I bet reviewer 2 is Graeme) and I responded something similar to option (2) so I'd rather go that way for consistency.
-
-% NC: I like this response. I dunno if it is Graeme, they use US spellings... Maybe Bapst?
-
 \item{\textcolor{blue}{Second, I do not see how Figure 3 supports the text (lines 435-447).
@@
 [...] Therefore, even if you pick a ``point in time'' and count the number of branches/ranges that cross that point, some of those are being captured because the resolution of the end points only allows us to know that the taxon started and/or ended somewhere in the interval encompassing that point in time.}}
-
-%TG: this is a fair point we should include in the discussion.
-%TG: though that is just a "practical" point for the vertebrate fossil record. Foraminiferas, for example, do not behave this way (or at a way finer scale).
 
@@
 %TG: good idea, we (I) can make a figure designating the three different strategies.
+We've added a figure describing the different time-sampling strategies and changed the relevant methods section to be clearer.
 
@@
 This is an edit that needs to be made only to the text; because the authors used functions from the Claddis package (see github documentation), they would have calculated Gower's distance correctly.}}
+
+We thank the reviewer for spotting this mistake and fixed the relevant section as follows:
+
+\textit{We calculated the $k$$\times$$k$ distances using the Gower distance (CITATION), i.e. the number of characters mismatched characters over the number of shared characters.}
 
@@
 
-%TG: I'll let you change the colours (I still don't want to touch ggplot!) how about the darkgrey/orange/blue palette?
-%Here are the hexadecimals:
-%colours_blind <- c("orange" = "#F65205", "blue" = "#3E9CBA", "darkgrey" = "#1C1C1C")
-%NC: The colours are colour blind friendly already, but I'll switch to using viridis.
-
 Interestingly the palettes in ggplot2 are chosen so that they are distinguishable by people that are colour blind.
@@
 Hopefully the new colour scheme is easier to discern.
+%TG: Does that not sound a bit too defensive? We can just say we changed them no?
 We have used the viridis colour palettes which are designed for precisely this purpose (https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html), and are colour-blind friendly too.

### 2018-02-07 manuscript/time-slice.tex
Updated comments
Add fig legend for new figure 2

@@
 To estimate the morphospaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
-We calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the Euclidean distance between two taxa divided by the number of shared characters. 
+We calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the number of characters mismatched characters over the number of shared characters.
 This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e.
@@
 \citealt{paradisape:2004}; but distinguished in \citealt{legendre2012numerical}) is an eigen decomposition of the distance matrix.
+
 Because we used Gower distances instead of raw Euclidean distances, negative eigenvalues can be calculated.
+
 To avoid this problem, we first transformed the distance matrices by applying the Cailliez correction \citep{cailliez1983analytical} which adds a constant $c^*$ to the values in a distance matrix (apart from the diagonal) so that all the Gower distances become Euclidean ($d_{Gower}+c^*=d_{Euclidean}$; \citealt{cailliez1983analytical}).
+
 We were then able to extract $n$ eigenvectors for each matrix (representing the $n$ dimensions of the morphospace) where $n$ is equal to $k-2$, i.e.
 the number of taxa in the matrix ($k$) minus the last two eigenvectors that are always null after applying the Cailliez correction.
+
 Contrary to previous studies \citep[e.g][]{brusatte50,cisneros2010,prentice2011,anderson2012using,Hughes20082013,bentonmodels2014}, we use all $n$ dimensions of our morphospaces and not a sub-sample representing the majority of the variance in the distance matrix (e.g.
@@
 
-To compare the time binning and time-slicing approaches we applied the methods as follows.
+To compare the time binning and time-slicing approaches we applied the methods as follows (see Figure \ref{figure:bins}).
+
+\begin{figure}[!htbp]
+    \centering
+    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure2-binning.pdf}
+    \caption[Outline of the three time sub-sampling methods]
+    {Outline of the three time sub-sampling methods. 
+    Stratigraphy: time sub-samples are defined by stratigraphic periods.
+    Here there are five stratigraphic periods in the 20 MY (million years) time frame of interest, i.e. five bins/slices with variable sizes/intervals.
+    Duration: time sub-samples are defined based on the mean duration of stratigraphic periods in the time frame of interest. 
+    Here, the mean duration of stratigraphic periods is 5 MY, so there are four bins/slices of 5 MY duration (or four slices with 5 MY intervals between them) in the 20 MY time frame of interest. 
+    Number: time sub-samples are defined based on the number of stratigraphic periods in the time frame of interest. 
+    Here, there are five stratigraphic periods, so there are five bins/slices of 4 MY duration (or five slices with 4 MY intervals between them) in the 20 MY time frame of interest.}
+    \label{fig:bins}
+\end{figure}
 
@@
     To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time bin by randomly resampling with replacement a new sub-sample of taxa from the observed taxa in the bin 100 times.
-    %TG: actually in the code the data is bootstrapped 100 times only.
-Should I change to 1000?
-To be fair, I think 1000 is too much (in terms of calculation and results) and 100 to few.
-A compromise could be 500.
-    %NC: 100 is fine, I was just guessing!
     We then calculated the median disparity value for each time bin along with the 50\% and 95\% confidence intervals.
@@
   \end{enumerate}
-
-%TG: well done! Nice description of the method! I figured out it was pretty obvious in our (my) head but it's actually more painful to write down.
-% NC: I know! It took me ages to get it logically described!
 

### 2018-02-07 analyses/04-table-datasets.Rmd
Typos

--- 2017-12-20 analyses/04-table-datasets.Rmd
@@
 age_ranges <- unlist(lapply(age_ranges, function(age) paste(age, collapse = " - ")))
+## Note that I've had to correct these for Bapst2016 (207.2) and Brusatte2014 (168.5) due to 
+## the rerooting that occurred when removing taxa
 
@@
 ## Table caption
-table_caption <- "Details of the datasets used in this study.
-Note that although Brusatte2014 and Bapst2016 cross the Cretaceous Palaeogene (K-Pg) mass extinction there are not enough taxa post K-Pg to investigate the effects of the mass extinction on disparity."
+table_caption <- "Details of the datasets used in this study. Age ranges are root time to most recent tip taxon."
 

### 2018-02-07 manuscript/time-slice.tex
Tidied up comments

@@
 Inapplicable and missing characters for any taxon were treated as ambiguous characters (i.e. any possible observed state for the character).
-% NC: I've left this one bit to make it clear but removed the explanation (this was all just copied from your thesis)
 To prevent poor ancestral state estimations from biasing our results, especially when a lot of error is associated with the estimations, we only included ancestral state estimations with a scaled Likelihood $\geq$ $0.95$.
@@
 \paragraph{Building morphospaces.} 
-% To explore disparity-through-time in our datasets we used a morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
-% This approach is similar to constructing a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010}, except a morphospace is an approximation of the morphospace based on discrete morphological data (i.e.
-the discrete morphological characters used to build a phylogenetic tree, often referred to as ``cladistic'' data).
-% %TG: about that, I'm actually less and less convinced we should call it caldisto-space: first, semantically, cladistic data means data used to make clades (phylos) so it could be nucleotides as well.
-Second, during the PhD, we used morphospace to distinguish it from the morpho-space but at that time I was naively thinking the the morpho-space is the morphometric-space (which it often is!) but, again, semantically, the morpho-space can also be simply the morphology-space...
-However, I like the sound of morphospace and most people will understand it the way we do.
-I'm just thinking about pedantic-palaeo-reviewers...
-% % NC: Good point. Let's leave it for now as I think this bit is nicely explained, but easy enough to edit if people don't like it.
-% Mathematically, a morphospace is an $n$ dimensional object that summarises the cladistic distances between the taxa present in a cladistic matrix.
-% Although empirically inter-taxon distances are the same in a morphospace or a morphospace \citep{foth2012different,hetherington2015cladistic}, we prefer the term morphospace to make it clear that this space is estimated using cladistic data and not morphometric data.
-
 To explore disparity-through-time in our datasets, we used a morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
@@
 This results in even-sampling of the morphospace across time and allows us to use different underlying models of character evolution (punctuated or gradual).
-
-%TG: Is it worth saying that Halliday and Goswami came up with a similar idea \citep{halliday2016eutherian} but that we came up with it first \cite{Guillerme2015} and we're the first to implement it \cite{dispRity.v02}?
-%NC: Yeah was going to ask about this - I've added a mention to the intro where I think it fits better, should we also cite the brusatte paper you said refers to it too?
 
@@
 We also recorded the number of taxa (or taxa and nodes for time-slicing methods) in each sub-sample as a proxy for taxonomic diversity.
-%(Supporting Information Appendix S2: Tables A1-A4). 
-% Tables are so huge I'm going to leave them out for now...
 
@@
 
-%Fig. or Figs
-%Fig width 80mm single column
-%110mm - 2/3 width
-%166mm full page
-
 \section{Discussion}
@@
 
-% Dataset details table
-%\begin{landscape}
-%\input{tables/table_datasets.tex} 
-%\end{landscape}
-
-% Wilcoxon results table
-%\input{tables/table_wilcox.tex}
-
 \end{document}

### 2018-02-07 manuscript/response_to_reviewers.tex
Some edits to response letter

--- 2018-02-06 manuscript/response_to_reviewers.tex
@@
 
-%TG: TODO: fix ages in table 1
-% NC: What are these? I converted this to word already.
-
 % ~~~~~~~~~~~~~~~~~~~~~~~~
@@
 
-%TG: tone done the statement? % NC: yeah, not sure what else we can do.
+We've toned down this statement by including the fact that bins can simply represent averages (as suggested by the reviewer).
+
+\textit{Second, the time binning approaches (whether bins are equally sized or not) favour punctuated equilibrium modes of evolution.
+Whether the disparity represents an average across the interval (with no interpretation of if or how it varied within the time bin), or it is effectively postulated to be constant, when analysing the changes in disparity-through-time, this method will only allow changes in disparity to occur between intervals rather than also allowing for gradual changes within intervals (a pattern that is fairly common in the fossil record, CITATION).} lines @@@
+
+%TG: tone done the statement? % NC: yeah, not sure what else we can do. %TG: Does that kinda work? NC: sounds ok to me, I tidied the english.
 
@@
 
-%TG: this is a fair point we should include in the discussion.
-%TG: though that is just a "practical" point for the vertebrate fossil record. Foraminiferas, for example, do not behave this way (or at a way finer scale).
+We've added the following caveat to the discussion:
+
+\textit{Finally, the difference between the time-binning and time-slicing results could also simply be due to the nature of the fossil record.
+Rates of sedimentation vary in time and space influencing the groups found within the rock record and their temporal distribution. 
+In this case, different beds could represent different ``packages'' of fauna through time separated by gaps, resulting in natural ``bins'' rather than slices of the data.
+Slicing through such strata will yield similar results no matter where in time the slice occurs.
+It is important to note however, that the time slicing method also includes ancestral estimations (either through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
+Additionally, this effect is likely to be most obvious in vertebrates and less problematic for groups like Foraminifera.} lines @@@
+
+% NC: Had to edit this quite a bit to make the English make sense. Does it still convey your meaning wrt forams?
 
@@
 
-%TG: good idea, we (I) can make a figure designating the three different strategies.
+We've added a figure describing the different time-sampling strategies and changed the relevant methods section to be clearer. % NC: Haven't changed the methods yet.
 
@@
 
-%TG: I'll let you change the colours (I still don't want to touch ggplot!) how about the darkgrey/orange/blue palette?
-%Here are the hexadecimals:
-%colours_blind <- c("orange" = "#F65205", "blue" = "#3E9CBA", "darkgrey" = "#1C1C1C")
-%NC: The colours are colour blind friendly already, but I'll switch to using viridis.
-
-Interestingly the palettes in ggplot2 are chosen so that they are distinguishable by people that are colour blind.
-But it's no problem to change the colours.
 Hopefully the new colour scheme is easier to discern.
@@
 % ~~~~~~~~~~~~~~~~~~~~~~~~
-
-%NC: It says in the instructions to respond to these too!
 
@@
 We have split the appendices into three files S1) additional details of the datasets; S2) additional figures; and S3) additional tables (now in .docx format rather than PDF).
-% NC: Just need to sort S2 when we have the new figure. I've added them as is to Dryad to get the reference link.
 

### 2018-02-12 manuscript/response_to_reviewers.tex
Added last comment from reviewer 2 + double checked reviewer 1's

@@
 
-\textit{This approach is similar to Brusatte et al. (2011) but uses model based estimations (rather than parsimony) allowing to control for ambiguous nodes (i.e. poorly estimated).} lines @@@
+\textit{This approach is similar to Brusatte et al. (2011) but uses model based estimations (rather than parsimony) allowing to control for ambiguous nodes (i.e. poorly estimated).} lines 141-144.
+%TG: checked and matched with manuscript.
 
@@
 \textit{To explore disparity-through-time in our datasets, we used a morphospace approach (CITATIONS).
-Morphospaces can be obtained from any multidimensional morphological data set but can differ in the data used (e.g.
-discrete or continuous), and whether they include phylogenetic data or not.
-% NC: do you mean cladistic data i.e.
-data used to build phylogenies?
+Morphospaces can be obtained from any multidimensional morphological data set but can differ in the data used (e.g. discrete or continuous), and whether they include phylogenetic data or not. 
 Although empirical morphospaces from discrete or continuous data have been shown to have similar properties (CITATIONS), our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis;} c.f.
 \textit{geometric morphometric data) and includes some phylogenetic information (see above).
-Mathematically, our morphospaces are $n$ dimensional objects that summarise the distances between discrete morphological characters of the taxa present and their ancestors.} lines @@@
+Mathematically, our morphospaces are $n$ dimensional objects that summarise the distances between discrete morphological characters of the taxa present and their ancestors.} lines 145-158
+%TG: checked and matched with manuscript.
 
@@
 (2016) and Beck \& Lee (2014) datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Fig.
-3; Supporting Information Appendix S2: Figs A3-A4).} lines @@@
+3; Supporting Information Appendix S2: Figs A3-A4).} lines 448-461
+%TG: checked and matched with manuscript.
 
@@
 proximity) except in the Beck \& Lee (2014) dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig.
-3; Supporting Information Appendix S2: Figs A3-A4).} lines @@@
+3; Supporting Information Appendix S2: Figs A3-A4).} lines 461-465
+%TG: checked and matched with manuscript.
 
@@
 In fact, when applying the \texttt{equal.split} model (or \texttt{gradual.splits}) on a decent number of bootstrap replicates (say 100), the ancestor's PCO1 value (0.04 in the reviewer's example) will be sampled 50\% of the time and the same for the descendant's value (0.03 - with the probability changing as a function of branch length for the \texttt{gradual.splits} model).
-The resulting metric, based on this value would be $0.5\times0.04 + 0.5\times0.03 = 0.035$ thus identical to what the reviewer suggests, without modifying the multidimensional space.
+The resulting metric would then be based on the following PCO1 value: $0.5\times0.04 + 0.5\times0.03 = 0.035$ thus identical to what the reviewer suggests, without modifying the multidimensional space.
 
@@
 By estimating this value rather than generating it (i.e.
-creating a new element mid-way along the branch that would be the average of the descendant and ancestor - $0.035$) we obtain the same results without modifying the morphospace properties.} lines @@@
+creating a new element mid-way along the branch that would be the average of the descendant and ancestor - $0.035$) we obtain the same results without modifying the morphospace properties.} lines 318-325
+%TG: checked and matched with manuscript.
 
@@
 
-\textit{Finally, the difference between the time-binning and time-slicing results could also simply be due to the nature of the fossil record.
+\textit{The difference between the time-binning and time-slicing results could also simply be due to the nature of the fossil record.
 Rates of sedimentation vary in time and space influencing the groups found within the rock record and their temporal distribution. 
@@
 It is important to note however, that the time slicing method also includes ancestral estimations (either through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
-Additionally, this effect is likely to be most obvious in vertebrates and less problematic for groups like Foraminifera.} lines @@@
+Additionally, this effect is likely to be most obvious in vertebrates or groups where the fossil is ``patchy'' but less problematic for groups like Foraminifera.} lines @@@
 
@@
 % NC: Yeah I'm not sure either, but maybe needs a paragraph in the discussion? Currently the discussion is short (and written by me in about an hour!) so would be good to add something there.
+
+We've added the following caveat to the discussion:
+
+\textit{Finally, Hunt et al 2015 found that time series are best characterized by gradual directional changes (biased random walks).
+In fact, homogeneous directional changes are more likely to be supported than heterogeneous ones (e.g. punctuated changes) in longer duration series with few samples per series.
+In our implementation of time-slicing, the models are not selected based on any fit value (e.g. AIC) but merely on researcher's assumptions.
+We thus suggest that both types of models (punctuated and gradual) are tested during analysis unless there could be strong independent support for one or the other.}
 
@@
 
-Done, and converted to docx.
+Done, and converted to \texttt{docx}.
+
+\bigskip
 
 We hope we have dealt with all these comments appropriately. 
+Please let us know if you need any further information,
 
-Please let us know if you need any further information,
+\bigskip
 

### 2018-02-12 manuscript/time-slice.tex
Added last comment from reviewer 2 + double checked reviewer 1's

@@
 To explore disparity-through-time in our datasets, we used a morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
-Morphospace can be obtained from any multidimensional morphological data set but can differ in the data used (e.g. discrete or continuous) or whether they include phylogenetic data or not.
-Although empirical morphospaces from discrete or continuous data have been shown to have similar properties \citep{foth2012different,hetherington2015cladistic}, we would like to highlight that our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis; \textit{c.f.} geometric morphometric data) and includes some phylogenetic information (see above).
-Mathematically, our morphospaces are $n$ dimensional objects that summarises the distances between discrete morphological characters of the taxa present and their ancestors.
+Morphospaces can be obtained from any multidimensional morphological data set but can differ in the data used (e.g. discrete or continuous), and whether they include phylogenetic data or not.
+Although empirical morphospaces from discrete or continuous data have been shown to have similar properties \citep{foth2012different,hetherington2015cladistic}, our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis; \textit{c.f.} geometric morphometric data) and includes some phylogenetic information (see above).
+Mathematically, our morphospaces are $n$ dimensional objects that summarise the distances between discrete morphological characters of the taxa present and their ancestors.
 
@@
 distance between species, variance of each axis, etc.) to remain constant.
-For example, using the ``equal.split'' model between an ancestor and a descendant with PCO1 values of respectively $0.04$ and $0.03$, after a sufficient number of bootstrap replicates (e.g.
+For example, using the ``equal.splits'' model on an ancestor and a descendant with PCO1 values of respectively $0.04$ and $0.03$, after a sufficient number of bootstrap replicates (e.g.
 100) the value along the branch will be close to $0.5\times0.04 + 0.5\times0.03 = 0.035$.
 By estimating this value rather than generating it (i.e.
-creating a new element midway along the branch that would be the average of the descendant and ancestor - $0.035$) we obtain the same results without modifying the morphospace properties.
+creating a new element mid-way along the branch that would be the average of the descendant and ancestor - $0.035$) we obtain the same results without modifying the morphospace properties.
 
@@
 \paragraph{Disparity peaks.}
-In the \cite{beckancient2014} and \cite{bapst2016topology} datasets, disparity peaks occur much at much older ages when time-slicing rather than time binning approaches are used (Fig.
+In the \cite{beckancient2014} and \cite{bapst2016topology}) datasets, disparity peaks occur much at much older ages when time-slicing rather than time binning approaches are used (Fig.
 \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
@@
 \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
-Across the three time binning methods, the \cite{brusatte2014gradual} dataset have similar disparity peaks whichever method is used, the \cite{wright2017bayesian} dataset only had variation in peaks when using un-equal time bins (stratigraphy) whereas in the \cite{bapst2016topology} and \cite{beckancient2014} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Fig.
+Across the three time binning methods, the \cite{brusatte2014gradual} dataset has similar disparity peaks whichever method is used, the \cite{wright2017bayesian} dataset only had variation in peaks when using unequal time bins (stratigraphy), whereas in the \cite{bapst2016topology} and \cite{beckancient2014} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Fig.
 \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
-Additionally, there seem to be low discrepancy within the time-slicing methods (gradual \textit{vs.} proximity) expect in \cite{beckancient2014} dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig.
-\ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
+Additionally, there seems to be small discrepancies within the time-slicing methods (gradual \textit{vs.} proximity) except in the \cite{beckancient2014} dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig.
+\ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4)
 
@@
 Care should be taken in interpreting these, as they are highly dependent on the data and models used for the estimations \citep{Slateretal2012,doi:10.1080/10635150801910451}.
-Finally, the difference between the time-binning and time-slicing methods could also simply be favoured towards time binning due to the nature of the fossil record itself.
+The difference between the time-binning and time-slicing methods could also simply be favoured towards time binning due to the nature of the fossil record itself.
 The rates of sedimentation varies in time a space and can be really sporadic, influencing the groups found within and their temporal distribution (e.g. vertebrates, c.f. foraminifera).
@@
 It is important to note however, that, as specified above, the time slicing method also includes ancestral estimations data (whether through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
+Finally, \cite{Hunt21042015} found that time series are best characterized by gradual directional changes (biased random walks).
+In fact, homogeneous directional changes are more likely to be supported than heterogeneous ones (e.g. punctuated changes) in longer duration series with few samples per series.
+In our implementation of time-slicing, the models are not selected based on any fit value (e.g. AIC) but merely on researcher's assumptions.
+We thus suggest that both types of models (punctuated and gradual) are tested during analysis unless there could be strong independent support for one or the other.
 
@@
   TG acknowledges support from the Australian Discovery Project Grant number DP170103227 awarded to Vera Weisbecker.
-  We thank Dave Bapst, Graeme Lloyd, April Wright and Davey Wright for assistance in gathering data for the analyses and/or discussions about the approach; and Steve Brusatte and one anonymous reviewer for helpful comments on the manuscript.
+  We thank Dave Bapst, Graeme Lloyd, April Wright and David Wright for assistance in gathering data for the analyses and/or discussions about the approach; and Steve Brusatte, Sally Thomas and one anonymous reviewer for helpful comments on the manuscript.
 

### 2018-02-12 manuscript/response_to_reviewers.tex
Double checked reviewer 2 comments

@@
 \textit{Second, the time binning approaches (whether bins are equally sized or not) favour punctuated equilibrium modes of evolution.
-Whether the disparity represents an average across the interval (with no interpretation of if or how it varied within the time bin), or it is effectively postulated to be constant, when analysing the changes in disparity-through-time, this method will only allow changes in disparity to occur between intervals rather than also allowing for gradual changes within intervals (a pattern that is fairly common in the fossil record, CITATION).} lines @@@
-
-%TG: tone done the statement? % NC: yeah, not sure what else we can do. %TG: Does that kinda work? NC: sounds ok to me, I tidied the english.
+Whether the disparity represents an average across the interval (with no interpretation of if or how it varied within the time bin), or it is effectively postulated to be constant, when analysing the changes in disparity-through-time, this method will only allow changes in disparity to occur between intervals rather than also allowing for gradual changes within intervals (a pattern that is fairly common in the fossil record, CITATION).} lines 69-77
+%TG: checked and matched with manuscript.
 
@@
 It is important to note however, that the time slicing method also includes ancestral estimations (either through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
-Additionally, this effect is likely to be most obvious in vertebrates or groups where the fossil is ``patchy'' but less problematic for groups like Foraminifera.} lines @@@
-
-% NC: Had to edit this quite a bit to make the English make sense. Does it still convey your meaning wrt forams?
+Additionally, this effect is likely to be most obvious in vertebrates or groups where the fossil is ``patchy'' but less problematic for groups with a more continuous one like Foraminifera.} lines 523-536
+%TG: checked and matched with manuscript.
 
@@
 In our implementation of time-slicing, the models are not selected based on any fit value (e.g. AIC) but merely on researcher's assumptions.
-We thus suggest that both types of models (punctuated and gradual) are tested during analysis unless there could be strong independent support for one or the other.}
+We thus suggest that both types of models (punctuated and gradual) are tested during analysis unless there could be strong independent support for one or the other.} lines 536-544
+%TG: checked and matched with manuscript. NC -> could you check for Thomasisms?
 
@@
 
-\textit{We calculated the $k$$\times$$k$ distances using the Gower distance (CITATION), i.e. the number of characters mismatched characters over the number of shared characters.}
+\textit{We calculated the $k$$\times$$k$ distances using the Gower distance (CITATION), i.e. the number of characters mismatched characters over the number of shared characters.} lines 160-162.
+%TG: checked and matched with manuscript.
 

### 2018-02-12 manuscript/time-slice.tex
Double checked reviewer 2 comments

@@
 Second, the time binning approaches (whether bins are equally sized or not) favour punctuated equilibrium modes of evolution.
-Whether the disparity represents an average across the interval (with no interpretation on if or how it varied within the time bin) or it is \textit{effectively} postulated to be constant, when analysing the changes in disparity through time, this method will only allow changes in disparity (in average or constant) to occur only \textit{between} intervals rather than also allowing for gradual changes \textit{within} intervals \citep[a pattern that is fairly common in the fossil record;][]{Hunt21042015}.
+Whether the disparity represents an average across the interval (with no interpretation of if or how it varied within the time bin), or it is \textit{effectively} postulated to be constant, when analysing the changes in disparity-through-time, this method will only allow changes in disparity to occur \textit{between} intervals rather than also allowing for gradual changes within intervals \citep[a pattern that is fairly common in the fossil record;][]{Hunt21042015}.
 Third, when investigating changes in disparity due to events at a specific time point (e.g.
@@
 Care should be taken in interpreting these, as they are highly dependent on the data and models used for the estimations \citep{Slateretal2012,doi:10.1080/10635150801910451}.
-The difference between the time-binning and time-slicing methods could also simply be favoured towards time binning due to the nature of the fossil record itself.
-The rates of sedimentation varies in time a space and can be really sporadic, influencing the groups found within and their temporal distribution (e.g. vertebrates, c.f. foraminifera).
-In this case, different beds could represent different ``packages'' of fauna through time separated by gaps thus effectively binning the data rather than slicing.
-For example, slicing through such strata will yield similar results whether where in time does the slice occur.
-It is important to note however, that, as specified above, the time slicing method also includes ancestral estimations data (whether through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
+The difference between the time-binning and time-slicing results could also simply be due to the nature of the fossil record.
+Rates of sedimentation vary in time and space influencing the groups found within the rock record and their temporal distribution. 
+In this case, different beds could represent different ``packages'' of fauna through time separated by gaps, resulting in natural ``bins'' rather than slices of the data.
+Slicing through such strata will yield similar results no matter where in time the slice occurs.
+It is important to note however, that the time slicing method also includes ancestral estimations (either through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
+Additionally, this effect is likely to be most obvious in vertebrates or groups where the fossil is ``patchy'' but less problematic for groups with a more continuous one like Foraminifera.
 Finally, \cite{Hunt21042015} found that time series are best characterized by gradual directional changes (biased random walks).

### 2018-02-12 manuscript/response_to_reviewers.tex
thorough check throughout the manuscript
Make sure the dates for Brusatte and Bapst are _n_ - **66** Mya rather
than 0 in the Table 1 (couldn’t find the word version).

@@
 
-\textit{Additionally, there seems to be small discrepancies within the time-slicing methods (gradual vs.
+\textit{Additionally, there seems to be small discrepancies within the time-slicing methods (gradual.split vs.
 proximity) except in the Beck \& Lee (2014) dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig.
@@
 
-\textit{We calculated the $k$$\times$$k$ distances using the Gower distance (CITATION), i.e. the number of characters mismatched characters over the number of shared characters.} lines 160-162.
+\textit{We calculated the $n$$\times$$n$ distances using the Gower distance (CITATION), i.e. the number of characters mismatched characters over the number of shared characters.} lines 160-162.
 %TG: checked and matched with manuscript.

### 2018-02-12 manuscript/time-slice.tex
thorough check throughout the manuscript
Make sure the dates for Brusatte and Bapst are _n_ - **66** Mya rather
than 0 in the Table 1 (couldn’t find the word version).

@@
 
-Disparity-through-time analyses can be used to determine how morphological diversity changes in response to mass extinctions, and to investigate the drivers of morphological change.
+Disparity-through-time analyses can be used to determine how morphological diversity changes in response to mass extinctions, or to investigate the drivers of morphological change.
 These analyses are routinely applied to palaeobiological datasets, yet although there is much discussion about how to best calculate disparity, there has been little consideration of how taxa should be sub-sampled through time.
@@
 To address these issues, we propose a ``time-slicing'' approach that takes advantage of the wealth of palaeontological datasets which now have associated phylogenies. 
-Time-slicing uses a phylogenetic tree and considers subsets of taxa at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time (a similar approach is outlined in \citep{halliday2016eutherian}).
+Time-slicing uses a phylogenetic tree and considers subsets of taxa at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time \citep[a similar approach is outlined in][]{halliday2016eutherian}.
 This results in even-sampling across time and permits us to define the underlying model of character evolution (punctuated or gradual).  
@@
 To test the different time sub-sampling methods, we followed the protocol below (Fig. \ref{fig:overview}). 
-All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (\url{https://github.com/nhcooper123/time-slice}.
+All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (\url{https://github.com/nhcooper123/time-slice}).
 % NC: Add proper reference to this before publication.
@@
     {Outline of the disparity-through-time pipeline.
-1) We use ancestral character estimation to infer nodal character states; 2) we measure the pairwise Gower distance between the tip character states and nodal character states; 3) we ordinate the distance matrix using principal coordinates analysis (PCoA); 4) we time sub-sample the PCoA matrix using time bins defined by stratigraphic periods, equally sized time bins and time-slices (using six methods to estimate ordination scores for branches); and finally 5) we measure disparity-through-time for each of these methods.}
+1) We use ancestral character estimation to infer nodal character states; 2) we measure the pairwise Gower distance between the tip character states and nodal character states; 3) we ordinate the distance matrix using principal coordinates analysis (PCoA/PCO); 4) we time sub-sample the PCoA matrix using time bins defined by stratigraphic periods, equally sized time bins and time-slices (using six methods to estimate ordination scores for branches); and finally 5) we measure disparity-through-time for each of these methods.}
     \label{fig:overview}
@@
 Table \ref{table:datasets} and the Supporting Information Appendix S1 provide more details. 
+%TG: TODO: I can't find the Table in Word format anymore. We have to make sure the dates are correct. 
 Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
@@
 \paragraph{Constructing distance matrices.}
-To estimate the morphospaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
-We calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the number of characters mismatched characters over the number of shared characters.
+To estimate the morphospaces for each of our datasets we first constructed pairwise distance matrices of length $n$, where $n$ is the total number of tips and nodes in the dataset.
+We calculated the $n$$\times$$n$ distances using the Gower distance \citep{Gower71}, i.e. the number of characters mismatched characters over the number of shared characters.
 This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e.
@@
 \citealt{paradisape:2004}; but distinguished in \citealt{legendre2012numerical}) is an eigen decomposition of the distance matrix.
-
 Because we used Gower distances instead of raw Euclidean distances, negative eigenvalues can be calculated.
-
 To avoid this problem, we first transformed the distance matrices by applying the Cailliez correction \citep{cailliez1983analytical} which adds a constant $c^*$ to the values in a distance matrix (apart from the diagonal) so that all the Gower distances become Euclidean ($d_{Gower}+c^*=d_{Euclidean}$; \citealt{cailliez1983analytical}).
-
-We were then able to extract $n$ eigenvectors for each matrix (representing the $n$ dimensions of the morphospace) where $n$ is equal to $k-2$, i.e.
-the number of taxa in the matrix ($k$) minus the last two eigenvectors that are always null after applying the Cailliez correction.
-
-Contrary to previous studies \citep[e.g][]{brusatte50,cisneros2010,prentice2011,anderson2012using,Hughes20082013,bentonmodels2014}, we use all $n$ dimensions of our morphospaces and not a sub-sample representing the majority of the variance in the distance matrix (e.g.
-selecting only $m$ dimensions that represent up to 90\% of the variance in the distance matrix; \citealt{Brusatte12092008,toljagictriassic-jurassic2013}).
-
+We were then able to extract $k$ eigenvectors for each matrix (representing the $k$ dimensions of the morphospace) where $k$ is equal to $n-2$, i.e.
+the number of taxa in the matrix ($n$) minus the last two eigenvectors that are always null after applying the Cailliez correction.
+Contrary to previous studies \citep[e.g][]{brusatte50,cisneros2010,prentice2011,anderson2012using,Hughes20082013,bentonmodels2014}, we use all $k$ dimensions of our morphospaces and not a sub-sample representing the majority of the variance in the distance matrix (e.g.
+selecting only $x$ dimensions that represent up to 90\% of the variance in the distance matrix; \citealt{Brusatte12092008,toljagictriassic-jurassic2013}).
 Note that our morphospaces represent an ordination of all possible morphologies coded in each study through time.
@@
 First, the range metrics are affected by the uneven sampling of the fossil record \citep{Butler2012}.
-Second, because we include all $n$ dimensions in the analysis (see above), the products of ranges and variances will tend towards zero since the scores of the last dimension are usually really close to zero themselves.
+Second, because we include all $k$ dimensions in the analysis (see above), the products of ranges and variances will tend towards zero since the scores of the last dimension are usually really close to zero themselves.
 We therefore use the sum of variances metric to estimate disparity here:
@@
 \begin{equation}
-\text{disparity} = \sum\sigma^{2}{n_i}
+\text{disparity} = \sum\sigma^{2}{k_i}
 \end{equation}
 
-Where $\sigma^{2}{n_i}$ is the variance for the $n_i^{th}$ dimension ranging from $k$ to $k-2$ with $k$ being the number of taxa in the dataset.
+\noindent where $\sigma^{2}{k_i}$ is the variance for the $k_i^{th}$ dimension ranging from $n$ to $n-2$ with $n$ being the number of taxa in the dataset.
 Note that there are still statistical issues with this metric (such as the co-variance between dimensions not being measured), but for the purposes of comparison with previous work we decided to use a standard metric for these analyses.
@@
     \setcounter{enumii}{4}
-      \item The ``equal splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with an equal probability
+      \item The ``equal splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with an equal probability:
           \begin{equation}
@@
 
-    \item The ``gradual splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with a probability function of the distance between the nodes/tip at the ends of the branch and the slice.
+    \item The ``gradual splits'' model (probabilistic), selects the ordination score from both the ancestor and the descendant with a probability function of the distance between the nodes/tip at the ends of the branch and the slice:
 
@@
           \end{equation}
-    \noindent Where $d(x,y)$ is the distance measured as branch length.
+    \noindent where $d(x,y)$ is the distance between the two elements $x$, $y$ (ancestor, slice or descendant) measured in units of branch length.
     \end{enumerate}
@@
     \item Time bins (unequal). 
-    We calculated disparity for the taxa in each stratigraphic period (age or epoch). 
+    We calculated disparity for the taxa in each stratigraphic period (stage or epoch). 
     To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time bin by randomly resampling with replacement a new sub-sample of taxa from the observed taxa in the bin 100 times.
@@
     \item Time-slices (non-equidistant).
-    We calculated disparity using our time-slicing approach with slices occurring at the midpoint of each stratigraphic period (age or epoch), and using all six time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits).
+    We calculated disparity using our time-slicing approach with slices occurring at the midpoint of each stratigraphic period (stage or epoch), and using all six time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits).
     To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement as described above for the stratigraphic time bins.
@@
     \item Time bins (equal). 
-    We calculated disparity for the taxa in each time bin where time bin size was defined by the mean \textit{duration} of the stratigraphic period (age or epoch), and bootstrapped the disparity values as described above.
+    We calculated disparity for the taxa in each time bin where time bin size was defined by the mean \textit{duration} of the stratigraphic period (stage or epoch), and bootstrapped the disparity values as described above.
     \item Time-slices (equidistant).
-    We calculated disparity using our time-slicing approach where the interval between slices, was defined by the mean \textit{duration} of the stratigraphic period (age or epoch).
+    We calculated disparity using our time-slicing approach where the interval between slices, was defined by the mean \textit{duration} of the stratigraphic period (stage or epoch).
     We used the six time-slicing methods and bootstrapped as described above.
@@
 \paragraph{Disparity peaks.}
-In the \cite{beckancient2014} and \cite{bapst2016topology}) datasets, disparity peaks occur much at much older ages when time-slicing rather than time binning approaches are used (Fig.
+In the \cite{beckancient2014} and \cite{bapst2016topology} datasets, disparity peaks occur much at much older ages when time-slicing rather than time binning approaches are used (Fig.
 \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
@@
 \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
-Additionally, there seems to be small discrepancies within the time-slicing methods (gradual \textit{vs.} proximity) except in the \cite{beckancient2014} dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig.
+Additionally, there seems to be small discrepancies within the time-slicing methods (gradual.split \textit{vs.} proximity) except in the \cite{beckancient2014} dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig.
 \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4)

### 2018-02-12 manuscript/tables/table_datasets.tex
Minor edits to tables
Correct dates on table 1

@@
 \begin{table}[!htbp]
+\caption{Details of the datasets used in this study. Age ranges are root time to most recent tip taxon.} 
 \centering
@@
   \# characters & 421 & 853 & 374 &  87 \\ 
-  Age range (MYA) & 171.8 - 0 & 168.5 - 0 & 207.2 - 0 & 485.4 - 372.2 \\ 
+  Age range (MYA) & 171.8 - 0 & 168.5 - 66 & 207.2 - 66 & 485.4 - 372.2 \\ 
   Mass extinction (MYA) & 66 (K-Pg) & NA & NA & 443 (O-S) \\ 
@@
 \end{tabular}
-\caption{Details of the datasets used in this study.
-Note that although Brusatte2014 and Bapst2016 cross the Cretaceous Palaeogene (K-Pg) mass extinction there are not enough taxa post K-Pg to investigate the effects of the mass extinction on disparity.
-Age ranges are root time to most recent tip taxon.}
 \label{table:datasets}  

### 2018-02-12 manuscript/time-slice.tex
Fix dryad refs and minor edits

@@
 Table \ref{table:datasets} and the Supporting Information Appendix S1 provide more details. 
-%TG: TODO: I can't find the Table in Word format anymore. We have to make sure the dates are correct. 
 Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
@@
     Stratigraphy: time sub-samples are defined by stratigraphic periods.
-    Here there are five stratigraphic periods in the 20 MY (million years) time frame of interest, i.e. five bins/slices with variable sizes/intervals.
+    Here there are five stratigraphic periods in the 20 My (million years) time frame of interest, i.e. five bins/slices with variable sizes/intervals.
     Duration: time sub-samples are defined based on the mean duration of stratigraphic periods in the time frame of interest. 
-    Here, the mean duration of stratigraphic periods is 5 MY, so there are four bins/slices of 5 MY duration (or four slices with 5 MY intervals between them) in the 20 MY time frame of interest. 
+    Here, the mean duration of stratigraphic periods is 5 My, so there are four bins/slices of 5 My duration (or four slices with 5 My intervals between them) in the 20 My time frame of interest. 
     Number: time sub-samples are defined based on the number of stratigraphic periods in the time frame of interest. 
-    Here, there are five stratigraphic periods, so there are five bins/slices of 4 MY duration (or five slices with 4 MY intervals between them) in the 20 MY time frame of interest.}
+    Here, there are five stratigraphic periods, so there are five bins/slices of 4 My duration (or five slices with 4 My intervals between them) in the 20 My time frame of interest.}
     \label{fig:bins}
@@
 % add proper link to zenodo
-Supporting information can be found on Dryad at % Add Dryad link
+Supporting information can be found on Dryad at https://doi.org/10.5061/dryad.vp4q518 \cite{data-time-slice}.
 
 \section{Supporting information}
-Additional Supporting information can be found on Dryad at:\\
+Additional Supporting information can be found on Dryad at: https://doi.org/10.5061/dryad.vp4q518 \cite{data-time-slice}.\\
 Appendix S1: Extra details of datasets. \\

### 2018-02-12 manuscript/time-slice.tex
Complete edits and reformat back to nice LaTeX layout

@@
 \usepackage{graphicx} 	
-\usepackage[nomarkers, nolists, figuresonly]{endfloat} 
+%\usepackage[nomarkers, nolists, figuresonly]{endfloat} 
 \usepackage{pdflscape}	
@@
 The nature of disparity (i.e. it is a diversity metric), means it cannot be calculated using a single individual, so some way of sub-sampling taxa is required.
-Changes in disparity-through-time are generally investigated by calculating the disparity of taxa present during specific time intervals or time bins \citep[e.g][]{cisneros2010,prentice2011,Hughes20082013,hopkinsdecoupling2013,bentonmodels2014,bensonfaunal2014}.
+Changes in disparity-through-time are generally investigated by calculating the disparity of taxa present during specific time intervals or time bins \citep[e.g.][]{cisneros2010,prentice2011,Hughes20082013,hopkinsdecoupling2013,bentonmodels2014,bensonfaunal2014}.
 These time bins are usually defined based on stratigraphy \citep[e.g.][]{cisneros2010,prentice2011,Hughes20082013,bentonmodels2014} but can also be arbitrarily chosen time bins of equal (or approximately equal) duration \citep{Butler2012,hopkinsdecoupling2013,bensonfaunal2014}.
@@
     \centering
-    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure1-outlinesvg.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure1-outlinesvg.pdf}
     \caption[Outline of the disparity-through-time pipeline]
@@
 % Dataset details table
-%\begin{landscape}
-%\input{tables/table_datasets.tex} 
+\begin{landscape}
+\input{tables/table_datasets.tex} 
   % Label in input or referencing doesn't work
-%\end{landscape}
+\end{landscape}
 
@@
 This allowed our results to be less dependent on the quality (or the absence thereof) of the ancestral state estimations methods, especially in parts of the datasets where data were sparse. 
-This approach is similar to \cite{brusatte2011phylogenetic} but uses model based estimations (rather than parsimony) allowing to control for ambiguous nodes (i.e. poorly estimated).
+This approach is similar to \cite{brusatte2011phylogenetic} but uses model based estimations (rather than parsimony) allowing us to control for ambiguous (i.e. poorly estimated) nodes.
 
@@
 Morphospaces can be obtained from any multidimensional morphological data set but can differ in the data used (e.g. discrete or continuous), and whether they include phylogenetic data or not.
-Although empirical morphospaces from discrete or continuous data have been shown to have similar properties \citep{foth2012different,hetherington2015cladistic}, our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis; \textit{c.f.} geometric morphometric data) and includes some phylogenetic information (see above).
+Although empirical morphospaces from discrete or continuous data have been shown to have similar properties \citep{foth2012different,hetherington2015cladistic}, our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis; \textit{c.f.} geometric morphometric data) and include some phylogenetic information (see above).
 Mathematically, our morphospaces are $n$ dimensional objects that summarise the distances between discrete morphological characters of the taxa present and their ancestors.
@@
 To estimate the morphospaces for each of our datasets we first constructed pairwise distance matrices of length $n$, where $n$ is the total number of tips and nodes in the dataset.
-We calculated the $n$$\times$$n$ distances using the Gower distance \citep{Gower71}, i.e. the number of characters mismatched characters over the number of shared characters.
+We calculated the $n$$\times$$n$ distances using the Gower distance \citep{Gower71}, i.e. the number of mismatched characters over the number of shared characters.
 This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e.
@@
 
-To compare the time binning and time-slicing approaches we applied the methods as follows (see Figure \ref{figure:bins}).
+To compare the time binning and time-slicing approaches we applied the methods as follows (see Fig. \ref{figure:bins}).
 
@@
     \centering
-    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure2-binning.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure2-binning.pdf}
     \caption[Outline of the three time sub-sampling methods]
@@
 \begin{enumerate}
-  \item Stratigraphy: time sub-samples defined by stratigraphic periods. 
+  \item Stratigraphy: time sub-samples defined by stratigraphic periods (Fig. \ref{fig:bins}). 
   \begin{enumerate}[(i)]
@@
 
-  \item Duration: time sub-samples defined by the duration of stratigraphic periods. 
+  \item Duration: time sub-samples defined by the duration of stratigraphic periods (Fig. \ref{fig:bins}). 
   \begin{enumerate}[(i)]
@@
 
-  \item Number: time sub-samples defined by the number of stratigraphic periods. 
+  \item Number: time sub-samples defined by the number of stratigraphic periods (Fig. \ref{fig:bins}). 
   \begin{enumerate}[(i)]
@@
     \centering
-    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure2-fig-dtt-epoch-revision.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure3-fig-dtt-epoch-revision.pdf}
     \caption[Relative disparity-through-time.]
@@
     Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
-    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset.
-    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
+    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the mean duration of stratigraphic epochs in the time frame of the dataset.
+    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the mean number of stratigraphic epochs in the time frame of the dataset. 
     In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
@@
 For example, the \cite{brusatte2014gradual}, \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, show significant differences when using bins versus time-slices defined by stratigraphy, but the \cite{beckancient2014} dataset appears robust to these different approaches.
-Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} datasets have different disparities when the number of bins or slices is the average number of stratigraphic periods, but this is not seen in the \cite{wright2017bayesian} dataset.
+Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} datasets have different disparities when the number of bins or slices is the mean number of stratigraphic periods, but this is not seen in the \cite{wright2017bayesian} dataset.
 Note that for epochs, we find fewer significant differences simply because the smaller numbers of bins and slices being compared means we have low power to detect a significant difference.
@@
 % Wilcoxon results table
-%\input{tables/table_wilcox.tex}
+\input{tables/table_wilcox.tex}
 
@@
 \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
-Additionally, there seems to be small discrepancies within the time-slicing methods (gradual.split \textit{vs.} proximity) except in the \cite{beckancient2014} dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig.
+Additionally, there seem to be small discrepancies within the time-slicing methods (gradual.split \textit{vs.} proximity) except in the \cite{beckancient2014} dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig.
 \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4)
@@
     \centering
-    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure3-fig-peaks-epoch-revision.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure4-fig-peaks-epoch-revision.pdf}
     \caption[Timing of peak disparity for four example datasets.]
@@
     Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
-    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset.
-    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
+    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the mean duration of stratigraphic epochs in the time frame of the dataset.
+    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the mean number of stratigraphic epochs in the time frame of the dataset. 
     For time bins the points indicate the maximum and minimum ages of the time bin within which peak disparities appeared.
@@
     \centering
-    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure4-fig-extinction-effect.pdf}
+    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure5-fig-extinction-effect.pdf}
     \caption[Extinction example datasets.]
@@
 This is particularly evident when comparing stratigraphic time bins to time slicing methods, where for most of our datasets we see a much later time to peak disparity. 
-This could have major implications for our understanding of how morphological diversity changes through time, for example in response to climate effects.
+This could have major implications for our understanding of how morphological diversity changes through time, for example in response to climate.
 These results highlight the sensitivity of disparity-through-time analyses to the choice of time sub-sampling method. 
@@
 Of course the method also has limitations.
-The main one of these is a practical one; it requires a time-calibrated phylogeny which are not available for all palaeontological datasets.
+The main one of these is a practical one; it requires a time-calibrated phylogeny and these are not available for all palaeontological datasets.
 Furthermore, like most phylogeny based methods, time-slicing depends on ancestral state estimations.
@@
 It is important to note however, that the time slicing method also includes ancestral estimations (either through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
-Additionally, this effect is likely to be most obvious in vertebrates or groups where the fossil is ``patchy'' but less problematic for groups with a more continuous one like Foraminifera.
+Additionally, this effect is likely to be most obvious in groups where the fossil record is ``patchy'', e.g.
+vertebrates, but less problematic for groups with a more continuous record like Foraminifera.
 Finally, \cite{Hunt21042015} found that time series are best characterized by gradual directional changes (biased random walks).
-In fact, homogeneous directional changes are more likely to be supported than heterogeneous ones (e.g. punctuated changes) in longer duration series with few samples per series.
-In our implementation of time-slicing, the models are not selected based on any fit value (e.g. AIC) but merely on researcher's assumptions.
-We thus suggest that both types of models (punctuated and gradual) are tested during analysis unless there could be strong independent support for one or the other.
+In fact, homogeneous directional changes are more likely to be supported than heterogeneous ones (e.g. punctuated changes) in longer duration series with few samples in each series.
+In our implementation of time-slicing, the models are not selected based on any model fit criterion (e.g. AIC) but merely on researcher assumptions.
+We thus suggest that both types of models (punctuated and gradual) are tested during analysis, unless there is strong independent support for one or the other.
 
@@
 % add proper link to zenodo
-Supporting information can be found on Dryad at https://doi.org/10.5061/dryad.vp4q518 \cite{data-time-slice}.
+Supporting information can be found on Dryad at https://doi.org/10.5061/dryad.vp4q518 \citep{data-time-slice}.
 
 \section{Supporting information}
-Additional Supporting information can be found on Dryad at: https://doi.org/10.5061/dryad.vp4q518 \cite{data-time-slice}.\\
+Additional Supporting information can be found on Dryad at: https://doi.org/10.5061/dryad.vp4q518 \citep{data-time-slice}.\\
 Appendix S1: Extra details of datasets. \\
@@
 
-\newpage
-\section{Figure legends}
-
 \end{document}

### 2018-02-12 manuscript/response_to_reviewers.tex
final edits to line numbers etc.

@@
 We are very grateful to the referees for their helpful and constructive comments. 
-Below we provide their comments in blue text with our responses.
+Below we provide their comments in blue text with our responses below.
 
@@
 
-\textit{This approach is similar to Brusatte et al. (2011) but uses model based estimations (rather than parsimony) allowing to control for ambiguous nodes (i.e. poorly estimated).} lines 141-144.
-%TG: checked and matched with manuscript.
+\textit{This approach is similar to Brusatte et al. (2011) but uses model based estimations (rather than parsimony) allowing us to control for ambiguous (i.e. poorly estimated) nodes.} lines 132-134.
 
@@
 Although empirical morphospaces from discrete or continuous data have been shown to have similar properties (CITATIONS), our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis;} c.f.
-\textit{geometric morphometric data) and includes some phylogenetic information (see above).
-Mathematically, our morphospaces are $n$ dimensional objects that summarise the distances between discrete morphological characters of the taxa present and their ancestors.} lines 145-158
-%TG: checked and matched with manuscript.
+\textit{geometric morphometric data) and include some phylogenetic information (see above).
+Mathematically, our morphospaces are $n$ dimensional objects that summarise the distances between discrete morphological characters of the taxa present and their ancestors.} lines 138-147.
 
@@
 (2016) and Beck \& Lee (2014) datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Fig.
-3; Supporting Information Appendix S2: Figs A3-A4).} lines 448-461
-%TG: checked and matched with manuscript.
+4; Supporting Information Appendix S2: Figs A3-A4).} lines 406-417.
 
@@
 proximity) except in the Beck \& Lee (2014) dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig.
-3; Supporting Information Appendix S2: Figs A3-A4).} lines 461-465
-%TG: checked and matched with manuscript.
+3; Supporting Information Appendix S2: Figs A3-A4).} lines 417-421.
 
@@
 By estimating this value rather than generating it (i.e.
-creating a new element mid-way along the branch that would be the average of the descendant and ancestor - $0.035$) we obtain the same results without modifying the morphospace properties.} lines 318-325
-%TG: checked and matched with manuscript.
+creating a new element mid-way along the branch that would be the average of the descendant and ancestor - $0.035$) we obtain the same results without modifying the morphospace properties.} lines 290-296.
 
@@
 It is important to note however, that the time slicing method also includes ancestral estimations (either through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
-Additionally, this effect is likely to be most obvious in vertebrates or groups where the fossil is ``patchy'' but less problematic for groups with a more continuous one like Foraminifera.} lines 523-536
-%TG: checked and matched with manuscript.
+Additionally, this effect is likely to be most obvious in groups where the fossil record is ``patchy'' e.g.
+vertebrates, but less problematic for groups with a more continuous record like Foraminifera.} lines 473-484.
 
@@
 
-%TG: not sure where the reviewer want us to go with that. Is that the same nuancing point as the comment before? In that case we can simply merge them and make a caveat paragraph. 
-% NC: Yeah I'm not sure either, but maybe needs a paragraph in the discussion? Currently the discussion is short (and written by me in about an hour!) so would be good to add something there.
-
 We've added the following caveat to the discussion:
@@
 \textit{Finally, Hunt et al 2015 found that time series are best characterized by gradual directional changes (biased random walks).
-In fact, homogeneous directional changes are more likely to be supported than heterogeneous ones (e.g. punctuated changes) in longer duration series with few samples per series.
-In our implementation of time-slicing, the models are not selected based on any fit value (e.g. AIC) but merely on researcher's assumptions.
-We thus suggest that both types of models (punctuated and gradual) are tested during analysis unless there could be strong independent support for one or the other.} lines 536-544
-%TG: checked and matched with manuscript. NC -> could you check for Thomasisms?
+In fact, homogeneous directional changes are more likely to be supported than heterogeneous ones (e.g. punctuated changes) in longer duration series with few samples in each series.
+In our implementation of time-slicing, the models are not selected based on any model fit criterion (e.g. AIC) but merely on researcher assumptions.
+We thus suggest that both types of models (punctuated and gradual) are tested during analysis, unless there is strong independent support for one or the other.} lines 484-491.
 
@@
 
-We've added a figure describing the different time-sampling strategies and changed the relevant methods section to be clearer. % NC: Haven't changed the methods yet.
+We've added a figure describing the different time-sampling strategies which we hope makes this section clearer.
 
@@
 
-\textit{We calculated the $n$$\times$$n$ distances using the Gower distance (CITATION), i.e. the number of characters mismatched characters over the number of shared characters.} lines 160-162.
-%TG: checked and matched with manuscript.
+\textit{We calculated the $n$$\times$$n$ distances using the Gower distance (CITATION), i.e. the number of mismatched characters over the number of shared characters.} lines 151-152.
 
@@
 Done.
-% NC: To do when uploaded
 
@@
 
-We have now archived the code on Zenodo, so provide the citation for this.
-% NC: To do
+We have now archived the code on Zenodo, so provide the citation for this:
+
+Guillerme T., \& Cooper N. 2018. GitHub: nhcooper123/time-slice: Release for publication. Zenodo. http://doi.org/10.5281/zenodo.1172000.
 
@@
 We have fixed this.
-% NC: Fix at converting to Word stage, hopefully there will be a paper to cite by then...
-for now I've used the v 0.5 zenodo citation.
-I've changed the citation of Claddis to Graeme's Biol Linn soc paper.
+The paper accompanying the dispRity package is currently in review so hopefully we will have a better citation to provide at the proof stage.
+For now we have added the Zenodo doi and reference.
+We have changed the Lloyd 2015 reference to the reference for the the paper that introduces the package.
 
